---
title: "Team colors and themes"
description: >
  Ten recipes for team colors: a league's palette, colors per row, team fill
  and outline scales, clashing colors, readable text on team fills, named
  palettes, conference colors, dark backgrounds, and choosing a gt theme.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Team colors and themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

Ten recipes for team colors in ggplot2 and gt: a league's palette at a
glance, one color per row of your data, team fill and outline scales, colors
that clash, readable text on team-colored fills, named palettes, conference
colors, dark backgrounds with dark logos, and picking a table theme. The data
is one season each from the NHL (fastRhockey), the NBA, the WNBA and women's
college basketball (hoopR and wehoop), MLB (baseballr) and college football
(cfbfastR), all read from SportsDataverse release files except MLB's final
standings, which come from the MLB Stats API.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
library(gt)

season <- 2026 # NBA, NHL and college basketball: the 2025-26 season; MLB and WNBA: 2026
```

## 1. See a league's palette

`team_reference()` holds every team's two colors (`color1`, `color2`), with
its conference and division. Laid out by division, with each team's logo above
its two swatches, it is a quick check of what a chart will look like:

```{r palette, fig.width = 8, fig.height = 6.5, fig.alt = "The 32 NHL teams in four rows, one per division. Each team's logo sits above two squares filled with its primary and secondary colors, with the hex codes underneath."}
nhl_ref <- team_reference("nhl") |>
  arrange(division, team_abbr) |>
  mutate(slot = row_number(), .by = division)

swatches <- bind_rows(
  transmute(nhl_ref, division, slot, x = slot - 0.2, color = color1),
  transmute(nhl_ref, division, slot, x = slot + 0.2, color = color2)
)

ggplot(nhl_ref, aes(x = slot)) +
  geom_tile(aes(x = x, y = 1, fill = color), data = swatches,
    width = 0.38, height = 0.5, colour = "grey70", linewidth = 0.2
  ) +
  geom_text(aes(x = x, y = 0.6, label = color), data = swatches, size = 1.8) +
  geom_sdv_logos(aes(y = 1.75, team = team_abbr), sport = "nhl", height = 0.32) +
  scale_fill_identity() +
  facet_wrap(~division, ncol = 1) +
  scale_y_continuous(limits = c(0.4, 2.15)) +
  labs(title = "NHL team colors by division", caption = "Colors: sdvplotR::team_reference()") +
  theme_void() +
  theme(strip.text = element_text(face = "bold", hjust = 0), plot.margin = margin(8, 8, 8, 8))
```

`sdv_team_colors("nhl", type = "all")` returns the same pairs as
`"primary, secondary"` strings, named by team.

## 2. One color per row of your data

`sdv_team_colors(sport, team)` returns one color per value, in the order
given, so it drops straight into `mutate()`; `type = "secondary"` gives the
other color. Any spelling `clean_team_abbrs()` understands works, here
fastRhockey's NHL codes (`LAK`, `TBL`). With color columns in the data, the
identity scales use them as they are:

```{r per-row, fig.alt = "Scatter of the 32 NHL teams' goals for against goals against per game in the 2025-26 regular season. Each team is a circle filled with its primary color and outlined in its secondary color, labelled with its abbreviation."}
nhl_games <- fastRhockey::load_nhl_team_box(season) |>
  as.data.frame() |>
  filter(substr(game_id, 5, 6) == "02") # 02 = regular season

nhl_teams <- nhl_games |>
  summarise(gf = mean(goals), ga = mean(goals_against), .by = team_abbrev) |>
  arrange(desc(gf), team_abbrev) |>
  mutate(
    primary = unname(sdv_team_colors("nhl", team_abbrev)),
    secondary = unname(sdv_team_colors("nhl", team_abbrev, type = "secondary"))
  )
head(nhl_teams, 5)

ggplot(nhl_teams, aes(x = gf, y = ga)) +
  geom_point(aes(fill = primary, colour = secondary), shape = 21, size = 6, stroke = 1.5) +
  ggrepel::geom_text_repel(aes(label = team_abbrev), size = 2.5, point.size = 6, seed = 1) +
  scale_fill_identity() +
  scale_colour_identity() +
  scale_y_reverse() +
  labs(
    title = "NHL goals for and against per game, 2025-26",
    x = "Goals for per game", y = "Goals against per game (reversed)",
    caption = "Data: fastRhockey::load_nhl_team_box()"
  ) +
  theme_minimal()
```

The labels come from ggrepel, which keeps them off each other and off the
points. A team that doesn't resolve gets `NA`, which the identity scales draw
grey; check with `valid_team_names()` before you publish. A white secondary,
like the Lightning's, disappears as an outline on a white chart (recipe 4).

## 3. Fill with the primary, outline with the secondary

`scale_fill_sdv()` and `scale_colour_sdv()` map a team column straight to
colors, no join needed. Fill with the primary and outline with the secondary
(`type = "secondary"`) for the two-color look of a team's uniform. The WNBA's
2026 regular season by point differential. ESPN files the All-Star Game and
the Commissioner's Cup final as regular-season games too, so keep the games
the schedule marks as standard (`type_abbreviation == "STD"`):

```{r fill-outline, fig.alt = "Bar chart of the 15 WNBA teams' average point differential per game in the 2026 regular season, sorted, each bar filled with the team's primary color and outlined in its secondary color, with the team's logo past the end of the bar."}
wnba_regular <- wehoop::load_wnba_schedule(season) |>
  as.data.frame() |>
  filter(season_type == 2, type_abbreviation == "STD") |>
  pull(game_id)
wnba <- wehoop::load_wnba_team_box(season) |>
  as.data.frame() |>
  filter(game_id %in% wnba_regular) |>
  summarise(
    games = n(),
    pf = mean(team_score),
    pa = mean(opponent_team_score),
    .by = team_abbreviation
  ) |>
  mutate(diff = pf - pa, team = team_abbreviation) |>
  arrange(desc(diff), team) |>
  mutate(team_f = factor(team, levels = team))

ggplot(wnba, aes(x = team_f, y = diff)) +
  geom_col(aes(fill = team, colour = team), width = 0.7, linewidth = 1.2) +
  geom_hline(yintercept = 0, colour = "grey30") +
  geom_sdv_logos(aes(y = diff + if_else(diff >= 0, 1.4, -1.4), team = team), sport = "wnba", width = 0.05) +
  scale_fill_sdv(sport = "wnba") +
  scale_colour_sdv(sport = "wnba", type = "secondary") +
  labs(
    title = "WNBA point differential, 2026 regular season",
    x = NULL, y = "Average point differential per game",
    caption = "Data: wehoop::load_wnba_team_box()"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid.major.x = element_blank())
```

`alpha =` on either scale fades the colors, which helps when a logo sits on
the bar.

## 4. When two teams' colors clash

Some rivals share a color: the Lakers', Kings' and Suns' primaries are all
purple, and all three play in the NBA's Pacific Division. Measure the gap
between two colors (a plain RGB distance does the job) and fall back to a
team's secondary when its primary is too close to one already used. Walk the
teams best record first, so the leaders keep their primaries:

```{r clash, fig.alt = "Line chart of the five NBA Pacific Division teams' cumulative wins across the 2025-26 regular season. Lines use each team's primary color, except where it was too close to a color already used, which switches to the team's secondary; each line ends in the team's logo."}
color_gap <- function(a, b) {
  m <- grDevices::col2rgb(c(a, b)) / 255
  sqrt(sum((m[, 1] - m[, 2])^2)) / sqrt(3) # 0 = same color, 1 = black vs white
}

pacific <- c("GS", "LAC", "LAL", "PHX", "SAC") # the NBA's Pacific Division
nba_regular <- hoopR::load_nba_schedule(season) |>
  as.data.frame() |>
  filter(season_type == 2, type_abbreviation == "STD") |> # no All-Star or NBA Cup final
  pull(game_id)
runs <- hoopR::load_nba_team_box(season) |>
  as.data.frame() |>
  filter(game_id %in% nba_regular, team_abbreviation %in% pacific) |>
  arrange(team_abbreviation, game_date, game_id) |>
  mutate(game_no = row_number(), wins = cumsum(team_winner), .by = team_abbreviation) |>
  rename(team = team_abbreviation)
finish <- runs |>
  slice_max(game_no, n = 1, by = team) |>
  arrange(desc(wins), team)

chosen <- character()
for (team in finish$team) {
  primary <- sdv_team_colors("nba", team)
  close <- length(chosen) > 0 && min(vapply(chosen, color_gap, numeric(1), b = primary)) < 0.15
  chosen[team] <- if (close) sdv_team_colors("nba", team, type = "secondary") else primary
}
data.frame(
  team = finish$team,
  primary = unname(sdv_team_colors("nba", finish$team)),
  used = unname(chosen[finish$team])
)

end <- max(finish$game_no)
# keep line-end logos one logo height apart (see the logos cookbook)
finish <- arrange(finish, wins, team)
finish$spot <- Reduce(function(below, y) max(y, below + 5), finish$wins, accumulate = TRUE)

ggplot(runs, aes(x = game_no, y = wins)) +
  geom_line(aes(colour = team), linewidth = 1) + # colour here only: on the logos it would tint them
  geom_segment(aes(x = game_no, xend = end + 2, y = wins, yend = spot),
    data = finish, colour = "grey60", linewidth = 0.3
  ) +
  geom_sdv_logos(aes(x = end + 5, y = spot, team = team), data = finish, sport = "nba", height = 0.08) +
  scale_colour_manual(values = chosen, guide = "none") +
  scale_x_continuous(limits = c(0, end + 8)) +
  labs(
    title = "The Pacific Division's 2025-26 race",
    x = "Game", y = "Wins",
    caption = "Data: hoopR::load_nba_team_box()"
  ) +
  theme_minimal()
```

A secondary is not always safe either: the Celtics' is white, which vanishes
on a white chart. Run the same check against the background.

## 5. Readable text on team-colored fills

Text on a team color needs the right ink: white on navy, black on gold. Pick
it from the fill's relative luminance (the WCAG formula): above 0.179, black
text contrasts more than white. MLB's 2026 regular season, wins printed inside
each bar:

```{r ink, fig.height = 7, fig.alt = "Horizontal bar chart of the 30 MLB teams' wins in the 2026 regular season, sorted, each bar filled with the team's primary color and the win total printed inside the end of the bar in black or white, whichever reads better on that color."}
ink <- function(fill) {
  rgb <- grDevices::col2rgb(fill) / 255
  linear <- ifelse(rgb <= 0.04045, rgb / 12.92, ((rgb + 0.055) / 1.055)^2.4)
  luminance <- colSums(linear * c(0.2126, 0.7152, 0.0722))
  ifelse(luminance > 0.179, "black", "white")
}

mlb_ids <- baseballr::mlb_teams(season = season, sport_ids = 1) |>
  as.data.frame() |>
  select(team_records_team_id = team_id, team_abbreviation)
mlb <- baseballr::mlb_standings(season = season, league_id = "103,104") |>
  as.data.frame() |>
  inner_join(mlb_ids, by = "team_records_team_id") |>
  transmute(
    team = clean_team_abbrs(team_abbreviation, sport = "mlb"),
    wins = team_records_wins,
    games = team_records_games_played
  ) |>
  arrange(wins, desc(team)) |>
  mutate(
    team_f = factor(team, levels = team),
    fill = unname(sdv_team_colors("mlb", team)),
    text = ink(fill)
  )
range(mlb$games)

ggplot(mlb, aes(x = wins, y = team_f)) +
  geom_col(aes(fill = fill), width = 0.75) +
  geom_text(aes(label = wins, colour = text), hjust = 1.3, size = 3.2, fontface = "bold") +
  scale_fill_identity() +
  scale_colour_identity() +
  scale_y_discrete(labels = NULL) +
  geom_sdv_logos(aes(x = -4, team = team), sport = "mlb", height = 0.028) +
  labs(
    title = "MLB wins, 2026 regular season",
    x = "Wins", y = NULL,
    caption = "Data: baseballr::mlb_standings() (MLB Stats API)"
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

In a gt table, `gt::data_color()` picks the ink for you (`autocolor_text`,
on by default), and `gt_theme_sdv_team()` does the same for its header (see
recipe 10).

## 6. A named palette for a manual scale

`sdv_color_palette(sport, teams)` returns a named vector of hex codes, the
shape `scale_colour_manual(values = )` wants. The SEC's 2025 season, one dot
per game by final margin, teams ordered by their average:

```{r named-palette, fig.height = 6, fig.alt = "Strip plot of the 16 SEC football teams' 2025 games, one dot per game at its final scoring margin, colored in the team's primary color, with a black tick at each team's average margin. Teams are sorted by average margin, best at the top."}
sec <- team_reference("cfb") |>
  filter(conference == "SEC") |>
  pull(team_abbr)

cfb <- cfbfastR::load_espn_cfb_schedules(2025) |> as.data.frame()
margins <- bind_rows(
  transmute(cfb, game_id, team = home_abbreviation, margin = home_score - away_score),
  transmute(cfb, game_id, team = away_abbreviation, margin = away_score - home_score)
) |>
  filter(team %in% sec)
order <- margins |>
  summarise(avg = mean(margin), .by = team) |>
  arrange(avg, desc(team))
margins <- mutate(margins, team = factor(team, levels = order$team))

ggplot(margins, aes(x = margin, y = team, colour = team)) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_point(size = 2.4, alpha = 0.8, position = position_jitter(width = 0, height = 0.15, seed = 1)) +
  geom_point(aes(x = avg), data = order, shape = "|", size = 6, colour = "black") +
  scale_colour_manual(values = sdv_color_palette("cfb", sec), guide = "none") +
  labs(
    title = "Every SEC game of 2025, by final margin",
    subtitle = "Black tick: each team's average margin",
    x = "Final margin (points)", y = NULL,
    caption = "Data: cfbfastR::load_espn_cfb_schedules()"
  ) +
  theme_minimal()
```

Several SEC primaries are near-identical crimsons and blues; across this many
teams, the axis labels carry the identity and the colors only group the dots.

## 7. Conference colors

Conferences are rows in sdvplotR's team data too, so `sdv_team_colors()` and
`geom_sdv_logos()` take a conference's name. Women's college basketball,
2025-26: how each conference's teams did against teams from other D-I
conferences, with each season's membership from wehoop's standings:

```{r conference-colors, fig.height = 5.5, fig.alt = "Horizontal bar chart of the twelve women's college basketball conferences with the best non-conference winning percentage in the 2025-26 regular season, each bar filled with the conference's color and ending in its logo."}
wbb_conf <- wehoop::load_wbb_standings(season) |>
  as.data.frame() |>
  distinct(team_id, conference = group_name)

wbb_games <- wehoop::load_wbb_team_box(season) |>
  as.data.frame() |>
  filter(season_type == 2) |>
  inner_join(wbb_conf, by = "team_id") |>
  inner_join(rename(wbb_conf, opponent_team_id = team_id, opp_conference = conference),
    by = "opponent_team_id"
  ) |>
  filter(conference != opp_conference)

nonconf <- wbb_games |>
  summarise(games = n(), win_pct = mean(team_winner), .by = conference) |>
  arrange(desc(win_pct), conference) |>
  slice_head(n = 12) |>
  mutate(
    label = sub(" Conference$", "", conference),
    label = factor(label, levels = rev(label)),
    fill = unname(sdv_team_colors("wbb", conference))
  )

ggplot(nonconf, aes(x = win_pct, y = label)) +
  geom_col(aes(fill = fill), width = 0.65) +
  geom_sdv_logos(aes(x = win_pct + 0.05, team = conference), sport = "wbb", height = 0.06) +
  scale_fill_identity() +
  scale_x_continuous(labels = scales::label_percent(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Non-conference record by conference, 2025-26",
    subtitle = "Women's college basketball, regular-season games against other D-I conferences",
    x = "Win %", y = NULL,
    caption = "Data: wehoop::load_wbb_team_box(), wehoop::load_wbb_standings()"
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(), plot.title.position = "plot")
```

`team_reference(sport, include_conferences = TRUE)` lists the conference rows
with their colors and logo URLs.

## 8. A dark background, dark logos and a matching table

Many logos have a dark version for dark backgrounds, in `team_reference()`'s
`logo_dark_url`; draw it with `geom_from_path()`. `theme_bg` holds the
background color of every gt theme, so a chart can sit on the same canvas as a
table in that theme. The WNBA on `gt_theme_midnight()`'s background:

```{r dark, fig.alt = "Scatter of the 15 WNBA teams' points scored against points allowed per game in the 2026 regular season on a near-black background, each team drawn with its dark-background logo, with light grey text and grid lines."}
bg <- theme_bg$bg[theme_bg$theme == "gt_theme_midnight"]
bg

dark <- wnba |>
  inner_join(select(team_reference("wnba"), team = team_abbr, logo_dark_url), by = "team")

ggplot(dark, aes(x = pf, y = pa)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey45", linetype = "dashed") +
  geom_from_path(aes(path = logo_dark_url), width = 0.065) +
  scale_x_continuous(expand = expansion(mult = 0.1)) +
  scale_y_continuous(expand = expansion(mult = 0.1)) +
  labs(
    title = "WNBA scoring, 2026 regular season",
    subtitle = "Points per game scored and allowed; teams below the dashed line outscored opponents",
    x = "Points scored per game", y = "Points allowed per game",
    caption = "Data: wehoop::load_wnba_team_box()"
  ) +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = bg, colour = NA),
    text = element_text(colour = "grey90"),
    axis.text = element_text(colour = "grey75"),
    panel.grid = element_line(colour = "grey25")
  )
```

```{r dark-table}
dark |>
  arrange(desc(diff), team) |>
  head(5) |>
  select(team, pf, pa, diff) |>
  gt() |>
  gt_sdv_logos(columns = team, sport = "wnba") |>
  fmt_number(columns = c(pf, pa, diff), decimals = 1) |>
  cols_label(team = "", pf = "Scored", pa = "Allowed", diff = "Margin") |>
  tab_header(title = "The WNBA's best margins, 2026") |>
  gt_theme_midnight()
```

`gt_sdv_logos()` draws the regular logos; in a reactable table,
`reactable_sdv_logos(variant = "dark")` picks the dark ones.

## 9. Choose a gt theme

`gt_theme_preview()` renders the same rows in several themes side by side, so
you can pick one by eye before styling the full table. The NHL's top scoring
teams in six themes:

```{r theme-preview}
nhl_top <- nhl_teams |>
  transmute(Team = team_abbrev, `GF/G` = round(gf, 2), `GA/G` = round(ga, 2)) |>
  head(5)

gt_theme_preview(
  nhl_top,
  themes = c(
    "gt_theme_sdv", "gt_theme_athletic", "gt_theme_kenpom",
    "gt_theme_savant", "gt_theme_broadsheet", "gt_theme_midnight"
  ),
  ncol = 3
)
```

Leave `themes` out to see every theme in the package; `file =` saves the grid
as one PNG (it needs a headless Chrome, through webshot2).

## 10. A table in a team's colors

`gt_theme_sdv_team()` dresses a table in one team's colors, with header text
in whichever ink reads on them. The team with the NBA's best point
differential, and its top scorers:

```{r team-table}
players <- hoopR::load_nba_player_box(season) |>
  as.data.frame() |>
  filter(game_id %in% nba_regular, !did_not_play)

best <- players |>
  distinct(game_id, team_abbreviation, team_score, opponent_team_score) |>
  summarise(diff = mean(team_score - opponent_team_score), .by = team_abbreviation) |>
  slice_max(diff, n = 1, with_ties = FALSE) |>
  pull(team_abbreviation)

players |>
  filter(team_abbreviation == best) |>
  summarise(
    games = n(), ppg = mean(points), rpg = mean(rebounds), apg = mean(assists),
    .by = c(athlete_id, athlete_display_name)
  ) |>
  filter(games >= 20) |>
  arrange(desc(ppg), athlete_id) |>
  head(6) |>
  gt() |>
  gt_sdv_headshots(columns = athlete_id, sport = "nba") |>
  fmt_number(columns = c(ppg, rpg, apg), decimals = 1) |>
  cols_label(
    athlete_id = "", athlete_display_name = "Player", games = "GP",
    ppg = "PTS", rpg = "REB", apg = "AST"
  ) |>
  tab_header(
    title = paste(team_reference("nba")$team_name[team_reference("nba")$team_abbr == best], "top scorers"),
    subtitle = "2025-26 regular season, per game, 20+ games"
  ) |>
  tab_source_note("Data: hoopR::load_nba_player_box()") |>
  gt_theme_sdv_team(team = best, sport = "nba")
```

`gt_theme_sdv()` is the house look in light or dark (`style = "dark"`), for
tables that belong to no one team.
