---
title: "Getting Started with sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

`sdvplotR` provides a unified interface for visualizing sports data across multiple leagues in the SportsDataverse ecosystem. This vignette walks you through the core features of the package.

```{r setup}
library(sdvplotR)
library(ggplot2)
```

## Valid Team Names

Before plotting, you need to know which team abbreviations are valid:

```{r valid-names}
# Get all NFL team abbreviations
nfl_teams <- valid_team_names("nfl")
head(nfl_teams)

# Get team names instead of abbreviations
nfl_team_names <- valid_team_names("nfl", type = "name")
head(nfl_team_names)

# Check supported sports
supported_sports()
```

## Plotting Team Logos

The `geom_sdv_logos()` function plots team logos at specified coordinates:

```{r logos-basic, eval=FALSE}
# Create sample data
df <- data.frame(
  x = rep(1:4, 4),
  y = sort(rep(1:4, 4), decreasing = TRUE),
  team = valid_team_names("nfl")[1:16]
)

# Plot logos
ggplot(df, aes(x = x, y = y)) +
  geom_sdv_logos(aes(team = team), sport = "nfl", width = 0.075) +
  theme_void()
```

## Using Team Colors

Map team names to their official colors using the color scales:

```{r colors, eval=FALSE}
df <- data.frame(
  team = valid_team_names("nfl")[1:8],
  value = runif(8, 0, 1)
)

ggplot(df, aes(x = team, y = value)) +
  geom_col(aes(color = team, fill = team), width = 0.5) +
  scale_color_sdv(sport = "nfl", type = "secondary") +
  scale_fill_sdv(sport = "nfl", alpha = 0.4) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Theme Elements

Replace axis labels with logos using theme elements:

```{r theme-elements, eval=FALSE}
df <- data.frame(
  team = valid_team_names("nfl")[1:8],
  value = runif(8, 0, 1)
)

ggplot(df, aes(x = team, y = value)) +
  geom_col(aes(fill = team), width = 0.5) +
  scale_fill_sdv(sport = "nfl", alpha = 0.4) +
  theme_minimal() +
  theme(axis.text.x = element_sdv_logo(sport = "nfl"))
```

## Player Headshots

Plot player headshots using their sport-specific IDs:

```{r headshots, eval=FALSE}
# NFL player GSIS IDs
df_players <- data.frame(
  x = 1:3,
  y = 3:1,
  player_id = c("00-0033873", "00-0026498", "00-0035228"),
  player_name = c("P. Mahomes", "M. Stafford", "K. Murray")
)

ggplot(df_players, aes(x = x, y = y)) +
  geom_sdv_headshots(aes(player_id = player_id), sport = "nfl", height = 0.2) +
  geom_label(aes(label = player_name), nudge_y = -0.35) +
  theme_void()
```

## gt Tables

Add logos to gt tables:

```{r gt-tables, eval=FALSE}
library(gt)

df_table <- data.frame(
  team = valid_team_names("nfl")[1:8],
  logo = valid_team_names("nfl")[1:8],
  wins = sample(1:16, 8)
)

df_table |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl")
```

## Team Tier Plots

Create tier plots for any sport:

```{r tier-plots, eval=FALSE}
team_abbr <- sample(valid_team_names("nfl"))
df_tiers <- data.frame(
  tier_no = sample(1:5, length(team_abbr), replace = TRUE),
  team = team_abbr
)

sdv_team_tiers(df_tiers, sport = "nfl")
```

## Working with Multiple Sports

The same functions work across all supported sports:

```{r multi-sport, eval=FALSE}
# NBA logos
nba_df <- data.frame(
  x = 1:5,
  y = 5:1,
  team = valid_team_names("nba")[1:5]
)

ggplot(nba_df, aes(x = x, y = y)) +
  geom_sdv_logos(aes(team = team), sport = "nba", width = 0.075) +
  theme_void()

# CFB logos
cfb_df <- data.frame(
  x = 1:5,
  y = 5:1,
  team = valid_team_names("cfb")[1:5]
)

ggplot(cfb_df, aes(x = x, y = y)) +
  geom_sdv_logos(aes(team = team), sport = "cfb", width = 0.075) +
  theme_void()
```

## Next Steps

- Explore the function documentation with `?geom_sdv_logos`
- Check out the [README](https://github.com/sportsdataverse/sdvplotR) for more examples
- Report issues on [GitHub](https://github.com/sportsdataverse/sdvplotR/issues)

## Companion Vignettes

Each sport has its own dedicated vignette showing integration with the
relevant SportsDataverse data package:

| Sport | Vignette | Companion Package(s) |
|-------|----------|----------------------|
| **NFL** | [NFL Visualizations](nfl-viz.html) | `nflfastR`, `nflreadr` |
| **CFB** | [CFB Visualizations](cfb-viz.html) | `cfbfastR`, `cfbseedR` |
| **NBA** | [NBA Visualizations](nba-viz.html) | `hoopR` |
| **WNBA** | [WNBA Visualizations](wnba-viz.html) | `wehoop` |
| **MLB** | [MLB Visualizations](mlb-viz.html) | `baseballr` |
| **NHL** | [NHL Visualizations](nhl-viz.html) | `fastRhockey` |
| **MBB** | [MBB Visualizations](mbb-viz.html) | `hoopR`, `cfbseedR` |
| **WBB** | [WBB Visualizations](wbb-viz.html) | `wehoop`, `cfbseedR` |

Cross-sport cookbook vignettes:

| Topic | Vignette |
|-------|----------|
| Social media posting patterns | [Social Posting](social-posting.html) |
| Leaderboard dashboards | [Leaderboard Dashboards](leaderboard-dashboards.html) |
| End-to-end workflows | [Workflows](workflows.html) |
