---
title: "Team Tables with the gt Toolkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Team Tables with the gt Toolkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

sdvplotR has two kinds of `gt` helpers, and this article uses both on one table.

- **The team layer** turns team keys into logos, headshots and team colors
  (`gt_sdv_logos()`, `gt_sdv_headshots()`, `gt_merge_stack_team_color()`). It
  resolves every abbreviation, alias and relocated franchise through
  `clean_team_abbrs()`.
- **The table toolkit**, ported from Andrew Weatherman's
  [gtUtils](https://github.com/andreweatherman/gtUtils), handles the editorial
  side: cut lines, colored ranks, captions, legends, grids of tables, themes, and
  saving images sized for posting.

Every one of these functions takes a `gt` table and returns one, so they chain
in any order. Having both in one package means one install, and one set of
conventions for how a sports table is built.

```{r setup, message = FALSE}
library(sdvplotR)
library(gt)
```

## The data

The final 2023 AFC standings, typed in so the article builds offline. The seven
playoff seeds come first in seed order, followed by the rest by record.

```{r data}
afc <- data.frame(
  team = c(
    "BAL", "BUF", "KC", "HOU", "CLE", "MIA", "PIT",
    "CIN", "IND", "JAX", "DEN", "LV", "NYJ", "TEN", "LAC", "NE"
  ),
  seed = c(1:7, rep(NA, 9)),
  w = c(13, 11, 11, 10, 11, 11, 10, 9, 9, 9, 8, 8, 7, 6, 5, 4)
)
afc$l <- 17 - afc$w
afc$pct <- afc$w / 17
```

## Step by step

Start with logos, labels and a box-score win percentage (`.765`, not `0.765`).

```{r base}
base <- function(df, conference) {
  gt(df) |>
    gt_sdv_logos(columns = "team", sport = "nfl", height = 24) |>
    cols_label(team = "", seed = "Seed", w = "W", l = "L", pct = "Win pct") |>
    sub_missing(columns = "seed", missing_text = "") |>
    fmt(columns = "pct", fns = function(x) sub("^0", "", sprintf("%.3f", x))) |>
    tab_header(paste(conference, "playoff picture"), "Final 2023 standings")
}
base(afc, "AFC")
```

A standings table is read around one question: who is in? `gt_cutline()` draws
the answer as a line after the seventh row, with a label, so readers don't have
to count seeds.

```{r cutline}
base(afc, "AFC") |>
  gt_cutline(after = 7, label = "Playoff line")
```

`gt_color_ranks()` colors a column by rank, which shows where the wins cluster
(four teams at 11 wins, three at 9) faster than reading the numbers;
`reverse = TRUE` puts the most wins at the green end. The cut line's label sits
in the row below the line and clears that row's cell colors, so with colored
cells, keep the line unlabeled and say what it means in the caption.
`gt_538_caption()` separates that note from the source line.
`gt_theme_sdv()` goes last so it styles everything added before it, and
`density` is passed through for the saved version below.

```{r full}
playoff_table <- function(df, conference, density = "comfortable") {
  base(df, conference) |>
    gt_cutline(after = 7) |>
    gt_color_ranks(columns = "w", reverse = TRUE) |>
    gt_538_caption(
      top_caption = "Above the line: playoff teams.",
      bottom_caption = "Data: nflverse"
    ) |>
    gt_theme_sdv(density = density)
}
playoff_table(afc, "AFC")
```

## Both conferences at once

`gt_grid()` lays several finished tables out under one shared title and source
line. Each table keeps its own theme and cut line; a note that applies to both
goes in the grid's `caption`, so each table's own caption stays short and the
tables stay narrow enough to sit side by side.

```{r grid}
nfc <- data.frame(
  team = c(
    "SF", "DAL", "DET", "TB", "PHI", "LA", "GB",
    "NO", "SEA", "ATL", "CHI", "MIN", "NYG", "ARI", "WAS", "CAR"
  ),
  seed = c(1:7, rep(NA, 9)),
  w = c(12, 12, 12, 9, 11, 10, 9, 9, 9, 7, 7, 7, 6, 4, 4, 2)
)
nfc$l <- 17 - nfc$w
nfc$pct <- nfc$w / 17

gt_grid(
  list(playoff_table(afc, "AFC"), playoff_table(nfc, "NFC")),
  ncol = 2,
  title = "The 2023 NFL playoff field",
  subtitle = "Final regular-season standings",
  caption = "Seeds 1-4 won their division; 5-7 are wild cards.",
  title_style = list(font = "Chivo", weight = 800)
)
```

On a narrow screen the grid scrolls sideways inside its own box, so the page
around it stays in place.

## Posting it

`gt_save_crop()` saves a single table, trimmed and padded; `gt_social_crop()`
pads it to a platform's aspect ratio. Use the theme's background from `theme_bg`
so the padding blends with the table. `gt_grid()` saves the whole grid when you
give it a `file`. Saving drives a headless Chrome through `webshot2`, so these
lines are not run here.

```{r save, eval = FALSE}
bg <- theme_bg$bg[theme_bg$theme == "gt_theme_sdv" & theme_bg$has_style == "light"]

playoff_table(afc, "AFC", density = "social") |>
  gt_social_crop("afc-playoffs.png", aspect_ratio = "4:5", bg = bg)

gt_grid(
  list(playoff_table(afc, "AFC"), playoff_table(nfc, "NFC")),
  title = "The 2023 NFL playoff field",
  file = "playoff-field.png"
)
```

## Where to go next

- [SportsDataverse Table Themes](sdv-table-themes.html) covers the house theme,
  the dark style and team colors.
- The gt table cookbooks go deeper on single tools:
  [styling headers, legends and captions](styling.html),
  [percentile bars and cut lines](delay_tables.html),
  [faceted tables](grid_tables.html),
  [tier lists](tier_list.html) and
  [saving and posting](saving_tables.html).
