---
title: "Leaderboard Dashboards with sdvplotR, gt, and Quarto"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Leaderboard Dashboards with sdvplotR, gt, and Quarto}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to create interactive leaderboard dashboards using
sdvplotR, gt tables, and Quarto. These patterns are perfect for tracking
standings, player stats, and team performance in real-time.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(gt)

# Get valid team abbreviations
nfl_teams <- valid_team_names("nfl")
nba_teams <- valid_team_names("nba")
cfb_teams <- valid_team_names("cfb")
```

## Basic Leaderboard with Logos

Create a simple leaderboard table with team logos:

```{r basic-leaderboard}
# Sample standings data
standings <- data.frame(
  team = c("KC", "BUF", "SF", "PHI", "DAL", "MIA", "CIN", "BAL"),
  wins = c(9, 8, 8, 7, 7, 6, 6, 6),
  losses = c(2, 3, 3, 4, 4, 5, 5, 5),
  ties = c(0, 0, 0, 0, 0, 0, 0, 0),
  pct = c(0.818, 0.727, 0.727, 0.636, 0.636, 0.545, 0.545, 0.545)
) |>
  mutate(
    logo = team,
    rank = row_number()
  ) |>
  select(rank, logo, team, wins, losses, ties, pct)

standings |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = "pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team = "Abbrev",
    wins = "W",
    losses = "L",
    ties = "T",
    pct = "Win %"
  ) |>
  tab_header(
    title = "NFL Standings",
    subtitle = paste("Week", ceiling(runif(1, 1, 18)))
  ) |>
  tab_footnote(
    footnote = "Data: nflfastR | Viz: sdvplotR",
    locations = cells_title(groups = "title")
  )
```

## Advanced Leaderboard with Team Colors

Create a leaderboard with team-colored rows:

```{r colored-leaderboard}
# Add team colors to standings
standings_colored <- standings |>
  mutate(
    primary_color = sdv_team_colors("nfl", team, type = "primary"),
    secondary_color = sdv_team_colors("nfl", team, type = "secondary")
  )

standings_colored |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = "pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team = "Abbrev",
    wins = "W",
    losses = "L",
    ties = "T",
    pct = "Win %"
  ) |>
  tab_header(
    title = "NFL Standings",
    subtitle = paste("Week", ceiling(runif(1, 1, 18)))
  ) |>
  data_color(
    columns = "pct",
    palette = c("lightblue", "darkblue")
  )
```

## Multi-Sport Leaderboard

Create leaderboards for multiple sports in one table:

```{r multi-sport-leaderboard}
# Sample multi-sport data
multi_sport_standings <- data.frame(
  sport = c(rep("NFL", 4), rep("NBA", 4), rep("MLB", 4)),
  team = c("KC", "BUF", "SF", "PHI",
           "BOS", "DEN", "MIL", "PHX",
           "LAD", "ATL", "HOU", "BAL"),
  wins = c(9, 8, 8, 7, 25, 23, 22, 20, 95, 90, 88, 85),
  losses = c(2, 3, 3, 4, 8, 10, 11, 13, 67, 72, 74, 77)
) |>
  mutate(
    pct = wins / (wins + losses),
    rank = row_number()
  ) |>
  select(sport, rank, team, wins, losses, pct)

multi_sport_standings |>
  gt() |>
  gt_sdv_logos(columns = "team", sport = "nfl", height = 30) |>
  fmt_number(columns = "pct", decimals = 3) |>
  cols_label(
    sport = "Sport",
    rank = "#",
    team = "Team",
    wins = "W",
    losses = "L",
    pct = "Win %"
  ) |>
  tab_header(
    title = "Multi-Sport Leaderboard",
    subtitle = "Top Teams Across Sports"
  )
```

## Player Leaderboard with Headshots

Create a player stats table with headshots:

```{r player-leaderboard}
# Sample player data
player_stats <- data.frame(
  player_id = c("00-0033873", "00-0026498", "00-0035228", "00-0033869"),
  player_name = c("P. Mahomes", "M. Stafford", "K. Murray", "J. Allen"),
  team = c("KC", "LAR", "ARI", "BUF"),
  pass_yds = c(4200, 3800, 3500, 4100),
  pass_td = c(32, 28, 25, 30),
  pass_int = c(8, 12, 10, 9)
) |>
  mutate(
    rank = row_number()
  ) |>
  select(rank, player_id, player_name, team, pass_yds, pass_td, pass_int)

player_stats |>
  gt() |>
  gt_sdv_headshots(columns = "player_id", sport = "nfl", height = 40) |>
  fmt_number(columns = c("pass_yds", "pass_td", "pass_int")) |>
  cols_label(
    rank = "#",
    player_id = "Player",
    player_name = "Name",
    team = "Team",
    pass_yds = "Yards",
    pass_td = "TD",
    pass_int = "INT"
  ) |>
  tab_header(
    title = "NFL Quarterback Leaderboard",
    subtitle = "Top Passers by Yards"
  )
```

## Quartile-Based Leaderboard

Create a leaderboard with quartile indicators:

```{r quartile-leaderboard}
# Add quartile information
standings_quartile <- standings |>
  mutate(
    quartile = case_when(
      pct >= 0.75 ~ "Q1 (Elite)",
      pct >= 0.50 ~ "Q2 (Good)",
      pct >= 0.25 ~ "Q3 (Average)",
      TRUE ~ "Q4 (Struggling)"
    )
  )

standings_quartile |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = "pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team = "Abbrev",
    wins = "W",
    losses = "L",
    ties = "T",
    pct = "Win %",
    quartile = "Quartile"
  ) |>
  tab_header(
    title = "NFL Standings by Quartile",
    subtitle = paste("Week", ceiling(runif(1, 1, 18)))
  ) |>
  tab_row_group(
    group = "Elite Teams",
    rows = pct >= 0.75
  ) |>
  tab_row_group(
    group = "Playoff Teams",
    rows = pct >= 0.50 & pct < 0.75
  )
```

## Quarto Dashboard Integration

Create a Quarto dashboard with sdvplotR visualizations:

```{r quarto-dashboard, eval=FALSE}
# In a Quarto document (.qmd), use this structure:

# ---
# title: "NFL Dashboard"
# format: dashboard
# ---

# ## Standings

# ```{r}
# standings |>
#   gt() |>
#   gt_sdv_logos(columns = "logo", sport = "nfl")
# ```

# ## Team Performance

# ```{r}
# ggplot(standings, aes(x = reorder(team, pct), y = pct)) +
#   geom_col(aes(fill = team), width = 0.7) +
#   scale_fill_sdv(sport = "nfl") +
#   theme_minimal()
# ```
```

## Real-Time Leaderboard Updates

Create a leaderboard that updates automatically:

```{r real-time-updates}
# Function to fetch and display standings
update_standings <- function(sport = "nfl", week = NULL) {
  # In a real implementation, this would fetch live data
  # For demonstration, we'll use sample data

  message(paste("Updating", toupper(sport), "standings..."))

  # Sample standings (replace with real data fetch)
  standings <- data.frame(
    team = sample(valid_team_names(sport), 10),
    wins = sample(0:16, 10),
    losses = sample(0:16, 10)
  ) |>
    mutate(
      pct = wins / (wins + losses),
      logo = team,
      rank = row_number()
    ) |>
    arrange(desc(pct)) |>
    head(10)

  standings |>
    gt() |>
    gt_sdv_logos(columns = "logo", sport = sport, height = 30) |>
    fmt_number(columns = "pct", decimals = 3)
}

# Update standings
update_standings("nfl")
```

## Conditional Formatting

Apply conditional formatting based on performance:

```{r conditional-formatting}
# Add performance indicators
standings_conditional <- standings |>
  mutate(
    performance = case_when(
      pct >= 0.75 ~ "🔥 Hot",
      pct >= 0.50 ~ "✅ Good",
      pct >= 0.25 ~ "⚠️ Average",
      TRUE ~ "❌ Struggling"
    )
  )

standings_conditional |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = "pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team = "Abbrev",
    wins = "W",
    losses = "L",
    ties = "T",
    pct = "Win %",
    performance = "Status"
  ) |>
  tab_header(
    title = "NFL Standings with Status",
    subtitle = paste("Week", ceiling(runif(1, 1, 18)))
  )
```

## Integrating with oddsapiR

Combine standings with betting odds:

```{r odds-integration}
# Sample odds data (replace with real oddsapiR data)
odds_data <- data.frame(
  team = c("KC", "BUF", "SF", "PHI"),
  spread = c(-7.5, -3.5, -6.5, -10.5),
  moneyline = c(-350, -180, -280, -550),
  total = c(48.5, 51.5, 47.5, 45.5)
)

# Merge with standings
standings_with_odds <- standings |>
  filter(team %in% odds_data$team) |>
  left_join(odds_data, by = "team") |>
  mutate(logo = team)

standings_with_odds |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = c("pct", "spread", "total"), decimals = 1) |>
  fmt_number(columns = "moneyline", decimals = 0) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team = "Abbrev",
    wins = "W",
    losses = "L",
    pct = "Win %",
    spread = "Spread",
    moneyline = "ML",
    total = "O/U"
  ) |>
  tab_header(
    title = "NFL Standings with Betting Odds",
    subtitle = "Combining Performance and Odds"
  )
```

## Best Practices for Dashboards

1. **Keep it Simple**: Focus on key metrics that matter to your audience

2. **Use Consistent Branding**: Apply the same colors, fonts, and logos across all visualizations

3. **Make it Interactive**: Use Quarto dashboards for interactive exploration

4. **Update Regularly**: Set up automated updates to keep data fresh

5. **Optimize for Mobile**: Ensure dashboards work on mobile devices

6. **Add Context**: Include captions, footnotes, and data sources

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Workflows](workflows.html)
