---
title: "MLB Leaderboard"
description: >
  The latest MLB season, rebuilt every week: division standings, run
  differential with logos, and the league leaders with headshots.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MLB Leaderboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 96
)
```

```{r season, include = FALSE}
library(sdvplotR)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(gt)

today <- Sys.Date()
current <- as.integer(format(today, "%Y")) # the season runs inside one calendar year

# A season's standings from the MLB Stats API, or NULL before opening day: the
# API answers with no rows for a season it has no standings for, or with every
# team on zero games
standings_for <- function(season) {
  table <- tryCatch(
    baseballr::mlb_standings(season = season, league_id = "103,104"),
    error = function(e) NULL
  )
  if (NROW(table) == 0 || max(table$team_records_games_played) == 0) NULL else table
}

season <- current
raw_standings <- standings_for(season)
if (is.null(raw_standings)) {
  season <- current - 1
  raw_standings <- standings_for(season)
}
stopifnot(!is.null(raw_standings))

# The Stats API's calendar for the current season: when its regular season and
# postseason end
calendar <- baseballr::mlb_seasons(sport_id = 1)
regular_end <- as.Date(calendar$regular_season_end_date[calendar$season_id == current])
post_end <- as.Date(calendar$post_season_end_date[calendar$season_id == current])
games_per_team <- round(median(raw_standings$team_records_games_played))
if (season < current) {
  status <- sprintf(
    "the offseason, so this is the final %d regular season; the %d season has no games yet.",
    season, current
  )
  through <- "final regular season"
} else if (length(regular_end) == 1 && today <= regular_end) {
  status <- sprintf("the %d season to date, about %d games per team.", season, games_per_team)
  through <- paste("through", games_per_team, "games")
} else if (length(post_end) == 1 && today <= post_end) {
  status <- sprintf("the final %d regular season; the postseason is under way.", season)
  through <- "final regular season"
} else {
  status <- sprintf("the offseason, so this is the final %d regular season.", season)
  through <- "final regular season"
}
```

**Updated `r format(today)`:** `r status`

This page is rebuilt every week with the site. It builds the division
standings, charts run differential with logos and lists the league leaders with
their headshots, for the latest MLB season with games: the season to date from
opening day to the end of September, the final regular season after that. Data:
the MLB Stats API through [baseballr](https://BillPetti.github.io/baseballr/);
no key needed.

## Picking the season

The season runs inside one calendar year, but before opening day the Stats API
standings for the new season come back empty, or with every team on zero games.
The helper turns both into "not published" and the page steps back one season.
The Stats API's season calendar (`mlb_seasons()`) says whether the regular
season is still being played, so the status line can say what the numbers
cover.

```{r season-code, ref.label = "season", eval = FALSE}
```

The standings name each club by its nickname; `mlb_teams()` adds its
abbreviation, full name and division, joined on the Stats API team id.

```{r standings-data}
clubs <- baseballr::mlb_teams(season = season, sport_ids = 1) |>
  select(team_id, abbreviation = team_abbreviation, team_name = team_full_name, division = division_name)
standings <- raw_standings |>
  transmute(
    team_id = team_records_team_id,
    rank = as.integer(team_records_division_rank),
    w = team_records_wins,
    l = team_records_losses,
    pct = as.numeric(team_records_winning_percentage),
    gb = team_records_division_games_back,
    rs = team_records_runs_scored,
    ra = team_records_runs_allowed,
    diff = team_records_run_differential,
    streak = team_records_streak_streak_code
  )
stopifnot(identical(class(standings$team_id), class(clubs$team_id)))
standings <- inner_join(standings, clubs, by = "team_id") |>
  arrange(division, rank)
nrow(standings)
```

## 1. Division standings

Six tables in one, grouped by division. `gt_sdv_logos()` turns the Stats API
abbreviations into logos (sdvplotR maps its `AZ` and `CWS` to ESPN's `ARI` and
`CHW`), and `gt_color_pills()` draws the run differential on a scale centred at
zero.

```{r standings-table}
reach <- max(abs(standings$diff))
standings |>
  mutate(logo = abbreviation) |>
  select(division, logo, team_name, w, l, pct, gb, rs, ra, diff, streak) |>
  gt(groupname_col = "division", id = "mlb-standings") |>
  tab_header(
    title = paste("MLB standings,", season),
    subtitle = paste("By division,", through)
  ) |>
  fmt_number(pct, decimals = 3) |>
  cols_label(
    logo = "", team_name = "Team", w = "W", l = "L", pct = "Pct", gb = "GB",
    rs = "RS", ra = "RA", diff = "Diff", streak = "Streak"
  ) |>
  tab_source_note("Data: MLB Stats API via baseballr | Viz: sdvplotR") |>
  gt_color_pills(
    diff,
    palette = c("#c84630", "#f7f7f7", "#2a7ab9"), domain = c(-reach, reach),
    digits = 0, pill_height = 22
  ) |>
  gt_sdv_logos(columns = logo, sport = "mlb", height = 24) |>
  gt_theme_broadsheet()
```

## 2. Run differential

Every club's run differential as a bar in its primary color, best at the top,
with its logo at the end of the bar.

```{r run-differential, fig.height = 8, fig.alt = "Horizontal bar chart of every MLB club's run differential, best at the top, each bar in the club's primary color with its logo at the end of the bar."}
rd <- standings |>
  arrange(diff, desc(team_name)) |> # the name breaks a tie, so an unchanged week renders the same
  mutate(
    label = sprintf("%s (%+d)", team_name, diff),
    label = factor(label, levels = label),
    end = diff + sign(diff) * 0.06 * reach
  )

ggplot(rd, aes(y = label)) +
  geom_col(aes(x = diff, fill = abbreviation), width = 0.72) +
  geom_vline(xintercept = 0, colour = "grey20", linewidth = 0.4) +
  geom_sdv_logos(aes(x = end, team = abbreviation), sport = "mlb", height = 0.028) +
  scale_fill_sdv(sport = "mlb", guide = "none") +
  scale_x_continuous(limits = c(-1.15, 1.15) * reach) +
  labs(
    title = paste("MLB run differential,", season, through),
    subtitle = "Runs scored minus runs allowed",
    x = "Run differential", y = NULL,
    caption = "Data: MLB Stats API via baseballr | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## 3. League leaders with headshots

The Stats API's leaderboards, one category at a time (`mlb_stats_leaders()`;
the rate stats list qualified players only). Its player ids are MLBAM ids, which
`gt_sdv_headshots(id_type = "league")` turns into MLB's own headshots. Ties
share a rank, so a category can show more than three players.

```{r leaders-data}
categories <- tibble::tribble(
  ~category, ~stat, ~group, ~digits,
  "Home runs", "homeRuns", "hitting", 0,
  "Batting average", "battingAverage", "hitting", 3,
  "OPS", "onBasePlusSlugging", "hitting", 3,
  "Stolen bases", "stolenBases", "hitting", 0,
  "ERA", "earnedRunAverage", "pitching", 2,
  "Strikeouts", "strikeouts", "pitching", 0
)
leaders <- bind_rows(lapply(seq_len(nrow(categories)), function(i) {
  baseballr::mlb_stats_leaders(
    leader_categories = categories$stat[i], season = season, sport_id = 1,
    leader_game_types = "R", limit = 3
  ) |>
    filter(stat_group == categories$group[i], rank <= 3) |>
    transmute(
      category = categories$category[i], rank, person_id, player = person_full_name, team_id,
      value = formatC(as.numeric(value), format = "f", digits = categories$digits[i])
    )
})) |>
  left_join(select(clubs, team_id, abbreviation), by = "team_id") |>
  mutate(
    value = sub("^0\\.", ".", value), # .312, not 0.312, as baseball writes it
    headshot = person_id
  )
head(leaders)
```

```{r leaders-table}
leaders |>
  select(category, rank, headshot, player, abbreviation, value) |>
  gt(groupname_col = "category", id = "mlb-leaders") |>
  tab_header(
    title = paste("MLB leaders,", season),
    subtitle = paste0("Top three, ", through)
  ) |>
  cols_label(rank = "", headshot = "", player = "Player", abbreviation = "", value = "") |>
  cols_align("right", value) |>
  tab_source_note("Data: MLB Stats API via baseballr | Viz: sdvplotR") |>
  gt_sdv_headshots(columns = headshot, sport = "mlb", id_type = "league", height = 36) |>
  gt_sdv_logos(columns = abbreviation, sport = "mlb", height = 22) |>
  gt_theme_savant()
```

## Related

- [MLB](mlb-viz.html) walks through baseballr and sdvplotR more broadly.
- [NHL](leaderboard-nhl.html) builds the same three views for hockey.
- [Social graphics, automated](automation-social.html) posts leaderboards like
  these on a schedule.
