---
title: "NFL Weekly Leaderboard"
description: >
  The NFL season to date, rebuilt every week: a power table by EPA per play,
  offense against defense, and a quarterback leaderboard with headshots.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NFL Weekly Leaderboard}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 96
)
```

```{r season, include = FALSE}
library(sdvplotR)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(gt)

today <- Sys.Date()
# The season starts in September, so before then the calendar points at last season
current <- as.integer(format(today, "%Y")) - (format(today, "%m") < "09")

# A season's regular-season plays, or NULL when nflverse has not published any:
# load_pbp() errors for a season past nflreadr::most_recent_season()
regular_season <- function(season) {
  pbp <- tryCatch(nflreadr::load_pbp(season), error = function(e) NULL)
  if (is.null(pbp)) {
    return(NULL)
  }
  pbp <- filter(pbp, season_type == "REG")
  if (nrow(pbp) == 0) NULL else pbp
}

season <- current
pbp <- regular_season(season)
if (is.null(pbp)) {
  season <- current - 1
  pbp <- regular_season(season)
}
stopifnot(!is.null(pbp))

schedule <- nflreadr::load_schedules(season)
week <- max(pbp$week)
n_games <- n_distinct(pbp$game_id)
unplayed <- sum(schedule$game_type == "REG" & is.na(schedule$result))
super_bowl <- any(schedule$game_type == "SB" & !is.na(schedule$result))
through <- if (unplayed > 0) paste("through week", week) else "final regular season"
status <- if (unplayed > 0) {
  sprintf("the %d season through week %d (%d games).", season, week, n_games)
} else if (!super_bowl) {
  sprintf("the final %d regular season; the playoffs are under way.", season)
} else {
  sprintf(
    "the offseason, so this is the final %d regular season. The %d season starts in September.",
    season, season + 1
  )
}
```

**Updated `r format(today)`:** `r status`

This page is rebuilt every week with the site. It finds the latest NFL season
with play-by-play, ranks every team by EPA (expected points added) per play,
plots offense against defense, and lists the most efficient quarterbacks with
their headshots: the season to date while games are being played, the last full
regular season in the offseason. Data: nflverse play-by-play and schedules,
read with [nflreadr](https://nflreadr.nflverse.com).

## Picking the season

The season starts in September, so before then the calendar points at last
season. nflverse publishes a season's play-by-play file with its first games;
until then `nflreadr::load_pbp()` errors, and the page steps back one season
instead of failing. The status line above comes from the schedule: games still
to play mean the season to date, a played Super Bowl means the offseason.

```{r season-code, ref.label = "season", eval = FALSE}
```

## 1. Power table

Every team's record and point differential from the schedule, and its EPA per
pass or run play on offense and allowed on defense, from the play-by-play. Net
EPA per play is offense minus defense; the trend compares each team's last
three games with its season. `gt_merge_stack_team_color()` stacks the record
under the nickname in team colors, `gt_delta()` computes the trend, and
`gt_sdv_logos()` turns the nflverse abbreviations into logos.

```{r power-data}
plays <- pbp |>
  filter(pass == 1 | rush == 1, !is.na(epa), !is.na(posteam))
offense <- plays |>
  group_by(game_id, week, team = posteam) |>
  summarise(off = sum(epa), off_n = n(), .groups = "drop")
defense <- plays |>
  group_by(game_id, team = defteam) |>
  summarise(dfn = sum(epa), dfn_n = n(), .groups = "drop")
# a fixed row order: every sum below adds up in the same order each week
per_game <- inner_join(offense, defense, by = c("game_id", "team")) |>
  arrange(team, week)

per_play <- function(games) {
  games |>
    group_by(team) |>
    summarise(
      off_epa = sum(off) / sum(off_n),
      def_epa = sum(dfn) / sum(dfn_n),
      .groups = "drop"
    ) |>
    mutate(net = off_epa - def_epa)
}
last3 <- per_game |>
  group_by(team) |>
  slice_tail(n = 3) |>
  ungroup() |>
  per_play() |>
  select(team, last3 = net)

games <- filter(schedule, game_type == "REG", game_id %in% pbp$game_id)
record <- bind_rows(
  transmute(games, team = home_team, pf = home_score, pa = away_score),
  transmute(games, team = away_team, pf = away_score, pa = home_score)
) |>
  group_by(team) |>
  summarise(
    w = sum(pf > pa), l = sum(pf < pa), t = sum(pf == pa), diff = sum(pf - pa),
    .groups = "drop"
  ) |>
  mutate(record = if_else(t > 0, paste(w, l, t, sep = "-"), paste(w, l, sep = "-")))

nicknames <- nflreadr::load_teams() |>
  select(team = team_abbr, name = team_nick)

power <- per_play(per_game) |>
  inner_join(last3, by = "team") |>
  inner_join(record, by = "team") |>
  inner_join(nicknames, by = "team") |>
  arrange(desc(net), team) |> # the team breaks a tie, so an unchanged week renders the same
  mutate(rank = row_number()) |>
  select(rank, team, name, record, diff, off_epa, def_epa, net, last3)
head(power)
```

```{r power-table}
good_bad <- c("#c84630", "#f7f7f7", "#2e8b57")
reach <- max(abs(c(power$off_epa, power$def_epa))) # one symmetric scale for both sides

power |>
  gt(id = "nfl-power") |> # a fixed id: gt draws a random one otherwise
  tab_header(
    title = paste("NFL power table,", season),
    subtitle = paste("Ranked by net EPA per play,", through)
  ) |>
  fmt_number(c(off_epa, def_epa, net, last3), decimals = 3, force_sign = TRUE) |>
  fmt_number(diff, decimals = 0, force_sign = TRUE) |>
  data_color(off_epa, palette = good_bad, domain = c(-reach, reach)) |>
  data_color(def_epa, palette = rev(good_bad), domain = c(-reach, reach)) |>
  data_color(net, palette = good_bad, domain = c(-1, 1) * max(abs(power$net))) |>
  tab_spanner("EPA per play", c(off_epa, def_epa, net, last3)) |>
  cols_label(
    rank = "", team = "", name = "Team", diff = "Pt diff",
    off_epa = "Offense", def_epa = "Defense", net = "Net", last3 = "Last 3"
  ) |>
  tab_source_note(paste(
    "Data: nflverse play-by-play via nflreadr. Pass and run plays;",
    "defense is EPA allowed (lower is better). Viz: sdvplotR"
  )) |>
  gt_merge_stack_team_color(name, record, team, sport = "nfl") |>
  gt_delta(net, last3, column_label = "Trend", decimals = 3, arrows = TRUE) |>
  gt_sdv_logos(columns = team, sport = "nfl", height = 26) |>
  gt_theme_athletic()
```

`gt_save_crop(table, "nfl-power.png")` renders the same table to a trimmed
PNG, ready to post; see [Saving and Posting Tables](saving_tables.html).

## 2. Offense against defense

The same EPA per play as a scatter, one logo per team. The y axis is reversed so
the better defenses sit higher: the top-right corner is where good teams live.
The faint diagonals are lines of equal net EPA per play, 0.1 apart.

```{r off-def, fig.alt = "Every NFL team's logo placed by offensive EPA per play (horizontal) and defensive EPA allowed per play (vertical, reversed so better defenses sit higher), with dashed league-average lines and faint diagonals of equal net EPA."}
ggplot(power, aes(x = off_epa, y = def_epa)) +
  geom_abline(slope = 1, intercept = seq(-0.6, 0.6, by = 0.1), colour = "grey88", linewidth = 0.3) +
  geom_hline(yintercept = mean(power$def_epa), linetype = "dashed", colour = "grey55") +
  geom_vline(xintercept = mean(power$off_epa), linetype = "dashed", colour = "grey55") +
  geom_sdv_logos(aes(team = team), sport = "nfl", width = 0.06, alpha = 0.9) +
  # logos do not widen the limits, so leave room for the outermost ones
  scale_x_continuous(expand = expansion(mult = 0.07)) +
  scale_y_reverse(expand = expansion(mult = 0.07)) +
  labs(
    title = paste("NFL offense and defense,", season),
    subtitle = paste("EPA per pass or run play,", through),
    x = "Offense: EPA per play",
    y = "Defense: EPA allowed per play (reversed)",
    caption = "Data: nflverse via nflreadr | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

## 3. Quarterback leaderboard

EPA per dropback (passes, sacks and scrambles) for quarterbacks with at least 15
dropbacks per team game, with completion percentage over expected (CPOE). The
play-by-play's `id` column is the passer's GSIS id, which is what
`gt_sdv_headshots()` takes for the NFL; the season's rosters, keyed by the same
id, give the full names.

```{r qb-data}
team_games <- max(count(per_game, team)$n)
qbs <- pbp |>
  filter(qb_dropback == 1, !is.na(qb_epa), !is.na(id)) |>
  arrange(game_id, play_id) |>
  group_by(id) |>
  summarise(
    name = last(name),
    team = last(posteam),
    dropbacks = n(),
    epa = mean(qb_epa),
    cpoe = mean(cpoe, na.rm = TRUE),
    success = mean(success, na.rm = TRUE),
    td = sum(pass_touchdown, na.rm = TRUE),
    int = sum(interception, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(dropbacks >= 15 * team_games) |>
  arrange(desc(epa), id) |>
  mutate(rank = row_number(), headshot = id)
full_names <- nflreadr::load_rosters(season) |>
  distinct(id = gsis_id, full_name) |>
  filter(!is.na(id), !duplicated(id))
qbs <- qbs |>
  left_join(full_names, by = "id") |>
  mutate(name = coalesce(full_name, name))
nrow(qbs)
```

```{r qb-table}
qbs |>
  slice_head(n = 16) |>
  select(rank, headshot, name, team, dropbacks, epa, cpoe, success, td, int) |>
  gt(id = "nfl-qbs") |>
  tab_header(
    title = paste("NFL quarterbacks by EPA per dropback,", season),
    subtitle = paste0("Minimum ", 15 * team_games, " dropbacks, ", through)
  ) |>
  fmt_number(epa, decimals = 3, force_sign = TRUE) |>
  fmt_number(cpoe, decimals = 1, force_sign = TRUE) |>
  fmt_percent(success, decimals = 1) |>
  cols_label(
    rank = "", headshot = "", name = "Quarterback", team = "", dropbacks = "Dropbacks",
    epa = "EPA/db", cpoe = "CPOE", success = "Success", td = "TD", int = "INT"
  ) |>
  tab_source_note("Data: nflverse play-by-play via nflreadr | Viz: sdvplotR") |>
  gt_color_pills(
    epa,
    palette = good_bad, domain = c(-1, 1) * max(abs(qbs$epa)), digits = 3, pill_height = 22
  ) |>
  gt_sdv_headshots(columns = headshot, sport = "nfl", height = 34) |>
  gt_sdv_logos(columns = team, sport = "nfl", height = 22) |>
  gt_theme_sdv()
```

The same quarterbacks on the two axes, each drawn as his headshot. Up and to the
right is efficient and accurate.

```{r qb-chart, fig.height = 7, fig.alt = "Every qualifying NFL quarterback's headshot placed by completion percentage over expected (horizontal) and EPA per dropback (vertical), with dashed lines at the averages of the group."}
ggplot(qbs, aes(x = cpoe, y = epa)) +
  geom_hline(yintercept = mean(qbs$epa), linetype = "dashed", colour = "grey55") +
  geom_vline(xintercept = mean(qbs$cpoe), linetype = "dashed", colour = "grey55") +
  geom_sdv_headshots(aes(player_id = id), sport = "nfl", height = 0.08) +
  ggrepel::geom_text_repel(
    aes(label = sub("^\\S+\\s+", "", name)), # the last name
    size = 3, colour = "grey25", point.size = 13, box.padding = 0.35,
    min.segment.length = 0, segment.colour = "grey70", max.overlaps = Inf,
    seed = 2026 # a seed: the same layout each week
  ) +
  labs(
    title = paste("NFL quarterbacks,", season),
    subtitle = paste0("EPA per dropback against CPOE, minimum ", 15 * team_games, " dropbacks, ", through),
    x = "Completion percentage over expected",
    y = "EPA per dropback",
    caption = "Data: nflverse via nflreadr | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

## Related

- [Leaderboard Dashboards](leaderboard-dashboards.html) for layout patterns.
- [College football weekly](leaderboard-cfb-weekly.html), and the
  [NBA](leaderboard-nba.html), [WNBA](leaderboard-wnba.html),
  [MLB](leaderboard-mlb.html) and [NHL](leaderboard-nhl.html) leaderboards.
- [Social graphics, automated](automation-social.html) posts tables like these on
  a schedule.
