---
title: "MBB Visualizations with hoopR, cfbseedR, and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MBB Visualizations with hoopR, cfbseedR, and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to create rich MBB (Men's College Basketball)
visualizations by combining [hoopR](https://hoopR.sportsdataverse.org) for player
and team data, [cfbseedR](https://cfbseedR.sportsdataverse.org) for tournament
brackets, and [sdvplotR](https://sdvplotr.sportsdataverse.org) for team logos,
headshots, colors, and gt tables.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(hoopR)
library(cfbseedR)
library(dplyr)
library(gt)

# Get valid MBB team abbreviations
mbb_teams <- valid_team_names("mbb")
length(mbb_teams)  # ~350+ D1 teams
head(mbb_teams)
```

## Loading MBB Data

Use `hoopR` to load MBB player and team statistics:

```{r load-data}
# Load MBB team stats
team_stats <- hoopR::load_mbb_team_stats(
  seasons = hoopR::most_recent_mbb_season(),
  level = "team"
)

# Load MBB player stats
player_stats <- hoopR::load_mbb_player_stats(
  seasons = hoopR::most_recent_mbb_season(),
  season_type = "Regular Season"
)
```

## MBB Team Performance

Visualize team performance with team logos:

```{r team-performance}
# Calculate team metrics
team_perf <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    avg_points = mean(points, na.rm = TRUE),
    avg_rebounds = mean(total_rebounds, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10)

ggplot(team_perf, aes(x = avg_points, y = avg_rebounds)) +
  geom_sdv_logos(
    aes(team = team_abbreviation),
    sport = "mbb",
    width = 0.075
  ) +
  labs(
    title = "MBB Team Performance",
    subtitle = paste("Season", hoopR::most_recent_mbb_season()),
    x = "Average Points per Game",
    y = "Average Rebounds per Game",
    caption = "Data: hoopR | Viz: sdvplotR"
  ) +
  theme_minimal()
```

## MBB Team Colors

Use team colors to visualize win percentages:

```{r team-colors}
# Calculate win percentage
team_wins <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    wins = sum(win == 1, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10) |>
  mutate(win_pct = wins / games) |>
  arrange(desc(win_pct)) |>
  head(25)

ggplot(team_wins, aes(x = reorder(team_abbreviation, win_pct), y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.7) +
  scale_fill_sdv(sport = "mbb", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 25 MBB Teams by Win Percentage",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Player Headshots

Visualize top performers with player headshots:

```{r player-headshots}
# Top scorers
top_scorers <- player_stats |>
  filter(!is.na(athlete_id)) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    avg_points = mean(points, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 15) |>
  arrange(desc(avg_points)) |>
  head(8)

ggplot(top_scorers, aes(x = games, y = avg_points)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "mbb",
    height = 0.15
  ) +
  geom_label(
    aes(label = athlete_display_name),
    nudge_y = -1.5,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "Top 8 MBB Scorers",
    subtitle = paste("Season", hoopR::most_recent_mbb_season()),
    x = "Games Played",
    y = "Average Points per Game"
  ) +
  theme_minimal()
```

## March Madness Bracket with cfbseedR

Use `cfbseedR` to simulate NCAA Tournament brackets:

```{r march-madness}
# Get current MBB rankings
# rankings <- cfbseedR::cfb_rankings(
#   year = hoopR::most_recent_mbb_season(),
#   sport = "mbb"
# )

# Simulate tournament bracket
# bracket_sim <- cfbseedR::cfb_simulate(
#   rankings = rankings,
#   n_sims = 10000
# )

# For demonstration, create sample bracket data
bracket_data <- data.frame(
  seed = 1:16,
  team = c("HOU", "UCLA", "KANSAS", "PURDUE", "GONZAGA", "BAYLOR",
           "ARIZONA", "DUKE", "CREIGHTON", "MARQUETTE", "TEXAS",
           "AUBURN", "MICH ST", "TENN", "SDSU", "TCU")
)

ggplot(bracket_data, aes(x = seed, y = 1)) +
  geom_sdv_logos(
    aes(team = team),
    sport = "mbb",
    width = 0.075
  ) +
  scale_x_continuous(breaks = 1:16) +
  labs(
    title = "NCAA Tournament Seeds",
    subtitle = paste("Season", hoopR::most_recent_mbb_season()),
    x = "Seed",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank()
  )
```

## MBB Team Tiers

Create a tier plot ranking MBB teams:

```{r team-tiers}
# Top 25 teams by win percentage
top_25 <- team_wins |>
  mutate(
    tier_no = case_when(
      win_pct > 0.85 ~ 1,
      win_pct > 0.75 ~ 2,
      win_pct > 0.65 ~ 3,
      win_pct > 0.55 ~ 4,
      TRUE ~ 5
    )
  ) |>
  select(tier_no, team = team_abbreviation)

sdv_team_tiers(
  top_25,
  sport = "mbb",
  title = "MBB Power Rankings",
  subtitle = paste("Week", ceiling(runif(1, 1, 20)), "of", hoopR::most_recent_mbb_season()),
  tier_desc = c(
    "1" = "Elite",
    "2" = "Championship Contenders",
    "3" = "Top 25",
    "4" = "Bubble Teams",
    "5" = "Rebuilding"
  ),
  presort = TRUE
)
```

## MBB Conference Map

Visualize teams grouped by conference:

```{r conference-map}
# Get team info with conferences
mbb_info <- hoopR::mbb_teams() |>
  select(team_abbreviation, team_name, conference_name)

conference_map <- mbb_info |>
  filter(!is.na(conference_name)) |>
  mutate(
    conference_num = as.numeric(factor(conference_name))
  ) |>
  arrange(conference_num, team_name) |>
  mutate(
    team_rank = row_number(),
    .by = conference_num
  ) |>
  filter(conference_num <= 10)  # Top 10 conferences for readability

ggplot(conference_map, aes(x = conference_num, y = team_rank)) +
  geom_sdv_logos(
    aes(team = team_abbreviation),
    sport = "mbb",
    width = 0.05
  ) +
  scale_x_continuous(
    breaks = 1:length(unique(conference_map$conference_name)),
    labels = unique(conference_map$conference_name)
  ) +
  labs(
    title = "MBB Teams by Conference (Top 10)",
    x = "Conference",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    panel.grid = element_blank()
  )
```

## MBB Standings Table with Logos

Create a gt table with team logos:

```{r gt-table}
standings_table <- team_wins |>
  head(20) |>
  mutate(
    logo = team_abbreviation,
    rank = row_number()
  ) |>
  select(rank, logo, team_abbreviation, wins, games, win_pct)

standings_table |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "mbb", height = 30) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team_abbreviation = "Abbrev",
    wins = "Wins",
    games = "Games",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "MBB Top 20",
    subtitle = paste("Season", hoopR::most_recent_mbb_season())
  )
```

## Axis Labels with Logos

Replace axis labels with team logos:

```{r axis-logos}
top_8 <- team_wins |>
  head(8) |>
  mutate(team_abbreviation = factor(team_abbreviation, levels = team_abbreviation))

ggplot(top_8, aes(x = team_abbreviation, y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.6) +
  scale_fill_sdv(sport = "mbb", alpha = 0.7) +
  scale_x_sdv(sport = "mbb") +
  theme_minimal() +
  theme_x_sdv() +
  labs(
    title = "Top 8 MBB Teams by Win %",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme(legend.position = "none")
```

## Next Steps

- Explore [hoopR documentation](https://hoopR.sportsdataverse.org/)
- Try [cfbseedR](https://cfbseedR.sportsdataverse.org) for tournament simulations
- Combine with [oddsapiR](https://oddsapiR.sportsdataverse.org) for betting lines

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
