---
title: "MLB Visualizations with baseballr and sdvplotR"
description: >
  Ten worked examples on the 2026 MLB regular season: standings, run
  differential, Statcast, home run leaders, a spray chart, the division races,
  franchise moves and a tier list.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MLB Visualizations with baseballr and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Ten charts and tables from the 2026 MLB regular season: run differential in team
colors, division standings, Pythagorean luck, team and player Statcast numbers,
home run leaders with their headshots, a spray chart on the field, the division
races week by week, the franchises that moved, and the postseason field as a
tier list. The data comes from two public sources through
[baseballr](https://baseballr.sportsdataverse.org): the MLB Stats API and
Baseball Savant. Neither needs a key.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(gt)
```

## Load the season

Standings, every regular-season result, the home run leaders and the name and
abbreviation of nine relocated franchises in every season since 1901, from the
MLB Stats API:

```{r load-data, eval = FALSE}
season <- 2026 # the regular season ended on September 27

teams <- baseballr::mlb_teams(season = season, sport_ids = 1) |>
  select(team_id, abbreviation = team_abbreviation, team_name = team_full_name, division = division_name)

standings <- baseballr::mlb_standings(season = season, league_id = "103,104") |>
  transmute(
    team_id = team_records_team_id,
    rank = as.integer(team_records_division_rank),
    w = team_records_wins,
    l = team_records_losses,
    gb = team_records_games_back,
    wc_gb = team_records_wild_card_games_back,
    rs = team_records_runs_scored,
    ra = team_records_runs_allowed,
    diff = team_records_run_differential,
    strk = team_records_streak_streak_code,
    clinch = team_records_clinch_indicator
  ) |>
  inner_join(teams, by = "team_id")

games <- baseballr::mlb_schedule(season = season, level_ids = "1") |>
  filter(game_type == "R", status_detailed_state %in% c("Final", "Completed Early")) |>
  transmute(
    game_pk,
    date = as.Date(official_date),
    home_id = teams_home_team_id,
    away_id = teams_away_team_id,
    home_score = teams_home_score,
    away_score = teams_away_score
  )

hitters <- baseballr::mlb_stats(
  stat_type = "season", stat_group = "hitting", season = season, player_pool = "All",
  sort_stat = "homeRuns", order = "desc", limit = 40
) |>
  select(player_id, player = player_full_name, team_id, games_played, home_runs)

# Athletics, Braves, Dodgers, Giants, Orioles, Twins, Rangers, Nationals, Brewers
moved <- c(133, 144, 119, 137, 110, 142, 140, 120, 158)
identities <- lapply(1901:season, \(s) baseballr::mlb_teams(season = s, sport_ids = 1)) |>
  bind_rows() |>
  filter(team_id %in% moved) |>
  select(team_id, season, name = team_full_name, abbreviation = team_abbreviation)
```

Team expected statistics, every qualified hitter's quality of contact and the
home run leaders' batted balls, from Baseball Savant:

```{r load-savant, eval = FALSE}
team_xstats <- baseballr::statcast_leaderboards(
  leaderboard = "expected_statistics", year = season, player_type = "batter-team"
) |>
  select(team_id, woba, est_woba)

contact <- baseballr::statcast_leaderboards(
  leaderboard = "exit_velocity_barrels", year = season, player_type = "batter"
) |>
  select(player_id, name = `last_name, first_name`, attempts, avg_hit_speed, brl_percent)

# every home run the season's leader(s) hit, with Statcast's hit coordinates
leaders <- hitters$player_id[hitters$home_runs == max(hitters$home_runs)]
homers <- lapply(leaders, \(id) {
  baseballr::statcast_search_batters(paste0(season, "-03-01"), paste0(season, "-10-01"), batterid = id)
}) |>
  bind_rows() |>
  filter(events == "home_run", game_type == "R") |>
  select(batter, game_date, hc_x, hc_y, hit_distance_sc, launch_speed)
```

```{r load-data-fixture, include = FALSE}
fx <- readRDS("fixtures/mlb-viz/mlb_stats_api.rds")
season <- fx$season
standings <- fx$standings
games <- fx$games
hitters <- fx$hitters
identities <- fx$identities
sv <- readRDS("fixtures/mlb-viz/savant.rds")
team_xstats <- sv$team_xstats
contact <- sv$contact
homers <- sv$homers
```

The output below uses a snapshot of these calls taken on
`r format(fx$taken, "%B %d, %Y")` (`r fx$package`), because neither API is
called when this site is built; `data-raw/article_fixtures.R` takes it with the
same code.

The Stats API and Savant write a few team abbreviations differently from ESPN,
whose abbreviations sdvplotR keys its team colors on. `clean_team_abbrs()`
converts them once, at the start, so every later join and color scale agrees:

```{r clean-abbrs}
standings <- standings |>
  mutate(team = clean_team_abbrs(abbreviation, sport = "mlb"), pct = w / (w + l))

standings |>
  filter(team != abbreviation) |>
  select(team_name, abbreviation, team)

stats_api <- "Data: MLB Stats API via baseballr | Viz: sdvplotR"
savant <- "Data: Baseball Savant via baseballr | Viz: sdvplotR"
# baseball writes rates without the leading zero: .412
rate <- function(x) sub("^(-?)0\\.", "\\1.", sprintf("%.3f", x))
```

## 1. Run differential in team colors

`scale_fill_sdv()` fills each bar with its team's color, and `scale_x_sdv()`
with `theme_x_sdv()` puts the logos on the axis in place of the abbreviations.

```{r team-performance, fig.width = 9, fig.height = 5, fig.alt = "Bar chart of the 30 MLB teams' 2026 regular-season run differential, sorted from best to worst, each bar in its team's color with the team logo under it on the horizontal axis."}
run_diff <- standings |>
  arrange(desc(diff), team) |>
  mutate(team = factor(team, levels = team))

ggplot(run_diff, aes(team, diff, fill = team)) +
  geom_col(width = 0.75) +
  geom_hline(yintercept = 0, colour = "grey20", linewidth = 0.4) +
  scale_fill_sdv(sport = "mlb") +
  scale_x_sdv(sport = "mlb", size = 20) +
  labs(
    title = paste(season, "MLB run differential"),
    subtitle = "Regular season, runs scored minus runs allowed",
    x = NULL,
    y = "Run differential",
    caption = stats_api
  ) +
  theme_minimal(base_size = 12) +
  theme_x_sdv() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 2. Division standings table

A gt table grouped by division: `gt_sdv_logos()` turns the abbreviation column
into logos and `gt_theme_savant()` gives it the Baseball Savant look. The run
differential fill goes on after the theme, so the theme's styling cannot
replace it.

```{r standings-table}
divisions <- paste(
  rep(c("American League", "National League"), each = 3),
  c("East", "Central", "West")
)
lim <- max(abs(standings$diff))

standings |>
  mutate(division = factor(division, levels = divisions), logo = team) |>
  arrange(division, rank) |>
  select(division, logo, team_name, w, l, pct, gb, rs, ra, diff, strk) |>
  gt(groupname_col = "division") |>
  gt_sdv_logos(columns = "logo", sport = "mlb", height = 24) |>
  fmt(columns = pct, fns = rate) |>
  cols_label(
    logo = "", team_name = "Team", w = "W", l = "L", pct = "Pct", gb = "GB",
    rs = "RS", ra = "RA", diff = "Diff", strk = "Strk"
  ) |>
  tab_header(
    title = paste(season, "MLB standings"),
    subtitle = "Final regular-season standings by division"
  ) |>
  tab_source_note(stats_api) |>
  gt_theme_savant() |>
  data_color(columns = diff, palette = c("#c84630", "white", "#2a7ab9"), domain = c(-lim, lim))
```

## 3. Pythagorean wins: who beat their run differential

Expected wins from runs scored and allowed (Bill James's formula with the 1.83
exponent) against actual wins. Teams above the dashed line won more games than
their runs suggest.

```{r pythag, fig.width = 7, fig.height = 7, fig.alt = "Scatter plot of the 30 MLB teams in 2026, each drawn as its logo, placed by Pythagorean expected wins (horizontal) and actual wins (vertical), with a dashed line where the two are equal."}
pyth <- standings |>
  mutate(xw = rs^1.83 / (rs^1.83 + ra^1.83) * (w + l), luck = w - xw)
lims <- range(pyth$xw, pyth$w) + c(-3, 3)

ggplot(pyth, aes(xw, w)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  annotate(
    "text", x = lims[1] + 1, y = lims[2] - 1, hjust = 0, vjust = 1, size = 3.6,
    colour = "grey35", label = "Won more than\ntheir runs suggest"
  ) +
  annotate(
    "text", x = lims[2] - 1, y = lims[1] + 1, hjust = 1, vjust = 0, size = 3.6,
    colour = "grey35", label = "Won fewer"
  ) +
  geom_sdv_logos(aes(team = team), sport = "mlb", width = 0.065) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(
    title = paste(season, "MLB: actual against expected wins"),
    subtitle = "Expected wins from runs scored and allowed (Pythagorean, exponent 1.83)",
    x = "Expected wins",
    y = "Actual wins",
    caption = stats_api
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r pythag-extremes, include = FALSE}
lucky <- slice_max(pyth, luck, n = 1, with_ties = FALSE)
unlucky <- slice_min(pyth, luck, n = 1, with_ties = FALSE)
```

The `r lucky$team_name` won `r round(lucky$luck, 1)` games more than their run
differential predicts; the `r unlucky$team_name` won
`r round(-unlucky$luck, 1)` fewer.

## 4. Statcast: team wOBA against expected wOBA

Baseball Savant's expected statistics at the team level. Expected wOBA (xwOBA)
grades every batted ball by its exit velocity and launch angle, so a team above
the line got more from its contact than the contact's quality predicts.
Savant's `team_id` column holds abbreviations, which `clean_team_abbrs()`
converts like any other.

```{r statcast-teams, fig.width = 7, fig.height = 7, fig.alt = "Scatter plot of the 30 MLB team offenses in 2026, each drawn as its logo, placed by expected wOBA (horizontal) and actual wOBA (vertical), with a dashed line where the two are equal."}
xstats <- team_xstats |>
  mutate(team = clean_team_abbrs(team_id, sport = "mlb"))
lims <- range(xstats$woba, xstats$est_woba) + c(-0.004, 0.004)

ggplot(xstats, aes(est_woba, woba)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  geom_sdv_logos(aes(team = team), sport = "mlb", width = 0.055) +
  scale_x_continuous(labels = rate) +
  scale_y_continuous(labels = rate) +
  coord_equal(xlim = lims, ylim = lims) +
  labs(
    title = paste(season, "team offense: wOBA against expected wOBA"),
    subtitle = "Above the line: more production than the quality of contact predicts",
    x = "Expected wOBA (xwOBA)",
    y = "wOBA",
    caption = savant
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 5. Home run leaders with headshots

The Stats API and Savant identify players by their MLBAM ID, which is why the
headshots pass `id_type = "league"`: it draws MLBAM IDs from MLB's own image
CDN. Leave `id_type` unset for ESPN athlete IDs, such as those in baseballr's
`espn_mlb_*()` data. Both are plain numbers, so the wrong setting draws someone
else or nothing. Everyone tied for tenth makes the list.

```{r hr-leaders, fig.width = 8, fig.height = 6.5, fig.alt = "Horizontal bar chart of the 2026 MLB home run leaders, each bar in the player's team color, with the team logo at the left, the player's headshot at the end of the bar and the home run total beside it."}
hr_leaders <- hitters |>
  filter(min_rank(desc(home_runs)) <= 10) |>
  inner_join(select(standings, team_id, team), by = "team_id") |>
  arrange(home_runs, desc(player)) |>
  mutate(player = factor(player, levels = player))

ggplot(hr_leaders, aes(home_runs, player)) +
  geom_col(aes(fill = team), width = 0.72) +
  geom_sdv_logos(aes(x = -3, team = team), sport = "mlb", height = 0.065) +
  geom_sdv_headshots(
    aes(x = home_runs + 3.2, player_id = player_id),
    sport = "mlb", id_type = "league", height = 0.085
  ) +
  geom_text(aes(x = home_runs + 6.6, label = home_runs), hjust = 0, fontface = "bold", size = 4) +
  scale_fill_sdv(sport = "mlb") +
  scale_x_continuous(limits = c(-6, max(hr_leaders$home_runs) + 10), expand = expansion(0)) +
  labs(
    title = paste(season, "MLB home run leaders"),
    subtitle = "Regular season",
    x = "Home runs",
    y = NULL,
    caption = stats_api
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 6. A spray chart on the field

`sdv_surface("mlb")` draws a regulation infield with 'sportyR', and the data
goes on top as ordinary layers. Statcast's hit coordinates become feet from
home plate with the usual transform (baseballr's `mlbam_xy_transformation()`
uses the same one); `xlims` and `ylims` pass through to 'sportyR' to show the
outfield. The two co-leaders get a panel each.

```{r spray-chart, fig.width = 9, fig.height = 5.5, fig.alt = "Two baseball fields side by side, one per 2026 MLB home run co-leader, with a dot for every home run the player hit placed where it landed, colored by distance, and the player's team logo in the corner."}
spray <- homers |>
  mutate(x = 2.5 * (hc_x - 125.42), y = 2.5 * (198.27 - hc_y)) |>
  filter(!is.na(x)) |>
  inner_join(select(hitters, batter = player_id, player, team_id, home_runs), by = "batter") |>
  inner_join(select(standings, team_id, team), by = "team_id") |>
  mutate(player = paste0(player, " (", home_runs, " HR)"))
arc <- data.frame(t = seq(-pi / 4, pi / 4, length.out = 50)) |>
  mutate(x = 400 * sin(t), y = 400 * cos(t))

sdv_surface("mlb", xlims = c(-330, 330), ylims = c(-20, 480)) +
  geom_path(aes(x, y), data = arc, linetype = "dashed", colour = "white", alpha = 0.7) +
  annotate("text", x = 0, y = 390, label = "400 ft", colour = "white", size = 3, vjust = 1) +
  geom_point(
    aes(x, y, fill = hit_distance_sc),
    data = spray, shape = 21, size = 2.6, colour = "grey15", stroke = 0.3
  ) +
  geom_sdv_logos(
    aes(x = -265, y = 430, team = team),
    data = distinct(spray, player, team), sport = "mlb", width = 0.13
  ) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Distance (ft)") +
  facet_wrap(~player) +
  labs(
    title = paste("Every home run by the", season, "MLB co-leaders"),
    subtitle = "Regular season, placed where Statcast tracked each one landing",
    caption = "Data: Baseball Savant and MLB Stats API via baseballr | Viz: sdvplotR"
  ) +
  # sportyR paints the whole figure as the field, so the text goes white
  theme(
    plot.title = element_text(face = "bold", colour = "white"),
    plot.subtitle = element_text(colour = "white"),
    plot.caption = element_text(colour = "white"),
    strip.text = element_text(face = "bold", size = 12),
    legend.position = "bottom"
  )
```

## 7. Quality of contact: exit velocity and barrels

Every qualified hitter from Savant's exit velocity leaderboard as a grey point,
and the five highest barrel rates as headshots (a tie for fifth makes it six
hitters). A barrel is a batted ball with the exit velocity and launch angle
that produce at least a .500 batting average and 1.500 slugging. The names sit
beside the headshots, pushed apart vertically by ggrepel.

```{r contact-quality, fig.width = 8, fig.height = 6, fig.alt = "Scatter plot of qualified MLB hitters in 2026 by average exit velocity (horizontal) and barrels per batted ball (vertical); most are grey points, and the hitters with the five highest barrel rates are drawn as headshots with their names beside them."}
contact <- contact |>
  mutate(player = paste(sub(".*, ", "", name), sub(",.*", "", name))) # "Ohtani, Shohei"
barrels <- contact |>
  filter(min_rank(desc(brl_percent)) <= 5) |>
  # names to the left of the headshots, except at the right edge
  mutate(side = if_else(avg_hit_speed > 94.5, 1, -1))
name_labels <- function(side) {
  ggrepel::geom_text_repel(
    aes(label = player),
    data = filter(barrels, side == !!side), size = 3.3, point.size = 13, box.padding = 0.4,
    nudge_x = 0.9 * side, direction = "y", hjust = if (side > 0) 0 else 1,
    min.segment.length = 0, bg.colour = "white", bg.r = 0.15, seed = 1
  )
}

ggplot(contact, aes(avg_hit_speed, brl_percent)) +
  geom_point(colour = "grey70", size = 1.6) +
  geom_sdv_headshots(
    aes(player_id = player_id),
    data = barrels, sport = "mlb", id_type = "league", height = 0.1
  ) +
  name_labels(-1) +
  name_labels(1) +
  scale_x_continuous(expand = expansion(add = c(0.3, 2))) +
  labs(
    title = paste("Who hit the ball hardest in", season),
    subtitle = paste(nrow(contact), "qualified hitters; the five highest barrel rates as headshots"),
    x = "Average exit velocity (mph)",
    y = "Barrels per batted ball (%)",
    caption = savant
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 8. The division races, week by week

A bump chart: each team's place in its division at the end of every week, from
every regular-season result. A suspended game is listed on the day it started
and the day it ended; keeping the last date counts it once. The final week
matches the official standings.

Division rivals often share a color (four of the five AL Central clubs are
navy or black), so the lines take their colors from `sdv_team_colors()`: each
team keeps its primary color unless that is too close to a color already used
in its division, and then takes its secondary.

```{r bump-chart, fig.width = 9, fig.height = 9, fig.alt = "Six small line charts, one per MLB division, showing each team's place in its division at the end of every week of the 2026 season, lines in team colors, with the team logo at the end of each line."}
games <- slice_max(games, date, by = game_pk)
results <- bind_rows(
  transmute(games, date, team_id = home_id, win = home_score > away_score),
  transmute(games, date, team_id = away_id, win = away_score > home_score)
)
# every Sunday from the first full week to the last day of the season
weeks <- seq(as.Date(paste0(season, "-04-05")), max(games$date), by = "week")

race <- cross_join(tibble(week = weeks), distinct(results, team_id)) |>
  left_join(results, by = join_by(team_id, week >= date)) |>
  summarise(w = sum(win), l = sum(!win), .by = c(week, team_id)) |>
  inner_join(select(standings, team_id, team, division, official = rank), by = "team_id") |>
  arrange(week, desc(w), team) |>
  mutate(place = row_number(desc(w / (w + l))), .by = c(week, division))

final <- filter(race, week == max(weeks))
stopifnot(all(final$place == final$official))

# the primary, unless it is within 20 (CIE 2000 color difference) of the white
# background or of a color already taken, and the secondary is farther away
distinct_colours <- function(primary, secondary, min_gap = 20) {
  taken <- "#FFFFFF"
  gap <- function(x) {
    min(farver::compare_colour(
      farver::decode_colour(x), farver::decode_colour(taken), "rgb",
      method = "cie2000"
    ))
  }
  for (i in seq_along(primary)) {
    swap <- gap(primary[i]) < min_gap && gap(secondary[i]) > gap(primary[i])
    taken <- c(taken, if (swap) secondary[i] else primary[i])
  }
  taken[-1]
}
line_colours <- final |>
  arrange(division, place) |>
  inner_join(sdv_team_colors("mlb", type = "all"), by = c(team = "team_abbr")) |>
  mutate(colour = distinct_colours(primary, secondary), .by = division)

ggplot(race, aes(week, place, colour = team)) +
  geom_line(linewidth = 1) +
  geom_point(size = 1.3) +
  # inherit.aes = FALSE: a colour aesthetic would tint the logos
  geom_sdv_logos(
    aes(x = week + 10, y = place, team = team),
    data = final, sport = "mlb", width = 0.07, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = setNames(line_colours$colour, line_colours$team)) +
  scale_y_reverse(breaks = 1:5) +
  scale_x_date(date_labels = "%b", expand = expansion(add = c(3, 16))) +
  facet_wrap(~ factor(division, levels = divisions), ncol = 2) +
  labs(
    title = paste("The", season, "division races"),
    subtitle = "Place in the division at the end of each week",
    x = NULL,
    y = "Place",
    caption = stats_api
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## 9. Franchise moves since 1901

Nine franchises that changed cities, one bar per franchise, one shade per city.
The abbreviations change with the eras (PHA, KCA, OAK, ATH for the Athletics),
but the Stats API team id never does, so each franchise is keyed by its id and
drawn with today's logo on the axis (`scale_y_sdv()` with `theme_y_sdv()`).
Old abbreviations are not a reliable key:

```{r old-codes}
clean_team_abbrs(c("OAK", "KCA", "MON", "BRO"), sport = "mlb", keep_non_matches = FALSE)
```

sdvplotR maps the Oakland and Montreal codes to the clubs they became, but KCA,
the Kansas City Athletics, matches KC, today's Royals. The Stats API lists the
1969 expansion Expos and Pilots from 1968, the year they were founded.

```{r franchise-eras, fig.width = 9, fig.height = 5.5, fig.alt = "Timeline from 1901 to 2026 with one bar per MLB franchise that relocated (Athletics, Nationals, Rangers, Brewers, Braves, Twins, Dodgers, Giants and Orioles), each in its team color with a lighter or darker shade for each city it played in, the earlier cities written inside the bars, today's name at the right and today's logo on the axis."}
eras <- identities |>
  # the city is the name without its last word ("Athletics" alone has none)
  mutate(city = sub(" [^ ]+$", "", name)) |>
  arrange(team_id, season) |>
  mutate(era = consecutive_id(city), .by = team_id) |>
  summarise(
    from = min(season), to = max(season), city = first(city), name = last(name),
    .by = c(team_id, era)
  ) |>
  inner_join(select(standings, team_id, team), by = "team_id") |>
  mutate(
    last_move = max(from), current = era == max(era),
    # today's city in full color, earlier ones alternating lighter and darker
    dark = (max(era) - era) %% 2 == 0,
    width = to - from + 1, mid = from + width / 2,
    .by = team_id
  ) |>
  mutate(
    team = reorder(team, last_move),
    # earlier cities go inside their era where they fit
    label = if_else(!current & width >= 1.15 * nchar(city), city, NA)
  )

ggplot(eras, aes(y = team)) +
  geom_tile(aes(x = mid, width = width, fill = team, alpha = dark), height = 0.72, colour = "white") +
  geom_text(
    aes(x = mid, label = label, colour = if_else(dark, "white", "grey10")),
    size = 2.9, na.rm = TRUE
  ) +
  geom_text(
    aes(x = to + 3, label = name),
    data = filter(eras, current), hjust = 0, size = 3.3, fontface = "bold"
  ) +
  scale_fill_sdv(sport = "mlb") +
  scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.55)) +
  scale_colour_identity() +
  scale_x_continuous(
    breaks = c(1901, seq(1925, 2000, 25), season),
    limits = c(1901, season + 32), expand = expansion(0)
  ) +
  scale_y_sdv(sport = "mlb", size = 22) +
  labs(
    title = "Nine MLB franchises that moved",
    subtitle = "One shade per city, from 1901 or the franchise's first season",
    x = NULL,
    y = NULL,
    caption = stats_api
  ) +
  theme_minimal(base_size = 12) +
  theme_y_sdv() +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 10. The postseason field as a tier list

`sdv_team_tiers()` draws a tier list from a tier and a team column. The tiers
here come from the standings' clinch codes (`y` and `z` won a division, `w` a
wild card) and the games behind the last wild card; within a tier the teams go
by wins. The result is a ggplot, so a light background goes on top of its dark
one, where the navy and brown logos would disappear.

```{r team-tiers, fig.width = 9, fig.height = 6, fig.alt = "Tier list of the 30 MLB teams in 2026, drawn as logos in four rows: division winners, wild cards, teams that missed the postseason by five games or fewer, and teams that missed by more."}
field <- standings |>
  mutate(
    behind = suppressWarnings(as.numeric(wc_gb)), # "-" and "+2.0" belong to playoff teams
    tier_no = case_when(
      clinch %in% c("y", "z") ~ 1,
      clinch == "w" ~ 2,
      behind <= 5 ~ 3,
      .default = 4
    )
  ) |>
  arrange(tier_no, desc(w), team) |>
  mutate(tier_rank = row_number(), .by = tier_no)
stopifnot(sum(field$tier_no == 1) == 6, sum(field$tier_no == 2) == 6)

sdv_team_tiers(
  select(field, tier_no, tier_rank, team),
  sport = "mlb",
  title = paste("How the", season, "postseason field was built"),
  subtitle = "Regular season; teams ordered by wins within each tier",
  caption = stats_api,
  tier_desc = c(
    "1" = "Division winners",
    "2" = "Wild cards",
    "3" = "Missed by 5 games or fewer",
    "4" = "Missed by more"
  ),
  width = 0.06
) +
  theme(
    plot.background = element_rect(fill = "white", colour = "white"),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.title = element_text(colour = "grey10", face = "bold"),
    plot.subtitle = element_text(colour = "grey35"),
    plot.caption = element_text(colour = "grey35"),
    axis.text.y = element_text(colour = "grey10", face = "bold")
  )
```

## Related articles

- [Getting Started](getting-started.html)
- [NHL Visualizations](nhl-viz.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
