---
title: "NBA Visualizations with hoopR and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NBA Visualizations with hoopR and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to create rich NBA visualizations by combining
[hoopR](https://hoopR.sportsdataverse.org) for player and team data with
[sdvplotR](https://sdvplotr.sportsdataverse.org) for team logos, headshots,
colors, and gt tables.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(hoopR)
library(dplyr)
library(gt)

# Get valid NBA team abbreviations
nba_teams <- valid_team_names("nba")
head(nba_teams)
```

## Loading NBA Data

Use `hoopR` to load this season's ESPN team and player box scores:

```{r load-data}
season <- hoopR::most_recent_nba_season()

# One row per team per game (ESPN box scores), regular season only
team_stats <- hoopR::load_nba_team_box(seasons = season) |>
  filter(season_type == 2)

# One row per player per game, with ESPN athlete IDs; players who did not
# play are dropped so games played counts real games
player_stats <- hoopR::load_nba_player_box(seasons = season) |>
  filter(season_type == 2, !did_not_play)
```

## NBA Team Performance

Visualize team performance with team logos:

```{r team-performance}
# Calculate team metrics
team_perf <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    avg_points = mean(team_score, na.rm = TRUE),
    avg_rebounds = mean(total_rebounds, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10)

ggplot(team_perf, aes(x = avg_points, y = avg_rebounds)) +
  geom_sdv_logos(
    aes(team = team_abbreviation),
    sport = "nba",
    width = 0.075
  ) +
  labs(
    title = "NBA Team Performance",
    subtitle = paste("Season", hoopR::most_recent_nba_season()),
    x = "Average Points per Game",
    y = "Average Rebounds per Game",
    caption = "Data: hoopR | Viz: sdvplotR"
  ) +
  theme_minimal()
```

## NBA Team Colors

Use team colors to visualize win percentages:

```{r team-colors}
# Calculate win percentage
team_wins <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    wins = sum(team_winner, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10) |>
  mutate(win_pct = wins / games) |>
  arrange(desc(win_pct)) |>
  head(16)

ggplot(team_wins, aes(x = reorder(team_abbreviation, win_pct), y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.7) +
  scale_fill_sdv(sport = "nba", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 16 NBA Teams by Win Percentage",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Player Headshots

Visualize top performers with player headshots:

```{r player-headshots}
# Top scorers
top_scorers <- player_stats |>
  filter(!is.na(athlete_id)) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    avg_points = mean(points, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 20) |>
  arrange(desc(avg_points)) |>
  head(8)

ggplot(top_scorers, aes(x = games, y = avg_points)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "nba",
    height = 0.15
  ) +
  geom_label(
    aes(label = athlete_display_name),
    nudge_y = -1.5,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "Top 8 NBA Scorers",
    subtitle = paste("Season", hoopR::most_recent_nba_season()),
    x = "Games Played",
    y = "Average Points per Game"
  ) +
  theme_minimal()
```

### NBA Stats player IDs

The headshots above use ESPN athlete IDs, the `athlete_id` in hoopR's
`load_nba_*()` data. hoopR's `nba_*()` functions read stats.nba.com, which
numbers players differently (`PLAYER_ID`). Pass `id_type = "league"` to draw
those from the NBA's own image CDN. Both are plain numbers, so without it an
NBA Stats ID is read as an ESPN ID and draws someone else or nothing: Dirk
Nowitzki's NBA Stats ID, 1717, is ESPN's Jared Jeffries.

```{r nba-stats-headshots}
leaders <- hoopR::nba_leagueleaders(stat_category = "PTS")$LeagueLeaders |>
  mutate(across(c(GP, PTS), as.numeric)) |>
  head(8)

ggplot(leaders, aes(x = GP, y = PTS)) +
  geom_sdv_headshots(
    aes(player_id = PLAYER_ID),
    sport = "nba",
    id_type = "league",
    height = 0.15
  ) +
  labs(title = "NBA scoring leaders", x = "Games Played", y = "Points")
```

`id_type` works the same in `gt_sdv_headshots()`, `reactable_sdv_headshots()`,
the headshot axis scales and `element_sdv_headshot()`. The NBA's image CDN
refuses requests from datacenter IPs, so a plot rendered on CI or a server can
come back without these headshots; tables are unaffected, because the reader's
browser loads the images.

## NBA Team Tiers

Create a tier plot ranking NBA teams:

```{r team-tiers}
# Sample tier assignments
tier_data <- data.frame(
  tier_no = c(1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3, 4, 4, 4, 4, 5),
  team = c("BOS", "DEN", "MIL", "PHX", "LAL", "GSW", "MIA",
           "DAL", "PHI", "NYK", "CLE",
           "SAC", "MIN", "MEM", "BKN",
           "ATL")
)

sdv_team_tiers(
  tier_data,
  sport = "nba",
  title = "NBA Power Rankings",
  subtitle = paste("As of", Sys.Date()),
  tier_desc = c(
    "1" = "Championship Contenders",
    "2" = "Playoff Favorites",
    "3" = "Solid Playoffs",
    "4" = "Bubble Teams",
    "5" = "Rebuilding"
  )
)
```

## NBA Conference Map

Show each conference's teams in a column, using the conferences sdvplotR keeps
for every team:

```{r conference-map}
conference_map <- team_reference("nba") |>
  arrange(conference, team_name) |>
  group_by(conference) |>
  mutate(team_rank = row_number()) |>
  ungroup() |>
  mutate(conference_num = as.numeric(factor(conference)))

ggplot(conference_map, aes(x = conference_num, y = team_rank)) +
  geom_sdv_logos(
    aes(team = team_abbr),
    sport = "nba",
    width = 0.08
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(factor(conference_map$conference))),
    labels = levels(factor(conference_map$conference)),
    expand = expansion(add = 0.6)
  ) +
  scale_y_reverse() +
  labs(
    title = "NBA Teams by Conference",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank()
  )
```

## NBA Standings Table with Logos

Create a gt table with team logos:

```{r gt-table}
standings_table <- team_wins |>
  head(10) |>
  mutate(
    logo = team_abbreviation,
    rank = row_number()
  ) |>
  select(rank, logo, team_abbreviation, wins, games, win_pct)

standings_table |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nba", height = 35) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team_abbreviation = "Abbrev",
    wins = "Wins",
    games = "Games",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "NBA Top 10",
    subtitle = paste("Season", hoopR::most_recent_nba_season())
  )
```

## Player Performance Comparison

Compare the scoring leaders with the rebounding leaders:

```{r player-comparison}
per_game <- player_stats |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    points = mean(points, na.rm = TRUE),
    rebounds = mean(rebounds, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 20)

comparison <- bind_rows(
  per_game |>
    slice_max(points, n = 5, with_ties = FALSE) |>
    transmute(athlete_id, category = "Points per Game", value = points),
  per_game |>
    slice_max(rebounds, n = 5, with_ties = FALSE) |>
    transmute(athlete_id, category = "Rebounds per Game", value = rebounds)
) |>
  group_by(category) |>
  mutate(rank = row_number()) |>
  ungroup()

ggplot(comparison, aes(x = rank, y = value)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "nba",
    height = 0.2
  ) +
  facet_wrap(~category, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "NBA Top Performers",
    subtitle = paste("Season", season),
    x = "Rank",
    y = NULL
  ) +
  theme_minimal()
```

## Axis Labels with Logos

Replace axis labels with team logos:

```{r axis-logos}
top_8 <- team_wins |>
  head(8) |>
  mutate(team_abbreviation = factor(team_abbreviation, levels = team_abbreviation))

ggplot(top_8, aes(x = team_abbreviation, y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.6) +
  scale_fill_sdv(sport = "nba", alpha = 0.7) +
  scale_x_sdv(sport = "nba") +
  theme_minimal() +
  theme_x_sdv() +
  labs(
    title = "Top 8 NBA Teams by Win %",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme(legend.position = "none")
```

## Next Steps

- Explore [hoopR documentation](https://hoopR.sportsdataverse.org/)
- Try combining with [oddsapiR](https://oddsapiR.sportsdataverse.org) for betting lines
- Build weekly dashboard with [Quarto](https://quarto.org/)

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
