---
title: "NBA Visualizations with hoopR and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NBA Visualizations with hoopR and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Ten charts and tables from the 2025-26 NBA regular season: an offense-vs-defense
quadrant and net-rating bars with logos, a conference bump chart, a scoring
leaderboard with headshots, a shot chart on a team-colored court, tiers, a
standings table, faceted logos, a home/road dumbbell and a table of category
leaders. Every number comes from
[hoopR](https://hoopR.sportsdataverse.org)'s ESPN release files on GitHub
(`load_nba_*()`), so nothing here calls stats.nba.com.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)

season <- 2026 # the 2025-26 season: hoopR names a season by the year it ends
label <- "2025-26"
source_note <- "Data: hoopR (ESPN) | Viz: sdvplotR"

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      plot.caption = element_text(color = "grey40", size = 8),
      plot.caption.position = "plot",
      panel.grid.minor = element_blank()
    )
)
```

The team box score has one row per team per game. `season_type` 2 is the
regular season (3 is the playoffs, 5 the play-in). ESPN files the All-Star
games as regular-season games too, between teams that aren't NBA clubs, so an
inner join on the ESPN team id to `team_reference("nba")` drops them and adds
each team's sdvplotR abbreviation and conference in the same step.

```{r load-data}
nba_teams <- team_reference("nba") |>
  select(team_id = espn_team_id, team = team_abbr, conference)

box <- hoopR::load_nba_team_box(seasons = season) |>
  filter(season_type == 2) |>
  inner_join(nba_teams, by = "team_id")

box |>
  distinct(game_id) |>
  nrow()
```

That is one more than the 1,230 games of the schedule: the NBA Cup final
between New York and San Antonio is filed as a regular-season game too. It
stays in the box scores (it was played) but not in the standings, so those two
teams show 83 games here.

## 1. Offense vs defense, with logos

Points scored and allowed per 100 possessions put all 30 teams on one chart.
Possessions are estimated from the box score (FGA - OREB + TOV + 0.44 x FTA)
and averaged with the opponent's estimate for the same game. One Chicago game
is missing its team turnovers, so `total_turnovers` falls back to the
players' `turnovers` there; without that, Chicago and its opponent that night,
Orlando, would drop out of every chart below. The last line stops the build if
a team ever goes missing.

```{r ratings}
games <- box |>
  mutate(
    poss = field_goals_attempted - offensive_rebounds +
      coalesce(total_turnovers, turnovers) + 0.44 * free_throws_attempted
  )
games <- games |>
  inner_join(
    games |> select(game_id, opponent_team_id = team_id, opp_poss = poss),
    by = c("game_id", "opponent_team_id")
  ) |>
  mutate(game_poss = (poss + opp_poss) / 2)

ratings <- games |>
  group_by(team, conference) |>
  summarise(
    games = n(),
    wins = sum(team_winner),
    pace = mean(game_poss),
    ortg = 100 * sum(team_score) / sum(game_poss),
    drtg = 100 * sum(opponent_team_score) / sum(game_poss),
    fg3a_rate = sum(three_point_field_goals_attempted) / sum(field_goals_attempted),
    fg3_pct = sum(three_point_field_goals_made) / sum(three_point_field_goals_attempted),
    .groups = "drop"
  ) |>
  mutate(net = ortg - drtg) |>
  arrange(desc(net), team)
stopifnot(nrow(ratings) == 30, !anyNA(ratings$net))

ratings |>
  select(team, conference, wins, ortg, drtg, net) |>
  head(5)
```

Defensive rating is points allowed, so its axis is reversed: good defenses sit
at the top and the best teams in the top-right corner.

```{r team-performance, fig.alt = "NBA teams of the 2025-26 regular season drawn as their logos, placed by offensive rating (horizontal) and defensive rating (vertical, reversed so better defenses are higher), with dashed league-average lines and a label in each corner."}
x_lim <- range(ratings$ortg) + c(-1.5, 1.5)
y_lim <- range(ratings$drtg) + c(-2, 1.5)

ggplot(ratings, aes(ortg, drtg)) +
  geom_mean_lines(
    aes(x0 = ortg, y0 = drtg),
    color = "grey60", linetype = "dashed", linewidth = 0.4
  ) +
  geom_sdv_logos(aes(team = team), sport = "nba", width = 0.065) +
  annotate(
    "text",
    x = x_lim[c(2, 1, 2, 1)], y = y_lim[c(1, 1, 2, 2)],
    hjust = c(1, 0, 1, 0), vjust = c(1.3, 1.3, -0.4, -0.4),
    label = c("Good offense, good defense", "Defense first", "Offense first", "Struggling"),
    color = "grey45", fontface = "italic", size = 3.5
  ) +
  scale_x_continuous(expand = expansion(0)) +
  scale_y_reverse(expand = expansion(0)) +
  expand_limits(x = x_lim, y = y_lim) +
  labs(
    title = paste("NBA offense vs defense,", label, "regular season"),
    x = "Offensive rating (points per 100 possessions)",
    y = "Defensive rating (points allowed per 100)",
    caption = source_note
  )
```

## 2. Net rating, ranked, with logos on the axis

`scale_x_sdv()` turns a discrete axis of team abbreviations into logos, and
`theme_x_sdv()` lets the axis draw them. Order the factor first: the bars keep
that order and so do the logos.

```{r net-rating, fig.width = 10, fig.height = 5, fig.alt = "Bar chart of NBA net rating per 100 possessions in 2025-26, best to worst, each bar in the team's color with the team's logo under it on the horizontal axis."}
ratings |>
  mutate(team = factor(team, levels = team)) |>
  ggplot(aes(team, net, fill = team)) +
  geom_col(width = 0.75) +
  geom_hline(yintercept = 0, linewidth = 0.4) +
  scale_fill_sdv(sport = "nba") +
  scale_x_sdv(sport = "nba", size = 16) +
  labs(
    title = paste("NBA net rating,", label, "regular season"),
    x = NULL,
    y = "Net rating (per 100 possessions)",
    caption = source_note
  ) +
  theme_x_sdv() +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

## 3. The race in the West, as a bump chart

Rank each team inside its conference by win percentage at the end of every
week, then draw one line per team in its color with its logo at the finish.
Ties are broken by point differential, then by abbreviation, not by the NBA's
tiebreakers, so the same data always draw the same chart.

```{r bump-chart, fig.width = 9, fig.height = 6, fig.alt = "Bump chart of the 15 Western Conference teams' weekly conference rank through the 2025-26 regular season, one line per team in its color, with each team's logo at its final rank on the right."}
weekly <- box |>
  mutate(
    week = as.Date(cut(game_date, "week", start.on.monday = TRUE)),
    margin = team_score - opponent_team_score
  ) |>
  group_by(team, conference, week) |>
  summarise(wins = sum(team_winner), games = n(), margin = sum(margin), .groups = "drop") |>
  complete(nesting(team, conference), week, fill = list(wins = 0, games = 0, margin = 0)) |>
  arrange(team, week) |>
  group_by(team) |>
  mutate(across(c(wins, games, margin), cumsum)) |>
  ungroup() |>
  mutate(week_no = dense_rank(week)) |>
  filter(week_no >= 3) |> # the first two weeks hold only a game or two
  arrange(conference, week_no, desc(wins / games), desc(margin), team) |>
  group_by(conference, week_no) |>
  mutate(rank = row_number()) |>
  ungroup()

west <- filter(weekly, conference == "Western")
finish <- filter(west, week_no == max(week_no))

ggplot(west, aes(week_no, rank)) +
  geom_line(aes(color = team), linewidth = 1.1, alpha = 0.85) +
  # the logos get no color aesthetic: a colour would tint the whole image
  geom_sdv_logos(aes(x = week_no + 1, team = team), data = finish, sport = "nba", width = 0.04) +
  scale_color_sdv(sport = "nba") +
  scale_y_reverse(breaks = 1:15) +
  scale_x_continuous(breaks = seq(5, 30, 5), expand = expansion(add = c(0.5, 1.5))) +
  labs(
    title = paste("The race in the West,", label),
    subtitle = "Conference rank by win percentage at the end of each week",
    x = "Week of the season",
    y = "Western Conference rank",
    caption = source_note
  ) +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

Swap `"Western"` for `"Eastern"` to draw the other conference from the same
`weekly` frame.

## 4. A scoring leaderboard with headshots

The player box score carries ESPN athlete ids, which is what
`geom_sdv_headshots()` takes for the NBA. Only games in `box` count, so the
All-Star games stay out, and a player's team is the one he played for last.

```{r leaders}
players <- hoopR::load_nba_player_box(seasons = season) |>
  semi_join(distinct(box, game_id), by = "game_id") |>
  filter(!did_not_play)

per_game <- players |>
  arrange(game_date) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    games = n(),
    team = last(team_abbreviation),
    across(
      c(points, rebounds, assists, steals, blocks, three_point_field_goals_made),
      mean
    ),
    .groups = "drop"
  ) |>
  filter(games >= 50)

scorers <- per_game |>
  arrange(desc(points), athlete_id) |>
  head(10) |>
  mutate(athlete_display_name = factor(athlete_display_name, levels = rev(athlete_display_name)))
```

```{r player-headshots, fig.height = 6, fig.alt = "Horizontal bar chart of the ten NBA scoring leaders of 2025-26 (50 or more games), each bar in the player's team color, followed by the team logo, the player's headshot and his points per game."}
ggplot(scorers, aes(points, athlete_display_name)) +
  geom_col(aes(fill = team), width = 0.7) +
  geom_sdv_logos(aes(x = points + 1.6, team = team), sport = "nba", height = 0.07) +
  geom_sdv_headshots(aes(x = points + 4.6, player_id = athlete_id), sport = "nba", height = 0.09) +
  geom_text(aes(x = points + 7.4, label = sprintf("%.1f", points)), fontface = "bold", hjust = 0) +
  scale_fill_sdv(sport = "nba") +
  scale_x_continuous(limits = c(0, max(scorers$points) + 10), expand = expansion(0)) +
  labs(
    title = paste("NBA scoring leaders,", label),
    subtitle = "Points per game, 50 or more games",
    x = "Points per game",
    y = NULL,
    caption = source_note
  ) +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

These are ESPN athlete ids. stats.nba.com numbers players differently
(`PERSON_ID`), and the two can't be told apart by shape: an NBA Stats id read
as an ESPN id draws someone else or nothing. For data keyed by NBA Stats ids,
pass `id_type = "league"` to draw from the NBA's own image CDN:

```{r id-type, eval = FALSE}
geom_sdv_headshots(aes(player_id = person_id), sport = "nba", id_type = "league")
```

That CDN refuses requests from datacenter IPs, this site's build server among
them, so the articles here use ESPN ids. `id_type` works the same in
`gt_sdv_headshots()`, `reactable_sdv_headshots()` and the headshot axis scales.

## 5. A shot chart on a team-colored court

`load_nba_shots()` holds ESPN's shot locations. hoopR already converts them to
feet on a center-court frame (`coordinate_x` from -47 to 47 along the court,
`coordinate_y` from -25 to 25 across it), which is the frame
`sdv_surface()`'s sportyR court uses, so they plot as they are.
`sdv_court_coords()` is not needed here: it is for the stats.nba.com legacy
frame (`x_legacy`/`y_legacy` from `hoopR::load_nba_stats_shots()`, tenths of a
foot around the hoop). Shots at the right basket are rotated onto the left one
so a half court holds them all; free throws are left out.

```{r shot-chart, fig.width = 7, fig.height = 7, fig.alt = "Half-court shot chart of the 2025-26 NBA scoring leader's field goal attempts on a court painted in his team's colors: made shots as filled circles, missed shots as grey crosses."}
star <- scorers |>
  slice_max(points, n = 1, with_ties = FALSE)

shots <- hoopR::load_nba_shots(seasons = season) |>
  semi_join(distinct(box, game_id), by = "game_id") |>
  filter(athlete_id_1 == star$athlete_id, !grepl("Free Throw", type_text)) |>
  mutate(
    right = coordinate_x > 0,
    x = if_else(right, -coordinate_x, coordinate_x),
    y = if_else(right, -coordinate_y, coordinate_y),
    result = if_else(scoring_play, "Made", "Missed")
  )

made_color <- sdv_team_colors(star$team, sport = "nba", type = "secondary")

sdv_surface("nba", star$team, display_range = "defense") +
  geom_point(
    aes(x, y),
    data = filter(shots, result == "Missed"),
    shape = 4, color = "grey25", size = 1.4, alpha = 0.6
  ) +
  geom_point(
    aes(x, y),
    data = filter(shots, result == "Made"),
    shape = 21, fill = made_color, color = "black", stroke = 0.3, size = 1.8
  ) +
  labs(
    title = paste0(star$athlete_display_name, ": every field goal attempt, ", label),
    subtitle = sprintf(
      "%d made (filled), %d missed (crosses): %.1f%% from the field",
      sum(shots$scoring_play), sum(!shots$scoring_play), 100 * mean(shots$scoring_play)
    ),
    caption = source_note
  ) +
  theme(plot.title = element_text(face = "bold"))
```

## 6. Team tiers from net rating

`sdv_team_tiers()` draws a tier list from a frame with `tier_no` and `team`.
Here the tiers are cut from the net ratings above, best team first in each
tier, so the list is the season's numbers rather than an opinion.

```{r team-tiers, fig.width = 9, fig.height = 6, fig.alt = "NBA team logos in five tiers cut from 2025-26 net rating, from Contenders (plus 6 or better) at the top to Rebuilding (minus 6 or worse) at the bottom."}
tiers <- ratings |>
  mutate(
    # bins from worst to best, so the best bin (6 or better) becomes tier 1
    tier_no = 6L - cut(net, c(-Inf, -6, -2, 2, 6, Inf), labels = FALSE, right = FALSE)
  ) |>
  arrange(tier_no, desc(net)) |>
  select(tier_no, team)

sdv_team_tiers(
  tiers,
  sport = "nba",
  title = paste("NBA tiers by net rating,", label),
  subtitle = "Regular-season net rating per 100 possessions, cut at +6, +2, -2 and -6",
  caption = source_note,
  tier_desc = c(
    "1" = "Contenders",
    "2" = "Playoff teams",
    "3" = "The middle",
    "4" = "Lottery",
    "5" = "Rebuilding"
  )
)
```

## 7. A conference standings table with logos and cut lines

ESPN's standings come long, one row per team and stat, so pivot them wide.
`gt_sdv_logos()` turns the abbreviation column into logos and `gt_cutline()`
marks the playoff and play-in lines: seeds 1-6 make the playoffs, 7-10 the
play-in.

```{r standings-table}
standings <- hoopR::load_nba_standings(seasons = season) |>
  select(group_name, team_abbreviation, team_short_display_name, stat_name, display_value) |>
  pivot_wider(names_from = stat_name, values_from = display_value)

standings |>
  filter(group_name == "Eastern Conference") |>
  mutate(seed = as.integer(playoffSeed), logo = team_abbreviation) |>
  arrange(seed) |>
  select(
    seed, logo, team_short_display_name, wins, losses, winPercent, gamesBehind,
    Home, Road, `Last Ten Games`, differential
  ) |>
  gt() |>
  tab_header(
    title = paste("Eastern Conference standings,", label),
    subtitle = "Seeds 1-6 make the playoffs; 7-10 the play-in"
  ) |>
  cols_label(
    seed = "", logo = "", team_short_display_name = "Team", wins = "W", losses = "L",
    winPercent = "Pct", gamesBehind = "GB", `Last Ten Games` = "L10",
    differential = "Diff"
  ) |>
  tab_source_note(source_note) |>
  gt_sdv_logos(columns = logo, sport = "nba", height = 26) |>
  gt_theme_athletic() |>
  cols_align("left", columns = team_short_display_name) |>
  gt_cutline(after = c(6, 10), label = c("Playoffs", "Play-in"), label_position = "above")
```

## 8. Logos faceted by conference

`geom_sdv_logos()` is a ggplot2 layer, so it facets like any other geom, and
`geom_mean_lines()` draws each panel's own averages. How often teams shoot
threes against how well they make them, East vs West:

```{r three-point, fig.width = 10, fig.height = 5, fig.alt = "Two panels, Eastern and Western Conference, placing each NBA team's logo by its share of field goal attempts from three (horizontal) and its three-point percentage (vertical) in 2025-26, with each panel's average lines."}
ggplot(ratings, aes(fg3a_rate, fg3_pct)) +
  geom_mean_lines(aes(x0 = fg3a_rate, y0 = fg3_pct), color = "grey60", linewidth = 0.4) +
  geom_sdv_logos(aes(team = team), sport = "nba", width = 0.09) +
  facet_wrap(~conference, labeller = labeller(conference = function(x) paste(x, "Conference"))) +
  scale_x_continuous(labels = scales::percent, expand = expansion(mult = 0.12)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1), expand = expansion(mult = 0.12)) +
  labs(
    title = paste("Three-point volume vs accuracy,", label),
    x = "Share of field goal attempts from three",
    y = "Three-point percentage",
    caption = source_note
  ) +
  theme(strip.text = element_text(face = "bold", size = 11))
```

## 9. Home and road, as a dumbbell with logos on the axis

Home and road records come from the same standings, as strings like `"29-12"`.
One row per team, sorted by the gap between them; `scale_y_sdv()` and
`theme_y_sdv()` put the logos on the vertical axis.

```{r home-road, fig.width = 8, fig.height = 9, fig.alt = "Dumbbell chart of the 30 NBA teams' 2025-26 home and road win percentages, one row per team with its logo on the vertical axis, sorted by the size of the home edge."}
win_pct <- function(record) {
  parts <- strsplit(record, "-", fixed = TRUE)
  vapply(parts, function(p) as.numeric(p[1]) / sum(as.numeric(p)), numeric(1))
}

home_road <- standings |>
  transmute(team = team_abbreviation, home = win_pct(Home), road = win_pct(Road)) |>
  mutate(edge = home - road) |>
  arrange(edge, team) |>
  mutate(team = factor(team, levels = team))

ggplot(home_road, aes(y = team)) +
  geom_segment(aes(x = road, xend = home, yend = team), color = "grey70", linewidth = 1.2) +
  geom_point(aes(x = road), color = "#E15759", size = 3) +
  geom_point(aes(x = home), color = "#4E79A7", size = 3) +
  scale_y_sdv(sport = "nba", size = 14) +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1)) +
  labs(
    title = paste("Home and road win percentage,", label),
    subtitle = "<span style='color:#4E79A7'>**Home**</span> and <span style='color:#E15759'>**road**</span>, sorted by the home edge (largest at the top)",
    x = "Win percentage",
    y = NULL,
    caption = source_note
  ) +
  theme_y_sdv() +
  theme(plot.subtitle = ggtext::element_markdown(), panel.grid.major.y = element_blank())
```

## 10. Category leaders in a headshot table

`gt_sdv_headshots()` turns a column of ESPN athlete ids into headshots and
`gt_sdv_logos()` a column of abbreviations into logos. One row per category:
the per-game leader among players with 50 or more games.

```{r leaders-table}
categories <- c(
  points = "Points", rebounds = "Rebounds", assists = "Assists",
  steals = "Steals", blocks = "Blocks", three_point_field_goals_made = "3-pointers made"
)

per_game |>
  pivot_longer(names(categories), names_to = "stat", values_to = "value") |>
  arrange(desc(value), athlete_id) |>
  group_by(stat) |>
  slice(1) |>
  ungroup() |>
  arrange(match(stat, names(categories))) |>
  transmute(
    category = categories[stat],
    headshot = athlete_id,
    player = athlete_display_name,
    logo = team,
    value,
    games
  ) |>
  gt() |>
  tab_header(
    title = paste("NBA per-game leaders,", label),
    subtitle = "Regular season, 50 or more games"
  ) |>
  cols_label(category = "", headshot = "", player = "Player", logo = "Team", value = "Per game", games = "GP") |>
  fmt_number(value, decimals = 1) |>
  tab_source_note(source_note) |>
  gt_sdv_headshots(columns = headshot, sport = "nba", height = 44) |>
  gt_sdv_logos(columns = logo, sport = "nba", height = 28) |>
  gt_theme_sdv()
```

## Related articles

- [Getting Started](getting-started.html)
- [WNBA Visualizations](wnba-viz.html)
- [Men's College Basketball Visualizations](mbb-viz.html)
- [Social Posting Patterns](social-posting.html)
