---
title: "Neighbour Ranks"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Neighbour Ranks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A rank says where a team sits, but not how close the teams around it are. This
recipe puts one team in the middle of an 11-row table, the five teams ranked
just above it and the five just below, so the gap to its neighbours shows next
to the number.

It uses three sdvplotR helpers: `gt_sdv_logos()` for the logos and names,
`gt_color_ranks()` for the rank column and `gt_highlight_cells()` for the focal
row.

```{r setup, message = FALSE}
library(sdvplotR)
library(gt)
```

## The data

cfbfastR's `load_espn_cfb_team_summaries()` reads the SportsDataverse season
summary release: one row per FBS team, with each metric next to a `_rank`
column.

```{r data}
summaries <- as.data.frame(cfbfastR::load_espn_cfb_team_summaries(2025))
dim(summaries)
head(grep("_rank$", names(summaries), value = TRUE))
```

Not every metric has a rank, so check the three this recipe uses. Rank 1 is
the best team in each one: the highest net and offensive EPA per play, and the
fewest EPA per play allowed on defense.

```{r metrics}
metrics <- c(
  net_adj_epa = "Net adj. EPA/play",
  EPAplay_off = "Offense EPA/play",
  EPAplay_def = "Defense EPA/play"
)
stopifnot(all(paste0(names(metrics), "_rank") %in% names(summaries)))
```

## The window

Sort by the rank column and take the 11 rows centred on the team. Near the top
or bottom of the table there are not five teams on both sides, so the window
slides instead of shrinking: it stays 11 rows and the focal team moves off
centre.

```{r window}
neighbours <- function(data, team, metric, k = 5) {
  data <- data[order(data[[paste0(metric, "_rank")]]), ]
  i <- match(team, data$pos_team)
  first <- min(max(i - k, 1), nrow(data) - 2 * k)
  rows <- data[first:(first + 2 * k), ]
  data.frame(
    rank = rows[[paste0(metric, "_rank")]],
    team = rows$pos_team,
    value = rows[[metric]]
  )
}

focal <- "Illinois"
neighbours(summaries, focal, "net_adj_epa")
```

## The table

The rank colors use the whole league as their domain (`1` to the number of
teams), not the 11 rows on show, so a mid-table window reads as mid-table. The
focal team's name and value are filled with `gt_highlight_cells()`, which takes
a logical mask the same shape as the columns it fills.

```{r table}
neighbour_table <- function(data, focal_team, metric, label) {
  win <- neighbours(data, focal_team, metric)
  is_focal <- win$team == focal_team
  gt(win) |>
    tab_header(title = label) |>
    gt_sdv_logos(team, sport = "cfb", height = 20, include_name = TRUE) |>
    fmt_number(value, decimals = 3) |>
    cols_label(rank = "Rank", team = "Team", value = "Value") |>
    gt_theme_sdv_team(team = focal_team, sport = "cfb", density = "compact") |>
    gt_color_ranks(rank, domain = c(1, nrow(data))) |>
    gt_highlight_cells(c(team, value), cbind(is_focal, is_focal), bold = TRUE)
}

neighbour_table(summaries, focal, "net_adj_epa", metrics[["net_adj_epa"]])
```

## Three metrics side by side

The same function, once per metric, laid out with `gt_grid()`.

```{r grid}
tables <- lapply(names(metrics), function(m) {
  neighbour_table(summaries, focal, m, metrics[[m]])
})
gt_grid(
  tables,
  ncol = 3,
  gap = 12,
  title = paste(focal, "and its neighbours"),
  subtitle = "2025 FBS season, five teams either side by rank",
  caption = "Data: cfbfastR::load_espn_cfb_team_summaries()"
)
```

## Checking the window

A team more than five places from either end is row 6 of 11. Check it for every
team on all three metrics, then look at the rows the top and bottom teams land
on.

```{r check}
row_of <- function(team, metric) {
  match(team, neighbours(summaries, team, metric)$team)
}
n <- nrow(summaries)

for (m in names(metrics)) {
  ranks <- summaries[[paste0(m, "_rank")]]
  rows <- vapply(summaries$pos_team, row_of, integer(1), metric = m)
  stopifnot(all(rows[ranks > 5 & ranks <= n - 5] == 6))
}

ranks <- summaries$net_adj_epa_rank
edges <- summaries$pos_team[order(ranks)][c(1:6, (n - 5):n)]
data.frame(
  team = edges,
  rank = ranks[match(edges, summaries$pos_team)],
  row = unname(vapply(edges, row_of, integer(1), metric = "net_adj_epa"))
)
```
