---
title: "NFL Visualizations with nflverse and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NFL Visualizations with nflverse and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Ten charts and tables from the 2025 NFL season: offense against defense with
team logos, a ranked bar chart with logos on the axis, division standings,
small multiples in team colors, the Super Bowl's win probability with a logo in
the title, two quarterback leaderboards with headshots, three relocated
franchises in the logos of their eras, a field in a team's colors and a tier
list.

The play-by-play is nflfastR's, read from the nflverse releases with
[nflreadr](https://nflreadr.nflverse.com/); `nflfastR::load_pbp()` reads the
same file. sdvplotR takes nflverse's team abbreviations (`LA`, `LV`, `WAS`) and
GSIS player ids as they are, so nothing needs renaming on the way in.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
library(gt)

season <- 2025
caption <- "Data: nflverse via nflreadr | 2025 regular season"

pbp <- nflreadr::load_pbp(season)

# Regular-season scrimmage plays with an EPA value
plays <- pbp |>
  filter(season_type == "REG", play_type %in% c("pass", "run"), !is.na(epa))
nrow(plays)
```

A plain `theme_minimal()` with a bold, left-aligned title serves every chart
below:

```{r theme}
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      plot.caption = element_text(color = "grey45", size = 8)
    )
)
```

## 1. Offense vs defense EPA per play

The chart every NFL season ends with: each team's offensive EPA per play
against the EPA per play its defense allowed, with the team's logo as the point.
`geom_mean_lines()` draws the league averages, and the defense axis is reversed
so the best teams sit top right.

```{r epa-plot, fig.alt = "Scatter plot of the 32 NFL teams in the 2025 regular season, each drawn as its logo, with offensive EPA per play on the horizontal axis and defensive EPA per play allowed on the vertical axis, reversed so better defenses sit higher. Dashed lines mark the league averages."}
offense <- plays |>
  group_by(team = posteam) |>
  summarise(off_epa = mean(epa), off_sr = mean(success), .groups = "drop")
defense <- plays |>
  group_by(team = defteam) |>
  summarise(def_epa = mean(epa), def_sr = mean(success), .groups = "drop")
teams <- inner_join(offense, defense, by = "team")

ggplot(teams, aes(x = off_epa, y = def_epa)) +
  geom_mean_lines(aes(x0 = off_epa, y0 = def_epa), color = "grey50", linewidth = 0.4) +
  geom_sdv_logos(aes(team = team), sport = "nfl", width = 0.055, alpha = 0.9) +
  annotate(
    "text", x = max(teams$off_epa), y = min(teams$def_epa),
    label = "good offense, good defense", hjust = 1, vjust = 1, color = "grey45", size = 3.2
  ) +
  annotate(
    "text", x = min(teams$off_epa), y = max(teams$def_epa),
    label = "bad offense, bad defense", hjust = 0, vjust = 0, color = "grey45", size = 3.2
  ) +
  scale_y_reverse() +
  labs(
    title = "NFL offense vs defense, 2025",
    x = "Offense EPA per play",
    y = "Defense EPA per play allowed (better is up)",
    caption = caption
  )
```

## 2. A ranked bar chart with logos on the axis

Offensive success rate against the league average, sorted, with each bar in its
team's primary color from `scale_fill_sdv()`. The x axis keeps the
abbreviations as its labels, and `element_sdv_logo()` set on `axis.text.x`
draws them as logos.

```{r axis-logos, fig.height = 4.8, fig.alt = "Bar chart of each NFL team's offensive success rate in the 2025 regular season minus the league average, sorted from highest to lowest, with bars in team colors and each team's logo in place of its name on the horizontal axis."}
league_sr <- mean(plays$success)

ranked <- teams |>
  mutate(sr_diff = off_sr - league_sr) |>
  arrange(desc(sr_diff), team) |>
  mutate(team = factor(team, levels = team))

ggplot(ranked, aes(x = team, y = sr_diff)) +
  geom_col(aes(fill = team), width = 0.75) +
  geom_hline(yintercept = 0, color = "grey30", linewidth = 0.3) +
  scale_fill_sdv(sport = "nfl") +
  scale_y_continuous(labels = function(v) sub("^\\+0$", "0", sprintf("%+.0f", 100 * v))) +
  labs(
    title = "Offensive success rate, 2025",
    subtitle = sprintf("Percentage points above or below the league's %.1f%%", 100 * league_sr),
    x = NULL,
    y = NULL,
    caption = caption
  ) +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    axis.text.x = element_sdv_logo(sport = "nfl", size = 0.5)
  )
```

`element_sdv_logo()` draws the axis labels with ggpath. The alternative,
`scale_x_sdv()` with `theme_x_sdv()`, writes them as HTML `<img>` tags for
ggtext; the college football tutorial uses it on the y axis.

## 3. Division standings table

Records come from the schedule (`nflreadr::load_schedules()`), divisions and
nicknames from `nflreadr::load_teams()`. `gt_sdv_logos()` turns the
abbreviation column into logos, `data_color()` shades the point differential
and `gt_theme_sdv()` gives the table the SportsDataverse look.

```{r gt-table}
schedule <- nflreadr::load_schedules(season)

# One row per team per regular-season game
games <- schedule |>
  filter(game_type == "REG", !is.na(result)) |>
  nflreadr::clean_homeaway() |>
  select(week, team, opponent, team_score, opponent_score)

nfl_teams <- nflreadr::load_teams() |>
  select(team = team_abbr, name = team_name, nick = team_nick, division = team_division)

standings <- games |>
  group_by(team) |>
  summarise(
    W = sum(team_score > opponent_score),
    L = sum(team_score < opponent_score),
    T = sum(team_score == opponent_score),
    PF = sum(team_score),
    PA = sum(opponent_score),
    .groups = "drop"
  ) |>
  mutate(Diff = PF - PA, pct = (W + T / 2) / (W + L + T)) |>
  inner_join(nfl_teams, by = "team") |>
  # win percentage, then point differential: not the NFL's tiebreakers
  arrange(division, desc(pct), desc(Diff), team) |>
  select(division, logo = team, nick, W, L, T, PF, PA, Diff)

limit <- max(abs(standings$Diff)) # a color scale centered on zero

standings |>
  gt(groupname_col = "division") |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 22) |>
  cols_label(logo = "", nick = "Team") |>
  data_color(
    columns = "Diff",
    palette = c("#b2182b", "#f7f7f7", "#1b7837"),
    domain = c(-limit, limit)
  ) |>
  tab_header(title = "2025 NFL standings", subtitle = "Regular season, by division") |>
  tab_source_note(caption) |>
  gt_theme_sdv(density = "compact")
```

## 4. Small multiples by division

Each team's running point differential through the season, one panel per
division. `scale_color_sdv()` colors each line by its team, and a second
`geom_sdv_logos()` layer, given only each team's last week, puts the logo just
past the end of its line. Because that layer's data carries the `division`
column too, each logo lands in its own panel.

```{r division-lines, fig.height = 6.5, fig.alt = "Eight small line charts, one per NFL division, each tracing its four teams' running point differential by week through the 2025 regular season in team colors, with each team's logo at the end of its line."}
running <- games |>
  arrange(team, week) |>
  group_by(team) |>
  mutate(diff = cumsum(team_score - opponent_score)) |>
  ungroup() |>
  inner_join(select(nfl_teams, team, division), by = "team")

ends <- running |>
  group_by(team) |>
  slice_max(week, n = 1) |>
  ungroup()

ggplot(running, aes(x = week, y = diff, color = team)) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.3) +
  geom_line(linewidth = 0.8) +
  geom_sdv_logos(
    aes(x = week + 1.4, y = diff, team = team),
    data = ends, sport = "nfl", width = 0.12, inherit.aes = FALSE
  ) +
  scale_color_sdv(sport = "nfl") +
  scale_x_continuous(breaks = c(1, 6, 12, 18), limits = c(1, 20)) +
  facet_wrap(~division, ncol = 4) +
  labs(
    title = "Running point differential by division, 2025",
    x = "Week",
    y = "Point differential",
    caption = caption
  ) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

## 5. A win probability chart with a logo in the title

nflfastR's `home_wp` traced through the Super Bowl. `ggtitle_image()` puts the
winner's logo beside the title (with `theme_title_image()`, which renders the
title as markdown and centres the logo on the words). The fills are the teams' colors from `sdv_team_colors()`;
Seattle and New England share the same navy, so when the two primaries match
the away team switches to its secondary color.

```{r sb-wp, fig.height = 5, fig.alt = "Area chart of Seattle's win probability through Super Bowl LX by minutes played, shaded in Seattle green above 50 percent and New England navy below, with the Seahawks logo beside the title."}
sb <- schedule |> filter(game_type == "SB")
away <- sb$away_team
home <- sb$home_team
winner <- if (sb$away_score > sb$home_score) away else home

wp <- pbp |>
  filter(game_id == sb$game_id, !is.na(home_wp)) |>
  transmute(minute = (3600 - game_seconds_remaining) / 60, away_wp = 1 - home_wp)

away_color <- sdv_team_colors("nfl", away)
home_color <- sdv_team_colors("nfl", home)
if (away_color == home_color) away_color <- sdv_team_colors("nfl", away, type = "secondary")

nick <- function(abbr) nfl_teams$nick[match(abbr, nfl_teams$team)]
title <- sprintf(
  "Super Bowl %s: %s %d, %s %d",
  utils::as.roman(season - 1965), nick(away), sb$away_score, nick(home), sb$home_score
)

ggplot(wp, aes(x = minute)) +
  geom_ribbon(aes(ymin = 0.5, ymax = pmax(away_wp, 0.5)), fill = away_color, alpha = 0.85) +
  geom_ribbon(aes(ymin = pmin(away_wp, 0.5), ymax = 0.5), fill = home_color, alpha = 0.85) +
  geom_line(aes(y = away_wp), linewidth = 0.4) +
  geom_hline(yintercept = 0.5, color = "grey40", linewidth = 0.3) +
  scale_x_continuous(breaks = c(0, 15, 30, 45, 60), limits = c(0, 60), expand = c(0, 0)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  ggtitle_image(
    title_image = winner,
    title = title,
    subtitle = "nflfastR win probability, by minutes played",
    image_height = 30,
    sport = "nfl"
  ) +
  labs(
    x = "Minutes played",
    y = paste(nick(away), "win probability"),
    caption = "Data: nflverse via nflreadr"
  ) +
  theme_title_image(size = 15, face = "bold")
```

## 6. A quarterback leaderboard with headshots

nflverse identifies players by GSIS id; the `id` column of the play-by-play is
the quarterback on each dropback, scrambles included. `geom_sdv_headshots()`
takes those ids and draws each player's NFL.com headshot. The full names come
from `nflreadr::load_players()`, and the bars take each quarterback's team
color.

```{r qb-headshots, fig.height = 6.5, fig.alt = "Horizontal bar chart of the 16 NFL quarterbacks with the highest EPA per dropback in the 2025 regular season (300 or more dropbacks), each bar in the quarterback's team color with his headshot at the end of the bar."}
players <- nflreadr::load_players() |>
  select(id = gsis_id, player = display_name)

qbs <- pbp |>
  filter(season_type == "REG", qb_dropback == 1, !is.na(epa), !is.na(id)) |>
  arrange(game_id, play_id) |>
  group_by(id) |>
  summarise(
    team = last(posteam), # the team he finished the season with
    dropbacks = n(),
    epa = mean(epa),
    success = mean(success),
    cpoe = mean(cpoe, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(dropbacks >= 300) |>
  left_join(players, by = "id") |>
  arrange(desc(epa), id)

top_qbs <- head(qbs, 16)

ggplot(top_qbs, aes(x = epa, y = reorder(player, epa))) +
  geom_col(aes(fill = team), width = 0.7) +
  geom_sdv_headshots(aes(x = epa + 0.022, player_id = id), sport = "nfl", height = 0.058) +
  scale_fill_sdv(sport = "nfl") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = "Top 16 quarterbacks by EPA per dropback, 2025",
    subtitle = "Regular season, 300+ dropbacks",
    x = "EPA per dropback",
    y = NULL,
    caption = caption
  ) +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

## 7. The same leaderboard as a table

`gt_sdv_headshots()` is the table version: it turns a column of GSIS ids into
headshots, next to `gt_sdv_logos()` for the team. This one is styled with
`gt_theme_athletic()`.

```{r qb-table}
top_qbs |>
  head(12) |>
  mutate(rank = row_number(), headshot = id) |>
  select(rank, headshot, player, team, dropbacks, epa, success, cpoe) |>
  gt() |>
  gt_sdv_headshots(columns = "headshot", sport = "nfl", height = 36) |>
  gt_sdv_logos(columns = "team", sport = "nfl", height = 24) |>
  fmt_number(columns = "epa", decimals = 3) |>
  fmt_percent(columns = "success", decimals = 1) |>
  fmt_number(columns = "cpoe", decimals = 1, force_sign = TRUE) |>
  data_color(columns = "epa", palette = c("#f7f7f7", "#1b7837")) |>
  cols_label(
    rank = "#", headshot = "", player = "Quarterback", team = "Team",
    dropbacks = "Dropbacks", epa = "EPA/db", success = "Success", cpoe = "CPOE"
  ) |>
  tab_header(
    title = "The 2025 quarterback leaderboard",
    subtitle = "EPA per dropback, regular season, 300+ dropbacks"
  ) |>
  tab_source_note(caption) |>
  gt_theme_athletic(density = "compact")
```

## 8. Relocated franchises and their eras

Schedules use the abbreviation a team had that season: `OAK` until 2019, `SD`
until 2016, `STL` until 2015. `resolve_historical_abbr()` maps each to today's
franchise and passes anything else through, so it reads only the old codes.
`clean_team_abbrs()` applies the same mapping and every other spelling too, so
other sources' `LAR` and a full name become nflverse's `LA` as well:

```{r relocation-keys}
data.frame(abbr = c("OAK", "SD", "STL", "LAR", "Los Angeles Rams")) |>
  mutate(
    current = resolve_historical_abbr(abbr, sport = "nfl"),
    key = clean_team_abbrs(abbr, sport = "nfl")
  )
```

Grouping on that key keeps each franchise together across its move. The logos
take the season's own abbreviation plus a `season` aesthetic. sdvplotR keeps
the marks of relocated NFL identities, so the `STL` and `SD` seasons draw the
St. Louis Rams' and San Diego Chargers' logos; every season of a current
identity draws today's logo, including the Rams' and Chargers' first years in
Los Angeles before their 2020 redesigns. (The Raiders kept their shield.)

```{r relocations, fig.height = 6, fig.alt = "Three line charts, one each for the Rams, Chargers and Raiders, of regular-season wins from 2012 to 2025, each season drawn as a team logo, with the St. Louis Rams and San Diego Chargers logos before their moves and a dashed line at the season each franchise moved."}
history <- nflreadr::load_schedules(2012:season) |>
  filter(game_type == "REG") |>
  nflreadr::clean_homeaway() |>
  filter(team %in% c("OAK", "LV", "SD", "LAC", "STL", "LA"))

wins <- history |>
  group_by(season, team) |>
  summarise(wins = sum(team_score > opponent_score), .groups = "drop") |>
  mutate(franchise = clean_team_abbrs(team, sport = "nfl"))

moves <- data.frame(
  franchise = c("LV", "LAC", "LA"),
  moved = c(2020, 2017, 2016) # the first season in the new city
) |>
  mutate(label = paste0(nick(franchise), " move"))

ggplot(wins, aes(x = season, y = wins)) +
  geom_vline(aes(xintercept = moved - 0.5), data = moves, linetype = "dashed", color = "grey50") +
  geom_line(aes(color = franchise), linewidth = 0.8) +
  geom_sdv_logos(aes(team = team, season = season), sport = "nfl", height = 0.22) +
  geom_text(
    aes(x = moved - 0.6, y = 16.5, label = label),
    data = moves, hjust = 1, vjust = 1, size = 3, color = "grey40"
  ) +
  scale_color_sdv(sport = "nfl") +
  scale_x_continuous(breaks = seq(2012, season, 2)) +
  scale_y_continuous(limits = c(-1, 17), breaks = c(0, 4, 8, 12, 16)) +
  facet_wrap(~franchise, ncol = 1, labeller = as_labeller(\(x) nick(x))) +
  labs(
    title = "Three relocated franchises, 2012-2025",
    subtitle = "Regular-season wins; St. Louis and San Diego seasons in those identities' logos",
    x = NULL,
    y = "Wins",
    caption = "Data: nflverse via nflreadr | regular season"
  ) +
  theme(legend.position = "none", panel.grid.minor = element_blank())
```

## 9. A field in team colors

`sdv_surface("nfl", team)` draws a regulation field with
[sportyR](https://sportyr.sportsdataverse.org), its end zones in the team's
colors, and `center_logo = TRUE` puts the logo at midfield. sportyR's field is
centered on midfield in yards, so a line of scrimmage `yardline_100` yards from
the goal is at `x = 50 - yardline_100`. On top: every Seattle touchdown from
scrimmage, from the line of scrimmage into the end zone, placed by the side of
the field the play went to.

```{r field, fig.height = 4.6, fig.alt = "An NFL field with Seattle-colored end zones and the Seahawks logo at midfield, with one arrow per Seattle touchdown from scrimmage in the 2025 regular season running from the line of scrimmage into the right end zone, green for passes and white for runs."}
tds <- pbp |>
  filter(
    season_type == "REG", posteam == "SEA", td_team == "SEA",
    play_type %in% c("pass", "run")
  ) |>
  arrange(game_id, play_id) |>
  mutate(side = coalesce(pass_location, run_location))

set.seed(1) # the jitter that keeps arrows in one lane apart
lane <- c(left = 15, middle = 0, right = -15)
tds <- tds |>
  mutate(
    y = coalesce(lane[side], 0) + runif(n(), -6, 6),
    x = 50 - yardline_100
  )

sdv_surface("nfl", "SEA", center_logo = TRUE, display_range = "in_bounds_only") +
  geom_segment(
    aes(x = x, xend = 53, y = y, yend = y, color = play_type),
    data = tds,
    arrow = arrow(length = unit(0.12, "cm")), linewidth = 0.6
  ) +
  scale_color_manual(
    values = c(pass = unname(sdv_team_colors("nfl", "SEA", type = "secondary")), run = "white")
  ) +
  labs(
    title = sprintf("Seattle's %d touchdowns from scrimmage, 2025", nrow(tds)),
    subtitle = "From the line of scrimmage into the end zone: green arrows are passes, white arrows runs",
    caption = caption
  ) +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.caption = element_text(color = "grey45", size = 8),
    plot.margin = margin(8, 8, 8, 8),
    legend.position = "none"
  )
```

## 10. Team tiers

A tier list from net EPA per play (offense EPA per play minus defense EPA per
play allowed), drawn by `sdv_team_tiers()`. It needs only `tier_no` and `team`;
the order within each tier is the order of the rows.

```{r team-tiers, fig.height = 6, fig.alt = "NFL team logos in five labeled tiers, from Contenders to Rebuilding, ranked by net EPA per play in the 2025 regular season, best teams first within each tier."}
sizes <- c(5, 7, 7, 7, 6) # teams per tier, top to bottom

tiers <- teams |>
  mutate(net = off_epa - def_epa) |>
  arrange(desc(net), team) |>
  mutate(tier_no = rep(seq_along(sizes), sizes)) |>
  select(tier_no, team)

sdv_team_tiers(
  tiers,
  sport = "nfl",
  title = "NFL tiers by net EPA per play, 2025",
  subtitle = "Offense EPA per play minus defense EPA per play allowed",
  caption = caption,
  tier_desc = c(
    "1" = "Contenders",
    "2" = "Good",
    "3" = "Middle",
    "4" = "Flawed",
    "5" = "Rebuilding"
  )
)
```

## Next steps

- [nflfastR](https://www.nflfastr.com/) documents every play-by-play column used
  here, and [nflreadr](https://nflreadr.nflverse.com/) the other loaders.
- The [college football tutorial](cfb-viz.html) covers 130+ teams, conference
  logos and logos on the y axis.
- [Social Posting](social-posting.html) sizes and saves charts like these for
  social media.
