---
title: "NHL Visualizations with fastRhockey and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHL Visualizations with fastRhockey and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to create rich NHL visualizations by combining
[fastRhockey](https://fastRhockey.sportsdataverse.org) for player and team data with
[sdvplotR](https://sdvplotr.sportsdataverse.org) for team logos, headshots,
colors, and gt tables.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(fastRhockey)
library(dplyr)
library(gt)

# Get valid NHL team abbreviations
nhl_teams <- valid_team_names("nhl")
head(nhl_teams)
```

## Loading NHL Data

Use `fastRhockey` to load team and skater statistics from the NHL Stats API:

```{r load-data}
# fastRhockey keys seasons by their ending year: 2026 is 2025-26
season <- fastRhockey::most_recent_nhl_season()
season_id <- paste0(season - 1, season)

# One row per team; clean_team_abbrs() turns the NHL's full names into
# sdvplotR's abbreviations
team_stats <- fastRhockey::nhl_stats_teams(season = season_id) |>
  mutate(team_abbr = clean_team_abbrs(team_full_name, sport = "nhl"))

# Every skater (limit = -1), keyed by NHL API player ID, so the goal and assist
# leaderboards below rank the full list
player_stats <- fastRhockey::nhl_stats_skaters(season = season_id, limit = -1)
```

## NHL Team Performance

Plot every team's goals for against goals against, with team logos:

```{r team-performance}
ggplot(team_stats, aes(x = goals_for_per_game, y = goals_against_per_game)) +
  geom_sdv_logos(
    aes(team = team_abbr),
    sport = "nhl",
    width = 0.06
  ) +
  scale_y_reverse() +
  labs(
    title = "NHL Goals For and Against",
    subtitle = paste0("Season ", season - 1, "-", substr(season, 3, 4)),
    x = "Goals For per Game",
    y = "Goals Against per Game (fewer is better)",
    caption = "Data: fastRhockey | Viz: sdvplotR"
  ) +
  theme_minimal()
```

## NHL Team Colors

Use team colors to show points percentage:

```{r team-colors}
team_wins <- team_stats |>
  arrange(desc(point_pct)) |>
  head(20)

ggplot(team_wins, aes(x = reorder(team_abbr, point_pct), y = point_pct)) +
  geom_col(aes(fill = team_abbr), width = 0.7) +
  scale_fill_sdv(sport = "nhl", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Top 20 NHL Teams by Points Percentage",
    x = NULL,
    y = "Points Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Player Headshots

Plot the goal scoring leaders with their headshots:

```{r player-headshots}
top_goals <- player_stats |>
  slice_max(goals, n = 8, with_ties = FALSE)

ggplot(top_goals, aes(x = games_played, y = goals)) +
  geom_sdv_headshots(
    aes(player_id = player_id),
    sport = "nhl",
    id_type = "league",
    height = 0.15
  ) +
  geom_label(
    aes(label = skater_full_name),
    nudge_y = -2,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "NHL Goal Scoring Leaders",
    subtitle = paste0("Season ", season - 1, "-", substr(season, 3, 4)),
    x = "Games Played",
    y = "Goals"
  ) +
  theme_minimal()
```

### NHL API player IDs

The NHL's own APIs, which fastRhockey's `nhl_*()` functions and its
`load_nhl_*()` datasets come from, identify players by their NHL API ID
(McDavid is 8478402), not their ESPN athlete ID, which is why these headshots
pass `id_type = "league"`: it draws NHL API IDs from the NHL's image CDN. Leave
`id_type` unset for ESPN athlete IDs, such as those in fastRhockey's
`espn_nhl_*()` data. Both are plain
numbers, so the wrong setting draws someone else or nothing.

## NHL Team Tiers

Create a tier plot ranking NHL teams:

```{r team-tiers}
# Sample tier assignments
tier_data <- data.frame(
  tier_no = c(1, 1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5, 5, 5),
  team = c("COL", "BOS", "VGK", "TOR", "EDM", "CAR",
           "NYR", "DAL", "TB",
           "FLA", "NJ", "MIN",
           "LA", "SEA", "PIT")
)

sdv_team_tiers(
  tier_data,
  sport = "nhl",
  title = "NHL Power Rankings",
  subtitle = paste("As of", Sys.Date()),
  tier_desc = c(
    "1" = "Stanley Cup Favorites",
    "2" = "Contenders",
    "3" = "Playoff Teams",
    "4" = "Bubble Teams",
    "5" = "Rebuilding"
  )
)
```

## NHL Conference Map

Show each division's teams in a column, using the conference and division
sdvplotR keeps for every team:

```{r conference-map}
conference_map <- team_reference("nhl") |>
  mutate(conf_div = paste(conference, division)) |>
  arrange(conf_div, team_name) |>
  group_by(conf_div) |>
  mutate(team_rank = row_number()) |>
  ungroup() |>
  mutate(conf_div_num = as.numeric(factor(conf_div)))

ggplot(conference_map, aes(x = conf_div_num, y = team_rank)) +
  geom_sdv_logos(
    aes(team = team_abbr),
    sport = "nhl",
    width = 0.075
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(factor(conference_map$conf_div))),
    labels = levels(factor(conference_map$conf_div))
  ) +
  scale_y_reverse() +
  labs(
    title = "NHL Teams by Conference & Division",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_blank(),
    panel.grid = element_blank()
  )
```

## NHL Standings Table with Logos

Create a gt table with team logos:

```{r gt-table}
team_wins |>
  head(15) |>
  mutate(rank = row_number(), logo = team_abbr) |>
  select(rank, logo, team_full_name, wins, losses, ot_losses, points, point_pct) |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nhl", height = 35) |>
  fmt_number(columns = "point_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "",
    team_full_name = "Team",
    wins = "W",
    losses = "L",
    ot_losses = "OTL",
    points = "Pts",
    point_pct = "Pts%"
  ) |>
  tab_header(
    title = "NHL Top 15",
    subtitle = paste0("Season ", season - 1, "-", substr(season, 3, 4))
  )
```

## Player Performance Comparison

Compare the goal scoring leaders with the assist leaders:

```{r player-comparison}
comparison <- bind_rows(
  player_stats |>
    slice_max(goals, n = 5, with_ties = FALSE) |>
    transmute(player_id, category = "Goals", value = goals),
  player_stats |>
    slice_max(assists, n = 5, with_ties = FALSE) |>
    transmute(player_id, category = "Assists", value = assists)
) |>
  group_by(category) |>
  mutate(rank = row_number()) |>
  ungroup()

ggplot(comparison, aes(x = rank, y = value)) +
  geom_sdv_headshots(
    aes(player_id = player_id),
    sport = "nhl",
    id_type = "league",
    height = 0.2
  ) +
  facet_wrap(~category, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "NHL Top Performers",
    subtitle = paste0("Season ", season - 1, "-", substr(season, 3, 4)),
    x = "Rank",
    y = NULL
  ) +
  theme_minimal()
```

## Axis Labels with Logos

Replace axis labels with team logos (`theme_x_sdv()` needs the `ggtext`
package, and goes after any complete theme such as `theme_minimal()`):

```{r axis-logos}
top_8 <- team_wins |>
  head(8) |>
  mutate(team_abbr = factor(team_abbr, levels = team_abbr))

ggplot(top_8, aes(x = team_abbr, y = point_pct)) +
  geom_col(aes(fill = team_abbr), width = 0.6) +
  scale_fill_sdv(sport = "nhl", alpha = 0.7) +
  scale_x_sdv(sport = "nhl") +
  theme_minimal() +
  theme_x_sdv() +
  labs(
    title = "Top 8 NHL Teams by Points Percentage",
    x = NULL,
    y = "Points Percentage"
  ) +
  theme(legend.position = "none")
```

## Next Steps

- Explore [fastRhockey documentation](https://fastRhockey.sportsdataverse.org/)
- Try combining with [oddsapiR](https://oddsapiR.sportsdataverse.org) for betting lines
- Build weekly dashboard with [Quarto](https://quarto.org/)

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
