---
title: "NHL Visualizations with fastRhockey and sdvplotR"
description: >
  Ten worked examples on the 2025-26 NHL season: standings, goal differential,
  expected goals, a shot map on the rink, scoring and goaltending leaders, the
  Coyotes-to-Utah lineage, the points race and the playoff tiers.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NHL Visualizations with fastRhockey and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Ten charts and tables from the 2025-26 NHL season: goal differential in team
colors, the final standings, five-on-five expected goals, the champion's shot
map on a rink, finishing and goaltending against expected goals, the scoring
leaders with their headshots, one franchise's marks from Winnipeg to Utah, the
points race and the playoffs as a tier list. The data comes through
[fastRhockey](https://fastRhockey.sportsdataverse.org): its release files
(play-by-play with an expected-goals value on every unblocked shot, and box
scores), and the NHL's own APIs for the official standings and the franchise
records.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(gt)
```

## Load the season

The release files are read when this page is built. Each `load_nhl_*()` call
downloads one season from the SportsDataverse data releases; the columns kept
here are the ones the examples use.

```{r load-data}
season <- 2026 # the 2025-26 season, named for the year it ends

pbp <- fastRhockey::load_nhl_pbp_lite(seasons = season) |>
  select(
    game_id, season_type, period, event_type, event_team_abbr, event_team_type,
    home_abbr, away_abbr, strength_state, x_fixed, y_fixed, xg,
    event_goalie_id, event_goalie_name
  )

# game_id 2025020001 is a regular-season game (02); 2025030416 a playoff game
# (03): round 4, series 1, game 6
team_box <- fastRhockey::load_nhl_team_box(seasons = season) |>
  select(game_id, game_date, team_abbrev, goals, goals_against) |>
  mutate(game_type = game_id %/% 10000 %% 100)

skater_box <- fastRhockey::load_nhl_skater_box(seasons = season) |>
  filter(game_id %/% 10000 %% 100 == 2) |>
  select(game_id, team_abbrev, player_id, player_name, goals, assists, points)

c(plays = nrow(pbp), team_games = nrow(team_box), skater_games = nrow(skater_box))
```

The official standings come from api-web.nhle.com and the franchise records
from records.nhl.com:

```{r load-api, eval = FALSE}
standings <- fastRhockey::nhl_standings(date = "2026-04-16") |> # the last day of the regular season
  select(
    team_abbr, team_name, conference_name, division_name, division_sequence, games_played,
    wins, losses, ot_losses, points, point_pctg, regulation_wins, goals_for, goals_against,
    goal_differential, streak_code, streak_count
  )

# one franchise's every season: the Winnipeg Jets (1979), the Coyotes, Utah
lineage <- fastRhockey::nhl_records_franchise_season_results() |>
  filter(game_type_id == 2, team_id %in% c(33, 27, 53, 59, 68), season_id <= 20252026) |>
  select(season_id, team_id, tri_code, team_name, games_played, points)
```

```{r load-api-fixture, include = FALSE}
fx <- readRDS("fixtures/nhl-viz/nhl_api.rds")
standings <- fx$standings
lineage <- fx$lineage
```

These two calls are read from a snapshot taken on
`r format(fx$taken, "%B %d, %Y")` (`r fx$package`), because the NHL's APIs are
not called when this site is built; `data-raw/article_fixtures.R` takes it with
the same code.

A few of the NHL's three-letter codes differ from the ESPN abbreviations
sdvplotR keys its team colors on. `clean_team_abbrs()` converts them once, so
every join and color scale below agrees:

```{r clean-abbrs}
standings <- standings |>
  mutate(team = clean_team_abbrs(team_abbr, sport = "nhl"))

standings |>
  filter(team != team_abbr) |>
  select(team_name, team_abbr, team)

api <- "Data: NHL API via fastRhockey | Viz: sdvplotR"
release <- "Data: fastRhockey play-by-play and box score release | Viz: sdvplotR"
```

## 1. Goal differential in team colors

Bars in each team's color from `scale_fill_sdv()`, with the logo at the end of
each bar from `geom_sdv_logos()`.

```{r team-performance, fig.width = 9, fig.height = 5, fig.alt = "Bar chart of the 32 NHL teams' 2025-26 regular-season goal differential, sorted from best to worst, each bar in its team's color with the team logo at the end of the bar."}
goal_diff <- standings |>
  arrange(desc(goal_differential), team) |>
  mutate(
    team = factor(team, levels = team),
    logo_y = goal_differential + if_else(goal_differential >= 0, 9, -9)
  )

ggplot(goal_diff, aes(team, goal_differential)) +
  geom_col(aes(fill = team), width = 0.75) +
  geom_hline(yintercept = 0, colour = "grey20", linewidth = 0.4) +
  geom_sdv_logos(aes(y = logo_y, team = team), sport = "nhl", width = 0.027) +
  scale_fill_sdv(sport = "nhl") +
  labs(
    title = "NHL goal differential, 2025-26",
    subtitle = "Regular season, goals for minus goals against",
    x = NULL,
    y = "Goal differential",
    caption = api
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    axis.text.x = element_blank(),
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 2. Final standings by division

A gt table grouped by division, with `gt_sdv_logos()` for the logos and
`gt_theme_athletic()` for the look. The goal differential fill goes on after
the theme so the theme cannot replace it.

```{r standings-table}
nhl_divisions <- c("Atlantic", "Metropolitan", "Central", "Pacific")
lim <- max(abs(standings$goal_differential))

standings |>
  mutate(
    division = factor(division_name, levels = nhl_divisions),
    logo = team,
    streak = paste0(streak_code, streak_count)
  ) |>
  arrange(division, division_sequence) |>
  select(
    division, logo, team_name, games_played, wins, losses, ot_losses, points,
    point_pctg, regulation_wins, goals_for, goals_against, goal_differential, streak
  ) |>
  gt(groupname_col = "division") |>
  gt_sdv_logos(columns = "logo", sport = "nhl", height = 22) |>
  fmt_number(columns = point_pctg, decimals = 3) |>
  cols_align(align = "left", columns = team_name) |>
  cols_label(
    logo = "", team_name = "Team", games_played = "GP", wins = "W", losses = "L",
    ot_losses = "OTL", points = "PTS", point_pctg = "PTS%", regulation_wins = "RW",
    goals_for = "GF", goals_against = "GA", goal_differential = "DIFF", streak = "STRK"
  ) |>
  tab_header(
    title = "NHL standings, 2025-26",
    subtitle = "Final regular season, grouped by division"
  ) |>
  tab_source_note(api) |>
  gt_theme_athletic() |>
  data_color(
    columns = goal_differential,
    palette = c("#c84630", "white", "#2a7ab9"),
    domain = c(-lim, lim)
  )
```

## 3. Five-on-five expected goals

Expected goals (xG) weigh every unblocked shot by its chance of scoring, so
summed for and against each team at five-on-five they show who drives play,
with less luck in them than goals. The y axis is reversed, so the top right is
where you want to be; `geom_mean_lines()` marks the league averages.

```{r xg-scatter, fig.width = 7.5, fig.height = 6.5, fig.alt = "Scatter plot of the 32 NHL teams in 2025-26, each drawn as its logo, placed by five-on-five expected goals for per game (horizontal) and against per game (vertical, reversed), with dashed lines at the league averages."}
five <- pbp |>
  filter(season_type == "R", strength_state == "5v5", !is.na(xg)) |>
  mutate(defending = if_else(event_team_abbr == home_abbr, away_abbr, home_abbr))
xg_for <- five |>
  summarise(xgf = sum(xg), .by = event_team_abbr) |>
  rename(team_abbrev = event_team_abbr)
xg_against <- five |>
  summarise(xga = sum(xg), .by = defending) |>
  rename(team_abbrev = defending)
games_played <- team_box |>
  filter(game_type == 2) |>
  count(team_abbrev, name = "gp")

xg <- xg_for |>
  inner_join(xg_against, by = "team_abbrev") |>
  inner_join(games_played, by = "team_abbrev") |>
  mutate(
    xgf_pg = xgf / gp,
    xga_pg = xga / gp,
    team = clean_team_abbrs(team_abbrev, sport = "nhl")
  )
stopifnot(nrow(xg) == 32)

ggplot(xg, aes(xgf_pg, xga_pg)) +
  geom_mean_lines(aes(x0 = xgf_pg, y0 = xga_pg), colour = "grey60") +
  geom_sdv_logos(aes(team = team), sport = "nhl", width = 0.06) +
  scale_y_reverse() +
  labs(
    title = "Who drove play at five-on-five, 2025-26",
    subtitle = "Expected goals per game, regular season",
    x = "5v5 xG for per game",
    y = "5v5 xG against per game (reversed)",
    caption = release
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

## 4. The champion's shot map on the rink

`sdv_surface("nhl", team)` draws a regulation rink with 'sportyR', its center
line, center faceoff circle and boards in the team's colors, and the shots go
on top as ordinary layers. The champion comes from the data: the team with four
wins in round 4. In the release, `x_fixed` has the home team shooting right and
the away team left, so flipping the away shots (both x and y) puts every shot
in one attacking end. `ggtitle_image()` puts the logo in the title.

```{r shot-map, fig.width = 7.5, fig.height = 7.5, fig.alt = "Offensive half of a hockey rink in the Stanley Cup champion's colors, with every unblocked shot the champion took in 2025-26 as a faint grey dot and every goal as a dot in the team's color, clustered in front of the net."}
champion <- team_box |>
  filter(game_type == 3, game_id %/% 100 %% 10 == 4) |>
  summarise(wins = sum(goals > goals_against), .by = team_abbrev) |>
  filter(wins == 4) |>
  pull(team_abbrev)
champ <- clean_team_abbrs(champion, sport = "nhl")
champ_name <- standings$team_name[standings$team == champ]

shots <- pbp |>
  filter(
    event_team_abbr == champion,
    event_type %in% c("SHOT", "MISSED_SHOT", "GOAL"),
    !is.na(x_fixed)
  ) |>
  mutate(
    flip = if_else(event_team_type == "away", -1, 1),
    x = x_fixed * flip,
    y = y_fixed * flip,
    goal = event_type == "GOAL"
  )

sdv_surface("nhl", champ, display_range = "offense") +
  geom_point(aes(x, y), data = filter(shots, !goal), colour = "grey35", alpha = 0.2, size = 0.9) +
  geom_point(
    aes(x, y),
    data = filter(shots, goal), shape = 21, size = 2.2, stroke = 0.3,
    colour = "white", fill = sdv_team_colors("nhl", champ)
  ) +
  ggtitle_image(
    title_image = champ,
    title = paste(champ_name, "shots, 2025-26"),
    subtitle = paste0(
      format(sum(!shots$goal), big.mark = ","), " shots and misses (grey), ",
      sum(shots$goal), " goals (color); regular season and playoffs"
    ),
    image_height = 20,
    sport = "nhl"
  ) +
  labs(caption = release) +
  theme_title_image(size = 14, face = "bold") +
  # the rink layers leave the background transparent
  theme(plot.background = element_rect(fill = "white", colour = NA))
```

## 5. Finishing: goals above expected

Goals minus expected goals in all situations, shootouts left out. The release's
xG model was fit on earlier seasons and expects more goals than 2025-26
produced, so every team's xG is first scaled by the league's goals-to-xG ratio;
what is left is finishing relative to the league. `scale_x_sdv()` and
`theme_x_sdv()` put the logos on the axis.

```{r finishing, fig.width = 9, fig.height = 5, fig.alt = "Bar chart of the 32 NHL teams' 2025-26 goals above expected, sorted from best to worst finishers, each bar in its team's color with the team logo under it on the horizontal axis."}
attempts <- pbp |>
  filter(season_type == "R", !is.na(xg), period <= 4) # period 5 is the shootout
ratio <- sum(attempts$event_type == "GOAL") / sum(attempts$xg)

finishing <- attempts |>
  summarise(goals = sum(event_type == "GOAL"), xg = sum(xg) * ratio, .by = event_team_abbr) |>
  mutate(gax = goals - xg, team = clean_team_abbrs(event_team_abbr, sport = "nhl")) |>
  arrange(desc(gax), team) |>
  mutate(team = factor(team, levels = team))

ggplot(finishing, aes(team, gax, fill = team)) +
  geom_col(width = 0.75) +
  geom_hline(yintercept = 0, colour = "grey20", linewidth = 0.4) +
  scale_fill_sdv(sport = "nhl") +
  scale_x_sdv(sport = "nhl", size = 20) +
  labs(
    title = "Who finished their chances, 2025-26",
    subtitle = "Goals minus expected goals (scaled to the league), regular season",
    x = NULL,
    y = "Goals above expected",
    caption = release
  ) +
  theme_minimal(base_size = 12) +
  theme_x_sdv() +
  theme(
    legend.position = "none",
    panel.grid.major.x = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

```{r finishing-extremes, include = FALSE}
team_names <- setNames(standings$team_name, standings$team)
best <- finishing[1, ]
worst <- finishing[nrow(finishing), ]
```

The model expects `r round(100 * (1 / ratio - 1))`% more goals than the league
scored. After scaling, the `r team_names[[as.character(best$team)]]` finished
best (`r sprintf("%+.1f", best$gax)` goals) and the
`r team_names[[as.character(worst$team)]]` worst
(`r sprintf("%+.1f", worst$gax)`).

## 6. Scoring leaders with headshots

The release's box scores identify players by their NHL API ID (McDavid is
8478402), not their ESPN athlete ID, which is why the headshots pass
`id_type = "league"`: it draws NHL API IDs from the NHL's image CDN. Leave
`id_type` unset for ESPN athlete IDs, such as those in fastRhockey's
`espn_nhl_*()` data. Both are plain numbers, so the wrong setting draws someone
else or nothing. Goals and assists stack into points, and each player's team is
the one he played his last game for.

```{r points-leaders, fig.width = 8.5, fig.height = 6.5, fig.alt = "Horizontal stacked bar chart of the ten NHL points leaders of the 2025-26 regular season, goals in dark blue and assists in light blue, with each player's headshot at the left, his point total and his team logo at the end of the bar."}
leaders <- skater_box |>
  arrange(game_id) |>
  summarise(
    name = last(player_name), team_abbrev = last(team_abbrev),
    goals = sum(goals), assists = sum(assists), points = sum(points),
    .by = player_id
  ) |>
  filter(min_rank(desc(points)) <= 10) |>
  mutate(team = clean_team_abbrs(team_abbrev, sport = "nhl")) |>
  arrange(points, desc(name)) |>
  mutate(name = factor(name, levels = name))
bars <- bind_rows(
  transmute(leaders, name, stat = "Goals", n = goals),
  transmute(leaders, name, stat = "Assists", n = assists)
) |>
  mutate(stat = factor(stat, levels = c("Assists", "Goals"))) # goals stack from zero

ggplot(leaders, aes(y = name)) +
  geom_col(aes(x = n, fill = stat), data = bars, width = 0.7) +
  geom_sdv_headshots(
    aes(x = -12, player_id = player_id),
    sport = "nhl", id_type = "league", height = 0.085
  ) +
  geom_text(aes(x = points + 3, label = points), hjust = 0, fontface = "bold", size = 4) +
  geom_sdv_logos(aes(x = points + 21, team = team), sport = "nhl", height = 0.065) +
  scale_fill_manual(values = c(Goals = "#1f3b73", Assists = "#9fb4d8"), breaks = c("Goals", "Assists")) +
  scale_x_continuous(
    breaks = seq(0, 150, 25),
    limits = c(-20, max(leaders$points) + 28),
    expand = expansion(0)
  ) +
  labs(
    title = "NHL points leaders, 2025-26",
    subtitle = "Regular season",
    x = "Points",
    y = NULL,
    fill = NULL,
    caption = release
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 7. Goaltending: goals saved above expected

The play-by-play names the goalie in net for every unblocked shot, so the same
expected goals grade the goalies: goals saved above expected (GSAx) is the
scaled xG a goalie faced minus the goals he allowed. Among the goalies who
faced 1,000 unblocked shots, the five best and five worst, in bars of their
team's color (the team he faced his last shot for) with the headshot at the
end.

```{r goalies, fig.width = 8.5, fig.height = 6, fig.alt = "Diverging horizontal bar chart of goals saved above expected in the 2025-26 NHL regular season for the five best and five worst goalies who faced at least 1,000 unblocked shots, bars in team colors, with each goalie's headshot and value at the end of his bar."}
faced <- pbp |>
  filter(
    season_type == "R", event_type %in% c("SHOT", "MISSED_SHOT", "GOAL"),
    !is.na(xg), !is.na(event_goalie_id), period <= 4
  ) |>
  mutate(goalie_team = if_else(event_team_abbr == home_abbr, away_abbr, home_abbr))
ratio_faced <- sum(faced$event_type == "GOAL") / sum(faced$xg)

goalies <- faced |>
  arrange(game_id) |>
  summarise(
    shots = n(), goals = sum(event_type == "GOAL"), xga = sum(xg) * ratio_faced,
    team_abbrev = last(goalie_team),
    .by = c(event_goalie_id, event_goalie_name)
  ) |>
  mutate(gsax = xga - goals) |>
  filter(shots >= 1000)
ends <- goalies |>
  filter(min_rank(desc(gsax)) <= 5 | min_rank(gsax) <= 5) |>
  arrange(gsax) |>
  mutate(
    name = factor(event_goalie_name, levels = event_goalie_name),
    team = clean_team_abbrs(team_abbrev, sport = "nhl"),
    side = sign(gsax)
  )

ggplot(ends, aes(gsax, name)) +
  geom_col(aes(fill = team), width = 0.7) +
  geom_vline(xintercept = 0, colour = "grey20") +
  geom_sdv_headshots(
    aes(x = gsax + 3.5 * side, player_id = event_goalie_id),
    sport = "nhl", id_type = "league", height = 0.085
  ) +
  geom_text(
    aes(x = gsax + 7.5 * side, label = sprintf("%+.1f", gsax), hjust = if_else(side > 0, 0, 1)),
    fontface = "bold", size = 3.8
  ) +
  scale_fill_sdv(sport = "nhl") +
  scale_x_continuous(expand = expansion(add = 13)) +
  labs(
    title = "Goals saved above expected, 2025-26",
    subtitle = paste(nrow(goalies), "goalies faced 1,000+ unblocked shots; the five best and five worst, regular season"),
    x = "Goals saved above expected",
    y = NULL,
    caption = release
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(face = "bold")
  )
```

## 8. One franchise, many marks: Winnipeg, Phoenix, Arizona, Utah

The original Winnipeg Jets moved to Phoenix in 1996, became the Arizona Coyotes
in 2014, and their hockey operations moved to Salt Lake City in 2024 as the
Utah Hockey Club, the Utah Mammoth since 2025. sdvplotR maps the Coyotes'
codes to today's club, so a Coyotes-era dataset still gets a logo and a color,
but `WIN` matches today's Winnipeg Jets, the former Atlanta Thrashers. An old
code is not a reliable franchise key:

```{r old-codes}
clean_team_abbrs(c("WIN", "PHX", "ARI", "UTA"), sport = "nhl")
```

Given a `season`, `geom_sdv_logos()` draws the mark a club wore that season
instead. Historical NHL marks are keyed by the NHL's code for that identity
(`WIN`, `PHX`, `ARI`), and seasons are named for the year they end. One mark
per logo era, with the club's name that season from the NHL's records:

```{r relocation-marks, fig.width = 9, fig.height = 2.4, fig.alt = "A row of eight logos worn by one NHL franchise, left to right: two Winnipeg Jets marks, two Phoenix Coyotes marks, two Arizona Coyotes marks, the Utah Hockey Club mark and the Utah Mammoth mark, each labelled with its season and team name."}
marks <- tibble(
  key = c("WIN", "WIN", "PHX", "PHX", "ARI", "ARI", "UTA", "UTA"),
  season = c(1980, 1991, 1997, 2004, 2015, 2022, 2025, 2026)
) |>
  inner_join(transmute(lineage, season = season_id %% 10000, team_name), by = "season") |>
  mutate(
    x = row_number(),
    # the records write "Winnipeg Jets (1979)" to tell the two Jets apart
    label = paste0(season - 1, "-", substr(season, 3, 4), "\n", sub(" \\(\\d{4}\\)$", "", team_name))
  )

ggplot(marks, aes(x, 1)) +
  geom_sdv_logos(aes(team = key, season = season), sport = "nhl", width = 0.08) +
  geom_text(aes(y = 0.55, label = label), size = 3, lineheight = 0.95, vjust = 1) +
  scale_x_continuous(expand = expansion(add = 0.6)) +
  scale_y_continuous(limits = c(0.2, 1.4)) +
  labs(title = "The same franchise, by season") +
  theme_void(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), plot.margin = margin(8, 8, 8, 8))
```

The NHL's records list the Utah club as a new franchise from 2024-25, so its
two seasons are picked by team ID along with the earlier ones. The franchise's
points percentage across all 46 seasons, one shade per identity:

```{r relocation-points, fig.width = 9, fig.height = 4.8, fig.alt = "Line chart of the Winnipeg Jets, Phoenix Coyotes, Arizona Coyotes and Utah franchise's points percentage in every season from 1979-80 to 2025-26, on shaded bands for each identity with the identity's logo at the top of its band and a dashed line at .500."}
lineage <- lineage |>
  mutate(season = season_id %% 10000, pts_pct = points / (2 * games_played)) |>
  arrange(season)
bands <- lineage |>
  summarise(from = min(season) - 0.5, to = max(season) + 0.5, .by = tri_code) |>
  mutate(shade = row_number() %% 2 == 1, mid = (from + to) / 2, mark_season = to - 0.5)

ggplot(lineage, aes(season, pts_pct)) +
  geom_rect(
    aes(xmin = from, xmax = to, ymin = -Inf, ymax = Inf, fill = shade),
    data = bands, inherit.aes = FALSE
  ) +
  geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  geom_line(colour = sdv_team_colors("nhl", "UTAH"), linewidth = 0.9) +
  geom_point(colour = sdv_team_colors("nhl", "UTAH"), size = 1.6) +
  geom_sdv_logos(
    aes(x = mid, y = 0.8, team = tri_code, season = mark_season),
    data = bands, sport = "nhl", height = 0.13, inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c(`TRUE` = "grey93", `FALSE` = NA), guide = "none") +
  scale_x_continuous(breaks = c(1980, 1990, 2000, 2010, 2020, 2026)) +
  scale_y_continuous(labels = scales::label_number(accuracy = 0.001), limits = c(0.15, 0.87)) +
  labs(
    title = "Winnipeg to Phoenix to Arizona to Utah",
    subtitle = "Regular-season points percentage; each band's logo is the identity's last mark",
    x = "Season (named for the year it ends)",
    y = "Points percentage",
    caption = "Data: NHL records API via fastRhockey | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), plot.title = element_text(face = "bold"))
```

## 9. The points race, by division

Standings points game by game (two for a win, one for an overtime or shootout
loss), measured against a .500 pace of one point a game so the lines spread
apart instead of all climbing together. The box scores have the score, the
play-by-play says which games went past regulation, and the totals match the
official standings for every team.

Division rivals often share a color, so each line takes its team's primary
color from `sdv_team_colors()` unless that is too close to one already used in
its division, and then its secondary.

```{r points-race, fig.width = 9, fig.height = 7, fig.alt = "Four small line charts, one per NHL division, showing each team's standings points above a .500 pace game by game through the 2025-26 regular season, lines in team colors, with the team logo at the end of each line."}
last_period <- summarise(pbp, last_period = max(period), .by = game_id)
stopifnot(identical(class(team_box$game_id), class(last_period$game_id)))

race <- team_box |>
  filter(game_type == 2) |>
  inner_join(last_period, by = "game_id") |>
  mutate(pts = case_when(goals > goals_against ~ 2, last_period > 3 ~ 1, .default = 0)) |>
  arrange(game_date, game_id) |>
  mutate(game = row_number(), points = cumsum(pts), .by = team_abbrev) |>
  inner_join(
    select(standings, team_abbr, team, division_name, division_sequence, official = points),
    by = c(team_abbrev = "team_abbr")
  ) |>
  mutate(above = points - game, division = factor(division_name, levels = nhl_divisions))
final <- filter(race, game == 82)
stopifnot(nrow(final) == 32, all(final$points == final$official))

# the primary, unless it is within 20 (CIE 2000 color difference) of the white
# background or of a color already taken, and the secondary is farther away
distinct_colours <- function(primary, secondary, min_gap = 20) {
  taken <- "#FFFFFF"
  gap <- function(x) {
    min(farver::compare_colour(
      farver::decode_colour(x), farver::decode_colour(taken), "rgb",
      method = "cie2000"
    ))
  }
  for (i in seq_along(primary)) {
    swap <- gap(primary[i]) < min_gap && gap(secondary[i]) > gap(primary[i])
    taken <- c(taken, if (swap) secondary[i] else primary[i])
  }
  taken[-1]
}
line_colours <- final |>
  arrange(division, division_sequence) |>
  inner_join(sdv_team_colors("nhl", type = "all"), by = c(team = "team_abbr")) |>
  mutate(colour = distinct_colours(primary, secondary), .by = division)

# teams that finish close together would stack their logos, so spread the logos
# at least `gap` points apart (keeping their order and average) and draw a
# connector from each line's end to its logo
spread <- function(y, gap) {
  o <- order(y)
  s <- y[o]
  for (i in seq_along(s)[-1]) s[i] <- max(s[i], s[i - 1] + gap)
  y[o] <- s - mean(s) + mean(y)
  y
}
final <- mutate(final, logo_y = spread(above, gap = 6), .by = division)

ggplot(race, aes(game, above, colour = team)) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_line(linewidth = 0.8) +
  geom_segment(aes(x = 82, xend = 86, yend = logo_y), data = final, linewidth = 0.3) +
  # inherit.aes = FALSE: a colour aesthetic would tint the logos
  geom_sdv_logos(
    aes(x = 89, y = logo_y, team = team),
    data = final, sport = "nhl", height = 0.075, inherit.aes = FALSE
  ) +
  scale_colour_manual(values = setNames(line_colours$colour, line_colours$team)) +
  scale_x_continuous(breaks = c(1, 20, 40, 60, 82), limits = c(1, 92)) +
  facet_wrap(~division, ncol = 2) +
  labs(
    title = "The 2025-26 points race, by division",
    subtitle = "Standings points above a .500 pace (one point a game)",
    x = "Game",
    y = "Points above .500 pace",
    caption = release
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold")
  )
```

## 10. The playoffs as a tier list

How far every team went, from the playoff game IDs (the seventh digit is the
round), with regular-season points deciding the order within a tier.
`sdv_team_tiers()` returns a ggplot, so a light background goes on top of its
dark one, where the black logos would disappear.

```{r playoff-tiers, fig.width = 10, fig.height = 7, fig.alt = "Tier list of the 32 NHL teams in 2025-26, drawn as logos in six rows: the Stanley Cup champion, the other finalist, the conference finalists, the second-round teams, the first-round teams and the teams that missed the playoffs."}
reached <- team_box |>
  filter(game_type == 3) |>
  summarise(round = max(game_id %/% 100 %% 10), .by = team_abbrev)

tiers <- standings |>
  left_join(reached, by = c(team_abbr = "team_abbrev")) |>
  mutate(tier_no = case_when(
    team_abbr == champion ~ 1,
    is.na(round) ~ 6,
    .default = 6 - round
  )) |>
  arrange(tier_no, desc(points), team) |>
  mutate(tier_rank = row_number(), .by = tier_no)
stopifnot(sum(tiers$tier_no < 6) == 16)

sdv_team_tiers(
  select(tiers, tier_no, tier_rank, team),
  sport = "nhl",
  title = "The 2026 Stanley Cup playoffs: how far everyone got",
  subtitle = paste(champ_name, "won the Cup; teams ordered by regular-season points within each tier"),
  caption = release,
  tier_desc = c(
    "1" = "Champion",
    "2" = "Lost the Final",
    "3" = "Conference final",
    "4" = "Second round",
    "5" = "First round",
    "6" = "Missed the playoffs"
  ),
  width = 0.05
) +
  theme(
    plot.background = element_rect(fill = "white", colour = "white"),
    panel.background = element_rect(fill = "white", colour = "white"),
    plot.title = element_text(colour = "grey10", face = "bold"),
    plot.subtitle = element_text(colour = "grey35"),
    plot.caption = element_text(colour = "grey35"),
    axis.text.y = element_text(colour = "grey10", face = "bold")
  )
```

## Related articles

- [Getting Started](getting-started.html)
- [MLB Visualizations](mlb-viz.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
