---
title: "WNBA Scoring Leaders Card"
description: >
  Make a WNBA scoring-leaders card with headshots, team logos and team-color
  bars, exported at 1080 x 1080 for Instagram and as a top-five 1200 x 675 cut
  for X.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WNBA Scoring Leaders Card}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7, fig.height = 5, dpi = 96
)
```

**The brief:** the regular season just ended, and the social team wants a
scoring-leaders card: the top ten in points per game with each player's face
and team, as a 1080 x 1080 image for Instagram, plus a top-five cut at
1200 x 675 for X and Bluesky. The box scores are ESPN's through wehoop; the
headshots come from ESPN's CDN through `geom_sdv_headshots()`.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2026
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

One row per player per game. ESPN files the All-Star Game and the Commissioner's
Cup final as regular-season games, but neither counts in the official stats;
the schedule marks the standard games with `type_abbreviation` `"STD"`, so a
semi join keeps only those. Players who did not play are dropped, the qualifier
is 30 games (about 70% of the 44-game schedule), and a player traded mid-season
is listed with her last team.

```{r data}
regular <- wehoop::load_wnba_schedule(season) |> filter(season_type == 2)
count(regular, type_abbreviation) # STD, plus the All-Star Game and the Cup final
standard <- regular |>
  filter(type_abbreviation == "STD") |>
  select(game_id)

box <- wehoop::load_wnba_player_box(season)
stopifnot(identical(class(box$game_id), class(standard$game_id))) # one type on both sides of the key

leaders <- box |>
  semi_join(standard, by = "game_id") |>
  filter(!did_not_play) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    games = n(),
    ppg = mean(points),
    team = last(team_abbreviation, order_by = game_date),
    .groups = "drop"
  ) |>
  filter(games >= 30) |>
  arrange(desc(ppg), athlete_display_name) |> # the name breaks a tie the same way every run
  slice_head(n = 10) |>
  mutate(rank = row_number())
leaders
```

## 2. The first draft

A horizontal bar chart is the right shape for a ranked list of names.

```{r draft, fig.alt = "First draft: grey horizontal bars of points per game for the top ten WNBA scorers, in alphabetical order with no values."}
ggplot(leaders, aes(ppg, athlete_display_name)) +
  geom_col()
```

The leader is buried (ggplot2 sorts a text axis alphabetically, bottom up),
every bar is the same grey, the exact values are missing, and nothing says what
or when.

## 3. Order, team colors and values

`reorder()` sorts the names by scoring, so No. 1 is on top. Each bar takes its
team's color from `sdv_team_colors()`, with one catch: a few primaries (the Aces'
silver, the Liberty's seafoam) are too light to carry white text or to stand out
on a light card, so a small WCAG luminance check swaps those to the team's
secondary color. The values go at the end of each bar, which makes the x axis
unnecessary.

```{r colors, fig.alt = "The top ten WNBA scorers as horizontal bars sorted by points per game, each in a team color, with the value printed at the end of each bar."}
luminance <- function(hex) {
  rgb <- grDevices::col2rgb(hex) / 255
  rgb <- ifelse(rgb <= 0.03928, rgb / 12.92, ((rgb + 0.055) / 1.055)^2.4)
  drop(c(0.2126, 0.7152, 0.0722) %*% rgb) # WCAG relative luminance, 0 (black) to 1 (white)
}

primary <- sdv_team_colors("wnba", leaders$team)
secondary <- sdv_team_colors("wnba", leaders$team, type = "secondary")
leaders <- leaders |>
  mutate(color = unname(if_else(luminance(primary) < 0.3, primary, secondary)))

ggplot(leaders, aes(ppg, reorder(athlete_display_name, ppg))) +
  geom_col(aes(fill = color), width = 0.7) +
  geom_text(aes(label = sprintf("%.1f", ppg)), hjust = -0.2, fontface = "bold", size = 3.4) +
  scale_fill_identity() +
  scale_x_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_blank(), panel.grid = element_blank())
```

## 4. Faces and logos

A card like this sells on faces. Moving to a blank canvas (an x axis fixed at 0
to 100, one unit per row on a reversed y axis) makes room for a column of
headshots and a small team logo under each name. `geom_sdv_headshots()` takes
ESPN athlete ids as they come, and both image geoms size their images with
`height`, a share of the panel's height, so with ten rows a headshot of 0.085
fills most of a row.

```{r faces, fig.alt = "The top ten scorers on a blank canvas: rank, headshot, name, a small team logo with the team abbreviation, and a team-color bar with the value inside it."}
rows <- function(data, bar_from = 47, bar_to = 95) {
  n <- nrow(data)
  scale <- (bar_to - bar_from) / max(data$ppg)
  ggplot(data, aes(y = rank)) +
    geom_text(aes(x = 3, label = rank), size = 5.5, fontface = "bold", color = "grey60") +
    geom_sdv_headshots(aes(x = 12, player_id = athlete_id), sport = "wnba", height = 0.85 / n) +
    geom_text(aes(x = 20, y = rank - 0.17, label = athlete_display_name), hjust = 0, size = 4, fontface = "bold") +
    geom_sdv_logos(aes(x = 21.6, y = rank + 0.2, team = team), sport = "wnba", height = 0.36 / n) +
    geom_text(aes(x = 24, y = rank + 0.2, label = team), hjust = 0, size = 3, color = "grey40") +
    geom_rect(
      aes(xmin = bar_from, xmax = bar_from + ppg * scale, ymin = rank - 0.28, ymax = rank + 0.28, fill = color)
    ) +
    geom_text(
      aes(x = bar_from + ppg * scale - 1, label = sprintf("%.1f", ppg)),
      hjust = 1, size = 4, fontface = "bold", color = "white"
    ) +
    scale_fill_identity() +
    scale_y_reverse() +
    coord_cartesian(xlim = c(0, 100), ylim = c(0.5, n + 0.5), expand = FALSE) +
    theme_void()
}

rows(leaders)
```

## 5. Make it a card

The finishing pass is the frame: a warm off-white background, the headline as
the title (the stat goes in the subtitle), the qualifier stated, and a footer
with the source. The frame is a theme on top of the rows, so the same function
draws the square post and a wider one.

```{r card, fig.width = 6, fig.height = 6, fig.alt = "The finished scoring-leaders card: an off-white square with the headline, the top ten scorers with headshots, logos and team-color bars, and a source footer."}
bg <- "#f6f4ef"
leader <- leaders[1, ]

card <- function(data) {
  rows(data) +
    labs(
      title = paste(leader$athlete_display_name, "led the WNBA in scoring"),
      subtitle = paste0("Points per game, ", season, " WNBA regular season (minimum 30 games)"),
      caption = "Data: ESPN via wehoop  |  #WNBA  |  made with sdvplotR"
    ) +
    theme(
      plot.background = element_rect(fill = bg, color = NA),
      plot.title = element_text(face = "bold", size = 19, color = "#1d1d1d", margin = margin(b = 4)),
      plot.subtitle = element_text(size = 10.5, color = "grey40", margin = margin(b = 10)),
      plot.caption = element_text(size = 8, color = "grey40", hjust = 0, margin = margin(t = 8)),
      plot.margin = margin(20, 20, 14, 20)
    )
}

card(leaders)
```

## 6. Export for Instagram and X

The square holds all ten. The 16:9 post for X and Bluesky is too short for ten
readable rows, so it gets the top five from the same function: changing the
content to fit the format beats shrinking the type. Inches times `dpi` gives the
exact pixels, and `bg` matches the card so no white edge shows.

```{r export, fig.alt = "The scoring-leaders card exported at 1080 by 1080 pixels for Instagram."}
square <- file.path(out_dir, "wnba_scoring_1080x1080.png")
wide <- file.path(out_dir, "wnba_scoring_top5_1200x675.png")
ggsave(square, card(leaders), width = 7.2, height = 7.2, dpi = 150, bg = bg)
ggsave(wide, card(head(leaders, 5)), width = 8, height = 4.5, dpi = 150, bg = bg)

magick::image_info(magick::image_read(c(square, wide)))[c("width", "height")]
magick::image_read(square)
```

The top-five cut for X and Bluesky:

```{r export-wide, fig.alt = "The top-five version of the card exported at 1200 by 675 pixels for X and Bluesky."}
magick::image_read(wide)
```
