---
title: "Rolling Form"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rolling Form}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A season average is slow to move. A team's form changes faster: the offense
that struggled in September can look different over its last two games or so.
This recipe compares each team's last 150 offensive plays with the 150 before
them, then lists the five biggest risers and the five biggest fallers.

There is no NBA rolling-window release yet, so this recipe is college football
only.

```{r setup, message = FALSE}
library(sdvplotR)
library(gt)
library(dplyr, warn.conflicts = FALSE)
```

## The data

The `cfb_rolling_windows` release has one file per season and no R loader, so
read it by URL. `nflreadr::load_from_url()` reads the `.rds` copy with packages
the site already uses. The `.parquet` and `.csv` copies hold the same rows.

```{r data}
url <- paste0(
  "https://github.com/sportsdataverse/sportsdataverse-data/releases/download/",
  "cfb_rolling_windows/rolling_windows_2026.rds"
)
windows <- as.data.frame(nflreadr::load_from_url(url))
stopifnot(nrow(windows) > 0)
dim(windows)
names(windows)
```

Each row is one player or team, one metric and one window size.

- `cur` is the metric over the last `window_n` events (plays, carries,
  dropbacks or targets, named in `window_unit`).
- `prev` is the metric over the `window_n` events before those.
- `delta_prev` is `cur - prev`, and `delta_prev_rank` ranks it (1 is the
  biggest rise).
- `n` is how many events the last window holds. A row is `qualified` once that
  window is full.

```{r shape}
count(windows, entity_type, metric, window_unit, window_n)
```

This file is the season in progress. The date shows how current it is:

```{r as-of}
as_of <- unique(windows$as_of_date)
stopifnot(length(as_of) == 1)
as_of
```

## Risers and fallers

Keep teams, EPA per play and the 150-play window, and only qualified rows that
also have a previous window: a program with no earlier plays has no change to
rank. The percentile places each team's last 150 plays among every team in the
same pool, so a big riser can still be below average.

```{r pool}
pool <- windows |>
  filter(
    entity_type == "team", metric == "epa", window_n == 150,
    qualified, !is.na(prev)
  ) |>
  mutate(pct = 100 * percent_rank(cur))
stopifnot(all(pool$n == 150), !anyDuplicated(pool$entity_id))
nrow(pool)

movers <- bind_rows(
  Risers = slice_max(pool, delta_prev, n = 5, with_ties = FALSE),
  Fallers = slice_min(pool, delta_prev, n = 5, with_ties = FALSE),
  .id = "group"
) |>
  select(group, entity_name, prev, cur, n, pct, delta_prev_rank)
movers
```

The pool holds every team the release covers, FCS programs as well as FBS.

## The table

`gt_delta()` computes the change from the two rate columns, so the table shows
the arithmetic it ranks. Every rate shows its sample: the earlier window is 150
plays by definition, and the last window's `n` sits next to its rate. The
change column's green and red are not the only cue, since every change also
carries its sign. Pass `color_positive` and `color_negative` to `gt_delta()`
for a pair that does not rely on red-green vision.

```{r table}
movers |>
  gt(groupname_col = "group") |>
  gt_sdv_logos(entity_name, sport = "cfb", height = 22, include_name = TRUE) |>
  gt_delta(prev, cur, column_label = "Change", decimals = 3) |>
  fmt_number(c(prev, cur), decimals = 3) |>
  cols_merge(c(cur, n), pattern = "{1} (n = {2})") |>
  gt_percentile_bar(pct, width = 160) |>
  cols_label(
    entity_name = "Team",
    prev = "Previous 150 plays",
    cur = "Last 150 plays",
    pct = "Last 150, percentile",
    delta_prev_rank = "Change rank"
  ) |>
  tab_header(
    title = "Offensive form: last 150 plays vs the 150 before",
    subtitle = paste("EPA per play, 2026 season, as of", as_of)
  ) |>
  tab_source_note("Data: sportsdataverse-data release cfb_rolling_windows") |>
  gt_theme_sdv(density = "compact")
```
