---
title: "SportsDataverse Table Themes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SportsDataverse Table Themes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

sdvplotR puts team logos, headshots and colors inside `gt` tables. The table
around them still needs a look, and two themes are built for that:

- `gt_theme_sdv()` is the SportsDataverse house table, in light or dark.
- `gt_theme_sdv_team()` dresses the same table in one team's colors.

Both follow the same rules. Titles and column labels are set in Chivo, and the
body in Lato with tabular figures, so digits line up down a column. Column
labels keep the case you give them. The one accent is a line under the column
labels, so logos and team colors stay the loudest thing in the table.

```{r setup, message = FALSE}
library(sdvplotR)
library(gt)
```

## The data

The final 2023 AFC West standings, typed in so this article builds offline.
The logos come from `gt_sdv_logos()`, which resolves the abbreviations through
`clean_team_abbrs()`.

```{r data}
standings <- data.frame(
  team = c("KC", "LV", "DEN", "LAC"),
  name = c("Kansas City Chiefs", "Las Vegas Raiders", "Denver Broncos", "Los Angeles Chargers"),
  w = c(11, 8, 8, 5),
  l = c(6, 9, 9, 12)
)
standings$pct <- standings$w / (standings$w + standings$l)

standings_table <- gt(standings) |>
  gt_sdv_logos(columns = "team", sport = "nfl", height = 26) |>
  cols_label(team = "", name = "Team", w = "W", l = "L", pct = "Win pct") |>
  # win percentage without the leading zero, as box scores print it (.647)
  fmt(columns = "pct", fns = function(x) sub("^0", "", sprintf("%.3f", x))) |>
  tab_header("AFC West standings", "2023 regular season") |>
  tab_source_note("Data: nflverse")
```

## The house theme

```{r house}
standings_table |> gt_theme_sdv()
```

The line under the column labels is the SportsDataverse gradient, the same one
in the package logo. It is the only decoration, drawn once across the whole
table rather than per cell.

## Dark

`style = "dark"` sets the same table on the SportsDataverse navy. It matches the
dark mode of this site, and suits posts that sit on a dark background.

```{r dark}
standings_table |> gt_theme_sdv(style = "dark")
```

## A team's colors

`gt_theme_sdv_team()` takes a team and a sport. It uses the colors
`sdv_team_colors()` returns: the title block is filled with the primary color,
the line under the labels uses the secondary color, and the labels are set in
the primary color. Any abbreviation, alias or full name `clean_team_abbrs()`
accepts works as `team`.

```{r team}
data.frame(
  player = c("Patrick Mahomes", "Travis Kelce", "Isiah Pacheco"),
  stat = c("Passing", "Receiving", "Rushing"),
  yards = c(4183, 984, 935)
) |>
  gt() |>
  cols_label(player = "Player", stat = "", yards = "Yards") |>
  fmt_number(columns = "yards", decimals = 0) |>
  tab_header("Chiefs yardage leaders", "2023 regular season") |>
  gt_theme_sdv_team(team = "KC", sport = "nfl")
```

### Colors that would not read

Team colors are chosen for jerseys, not for text on a white table, so the theme
checks contrast rather than using the colors as they come:

- Title text on the filled block is black or white, whichever contrasts more
  with the primary color.
- A secondary color too pale to see on white (Duke's is white) gives way to the
  primary color for the line.
- A primary color too light to read as text (the Saints' gold) gives way to the
  SportsDataverse navy for the column labels.

Each table below shows the team's own colors as `sdv_team_colors()` returns
them, dressed in those colors.

```{r fallbacks}
color_record <- function(sport, team) {
  data.frame(
    team = team,
    primary = unname(sdv_team_colors(sport, team, "primary")),
    secondary = unname(sdv_team_colors(sport, team, "secondary"))
  )
}

color_record("mbb", "DUKE") |>
  gt() |>
  tab_header("Duke Blue Devils", "A white secondary: the line takes the primary") |>
  gt_theme_sdv_team(team = "DUKE", sport = "mbb")

color_record("nfl", "NO") |>
  gt() |>
  tab_header("New Orleans Saints", "A gold primary: the labels go navy") |>
  gt_theme_sdv_team(team = "NO", sport = "nfl")
```

With `team = NULL` the table wears the SportsDataverse navy and cyan.

## Sizing for where it goes

Every theme takes `density`. `"comfortable"` is the default, `"compact"`
tightens type and row padding for long tables, and `"social"` scales both up to
the size `gt_save_crop()` and `gt_social_crop()` export at.

```{r compact}
standings_table |> gt_theme_sdv(density = "compact")
```

## Changing one thing

Anything in `...` goes to `gt::tab_options()` after the theme's own options, so
it overrides them. Center the heading and keep everything else:

```{r override}
standings_table |> gt_theme_sdv(heading.align = "center")
```

## Saving an image

`theme_bg` records the background each theme paints. Pass it to
`gt_save_crop()` so the padding around the saved image matches the table
instead of framing it in white.

```{r save, eval = FALSE}
bg <- theme_bg$bg[theme_bg$theme == "gt_theme_sdv" & theme_bg$has_style == "dark"]

standings_table |>
  gt_theme_sdv(style = "dark", density = "social") |>
  gt_save_crop("afc-west.png", bg = bg)
```

To see these next to the 18 other themes in the package on your own data, run
`gt_theme_preview(standings)`.
