---
title: "Women's College Basketball Visualizations with wehoop and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Women's College Basketball Visualizations with wehoop and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Division I women's basketball has more than 360 teams, and each shares its
ESPN id with the school's men's team. These ten examples chart the 2025-26
season: one id for two programs, the efficiency landscape of the best teams, a
conference table, conference strength with conference logos, a headshot
leaderboard, a shot chart on a college court, a ranked top 25, seed-line
tiers, the champion's tournament run and a conference race as a bump chart.
The data are [wehoop](https://wehoop.sportsdataverse.org)'s ESPN release
files on GitHub (`load_wbb_*()`): box scores, standings, schedules, player
box scores and shots. Nothing here calls stats.ncaa.org.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)

season <- 2026 # the 2025-26 season: wehoop names a season by the year it ends
label <- "2025-26"
source_note <- "Data: wehoop (ESPN) | Viz: sdvplotR"

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      plot.caption = element_text(color = "grey40", size = 8),
      plot.caption.position = "plot",
      panel.grid.minor = element_blank()
    )
)
```

## 1. One school id, two programs

ESPN gives a school one team id for its men's and women's teams. sdvplotR
keeps them as separate sports, so the `sport` argument picks the program:
2633 is the Tennessee Volunteers in `"mbb"` and the Lady Volunteers in
`"wbb"`, each with its own name and mark.

```{r one-id}
wbb_teams <- team_reference("wbb")
nrow(wbb_teams)

bind_rows(team_reference("mbb"), wbb_teams) |>
  filter(espn_team_id == 2633) |>
  select(sport, espn_team_id, team_abbr, team_name)
```

```{r one-id-logos, fig.width = 6, fig.height = 3, fig.alt = "The two Tennessee logos sdvplotR draws for ESPN team id 2633: the men's Volunteers mark for sport mbb on the left and the women's Lady Volunteers mark for sport wbb on the right."}
tenn <- data.frame(x = c(1, 2), y = 1, team = "TENN", label = c('sport = "mbb"', 'sport = "wbb"'))

ggplot(tenn, aes(x, y)) +
  geom_sdv_logos(aes(team = team), data = tenn[1, ], sport = "mbb", height = 0.6) +
  geom_sdv_logos(aes(team = team), data = tenn[2, ], sport = "wbb", height = 0.6) +
  geom_text(aes(y = 0.55, label = label), family = "mono", size = 4) +
  scale_x_continuous(limits = c(0.4, 2.6)) +
  scale_y_continuous(limits = c(0.45, 1.4)) +
  labs(title = "One ESPN team id (2633), two marks") +
  theme_void() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

Names resolve through the reference, nicknames do not, and ESPN ids are not
keys (join on `espn_team_id` instead):

```{r names}
clean_team_abbrs(
  c("Tennessee", "Tennessee Lady Volunteers", "Lady Vols", "2633"),
  sport = "wbb",
  keep_non_matches = FALSE
)
```

## 2. The efficiency landscape of the top 36

Box scores are enough for a simple rating: points scored and allowed per 100
possessions, counting only regular-season games between two Division I teams,
with possessions estimated as FGA - OREB + TOV + 0.475 x FTA (the usual
college weight on free throws) and averaged with the opponent's. It is a raw
rating, not adjusted for schedule. Each team's conference comes from the
season's standings; the full name resolves to the conference's own key
(`"American Conference"`, not `"American"`, which is American University).

```{r ratings}
standings <- wehoop::load_wbb_standings(seasons = season)
conferences <- standings |>
  distinct(team_id, conference = group_short_name, group_name) |>
  mutate(conference_key = clean_team_abbrs(group_name, sport = "wbb")) |>
  select(-group_name)
stopifnot(!anyDuplicated(conferences$team_id))

d1 <- wbb_teams$espn_team_id
box <- wehoop::load_wbb_team_box(seasons = season) |>
  filter(season_type == 2, team_id %in% d1, opponent_team_id %in% d1) |>
  mutate(
    poss = field_goals_attempted - offensive_rebounds +
      coalesce(total_turnovers, turnovers) + 0.475 * free_throws_attempted
  )
box <- box |>
  inner_join(
    box |> select(game_id, opponent_team_id = team_id, opp_poss = poss),
    by = c("game_id", "opponent_team_id")
  ) |>
  mutate(game_poss = (poss + opp_poss) / 2, margin = team_score - opponent_team_score)

ratings <- box |>
  group_by(team_id) |>
  summarise(
    games = n(),
    ortg = 100 * sum(team_score) / sum(game_poss),
    drtg = 100 * sum(opponent_team_score) / sum(game_poss),
    .groups = "drop"
  ) |>
  mutate(net = ortg - drtg) |>
  inner_join(select(wbb_teams, team_id = espn_team_id, team = team_abbr, team_name), by = "team_id") |>
  left_join(conferences, by = "team_id") |>
  arrange(desc(net), team)
nrow(ratings)
```

One way to keep logos legible is to zoom to the teams you care about: here the
axes hold only the 36 best teams by net rating, so every logo gets room. The
subtitle gives the Division I averages, far below and left of every logo.
Defense is points allowed, so its axis is reversed to put good defenses on top.

```{r team-performance, fig.width = 9, fig.height = 6, fig.alt = "The 36 best Division I women's basketball teams of 2025-26 by net rating drawn as logos, placed by offensive rating (horizontal) and defensive rating (vertical, reversed so better defenses are higher)."}
top36 <- head(ratings, 36)

ggplot(top36, aes(ortg, drtg)) +
  geom_sdv_logos(aes(team = team), sport = "wbb", width = 0.05) +
  scale_y_reverse(expand = expansion(mult = 0.08)) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  labs(
    title = paste("The efficiency landscape of the top 36,", label),
    subtitle = sprintf(
      "Division I averages: %.1f offense, %.1f defense; games between Division I teams only",
      mean(ratings$ortg), mean(ratings$drtg)
    ),
    x = "Offensive rating (points per 100 possessions)",
    y = "Defensive rating (allowed per 100)",
    caption = source_note
  )
```

## 3. A conference table

The SEC's 2025-26 standings, with home and road records and the record against
AP-ranked teams. College standings repeat every stat for each split (home,
road, conference, against ranked teams), so the unique key to pivot on is
`stat_type`. `gt_color_pills()` with a fixed `domain` keeps the margin colors
comparable from one table to the next.

```{r standings-table}
standings |>
  filter(group_short_name == "SEC") |>
  select(team_id, stat_type, display_value) |>
  pivot_wider(names_from = stat_type, values_from = display_value) |>
  inner_join(select(ratings, team_id, team, team_name), by = "team_id") |>
  mutate(
    seed = as.integer(playoffseed),
    logo = team,
    margin = as.numeric(avgpointsfor) - as.numeric(avgpointsagainst)
  ) |>
  arrange(seed, team) |>
  select(seed, logo, team_name, total, vsconf, home, road, vsaprankedteams, margin) |>
  gt() |>
  tab_header(
    title = paste("SEC standings,", label),
    subtitle = "Margin is points per game over the whole season"
  ) |>
  cols_label(
    seed = "", logo = "", team_name = "Team", total = "Overall", vsconf = "Conf",
    home = "Home", road = "Road", vsaprankedteams = "vs AP 25", margin = "Margin"
  ) |>
  tab_source_note(source_note) |>
  gt_sdv_logos(columns = logo, sport = "wbb", height = 26) |>
  gt_color_pills(margin, digits = 1, domain = c(-40, 40)) |>
  gt_theme_athletic() |>
  cols_align("left", columns = team_name)
```

## 4. Conference strength, with conference logos

Conferences are rows in sdvplotR too, with logos and colors, so a chart of
conferences works the same way as a chart of teams: `scale_y_sdv()` draws the
conference marks on the axis and `scale_fill_sdv()` fills each bar in the
conference's color. Each bar is the average net rating of the conference's
teams.

```{r conference-strength, fig.height = 9, fig.alt = "Horizontal bar chart of the 31 Division I women's basketball conferences of 2025-26 by the average net rating of their teams, strongest at the top, each bar in the conference's color with the conference logo on the vertical axis."}
ratings |>
  filter(!is.na(conference_key)) |>
  group_by(conference_key) |>
  summarise(teams = n(), net = mean(net)) |>
  arrange(desc(net), conference_key) |>
  mutate(conference_key = factor(conference_key, levels = rev(conference_key))) |>
  ggplot(aes(net, conference_key, fill = conference_key)) +
  geom_col(width = 0.75, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  geom_text(
    aes(label = sprintf("%+.1f", net), hjust = if_else(net > 0, -0.15, 1.15)),
    size = 3.2
  ) +
  scale_fill_sdv(sport = "wbb") +
  scale_y_sdv(sport = "wbb", size = 18) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  labs(
    title = paste("Conference strength,", label),
    subtitle = "Average net rating per 100 possessions of each conference's teams",
    x = "Average net rating",
    y = NULL,
    caption = source_note
  ) +
  theme_y_sdv() +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

## 5. A scoring leaderboard with headshots

Player box scores carry ESPN athlete ids: `gt_sdv_headshots()` turns them into
headshots and `gt_sdv_logos()` the team abbreviations into logos. A light
theme keeps dark logos readable. Only Division I teams count, a player's team
is the one she played for last, and leaders need at least 20 games.

```{r leaders-table}
players <- wehoop::load_wbb_player_box(seasons = season) |>
  filter(season_type == 2, !did_not_play, team_id %in% d1)

leaders <- players |>
  arrange(game_date) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    games = n(),
    team_id = last(team_id),
    headshot_url = last(athlete_headshot_href),
    ppg = mean(points, na.rm = TRUE),
    fg_pct = sum(field_goals_made, na.rm = TRUE) / sum(field_goals_attempted, na.rm = TRUE),
    fg3_pct = sum(three_point_field_goals_made, na.rm = TRUE) /
      sum(three_point_field_goals_attempted, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(games >= 20) |>
  arrange(desc(ppg), athlete_id) |>
  head(10) |>
  inner_join(select(ratings, team_id, team, team_name), by = "team_id")

leaders |>
  mutate(rank = row_number(), headshot = athlete_id, logo = team) |>
  select(rank, headshot, athlete_display_name, logo, team_name, games, ppg, fg_pct, fg3_pct) |>
  gt() |>
  tab_header(
    title = paste("Division I scoring leaders,", label),
    subtitle = "Regular season, 20 or more games"
  ) |>
  cols_label(
    rank = "", headshot = "", athlete_display_name = "Player", logo = "",
    team_name = "Team", games = "GP", ppg = "PPG", fg_pct = "FG%", fg3_pct = "3P%"
  ) |>
  fmt_number(ppg, decimals = 1) |>
  fmt_percent(c(fg_pct, fg3_pct), decimals = 1) |>
  tab_source_note(source_note) |>
  gt_sdv_headshots(columns = headshot, sport = "wbb", height = 42) |>
  gt_sdv_logos(columns = logo, sport = "wbb", height = 28) |>
  gt_theme_sofa()
```

## 6. A shot chart on a college court, with a headshot in the title

`load_wbb_shots()` holds ESPN's shot locations in feet from center court,
which wehoop has already converted to the frame `sdv_surface()`'s sportyR
court draws, so they need no conversion (`sdv_court_coords()` is only for the
stats-API legacy frame). Two cleanups first: free throws are placed at the
rim, and a shot with no location carries a placeholder of about ±215 million.
Then fold both halves onto one basket. `ggtitle_image()` puts the scorer's
headshot beside the title; it takes an image URL as well as a team. The
leader plays for Vanderbilt, whose black and gold match the painted area and
the wood, so the shots are drawn in white and red rather than team colors.

```{r shot-chart, fig.width = 7, fig.height = 7.4, fig.alt = "Half-court shot chart of the 2025-26 Division I women's scoring leader on a college court painted in her team's colors, made shots as white circles and missed shots as red crosses, with her headshot beside the title."}
star <- slice(leaders, 1)

shots <- wehoop::load_wbb_shots(seasons = season) |>
  filter(
    athlete_id_1 == star$athlete_id,
    !grepl("FreeThrow", type_text),
    abs(coordinate_x) <= 47, abs(coordinate_y) <= 25
  ) |>
  mutate(
    right = coordinate_x > 0,
    x = if_else(right, -coordinate_x, coordinate_x),
    y = if_else(right, -coordinate_y, coordinate_y)
  )

sdv_surface("wbb", star$team, display_range = "defense") +
  geom_point(
    aes(x, y),
    data = filter(shots, !scoring_play),
    shape = 4, color = "#D7191C", size = 1.7, stroke = 0.9
  ) +
  geom_point(
    aes(x, y),
    data = filter(shots, scoring_play),
    shape = 21, fill = "white", color = "black", stroke = 0.4, size = 2
  ) +
  ggtitle_image(
    title_image = star$headshot_url,
    title = paste0("**", star$athlete_display_name, "**, ", star$team_name, ", ", label),
    image_height = 48,
    subtitle = sprintf(
      "Every field goal attempt: %d made (white), %d missed (red crosses), %.1f%%",
      sum(shots$scoring_play), sum(!shots$scoring_play), 100 * mean(shots$scoring_play)
    ),
    sport = "wbb"
  ) +
  labs(caption = source_note) +
  theme_title_image(size = 14)
```

## 7. A ranked top 25, logos on the axis

The top 25 by net rating, as columns. `scale_x_sdv()` swaps the team
abbreviations on a discrete horizontal axis for logos; the factor order sets
the bar order.

```{r top-25, fig.width = 10, fig.height = 5.5, fig.alt = "Column chart of the 25 best Division I women's basketball teams of 2025-26 by net rating, best on the left, each column in the team's color with its logo under it on the horizontal axis."}
head(ratings, 25) |>
  mutate(team = factor(team, levels = team)) |>
  ggplot(aes(team, net, fill = team)) +
  geom_col(width = 0.75, color = "black", linewidth = 0.2) +
  geom_text(aes(label = sprintf("%.1f", net)), vjust = -0.4, size = 3) +
  scale_fill_sdv(sport = "wbb") +
  scale_x_sdv(sport = "wbb", size = 18) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(
    title = paste("The top 25 by net rating,", label),
    subtitle = "Points per 100 possessions, regular-season games between Division I teams",
    x = NULL,
    y = "Net rating",
    caption = source_note
  ) +
  theme_x_sdv() +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

## 8. Tiers as seed lines

The NCAA tournament seeds four teams to a line, so seed lines make natural
tiers: the top four by net rating on the 1 line, the next four on the 2 line,
and so on. This is a seed list from one rating, not the committee's bracket.
West Virginia's navy mark is hard to see on the dark Tiermaker theme, and the
result is a ggplot, so a few `theme()` settings make it light.

```{r team-tiers, fig.width = 8, fig.height = 5.5, fig.alt = "The 16 best Division I women's basketball teams of 2025-26 by net rating as logos in four tiers of four, labeled as the 1 to 4 seed lines."}
tiers <- head(ratings, 16) |>
  mutate(tier_no = (row_number() - 1) %/% 4 + 1) |>
  select(tier_no, team)

sdv_team_tiers(
  tiers,
  sport = "wbb",
  title = paste("Seed lines by net rating,", label),
  subtitle = "Four teams to a line, best first; not the selection committee's bracket",
  caption = source_note,
  tier_desc = c("1" = "1 seeds", "2" = "2 seeds", "3" = "3 seeds", "4" = "4 seeds"),
  alpha = 1,
  width = 0.1
) +
  theme(
    plot.background = element_rect(fill = "white", color = "white"),
    panel.background = element_rect(fill = "white", color = "white"),
    plot.title = element_text(color = "black", face = "bold"),
    axis.text.y = element_text(color = "black", face = "bold", size = rel(1.1))
  )
```

## 9. March: the champion's run

The schedule's `notes_headline` names each NCAA tournament game. Find the
title game's winner, then chart its six wins: the margin of each, the
opponent's logo on top, and the champion's logo beside the title through
`ggtitle_image()`, which takes a team abbreviation as well as a URL.

```{r champion-run, fig.width = 8, fig.height = 5.5, fig.alt = "Column chart of the 2026 women's national champion's six NCAA tournament wins in order, each column the winning margin in the champion's color with the opponent's logo above it, and the champion's logo beside the title."}
schedule <- wehoop::load_wbb_schedule(seasons = season)
ncaa <- schedule |>
  filter(grepl("^NCAA Women's Basketball Championship", notes_headline))
final <- filter(ncaa, grepl("National Championship", notes_headline))
champion_id <- if_else(final$home_winner, final$home_id, final$away_id)
champion <- wbb_teams$team_abbr[wbb_teams$espn_team_id == champion_id]

run <- ncaa |>
  filter(home_id == champion_id | away_id == champion_id) |>
  arrange(game_date) |>
  transmute(
    round = sub("National Championship", "Title game", sub(".* - ", "", notes_headline)),
    round = factor(round, levels = unique(round)),
    opponent_id = if_else(home_id == champion_id, away_id, home_id),
    margin = if_else(home_id == champion_id, home_score - away_score, away_score - home_score)
  ) |>
  inner_join(select(wbb_teams, opponent_id = espn_team_id, opponent = team_abbr), by = "opponent_id")

ggplot(run, aes(round, margin)) +
  geom_col(fill = sdv_team_colors(champion, sport = "wbb"), width = 0.65) +
  geom_text(aes(y = margin / 2, label = paste0("+", margin)), color = "white", fontface = "bold", size = 5) +
  geom_sdv_logos(aes(y = margin + 6, team = opponent), sport = "wbb", height = 0.13) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), limits = c(0, max(run$margin) + 11)) +
  ggtitle_image(
    title_image = champion,
    title = paste0("**The national champion's run, ", season, "**"),
    image_height = 40,
    subtitle = "Winning margin in each NCAA tournament game, opponent's logo on top",
    sport = "wbb"
  ) +
  labs(x = NULL, y = "Winning margin (points)", caption = source_note) +
  theme_title_image(size = 15) +
  theme(panel.grid.major.x = element_blank())
```

## 10. The SEC race, week by week

A conference race as a bump chart: rank the SEC's teams by conference win
percentage at the end of each week, then draw one line per team in its color
with its logo at the finish. The schedule's `conference_competition` flag
marks conference games, and a game with a `notes_headline` is a
conference-tournament game, so it is left out. Ties are broken by conference
point differential, then by abbreviation, so the same data always draw the
same chart.

```{r bump-chart, fig.width = 9, fig.height = 6.5, fig.alt = "Bump chart of the 16 SEC teams' weekly rank by conference win percentage through the 2025-26 regular season, one line per team in its color with its logo at its final rank."}
conference_games <- schedule |>
  filter(season_type == 2, conference_competition, is.na(notes_headline) | notes_headline == "") |>
  distinct(game_id)
sec <- filter(ratings, conference == "SEC")

weekly <- box |>
  semi_join(conference_games, by = "game_id") |>
  filter(team_id %in% sec$team_id) |>
  mutate(week = as.Date(cut(game_date, "week", start.on.monday = TRUE))) |>
  group_by(team_id, week) |>
  summarise(wins = sum(team_winner), games = n(), margin = sum(margin), .groups = "drop") |>
  complete(team_id, week, fill = list(wins = 0, games = 0, margin = 0)) |>
  arrange(team_id, week) |>
  group_by(team_id) |>
  mutate(across(c(wins, games, margin), cumsum)) |>
  ungroup() |>
  inner_join(select(sec, team_id, team), by = "team_id") |>
  filter(week >= sort(unique(week))[3]) |> # the first two weeks hold a game or two
  arrange(week, desc(wins / pmax(games, 1)), desc(margin), team) |>
  group_by(week) |>
  mutate(rank = row_number()) |>
  ungroup()
finish <- filter(weekly, week == max(week))

ggplot(weekly, aes(week, rank)) +
  geom_line(aes(group = team), color = "black", linewidth = 1.6) +
  geom_line(aes(color = team), linewidth = 1) +
  geom_sdv_logos(aes(x = week + 4, team = team), data = finish, sport = "wbb", width = 0.04) +
  scale_color_sdv(sport = "wbb") +
  scale_y_reverse(breaks = 1:16) +
  scale_x_date(date_labels = "%b %d", expand = expansion(add = c(1, 7))) +
  labs(
    title = paste("The SEC race,", label),
    subtitle = "Rank by conference win percentage at the end of each week",
    x = NULL,
    y = "SEC rank",
    caption = source_note
  ) +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

## Related articles

- [Getting Started](getting-started.html)
- [Men's College Basketball Visualizations](mbb-viz.html)
- [WNBA Visualizations](wnba-viz.html)
- [Social Posting Patterns](social-posting.html)
