---
title: "Winningest Franchises in Rolling Windows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Winningest Franchises in Rolling Windows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

This article builds one table: the NFL's five winningest franchises in every
ten-year window from 1999-2008 to 2016-2025, each franchise drawn as its logo
and shaded by how many Super Bowls it won inside that window. The games are
nflverse's schedule file, read with
[`nflreadr`](https://nflreadr.nflverse.com), regular season only.

Eighteen windows give a table six columns wide and eighteen rows tall, which
exports as a narrow strip with empty space on either side. `gt_snake()` folds it
into two side-by-side blocks that repeat the column labels. The shading is
per-cell, and it is far easier to reason about on the plain eighteen-row grid
than on the folded one, so we color first and fold second. Along the way the
article uses `gt_highlight_cells()`, `gt_legend_discrete()` and
`gt_538_caption()`.

`gt_snake()` and the other `gt_*` table functions here come from Andrew
Weatherman's [gtUtils](https://github.com/andreweatherman/gtUtils); this article
follows his rolling-window example, rebuilt on nflverse data.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr)
library(tidyr)
```

## 1. One row per team per game

`load_schedules()` holds one row per game. `clean_homeaway()` doubles it into one
row per team per game, renaming the `home_*` / `away_*` pairs to `team_*` /
`opponent_*`.

```{r games}
seasons <- 1999:2025
schedules <- nflreadr::load_schedules(seasons)

games <- schedules |>
  filter(game_type == "REG", !is.na(home_score)) |>
  select(season, home_team, away_team, home_score, away_score) |>
  nflreadr::clean_homeaway() |>
  mutate(franchise = clean_team_abbrs(team, sport = "nfl"))
```

The schedules keep each season's abbreviation, so the Raiders are `OAK` before
2020 and `LV` after. `clean_team_abbrs()` maps every relocated abbreviation to
today's franchise key, which is what lets one franchise's ten seasons add up.

```{r relocations}
clean_team_abbrs(c("OAK", "SD", "STL"), sport = "nfl")
```

Each franchise's record per season is then one summary:

```{r records}
records <- games |>
  summarise(
    wins = sum(team_score > opponent_score),
    losses = sum(team_score < opponent_score),
    ties = sum(team_score == opponent_score),
    .by = c(season, franchise)
  )
```

## 2. Rolling windows

A rolling window is a filter and a sum. This function keeps the ten seasons from
a start year and totals each franchise's record.

```{r windows}
calculate_windows <- function(start_year, data) {
  data |>
    filter(between(season, start_year, start_year + 9)) |>
    summarise(across(c(wins, losses, ties), sum), .by = franchise) |>
    mutate(
      win_pct = (wins + ties / 2) / (wins + losses + ties),
      years = paste(start_year, start_year + 9, sep = "-"),
      begin = start_year
    )
}

starts <- min(seasons):(max(seasons) - 9)

plot_data <- bind_rows(lapply(starts, calculate_windows, records)) |>
  arrange(begin, desc(wins), desc(win_pct), franchise) |>
  mutate(position = row_number(), .by = begin) |>
  filter(position <= 5)
```

The start years stop nine short of the last season, since a window that starts
in 2017 does not have ten seasons in it yet. Ranking happens inside each window
with `row_number()` after the `arrange()`: wins first, win percentage (a tie
counts as half a win) next, and the franchise key last, so a tie that survives
both still sorts the same way every time the site is built.

## 3. Counting titles

The shading comes from Super Bowls won inside each window. The same schedule
file holds the Super Bowls (`game_type == "SB"`), so the winners are one more
filter, and the windowing logic is the same.

```{r titles}
champions <- schedules |>
  filter(game_type == "SB") |>
  transmute(
    season,
    franchise = clean_team_abbrs(
      if_else(home_score > away_score, home_team, away_team),
      sport = "nfl"
    )
  )

titles <- bind_rows(lapply(starts, function(start_year) {
  champions |>
    filter(between(season, start_year, start_year + 9)) |>
    count(franchise, name = "titles") |>
    mutate(years = paste(start_year, start_year + 9, sep = "-"))
}))

plot_data <- plot_data |>
  left_join(titles, by = c("franchise", "years")) |>
  mutate(titles = case_when(
    is.na(titles) ~ "0",
    titles == 1 ~ "1",
    TRUE ~ "2+"
  ))
```

The join only produces a count for a franchise that won something, so every other
franchise-window comes back `NA`, which the `case_when()` turns into `"0"`.
Bucketing into `"0"`, `"1"` and `"2+"` keeps the key short.

## 4. Two pivots from one ordered frame

We want one row per window and five franchise columns in rank order. The cells
hold the franchise keys themselves; `gt_sdv_logos()` turns them into logos at the
end, so the data never carries an image URL.

`pivot_wider()` spreads one value column, so pivoting on the franchise drops the
title count. The way around it is to pivot twice from the same ordered frame,
once for what you draw and once for what you style by.

```{r pivots}
ordered <- arrange(plot_data, begin, position)

# years + five franchises
wide <- ordered |>
  pivot_wider(id_cols = years, names_from = position, values_from = franchise, names_sort = TRUE)

# the shading key, the same shape as the five franchise columns
title_grid <- ordered |>
  pivot_wider(id_cols = years, names_from = position, values_from = titles, names_sort = TRUE) |>
  select(-years)

mask_1 <- title_grid == "1"
mask_2 <- title_grid == "2+"
```

Sorting once, before either pivot, keeps the two frames in step: both inherit the
same row order and the same column order, so cell `[3, 2]` of `title_grid`
describes cell `[3, 2]` of `wide` with no matching afterwards. Each mask is then
a `TRUE`/`FALSE` grid the same shape as the franchise columns. `"0"` is the base
state, so it needs no mask of its own.

```{r peek}
head(wide, 3)
```

## 5. Color first, then snake

`gt_highlight_cells()` fills individual cells from a logical grid of the same
shape, which is what the masks are. We run it on the plain, un-snaked table, then
fold it.

```{r snake}
wide |>
  gt(id = "windows-draft") |>
  gt_highlight_cells(-years, condition = mask_1, fill = "#F4D7A4") |>
  gt_highlight_cells(-years, condition = mask_2, fill = "#FFA970") |>
  gt_snake(n_cols = 2) |>
  gt_theme_athletic()
```

`gt_snake()` rebuilds the table with the rows cut into two blocks side by side,
and it moves any body-cell styling to wherever that cell ended up. A fill on row
10 of the long grid lands on row one of the second block, because it is the same
cell.

Styles carry through the reshape, but formatting and content transforms do not:
anything from the `fmt_*()` family, `text_transform()`, or the logo helpers has
to come after the snake, and so does the theme. For a parallel frame you cannot
apply before folding, `gt_snake_align()` reshapes it into the same blocks so it
still lines up.

`gt_snake()` puts a real empty column between the blocks rather than padding the
seam, because padding shifts the body cells without shifting the column labels.
`clean_gaps = TRUE`, the default, strips that spacer's borders and background so
the gap stays blank under any theme.

## 6. Logos after the fold

After the fold each column carries its block number as a suffix, so the franchise
columns run `1_1` through `5_2` and the years are `years_1` and `years_2`. One
`gt_sdv_logos()` call with a `matches()` selector takes all ten franchise
columns; each logo's `alt` text is the team's full name.

```{r logos, eval = FALSE}
# ... |>
gt_sdv_logos(columns = matches("^[1-5]_[0-9]+$"), sport = "nfl", height = 34)
```

The pattern picks the snaked columns out by name rather than listing ten of them,
so the code survives a change to `n_cols`.

## 7. Spanners, widths and rules

The rest is ordinary `gt`: a spanner over each block, a label and a fixed width
for the years, and a dotted rule between the logos.

```{r rules, eval = FALSE}
# ... |>
cols_align(columns = everything(), "center") |>
  tab_spanner(columns = matches("^[1-5]_1$"), label = "Most wins in the window", id = "block_1") |>
  tab_spanner(columns = matches("^[1-5]_2$"), label = "Most wins in the window", id = "block_2") |>
  cols_label(starts_with("years") ~ "Years", matches("^[1-5]_") ~ "") |>
  cols_width(starts_with("years") ~ px(92), matches("^[1-5]_") ~ px(48)) |>
  tab_options(table_body.hlines.style = "solid", table_body.hlines.color = "black") |>
  tab_style(
    locations = cells_body(columns = c(starts_with("years"), matches("^[1-5]_"))),
    style = cell_borders(color = "black", sides = "right", weight = px(1.5), style = "dotted")
  )
```

The dotted rule is set to `px(1.5)` because `gt_theme_athletic()` already draws a
thin solid separator on every column, and a lighter rule loses to it.

## 8. The legend and the caption

`gt_legend_discrete()` takes a named vector of `label = color` and draws the key
in the header. `label_placement = "inside"` prints each label on its own swatch
and picks the text color per swatch for contrast, which keeps the white swatch
readable. `gt_538_caption()` takes a top line, which gets a rule under it, and a
bottom line for the source. Apply it after the theme, since it reads the themed
table to borrow a matching rule color.

## The finished table

```{r window-table}
title_key <- c(
  "No Super Bowls in window" = "#FFFFFF",
  "1 Super Bowl in window" = "#F4D7A4",
  "2+ Super Bowls in window" = "#FFA970"
)

wide |>
  gt(id = "windows") |>
  gt_highlight_cells(-years, condition = mask_1, fill = "#F4D7A4") |>
  gt_highlight_cells(-years, condition = mask_2, fill = "#FFA970") |>
  gt_snake(n_cols = 2) |>
  gt_theme_athletic() |>
  gt_sdv_logos(columns = matches("^[1-5]_[0-9]+$"), sport = "nfl", height = 34) |>
  cols_align(columns = everything(), "center") |>
  tab_spanner(columns = matches("^[1-5]_1$"), label = "Most wins in the window", id = "block_1") |>
  tab_spanner(columns = matches("^[1-5]_2$"), label = "Most wins in the window", id = "block_2") |>
  cols_label(starts_with("years") ~ "Years", matches("^[1-5]_") ~ "") |>
  cols_width(starts_with("years") ~ px(92), matches("^[1-5]_") ~ px(48)) |>
  tab_options(table_body.hlines.style = "solid", table_body.hlines.color = "black") |>
  tab_style(
    locations = cells_body(columns = c(starts_with("years"), matches("^[1-5]_"))),
    style = cell_borders(color = "black", sides = "right", weight = px(1.5), style = "dotted")
  ) |>
  gt_legend_discrete(
    title_key,
    location = "top", label_placement = "inside", shape = "square",
    gap = 8, swatch_size = 20, border_width = 1.5, border_color = "black",
    heading = "The NFL's winningest franchises",
    subtitle = paste0(
      "Regular-season wins in each 10-year window, ", min(seasons), "-", max(seasons),
      ". Ties in the ranking go to win percentage."
    ),
    heading_style = list(
      font = "Spline Sans Mono", size = 22, weight = 600,
      transform = "uppercase", margin_bottom = 2, align = "center"
    ),
    subtitle_style = list(
      font = "Spline Sans Mono", size = 12, margin_bottom = 6,
      align = "center", color = "black"
    ),
    label_style = list(font = "Spline Sans Mono", size = 12, margin_bottom = 12)
  ) |>
  gt_538_caption(
    "Relocated franchises count as one: the Raiders, Chargers and Rams across their moves.",
    "Data: nflverse schedules via nflreadr"
  )
```

Three franchises hold the top spot between them: the Colts in the first two
windows, the Patriots in every window from 2001-2010 to 2012-2021, and the
Chiefs in every window since.
