---
title: "WNBA Visualizations with wehoop and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WNBA Visualizations with wehoop and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 5.5,
  dpi = 96
)
```

Ten charts and tables from the 2026 WNBA regular season, the league's first
with 15 teams: an offense-vs-defense quadrant, the expansion teams' win races,
net-rating bars with logos on the axis, tiers, a scoring leaderboard with
headshots as axis labels, a binned shot chart, a standings table, running
point differential by conference, game-by-game margins and a season-long bump
chart. Every number comes from [wehoop](https://wehoop.sportsdataverse.org)'s
ESPN release files on GitHub (`load_wnba_*()`), so nothing here calls
stats.wnba.com.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(tidyr)
library(gt)

season <- 2026
source_note <- "Data: wehoop (ESPN) | Viz: sdvplotR"

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      plot.title.position = "plot",
      plot.caption = element_text(color = "grey40", size = 8),
      plot.caption.position = "plot",
      panel.grid.minor = element_blank()
    )
)
```

The team box score has one row per team per game; `season_type` 2 is the
regular season. 2025 is loaded too, for Golden State's first season. ESPN
files each All-Star Game as a regular-season game between two teams that
aren't WNBA clubs, so an inner join on the ESPN team id to
`team_reference("wnba")` drops them, and adds each team's sdvplotR
abbreviation and conference.

```{r load-data}
wnba_teams <- team_reference("wnba") |>
  select(team_id = espn_team_id, team = team_abbr, conference)

box_all <- wehoop::load_wnba_team_box(seasons = c(season - 1, season)) |>
  filter(season_type == 2) |>
  inner_join(wnba_teams, by = "team_id")
box <- filter(box_all, season == !!season)

count(box, team, name = "games") |>
  count(games, name = "teams")
```

The Commissioner's Cup final is filed as a regular-season game as well, which
is why two teams show 45 games. It stays in the box scores (it was played) but
does not count in the standings in example 7.

## 1. Offense vs defense, with logos

Points scored and allowed per 100 possessions, with possessions estimated from
the box score (FGA - OREB + TOV + 0.44 x FTA) and averaged with the opponent's
estimate for the same game. Defensive rating is points allowed, so its axis is
reversed to put good defenses on top.

```{r ratings}
games <- box |>
  mutate(
    poss = field_goals_attempted - offensive_rebounds +
      coalesce(total_turnovers, turnovers) + 0.44 * free_throws_attempted
  )
games <- games |>
  inner_join(
    games |> select(game_id, opponent_team_id = team_id, opp_poss = poss),
    by = c("game_id", "opponent_team_id")
  ) |>
  mutate(game_poss = (poss + opp_poss) / 2)

ratings <- games |>
  group_by(team, conference) |>
  summarise(
    games = n(),
    ortg = 100 * sum(team_score) / sum(game_poss),
    drtg = 100 * sum(opponent_team_score) / sum(game_poss),
    .groups = "drop"
  ) |>
  mutate(net = ortg - drtg) |>
  arrange(desc(net), team)
stopifnot(nrow(ratings) == 15, !anyNA(ratings$net))
```

```{r team-performance, fig.alt = "WNBA teams of the 2026 regular season drawn as their logos, placed by offensive rating (horizontal) and defensive rating (vertical, reversed so better defenses are higher), with dashed league-average lines and a label in each corner."}
x_lim <- range(ratings$ortg) + c(-1.5, 1.5)
y_lim <- range(ratings$drtg) + c(-2, 1.5)

ggplot(ratings, aes(ortg, drtg)) +
  geom_mean_lines(
    aes(x0 = ortg, y0 = drtg),
    color = "grey60", linetype = "dashed", linewidth = 0.4
  ) +
  geom_sdv_logos(aes(team = team), sport = "wnba", width = 0.075) +
  annotate(
    "text",
    x = x_lim[c(2, 1, 2, 1)], y = y_lim[c(1, 1, 2, 2)],
    hjust = c(1, 0, 1, 0), vjust = c(1.3, 1.3, -0.4, -0.4),
    label = c("Good offense, good defense", "Defense first", "Offense first", "Struggling"),
    color = "grey45", fontface = "italic", size = 3.5
  ) +
  scale_x_continuous(expand = expansion(0)) +
  scale_y_reverse(expand = expansion(0)) +
  expand_limits(x = x_lim, y = y_lim) +
  labs(
    title = paste("WNBA offense vs defense,", season, "regular season"),
    x = "Offensive rating (points per 100 possessions)",
    y = "Defensive rating (points allowed per 100)",
    caption = source_note
  )
```

## 2. The expansion teams, game by game

Golden State joined in 2025, Portland and Toronto in 2026. Cumulative wins by
game number put all four seasons on one chart; the dashed line is a .500 pace.
Portland's primary color is a pale ice blue, so every line is drawn over a
black one that outlines it on white.

```{r expansion, fig.width = 9, fig.alt = "Cumulative wins by game number for the WNBA's expansion teams: Golden State in 2025 and 2026, Portland and Toronto in 2026, each line in the team's color with a black outline, a logo and the final record at its end, and a dashed .500 pace line."}
runs <- box_all |>
  filter(team == "GS" | (team %in% c("POR", "TOR") & season == 2026)) |>
  arrange(season, team, game_date) |>
  group_by(season, team) |>
  mutate(game_no = row_number(), wins = cumsum(team_winner)) |>
  ungroup() |>
  mutate(run = paste(team, season))

ends <- runs |>
  group_by(run) |>
  slice_max(game_no, n = 1) |>
  ungroup()

ggplot(runs, aes(game_no, wins, group = run)) +
  annotate("segment", x = 0, y = 0, xend = 44, yend = 22, color = "grey55", linetype = "dashed") +
  geom_line(color = "black", linewidth = 2.4) +
  geom_line(aes(color = team), linewidth = 1.4) +
  geom_sdv_logos(aes(x = game_no + 2, team = team), data = ends, sport = "wnba", width = 0.045) +
  geom_text(
    aes(x = game_no + 3.6, label = sprintf("%s: %d-%d", season, wins, game_no - wins)),
    data = ends, hjust = 0, size = 3.6
  ) +
  scale_color_sdv(sport = "wnba") +
  scale_x_continuous(breaks = seq(0, 44, 11), expand = expansion(add = c(0.5, 10))) +
  labs(
    title = "The WNBA's expansion teams, game by game",
    subtitle = "Cumulative wins; the dashed line is a .500 pace",
    x = "Game of the season",
    y = "Wins",
    caption = source_note
  ) +
  theme(legend.position = "none")
```

## 3. Net rating, with logos on the axis

`scale_y_sdv()` turns a discrete vertical axis of abbreviations into logos and
`theme_y_sdv()` lets the axis draw them. The factor order sets the bar order:
here the best net rating sits on top.

```{r net-rating, fig.height = 6, fig.alt = "Horizontal bar chart of WNBA net rating per 100 possessions in 2026, best at the top, each bar in the team's color with a thin black outline and the team's logo on the vertical axis."}
ratings |>
  mutate(team = factor(team, levels = rev(team))) |>
  ggplot(aes(net, team, fill = team)) +
  geom_col(width = 0.72, color = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linewidth = 0.4) +
  geom_text(
    aes(label = sprintf("%+.1f", net), hjust = if_else(net > 0, -0.15, 1.15)),
    size = 3.4
  ) +
  scale_fill_sdv(sport = "wnba") +
  scale_y_sdv(sport = "wnba", size = 18) +
  scale_x_continuous(expand = expansion(mult = 0.12)) +
  labs(
    title = paste("WNBA net rating,", season, "regular season"),
    x = "Net rating (points per 100 possessions)",
    y = NULL,
    caption = source_note
  ) +
  theme_y_sdv() +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

## 4. Team tiers from net rating

`sdv_team_tiers()` draws a tier list from `tier_no` and `team`. The tiers here
are cut from the net ratings above, so they are the season's numbers. Its
default background is dark, and Toronto's dark plum mark all but disappears on
it, so `theme = "light"` draws the list on white and `alpha = 1` draws the
logos at full strength.

```{r team-tiers, fig.width = 9, fig.height = 5.5, fig.alt = "WNBA team logos in four tiers cut from 2026 net rating per 100 possessions: Title contenders, Playoff teams, Fringe and Rebuilding."}
tiers <- ratings |>
  mutate(tier_no = 5L - cut(net, c(-Inf, -5, 0, 5, Inf), labels = FALSE, right = FALSE)) |>
  arrange(tier_no, desc(net)) |>
  select(tier_no, team)

sdv_team_tiers(
  tiers,
  sport = "wnba",
  title = paste("WNBA tiers by net rating,", season),
  subtitle = "Regular-season net rating per 100 possessions, cut at +5, 0 and -5",
  caption = source_note,
  tier_desc = c("1" = "Title contenders", "2" = "Playoff teams", "3" = "Fringe", "4" = "Rebuilding"),
  alpha = 1,
  theme = "light"
)
```

## 5. Scoring leaders, with headshots as the axis labels

`scale_y_sdv_headshots()` draws a headshot for each ESPN athlete id on a
discrete axis, the way `scale_y_sdv()` draws logos. A thin black edge keeps
the pale team colors (Portland, New York) visible. Players need 30 games.

```{r leaders}
players <- wehoop::load_wnba_player_box(seasons = season) |>
  semi_join(distinct(box, game_id), by = "game_id") |>
  filter(!did_not_play)

scorers <- players |>
  arrange(game_date) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(games = n(), team = last(team_abbreviation), ppg = mean(points), .groups = "drop") |>
  filter(games >= 30) |>
  arrange(desc(ppg), athlete_id) |>
  head(10)
```

```{r player-headshots, fig.height = 6.5, fig.alt = "Horizontal bar chart of the ten WNBA scoring leaders of 2026 (30 or more games), each bar in the player's team color, with her headshot as the axis label and her name and points per game at the end of the bar."}
scorers |>
  mutate(athlete_id = factor(athlete_id, levels = rev(athlete_id))) |>
  ggplot(aes(ppg, athlete_id)) +
  geom_col(aes(fill = team), width = 0.7, color = "black", linewidth = 0.2) +
  geom_text(
    aes(x = ppg + 0.4, label = sprintf("%s  %.1f", athlete_display_name, ppg)),
    hjust = 0, size = 3.8
  ) +
  scale_fill_sdv(sport = "wnba") +
  scale_y_sdv_headshots(sport = "wnba", size = 34) +
  scale_x_continuous(limits = c(0, max(scorers$ppg) + 9), expand = expansion(0)) +
  labs(
    title = paste("WNBA scoring leaders,", season),
    subtitle = "Points per game, 30 or more games",
    x = "Points per game",
    y = NULL,
    caption = source_note
  ) +
  theme_y_sdv() +
  theme(legend.position = "none", panel.grid.major.y = element_blank())
```

## 6. A binned shot chart on a team-colored court

`load_wnba_shots()` holds ESPN's shot locations, which wehoop already converts
to feet on a center-court frame, the frame `sdv_surface()`'s sportyR court
uses. They need no conversion; `sdv_court_coords()` is only for the
stats.wnba.com legacy frame (`x_legacy`/`y_legacy`, tenths of a foot around
the hoop). Fold the right-basket shots onto the left one, then bin them into
3-foot squares: the size of each point is the attempts from there, its color
the field goal percentage.

```{r shot-chart, fig.width = 7, fig.height = 7.5, fig.alt = "Half-court chart of the 2026 WNBA scoring leader's field goal attempts binned into 3-foot squares on a court painted in her team's colors: point size shows attempts and color shows field goal percentage, blue below 45 percent and red above."}
star <- slice(scorers, 1)

zones <- wehoop::load_wnba_shots(seasons = season) |>
  semi_join(distinct(box, game_id), by = "game_id") |>
  filter(athlete_id_1 == star$athlete_id, !grepl("Free Throw", type_text)) |>
  mutate(
    right = coordinate_x > 0,
    x = if_else(right, -coordinate_x, coordinate_x),
    y = if_else(right, -coordinate_y, coordinate_y)
  ) |>
  group_by(x = round(x / 3) * 3, y = round(y / 3) * 3) |>
  summarise(attempts = n(), fg_pct = mean(scoring_play), .groups = "drop")

sdv_surface("wnba", star$team, display_range = "defense") +
  geom_point(
    aes(x, y, size = attempts, fill = fg_pct),
    data = zones, shape = 21, color = "grey15", stroke = 0.3
  ) +
  scale_size_area(max_size = 11, name = "Attempts") +
  scale_fill_gradient2(
    low = "#2C7BB6", mid = "white", high = "#D7191C", midpoint = 0.45,
    limits = c(0, 1), breaks = c(0, 0.5, 1), labels = scales::percent, name = "FG%"
  ) +
  labs(
    title = paste0(star$athlete_display_name, ": where she shot from, ", season),
    subtitle = sprintf("%d field goal attempts in 3-foot squares", sum(zones$attempts)),
    caption = source_note
  ) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))
```

## 7. A standings table with logos and a playoff line

ESPN's standings come long (one row per team and stat), so pivot them wide.
The `clincher` column marks the eight teams that clinched a playoff spot with
an `x`, and they are the eight best records, so one league-wide table with a
cut line after eighth tells the story.

```{r standings-table}
standings <- wehoop::load_wnba_standings(seasons = season) |>
  select(group_name, team_abbreviation, team_short_display_name, stat_name, display_value) |>
  pivot_wider(names_from = stat_name, values_from = display_value)

standings |>
  mutate(
    conf = sub(" Conference", "", group_name),
    logo = team_abbreviation,
    pct = as.numeric(winPercent)
  ) |>
  arrange(desc(pct), desc(as.numeric(differential)), team_abbreviation) |>
  mutate(rank = row_number()) |>
  select(rank, logo, team_short_display_name, conf, wins, losses, winPercent, Home, Road, `Last Ten Games`, differential) |>
  gt() |>
  tab_header(
    title = paste("WNBA standings,", season),
    subtitle = "The eight best records made the playoffs"
  ) |>
  cols_label(
    rank = "", logo = "", team_short_display_name = "Team", conf = "Conf", wins = "W",
    losses = "L", winPercent = "Pct", `Last Ten Games` = "L10",
    differential = "Diff"
  ) |>
  tab_source_note(source_note) |>
  gt_sdv_logos(columns = logo, sport = "wnba", height = 26) |>
  gt_theme_athletic() |>
  cols_align("left", columns = team_short_display_name) |>
  gt_cutline(after = 8, label = "Playoffs", label_position = "above")
```

## 8. The season as a running point differential, by conference

Running point differential through the season, one line per team in its color
(`scale_color_sdv()`), one panel per conference, with `geom_sdv_logos()`
marking where each team finished.

```{r running-differential, fig.width = 10, fig.height = 5.5, fig.alt = "Two panels, Eastern and Western Conference, with one line per WNBA team showing its running point differential by game number through the 2026 regular season, each line in the team's color with its logo at the end."}
running <- box |>
  arrange(team, game_date) |>
  group_by(team) |>
  mutate(game_no = row_number(), diff = cumsum(team_score - opponent_team_score)) |>
  ungroup()
finish <- running |>
  group_by(team) |>
  slice_max(game_no, n = 1) |>
  ungroup()

ggplot(running, aes(game_no, diff)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_line(aes(group = team), color = "black", linewidth = 1.6) +
  geom_line(aes(color = team), linewidth = 1) +
  geom_sdv_logos(aes(x = game_no + 3, team = team), data = finish, sport = "wnba", width = 0.07) +
  facet_wrap(~conference, labeller = labeller(conference = function(x) paste(x, "Conference"))) +
  scale_color_sdv(sport = "wnba") +
  scale_x_continuous(expand = expansion(add = c(1, 6))) +
  labs(
    title = paste("Running point differential,", season),
    x = "Game of the season",
    y = "Point differential",
    caption = source_note
  ) +
  theme(legend.position = "none", strip.text = element_text(face = "bold", size = 11))
```

## 9. Every game's margin, in team colors

A box plot shows the spread of a season; the points on top are the games, in
each team's color, and `scale_x_sdv()` labels the axis with logos. Teams are
ordered by their median margin. `position_jitter(seed = 1)` keeps the jitter
the same from one build to the next.

```{r game-margins, fig.width = 10, fig.height = 5.5, fig.alt = "Box plots of every 2026 WNBA game's final margin by team, ordered by median margin, with each game as a point in the team's color and team logos along the horizontal axis."}
margins <- box |>
  mutate(margin = team_score - opponent_team_score)
order <- margins |>
  group_by(team) |>
  summarise(med = median(margin), avg = mean(margin)) |>
  arrange(desc(med), desc(avg), team) |>
  pull(team)

margins |>
  mutate(team = factor(team, levels = order)) |>
  ggplot(aes(team, margin)) +
  geom_hline(yintercept = 0, color = "grey50", linewidth = 0.4) +
  geom_boxplot(outlier.shape = NA, fill = NA, color = "grey40", width = 0.6) +
  geom_point(
    aes(fill = team),
    shape = 21, color = "black", stroke = 0.2, size = 1.9, alpha = 0.85,
    position = position_jitter(width = 0.18, height = 0, seed = 1)
  ) +
  scale_fill_sdv(sport = "wnba") +
  scale_x_sdv(sport = "wnba", size = 20) +
  labs(
    title = paste("Every game's final margin,", season),
    subtitle = "Teams ordered by median margin",
    x = NULL,
    y = "Final margin (points)",
    caption = source_note
  ) +
  theme_x_sdv() +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

## 10. The standings race, as a bump chart

Rank all 15 teams by win percentage at the end of every week, then draw one
line per team in its color with its logo at the finish. Ties are broken by
point differential, then by abbreviation, so the same data always draw the
same chart.

```{r bump-chart, fig.width = 9, fig.height = 6.5, fig.alt = "Bump chart of the 15 WNBA teams' weekly league rank by win percentage through the 2026 regular season, one line per team in its color, with each team's logo at its final rank on the right."}
weekly <- box |>
  mutate(
    week = as.Date(cut(game_date, "week", start.on.monday = TRUE)),
    margin = team_score - opponent_team_score
  ) |>
  group_by(team, week) |>
  summarise(wins = sum(team_winner), games = n(), margin = sum(margin), .groups = "drop") |>
  complete(team, week, fill = list(wins = 0, games = 0, margin = 0)) |>
  arrange(team, week) |>
  group_by(team) |>
  mutate(across(c(wins, games, margin), cumsum)) |>
  ungroup() |>
  mutate(week_no = dense_rank(week)) |>
  filter(week_no >= 3) |> # the first two weeks hold only a game or two
  arrange(week_no, desc(wins / games), desc(margin), team) |>
  group_by(week_no) |>
  mutate(rank = row_number()) |>
  ungroup()
last_week <- filter(weekly, week_no == max(week_no))

ggplot(weekly, aes(week_no, rank)) +
  geom_line(aes(group = team), color = "black", linewidth = 1.7) +
  geom_line(aes(color = team), linewidth = 1.1) +
  geom_sdv_logos(aes(x = week_no + 0.9, team = team), data = last_week, sport = "wnba", width = 0.045) +
  scale_color_sdv(sport = "wnba") +
  scale_y_reverse(breaks = 1:15) +
  scale_x_continuous(breaks = seq(3, 30, 3), expand = expansion(add = c(0.4, 1.4))) +
  labs(
    title = paste("The WNBA standings race,", season),
    subtitle = "League rank by win percentage at the end of each week",
    x = "Week of the season",
    y = "League rank",
    caption = source_note
  ) +
  theme(legend.position = "none", panel.grid.major.x = element_blank())
```

## Related articles

- [Getting Started](getting-started.html)
- [NBA Visualizations](nba-viz.html)
- [Women's College Basketball Visualizations](wbb-viz.html)
- [Social Posting Patterns](social-posting.html)
