---
title: "WNBA Visualizations with wehoop and sdvplotR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WNBA Visualizations with wehoop and sdvplotR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates how to create rich WNBA visualizations by combining
[wehoop](https://wehoop.sportsdataverse.org) for player and team data with
[sdvplotR](https://sdvplotr.sportsdataverse.org) for team logos, headshots,
colors, and gt tables.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(wehoop)
library(dplyr)
library(gt)

# Get valid WNBA team abbreviations
wnba_teams <- valid_team_names("wnba")
head(wnba_teams)
```

## Loading WNBA Data

Use `wehoop` to load this season's ESPN team and player box scores:

```{r load-data}
season <- wehoop::most_recent_wnba_season()

# One row per team per game (ESPN box scores), regular season only
team_stats <- wehoop::load_wnba_team_box(seasons = season) |>
  filter(season_type == 2)

# One row per player per game, with ESPN athlete IDs; players who did not
# play are dropped so games played counts real games
player_stats <- wehoop::load_wnba_player_box(seasons = season) |>
  filter(season_type == 2, !did_not_play)
```

## WNBA Team Performance

Visualize team performance with team logos:

```{r team-performance}
# Calculate team metrics
team_perf <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    avg_points = mean(team_score, na.rm = TRUE),
    avg_rebounds = mean(total_rebounds, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10)

ggplot(team_perf, aes(x = avg_points, y = avg_rebounds)) +
  geom_sdv_logos(
    aes(team = team_abbreviation),
    sport = "wnba",
    width = 0.075
  ) +
  labs(
    title = "WNBA Team Performance",
    subtitle = paste("Season", wehoop::most_recent_wnba_season()),
    x = "Average Points per Game",
    y = "Average Rebounds per Game",
    caption = "Data: wehoop | Viz: sdvplotR"
  ) +
  theme_minimal()
```

## WNBA Team Colors

Use team colors to visualize win percentages:

```{r team-colors}
# Calculate win percentage
team_wins <- team_stats |>
  filter(!is.na(team_abbreviation)) |>
  group_by(team_abbreviation) |>
  summarise(
    wins = sum(team_winner, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 10) |>
  mutate(win_pct = wins / games) |>
  arrange(desc(win_pct)) |>
  head(12)

ggplot(team_wins, aes(x = reorder(team_abbreviation, win_pct), y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.7) +
  scale_fill_sdv(sport = "wnba", alpha = 0.8) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "WNBA Teams by Win Percentage",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Player Headshots

Visualize top performers with player headshots:

```{r player-headshots}
# Top scorers
top_scorers <- player_stats |>
  filter(!is.na(athlete_id)) |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    avg_points = mean(points, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 15) |>
  arrange(desc(avg_points)) |>
  head(8)

ggplot(top_scorers, aes(x = games, y = avg_points)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "wnba",
    height = 0.15
  ) +
  geom_label(
    aes(label = athlete_display_name),
    nudge_y = -1.0,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "Top 8 WNBA Scorers",
    subtitle = paste("Season", wehoop::most_recent_wnba_season()),
    x = "Games Played",
    y = "Average Points per Game"
  ) +
  theme_minimal()
```

### WNBA Stats player IDs

The headshots above use ESPN athlete IDs, the `athlete_id` in wehoop's
`load_wnba_*()` data. wehoop's `wnba_*()` functions read stats.wnba.com, which
numbers players differently (`PLAYER_ID`). Pass `id_type = "league"` to draw
those from the WNBA's own image CDN. Both are plain numbers, so without it a
WNBA Stats ID is read as an ESPN ID and draws someone else or nothing.

```{r wnba-stats-headshots}
leaders <- wehoop::wnba_leagueleaders()$LeagueLeaders |>
  mutate(across(c(GP, PTS), as.numeric)) |>
  head(8)

ggplot(leaders, aes(x = GP, y = PTS)) +
  geom_sdv_headshots(
    aes(player_id = PLAYER_ID),
    sport = "wnba",
    id_type = "league",
    height = 0.15
  ) +
  labs(title = "WNBA scoring leaders", x = "Games Played", y = "Points")
```

`id_type` works the same in `gt_sdv_headshots()`, `reactable_sdv_headshots()`,
the headshot axis scales and `element_sdv_headshot()`. The WNBA's image CDN
refuses requests from datacenter IPs, so a plot rendered on CI or a server can
come back without these headshots; tables are unaffected, because the reader's
browser loads the images.

## WNBA Team Tiers

Create a tier plot ranking WNBA teams:

```{r team-tiers}
# Sample tier assignments
tier_data <- data.frame(
  tier_no = c(1, 1, 2, 2, 2, 3, 3, 3, 4, 4, 4, 5),
  team = c("LV", "NY", "WAS", "DAL", "MIN",
           "SEA", "CHI", "ATL",
           "PHX", "LA", "CON",
           "IND")
)

sdv_team_tiers(
  tier_data,
  sport = "wnba",
  title = "WNBA Power Rankings",
  subtitle = paste("As of", Sys.Date()),
  tier_desc = c(
    "1" = "Championship Favorites",
    "2" = "Contenders",
    "3" = "Playoff Teams",
    "4" = "Bubble Teams",
    "5" = "Developing"
  )
)
```

## WNBA Conference Standings

Visualize teams by conference:

```{r conference-standings}
# Every team's winning percentage, with the conference sdvplotR keeps for it
conference_standings <- team_stats |>
  group_by(team_abbreviation) |>
  summarise(win_pct = mean(team_winner, na.rm = TRUE), .groups = "drop") |>
  mutate(team_abbr = clean_team_abbrs(team_abbreviation, sport = "wnba")) |>
  inner_join(select(team_reference("wnba"), team_abbr, conference), by = "team_abbr") |>
  arrange(conference, desc(win_pct)) |>
  group_by(conference) |>
  mutate(team_rank = row_number()) |>
  ungroup() |>
  mutate(conference_num = as.numeric(factor(conference)))

ggplot(conference_standings, aes(x = conference_num, y = team_rank)) +
  geom_sdv_logos(
    aes(team = team_abbr),
    sport = "wnba",
    width = 0.12
  ) +
  scale_x_continuous(
    breaks = seq_along(levels(factor(conference_standings$conference))),
    labels = levels(factor(conference_standings$conference)),
    expand = expansion(add = 0.6)
  ) +
  scale_y_reverse() +
  labs(
    title = "WNBA Teams by Conference, Best Record on Top",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_blank(),
    panel.grid = element_blank()
  )
```

## WNBA Standings Table with Logos

Create a gt table with team logos:

```{r gt-table}
standings_table <- team_wins |>
  mutate(
    logo = team_abbreviation,
    rank = row_number()
  ) |>
  select(rank, logo, team_abbreviation, wins, games, win_pct)

standings_table |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "wnba", height = 35) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team_abbreviation = "Abbrev",
    wins = "Wins",
    games = "Games",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "WNBA Standings",
    subtitle = paste("Season", wehoop::most_recent_wnba_season())
  )
```

## Player Performance Comparison

Compare the scoring leaders with the assist leaders:

```{r player-comparison}
per_game <- player_stats |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    points = mean(points, na.rm = TRUE),
    assists = mean(assists, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 15)

comparison <- bind_rows(
  per_game |>
    slice_max(points, n = 5, with_ties = FALSE) |>
    transmute(athlete_id, category = "Points per Game", value = points),
  per_game |>
    slice_max(assists, n = 5, with_ties = FALSE) |>
    transmute(athlete_id, category = "Assists per Game", value = assists)
) |>
  group_by(category) |>
  mutate(rank = row_number()) |>
  ungroup()

ggplot(comparison, aes(x = rank, y = value)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "wnba",
    height = 0.2
  ) +
  facet_wrap(~category, scales = "free_y") +
  scale_x_continuous(breaks = 1:5) +
  labs(
    title = "WNBA Top Performers",
    subtitle = paste("Season", season),
    x = "Rank",
    y = NULL
  ) +
  theme_minimal()
```

## Axis Labels with Logos

Replace axis labels with team logos:

```{r axis-logos}
top_8 <- team_wins |>
  head(8) |>
  mutate(team_abbreviation = factor(team_abbreviation, levels = team_abbreviation))

ggplot(top_8, aes(x = team_abbreviation, y = win_pct)) +
  geom_col(aes(fill = team_abbreviation), width = 0.6) +
  scale_fill_sdv(sport = "wnba", alpha = 0.7) +
  scale_x_sdv(sport = "wnba") +
  theme_minimal() +
  theme_x_sdv() +
  labs(
    title = "Top 8 WNBA Teams by Win %",
    x = NULL,
    y = "Win Percentage"
  ) +
  theme(legend.position = "none")
```

## Next Steps

- Explore [wehoop documentation](https://wehoop.sportsdataverse.org/)
- Try combining with [oddsapiR](https://oddsapiR.sportsdataverse.org) for betting lines
- Build weekly dashboard with [Quarto](https://quarto.org/)

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
