---
title: "End-to-End Workflows with sdvplotR and the SportsDataverse Ecosystem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End Workflows with sdvplotR and the SportsDataverse Ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates end-to-end workflows for ingesting data from
SportsDataverse packages, creating visualizations with sdvplotR, and exporting
results. Each workflow loads real data, processes it, draws it and saves it.
We'll cover common patterns for data engineering, visualization, and reporting.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(gt)

# Companion packages, called with pkg:: below:
# nflfastR / nflreadr (NFL), cfbfastR + cfbseedR (CFB), hoopR (NBA, MBB),
# wehoop (WNBA, WBB), baseballr (MLB), fastRhockey (NHL), oddsapiR (odds)

# Exports go to a temporary directory; point this at your own folder
out_dir <- tempdir()
```

## Workflow 1: NFL Weekly Recap

Complete workflow for creating an NFL passing recap:

```{r nfl-workflow}
# Step 1: Load data with nflfastR, for the last completed regular season (NFL
# seasons are named for the year they start and end in early January)
season <- as.integer(format(Sys.Date(), "%Y")) - 1 -
  (format(Sys.Date(), "%m-%d") < "01-15")
pbp <- nflfastR::load_pbp(seasons = season, file_type = "rds")

# Step 2: Process data
team_epa <- pbp |>
  filter(pass == 1, !is.na(epa), !is.na(posteam)) |>
  group_by(posteam) |>
  summarise(
    mean_epa = mean(epa, na.rm = TRUE),
    n_plays = n(),
    .groups = "drop"
  ) |>
  filter(n_plays >= 100) |>
  slice_max(mean_epa, n = 10, with_ties = FALSE)

# Step 3: Create visualization
plot <- ggplot(team_epa, aes(x = reorder(posteam, mean_epa), y = mean_epa)) +
  geom_col(aes(fill = posteam), width = 0.7) +
  scale_fill_sdv(sport = "nfl", alpha = 0.8) +
  labs(
    title = "Top 10 NFL Teams by Pass EPA",
    subtitle = paste("Season", season),
    x = NULL,
    y = "Mean EPA per Play",
    caption = "Data: nflfastR | Viz: sdvplotR"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16)
  )

# Step 4: Export
ggsave(file.path(out_dir, "nfl_weekly_recap.png"),
       plot = plot, width = 12, height = 8, dpi = 300)

plot
```

## Workflow 2: CFB Power Rankings

Complete workflow for CFB tiers from passing efficiency:

```{r cfb-workflow}
# Step 1: Load data with cfbfastR, for the last completed regular season (named
# for the year it starts; it ends in mid-December)
cfb_season <- as.integer(format(Sys.Date(), "%Y")) -
  (format(Sys.Date(), "%m-%d") < "12-15")
cfb_pbp <- cfbfastR::load_cfb_pbp(seasons = cfb_season)

# Step 2: Process data: EPA per pass play; clean_team_abbrs() turns school
# names into sdvplotR's abbreviations
team_perf <- cfb_pbp |>
  filter(pass == 1, !is.na(EPA), !is.na(pos_team)) |>
  mutate(team = clean_team_abbrs(pos_team, sport = "cfb")) |>
  # FBS offenses only: the play-by-play also covers FCS games
  filter(team %in% team_reference("cfb")$team_abbr[team_reference("cfb")$division == "FBS"]) |>
  group_by(team) |>
  summarise(mean_epa = mean(EPA), n_plays = n(), .groups = "drop") |>
  filter(n_plays >= 100) |>
  slice_max(mean_epa, n = 16, with_ties = FALSE)

# Step 3: Create tier plot (four tiers of four)
tier_data <- team_perf |>
  mutate(tier_no = rep(1:4, each = 4)[seq_len(n())]) |>
  select(tier_no, team)

tier_plot <- sdv_team_tiers(
  tier_data,
  sport = "cfb",
  title = "CFB Passing Offense Tiers",
  subtitle = paste("EPA per pass play, season", cfb_season),
  tier_desc = c(
    "1" = "Elite",
    "2" = "Contenders",
    "3" = "Top 25",
    "4" = "Bubble"
  )
)

# Step 4: Export
ggsave(file.path(out_dir, "cfb_power_rankings.png"),
       plot = tier_plot, width = 14, height = 10, dpi = 300)

tier_plot
```

## Workflow 3: NBA Player Comparison

Complete workflow for comparing NBA scorers:

```{r nba-workflow}
# Step 1: Load data with hoopR (ESPN player box scores) for the last completed
# regular season (hoopR names a season for the year it ends, in mid-April)
nba_season <- as.integer(format(Sys.Date(), "%Y")) -
  (format(Sys.Date(), "%m-%d") < "04-20")
# ESPN tags the All-Star games as regular season too; keeping the league's own
# teams drops them
nba_box <- hoopR::load_nba_player_box(seasons = nba_season) |>
  filter(season_type == 2, !did_not_play, team_abbreviation %in% team_reference("nba")$team_abbr)

# Step 2: Process data
# one row per player: a player traded mid-season has box scores for two teams
top_players <- nba_box |>
  group_by(athlete_id, athlete_display_name) |>
  summarise(
    avg_points = mean(points, na.rm = TRUE),
    games = n(),
    .groups = "drop"
  ) |>
  filter(games >= 20) |>
  slice_max(avg_points, n = 8, with_ties = FALSE)

# Step 3: Create visualization (athlete_id is an ESPN athlete ID, the default)
player_plot <- ggplot(top_players, aes(x = games, y = avg_points)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "nba",
    height = 0.15
  ) +
  geom_label(
    aes(label = athlete_display_name),
    nudge_y = -1.5,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "Top 8 NBA Scorers",
    subtitle = paste("Season", nba_season),
    x = "Games Played",
    y = "Average Points per Game",
    caption = "Data: hoopR | Viz: sdvplotR"
  ) +
  theme_minimal()

# Step 4: Export
ggsave(file.path(out_dir, "nba_top_scorers.png"),
       plot = player_plot, width = 12, height = 8, dpi = 300)

player_plot
```

## Workflow 4: MLB Standings Dashboard

Complete workflow for MLB standings:

```{r mlb-workflow}
# Step 1: Load data with baseballr (MLB Stats API; one string for both
# leagues, because mlb_standings() rejects a vector)
# the last completed regular season (it ends around October 1)
mlb_season <- as.integer(format(Sys.Date(), "%Y")) -
  (format(Sys.Date(), "%m-%d") < "10-05")
teams <- baseballr::mlb_teams(season = mlb_season, sport_ids = 1) |>
  select(team_id, team_abbreviation, team_name = team_full_name)

# Step 2: Process data
standings <- baseballr::mlb_standings(season = mlb_season, league_id = "103,104") |>
  transmute(
    team_id = team_records_team_id,
    wins = team_records_wins,
    losses = team_records_losses,
    win_pct = as.numeric(team_records_winning_percentage)
  ) |>
  inner_join(teams, by = "team_id") |>
  slice_max(win_pct, n = 15, with_ties = FALSE) |>
  mutate(rank = row_number(), logo = team_abbreviation) |>
  select(rank, logo, team_name, wins, losses, win_pct)

# Step 3: Create gt table
standings_table <- standings |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "mlb", height = 35) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "",
    team_name = "Team",
    wins = "W",
    losses = "L",
    win_pct = "Pct"
  ) |>
  tab_header(
    title = "MLB Top 15",
    subtitle = paste("Season", mlb_season)
  )

# Step 4: Export to HTML
gtsave(standings_table, file.path(out_dir, "mlb_standings.html"))

standings_table
```

## Workflow 5: NHL Team Performance

Complete workflow for NHL team analysis:

```{r nhl-workflow}
# Step 1: Load data with fastRhockey for the last completed regular season
# (named for the year it ends; the regular season ends in mid-April)
nhl_season <- as.integer(format(Sys.Date(), "%Y")) -
  (format(Sys.Date(), "%m-%d") < "04-20")
nhl_teams <- fastRhockey::nhl_stats_teams(season = paste0(nhl_season - 1, nhl_season))

# Step 2: Process data: the NHL's full team names become abbreviations
team_perf <- nhl_teams |>
  mutate(team = clean_team_abbrs(team_full_name, sport = "nhl"))

# Step 3: Create visualization
nhl_plot <- ggplot(team_perf, aes(x = goals_for_per_game, y = goals_against_per_game)) +
  geom_sdv_logos(
    aes(team = team),
    sport = "nhl",
    width = 0.06
  ) +
  scale_y_reverse() +
  labs(
    title = "NHL Goals For and Against",
    subtitle = paste0("Season ", nhl_season - 1, "-", substr(nhl_season, 3, 4)),
    x = "Goals For per Game",
    y = "Goals Against per Game (fewer is better)",
    caption = "Data: fastRhockey | Viz: sdvplotR"
  ) +
  theme_minimal()

# Step 4: Export
ggsave(file.path(out_dir, "nhl_team_performance.png"),
       plot = nhl_plot, width = 12, height = 8, dpi = 300)

nhl_plot
```

## Workflow 6: Betting Odds Integration

Combine standings with betting lines from
[oddsapiR](https://oddsapiR.sportsdataverse.org), which needs a free key from
The Odds API in `ODDS_API_KEY`:

```{r odds-workflow}
# Step 1: Load data: regular-season records from nflreadr's schedules
nfl_standings <- nflreadr::load_schedules(season) |>
  filter(game_type == "REG", !is.na(result)) |>
  nflreadr::clean_homeaway() |>
  group_by(team) |>
  summarise(
    wins = sum(team_score > opponent_score),
    win_pct = mean(team_score > opponent_score) + 0.5 * mean(team_score == opponent_score),
    .groups = "drop"
  ) |>
  slice_max(win_pct, n = 8, with_ties = FALSE)

# Step 2: Merge upcoming spreads when a key is set; odds name teams in full,
# and clean_team_abbrs() maps full names to abbreviations
if (nzchar(Sys.getenv("ODDS_API_KEY"))) {
  spreads <- oddsapiR::toa_sports_odds(sport_key = "americanfootball_nfl", markets = "spreads") |>
    mutate(team = clean_team_abbrs(outcomes_name, sport = "nfl")) |>
    group_by(team) |>
    summarise(spread = median(outcomes_point, na.rm = TRUE), .groups = "drop")
  nfl_standings <- left_join(nfl_standings, spreads, by = "team")
}

# Step 3: Create the table
odds_table <- nfl_standings |>
  mutate(logo = team, .before = 1) |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(logo = "", team = "Team", wins = "Wins", win_pct = "Win %") |>
  tab_header(
    title = "NFL Standings with Betting Lines",
    subtitle = paste("Season", season)
  )

# Step 4: Export
gtsave(odds_table, file.path(out_dir, "nfl_standings_odds.html"))

odds_table
```

## Workflow 7: Automated Reporting Pipeline

Wrap the draw-and-export steps in a function that takes any sport's
standings, so one pipeline serves every league:

```{r automated-pipeline}
generate_standings_report <- function(standings, sport, out_dir = tempdir()) {
  message("Generating ", toupper(sport), " standings report...")

  standings <- standings |>
    arrange(desc(win_pct)) |>
    mutate(rank = row_number(), logo = team)

  plot <- ggplot(standings, aes(x = reorder(team, win_pct), y = win_pct)) +
    geom_col(aes(fill = team), width = 0.7) +
    scale_fill_sdv(sport = sport, alpha = 0.8) +
    labs(title = paste(toupper(sport), "Standings"), x = NULL, y = "Win Percentage") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  table <- standings |>
    select(rank, logo, team, win_pct) |>
    gt() |>
    gt_sdv_logos(columns = "logo", sport = sport, height = 30) |>
    fmt_number(columns = "win_pct", decimals = 3)

  # Include a timestamp in exported files
  stamp <- format(Sys.time(), "%Y%m%d_%H%M%S")
  ggsave(file.path(out_dir, paste0(sport, "_standings_", stamp, ".png")),
         plot = plot, width = 12, height = 8, dpi = 300)
  gtsave(table, file.path(out_dir, paste0(sport, "_standings_", stamp, ".html")))
  message("Report generated: ", stamp)

  list(plot = plot, table = table)
}

# The same function for two leagues
nfl_report <- generate_standings_report(select(nfl_standings, team, win_pct), "nfl")
mlb_report <- generate_standings_report(
  transmute(standings, team = logo, win_pct),
  "mlb"
)
nfl_report$plot
```

## Workflow 8: Multi-Sport Comparison

Collect each league's best record into one table:

```{r multi-sport-comparison}
nba_best <- hoopR::load_nba_standings(seasons = nba_season) |>
  filter(stat_name == "winPercent") |>
  slice_max(value, n = 1, with_ties = FALSE)

multi_sport_comparison <- bind_rows(
  nfl_standings |> slice_max(win_pct, n = 1, with_ties = FALSE) |>
    transmute(sport = "NFL", top_team = team, win_pct),
  nba_best |> transmute(sport = "NBA", top_team = team_abbreviation, win_pct = value),
  standings |> slice_max(win_pct, n = 1, with_ties = FALSE) |>
    transmute(sport = "MLB", top_team = logo, win_pct),
  team_perf |> slice_max(point_pct, n = 1, with_ties = FALSE) |>
    transmute(sport = "NHL (points %)", top_team = team, win_pct = point_pct)
)

comparison_table <- multi_sport_comparison |>
  gt() |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    sport = "Sport",
    top_team = "Top Team",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "Multi-Sport Leaderboard",
    subtitle = "The best record in each league's latest season"
  ) |>
  tab_footnote(
    footnote = "Data: SportsDataverse packages | Viz: sdvplotR",
    locations = cells_title(groups = "title")
  )

comparison_table
```

## Best Practices for Workflows

1. **Modularize Code**: Break workflows into reusable functions

2. **Use Parameters**: Make workflows flexible with function parameters

3. **Error Handling**: Add error handling for data loading and processing

4. **Logging**: Use `message()` or `cli::cli_alert_*()` to track progress

5. **Versioning**: Include timestamps in exported files

6. **Automation**: Use cron jobs or GitHub Actions for scheduled runs

7. **Documentation**: Document each workflow step with comments

## Scheduling Automated Reports

Use cron or GitHub Actions for automated reporting:

```{r scheduling}
# Example: GitHub Actions workflow
# .github/workflows/weekly_report.yml

# name: Weekly Sports Report
# on:
#   schedule:
#     - cron: '0 12 * * 1'  # Every Monday at noon
#   workflow_dispatch:
# jobs:
#   generate-report:
#     runs-on: ubuntu-latest
#     steps:
#       - uses: actions/checkout@v5
#       - uses: r-lib/actions/setup-r@v2
#       - name: Install packages
#         run: Rscript -e "install.packages(c('sdvplotR', 'nflfastR', 'gt'))"
#       - name: Generate report
#         run: Rscript scripts/weekly_report.R
#       - name: Upload artifacts
#         uses: actions/upload-artifact@v4
#         with:
#           name: weekly-reports
#           path: reports/
```

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
- [NFL Visualizations](nfl-viz.html)
- [CFB Visualizations](cfb-viz.html)
- [NBA Visualizations](nba-viz.html)
