---
title: "End-to-End Workflows with sdvplotR and the SportsDataverse Ecosystem"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-End Workflows with sdvplotR and the SportsDataverse Ecosystem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

This vignette demonstrates end-to-end workflows for ingesting data from
SportsDataverse packages, creating visualizations with sdvplotR, and exporting
results. We'll cover common patterns for data engineering, visualization, and
reporting.

## Setup

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr)
library(gt)

# Load companion packages
# library(nflfastR)      # NFL
# library(cfbfastR)      # CFB
# library(hoopR)         # NBA, MBB
# library(wehoop)        # WNBA, WBB
# library(baseballr)     # MLB
# library(fastRhockey)   # NHL
# library(oddsapiR)      # Betting odds
# library(sportyR)       # General sports data
```

## Workflow 1: NFL Weekly Recap

Complete workflow for creating a weekly NFL recap:

```{r nfl-workflow}
# Step 1: Load data with nflfastR
# pbp <- nflfastR::load_pbp(
#   seasons = nflreadr::most_recent_season(),
#   file_type = "rds"
# )

# Step 2: Process data
# team_epa <- pbp |>
#   filter(pass == 1, !is.na(epa)) |>
#   group_by(posteam) |>
#   summarise(
#     mean_epa = mean(epa, na.rm = TRUE),
#     n_plays = n(),
#     .groups = "drop"
#   ) |>
#   filter(n_plays >= 100) |>
#   arrange(desc(mean_epa)) |>
#   head(10)

# Step 3: Create visualization
# For demonstration, use sample data
team_epa <- data.frame(
  posteam = c("KC", "BUF", "SF", "PHI", "DAL", "MIA", "CIN", "BAL", "DET", "JAX"),
  mean_epa = sort(runif(10, 0.1, 0.3), decreasing = TRUE),
  n_plays = sample(200:500, 10)
)

plot <- ggplot(team_epa, aes(x = reorder(posteam, mean_epa), y = mean_epa)) +
  geom_col(aes(fill = posteam), width = 0.7) +
  scale_fill_sdv(sport = "nfl", alpha = 0.8) +
  labs(
    title = "Top 10 NFL Teams by Pass EPA",
    subtitle = paste("Season", nflreadr::most_recent_season()),
    x = NULL,
    y = "Mean EPA per Play",
    caption = "Data: nflfastR | Viz: sdvplotR"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 16)
  )

# Step 4: Export
# ggsave("nfl_weekly_recap.png",
#        plot = plot,
#        width = 12, height = 8, dpi = 300)

plot
```

## Workflow 2: CFB Power Rankings

Complete workflow for CFB power rankings:

```{r cfb-workflow}
# Step 1: Load data with cfbfastR
# pbp <- cfbfastR::load_cfb_pbp(
#   seasons = cfbfastR::most_recent_cfb_season(),
#   epa_wpa = TRUE
# )

# Step 2: Process data
# team_perf <- pbp |>
#   filter(pass == 1, !is.na(EPA)) |>
#   group_by(posteam) |>
#   summarise(
#     mean_epa = mean(EPA, na.rm = TRUE),
#     n_plays = n(),
#     .groups = "drop"
#   ) |>
#   filter(n_plays >= 100) |>
#   arrange(desc(mean_epa)) |>
#   head(25)

# Step 3: Create tier plot
# For demonstration, use sample data
tier_data <- data.frame(
  tier_no = c(1, 1, 1, 2, 2, 2, 2, 3, 3, 3, 3, 3, 4, 4, 4, 4),
  team = c("UGA", "MICH", "WASH", "FSU", "OSU", "TEX", "ALA",
           "ORE", "PSU", "MIZ", "AZ", "LSU",
           "CLE", "GB", "LAR", "MIN")
)

tier_plot <- sdv_team_tiers(
  tier_data,
  sport = "cfb",
  title = "CFB Power Rankings",
  subtitle = paste("Week", ceiling(runif(1, 1, 15)), "of", cfbfastR::most_recent_cfb_season()),
  tier_desc = c(
    "1" = "Elite",
    "2" = "Contenders",
    "3" = "Top 25",
    "4" = "Bubble"
  )
)

# Step 4: Export
# ggsave("cfb_power_rankings.png",
#        plot = tier_plot,
#        width = 14, height = 10, dpi = 300)

tier_plot
```

## Workflow 3: NBA Player Comparison

Complete workflow for comparing NBA players:

```{r nba-workflow}
# Step 1: Load data with hoopR
# player_stats <- hoopR::load_nba_player_stats(
#   seasons = hoopR::most_recent_nba_season(),
#   season_type = "Regular Season"
# )

# Step 2: Process data
# top_players <- player_stats |>
#   filter(!is.na(athlete_id)) |>
#   group_by(athlete_id, athlete_display_name, team_abbreviation) |>
#   summarise(
#     avg_points = mean(points, na.rm = TRUE),
#     games = n(),
#     .groups = "drop"
#   ) |>
#   filter(games >= 20) |>
#   arrange(desc(avg_points)) |>
#   head(8)

# Step 3: Create visualization
# For demonstration, use sample data
top_players <- data.frame(
  athlete_id = c("1234", "5678", "9012", "3456", "7890", "2345", "6789", "0123"),
  athlete_display_name = c("LeBron James", "Stephen Curry", "Kevin Durant",
                          "Giannis Antetokounmpo", "Luka Doncic", "Jayson Tatum",
                          "Nikola Jokic", "Joel Embiid"),
  team_abbreviation = c("LAL", "GSW", "PHX", "MIL", "DAL", "BOS", "DEN", "PHI"),
  avg_points = sort(runif(8, 25, 35), decreasing = TRUE),
  games = sample(20:40, 8)
)

player_plot <- ggplot(top_players, aes(x = games, y = avg_points)) +
  geom_sdv_headshots(
    aes(player_id = athlete_id),
    sport = "nba",
    height = 0.15
  ) +
  geom_label(
    aes(label = athlete_display_name),
    nudge_y = -1.5,
    size = 3,
    alpha = 0.7
  ) +
  labs(
    title = "Top 8 NBA Scorers",
    subtitle = paste("Season", hoopR::most_recent_nba_season()),
    x = "Games Played",
    y = "Average Points per Game",
    caption = "Data: hoopR | Viz: sdvplotR"
  ) +
  theme_minimal()

# Step 4: Export
# ggsave("nba_top_scorers.png",
#        plot = player_plot,
#        width = 12, height = 8, dpi = 300)

player_plot
```

## Workflow 4: MLB Standings Dashboard

Complete workflow for MLB standings:

```{r mlb-workflow}
# Step 1: Load data with baseballr
# team_stats <- baseballr::get_team_stats(
#   season = baseballr::most_recent_mlb_season(),
#   team_id = NULL
# )

# Step 2: Process data
# standings <- team_stats |>
#   filter(!is.na(team_abbreviation)) |>
#   group_by(team_abbreviation) |>
#   summarise(
#     wins = sum(win == 1, na.rm = TRUE),
#     games = n(),
#     .groups = "drop"
#   ) |>
#   filter(games >= 10) |>
#   mutate(win_pct = wins / games) |>
#   arrange(desc(win_pct)) |>
#   head(15)

# Step 3: Create gt table
# For demonstration, use sample data
standings <- data.frame(
  team_abbreviation = c("LAD", "ATL", "HOU", "BAL", "TB", "NYY",
                        "PHI", "TEX", "TOR", "SEA", "SF", "CHC",
                        "BOS", "STL", "CIN"),
  wins = sort(sample(80:105, 15), decreasing = TRUE),
  games = rep(162, 15)
) |>
  mutate(
    win_pct = wins / games,
    logo = team_abbreviation,
    rank = row_number()
  ) |>
  select(rank, logo, team_abbreviation, wins, games, win_pct)

standings_table <- standings |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "mlb", height = 35) |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    rank = "#",
    logo = "Team",
    team_abbreviation = "Abbrev",
    wins = "Wins",
    games = "Games",
    win_pct = "Win %"
  ) |>
  tab_header(
    title = "MLB Top 15",
    subtitle = paste("Season", baseballr::most_recent_mlb_season())
  )

# Step 4: Export to HTML
# standings_table |>
#   gtsave("mlb_standings.html")

standings_table
```

## Workflow 5: NHL Team Performance

Complete workflow for NHL team analysis:

```{r nhl-workflow}
# Step 1: Load data with fastRhockey
# team_stats <- fastRhockey::load_nhl_team_stats(
#   seasons = fastRhockey::most_recent_nhl_season(),
#   level = "team"
# )

# Step 2: Process data
# team_perf <- team_stats |>
#   filter(!is.na(team_abbreviation)) |>
#   group_by(team_abbreviation) |>
#   summarise(
#     avg_goals = mean(goals, na.rm = TRUE),
#     avg_assists = mean(assists, na.rm = TRUE),
#     games = n(),
#     .groups = "drop"
#   ) |>
#   filter(games >= 10)

# Step 3: Create visualization
# For demonstration, use sample data
team_perf <- data.frame(
  team_abbreviation = sample(valid_team_names("nhl"), 16),
  avg_goals = runif(16, 2.5, 3.5),
  avg_assists = runif(16, 3.5, 4.5),
  games = sample(20:40, 16)
)

nhl_plot <- ggplot(team_perf, aes(x = avg_goals, y = avg_assists)) +
  geom_sdv_logos(
    aes(team = team_abbreviation),
    sport = "nhl",
    width = 0.075
  ) +
  labs(
    title = "NHL Team Performance",
    subtitle = paste("Season", fastRhockey::most_recent_nhl_season()),
    x = "Average Goals per Game",
    y = "Average Assists per Game",
    caption = "Data: fastRhockey | Viz: sdvplotR"
  ) +
  theme_minimal()

# Step 4: Export
# ggsave("nhl_team_performance.png",
#        plot = nhl_plot,
#        width = 12, height = 8, dpi = 300)

nhl_plot
```

## Workflow 6: Betting Odds Integration

Combine standings with betting odds using oddsapiR:

```{r odds-workflow}
# Step 1: Load data
# standings <- nflfastR::load_pbp(
#   seasons = nflreadr::most_recent_season(),
#   file_type = "rds"
# ) |>
#   filter(!is.na(result)) |>
#   group_by(posteam) |>
#   summarise(
#     wins = sum(result > 0, na.rm = TRUE),
#     games = n(),
#     .groups = "drop"
#   ) |>
#   filter(games >= 10) |>
#   mutate(win_pct = wins / games) |>
#   arrange(desc(win_pct))

# odds <- oddsapiR::get_odds(
#   sport = "americanfootball_nfl",
#   date = Sys.Date()
# )

# Step 2: Merge data
# For demonstration, use sample data
standings_with_odds <- data.frame(
  team = c("KC", "BUF", "SF", "PHI", "DAL"),
  wins = c(9, 8, 8, 7, 7),
  win_pct = c(0.818, 0.727, 0.727, 0.636, 0.636),
  spread = c(-7.5, -3.5, -6.5, -10.5, -4.5),
  moneyline = c(-350, -180, -280, -550, -220),
  logo = c("KC", "BUF", "SF", "PHI", "DAL")
)

odds_table <- standings_with_odds |>
  gt() |>
  gt_sdv_logos(columns = "logo", sport = "nfl", height = 35) |>
  fmt_number(columns = c("win_pct", "spread"), decimals = 1) |>
  fmt_number(columns = "moneyline", decimals = 0) |>
  cols_label(
    logo = "Team",
    team = "Abbrev",
    wins = "Wins",
    win_pct = "Win %",
    spread = "Spread",
    moneyline = "ML"
  ) |>
  tab_header(
    title = "NFL Standings with Betting Odds",
    subtitle = "Combining Performance and Odds"
  )

# Step 3: Export
# odds_table |>
#   gtsave("nfl_standings_odds.html")

odds_table
```

## Workflow 7: Automated Reporting Pipeline

Create an automated reporting pipeline:

```{r automated-pipeline}
# Define a function for automated reporting
generate_weekly_report <- function(sport, week = NULL) {
  message(paste("Generating", toupper(sport), "weekly report..."))

  # Step 1: Load data (sport-specific)
  # In real implementation, load from appropriate package

  # Step 2: Process data
  # For demonstration, use sample data
  standings <- data.frame(
    team = sample(valid_team_names(sport), 10),
    wins = sample(0:16, 10),
    losses = sample(0:16, 10)
  ) |>
    mutate(
      win_pct = wins / (wins + losses),
      logo = team,
      rank = row_number()
    ) |>
    arrange(desc(win_pct))

  # Step 3: Create visualizations
  plot <- ggplot(standings, aes(x = reorder(team, win_pct), y = win_pct)) +
    geom_col(aes(fill = team), width = 0.7) +
    scale_fill_sdv(sport = sport, alpha = 0.8) +
    labs(
      title = paste(toupper(sport), "Standings"),
      subtitle = ifelse(is.null(week), "Current", paste("Week", week)),
      x = NULL,
      y = "Win Percentage"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )

  table <- standings |>
    select(rank, logo, team, wins, losses, win_pct) |>
    gt() |>
    gt_sdv_logos(columns = "logo", sport = sport, height = 30) |>
    fmt_number(columns = "win_pct", decimals = 3)

  # Step 4: Export
  timestamp <- format(Sys.time(), "%Y%m%d_%H%M%S")

  # ggsave(paste0(sport, "_standings_", timestamp, ".png"),
  #        plot = plot,
  #        width = 12, height = 8, dpi = 300)

  # table |>
  #   gtsave(paste0(sport, "_standings_", timestamp, ".html"))

  message(paste("Report generated:", timestamp))

  list(plot = plot, table = table)
}

# Generate reports for multiple sports
# nfl_report <- generate_weekly_report("nfl", week = 10)
# nba_report <- generate_weekly_report("nba")
# cfb_report <- generate_weekly_report("cfb")
```

## Workflow 8: Multi-Sport Comparison

Create a multi-sport comparison dashboard:

```{r multi-sport-comparison}
# Load data from multiple sports
# nfl_data <- nflfastR::load_pbp(seasons = nflreadr::most_recent_season())
# nba_data <- hoopR::load_nba_team_stats(seasons = hoopR::most_recent_nba_season())
# mlb_data <- baseballr::get_team_stats(season = baseballr::most_recent_mlb_season())

# Create comparison table
# For demonstration, use sample data
multi_sport_comparison <- data.frame(
  sport = c("NFL", "NBA", "MLB", "NHL", "CFB"),
  top_team = c("KC", "BOS", "LAD", "COL", "UGA"),
  win_pct = c(0.818, 0.758, 0.586, 0.616, 0.917),
  conference = c("AFC West", "Atlantic", "NL West", "Central", "SEC")
)

comparison_table <- multi_sport_comparison |>
  gt() |>
  fmt_number(columns = "win_pct", decimals = 3) |>
  cols_label(
    sport = "Sport",
    top_team = "Top Team",
    win_pct = "Win %",
    conference = "Conference/Division"
  ) |>
  tab_header(
    title = "Multi-Sport Leaderboard",
    subtitle = "Top Teams Across All Sports"
  ) |>
  tab_footnote(
    footnote = "Data: SportsDataverse packages | Viz: sdvplotR",
    locations = cells_title(groups = "title")
  )

comparison_table
```

## Best Practices for Workflows

1. **Modularize Code**: Break workflows into reusable functions

2. **Use Parameters**: Make workflows flexible with function parameters

3. **Error Handling**: Add error handling for data loading and processing

4. **Logging**: Use `message()` or `cli::cli_alert_*()` to track progress

5. **Versioning**: Include timestamps in exported files

6. **Automation**: Use cron jobs or GitHub Actions for scheduled runs

7. **Documentation**: Document each workflow step with comments

## Scheduling Automated Reports

Use cron or GitHub Actions for automated reporting:

```{r scheduling}
# Example: GitHub Actions workflow
# .github/workflows/weekly_report.yml

# name: Weekly Sports Report
# on:
#   schedule:
#     - cron: '0 12 * * 1'  # Every Monday at noon
#   workflow_dispatch:
# jobs:
#   generate-report:
#     runs-on: ubuntu-latest
#     steps:
#       - uses: actions/checkout@v3
#       - uses: r-lib/actions/setup-r@v2
#       - name: Install packages
#         run: Rscript -e "install.packages(c('sdvplotR', 'nflfastR', 'gt'))"
#       - name: Generate report
#         run: Rscript scripts/weekly_report.R
#       - name: Upload artifacts
#         uses: actions/upload-artifact@v3
#         with:
#           name: weekly-reports
#           path: reports/
```

## Related Vignettes

- [Getting Started](getting-started.html)
- [Social Posting Patterns](social-posting.html)
- [Leaderboard Dashboards](leaderboard-dashboards.html)
- [NFL Visualizations](nfl-viz.html)
- [CFB Visualizations](cfb-viz.html)
- [NBA Visualizations](nba-viz.html)
