---
title: "Surfaces and shot charts"
description: >
  Ten recipes for playing surfaces: every sport's surface, team colors and a
  center logo, NBA and WNBA shot charts with sdv_court_coords(), a rotated half
  court, a college shot chart, a hockey goal map, a baseball field and a
  football drive chart.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

Ten recipes for playing surfaces and the charts drawn on them:
`sdv_surface()` for every sport it supports, team colors and a center logo,
NBA and WNBA shot charts from the stats-API shot files (moved onto the court
with `sdv_court_coords()`), a vertical half court, a college shot chart from
ESPN's locations, a hockey goal map, a baseball field and a football drive
chart. The surfaces are drawn by [sportyR](https://sportyR.sportsdataverse.org).
The data is one season each from the NFL (nflreadr), the NBA, WNBA and
college basketball (hoopR and wehoop), the NHL (fastRhockey) and MLB
(baseballr). The NBA and WNBA shots are the stats-API shot files the
SportsDataverse publishes as GitHub releases, so nothing here calls
stats.nba.com or stats.wnba.com.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

# sportyR's surfaces have no plot margin, which clips a title; give it room
titled <- theme(plot.margin = margin(12, 10, 8, 10))
```

## 1. Every surface sdvplotR can draw

`sdv_surface(sport)` picks sportyR's regulation surface for the league: a
football field for the NFL and college football (their hash marks differ), a
court for the NBA, the WNBA and college basketball, a rink and a baseball
field. Surfaces use sportyR's coordinates: the origin at the center, in feet
(yards for football), and the baseball field's origin at home plate.

```{r every-surface, fig.width = 6, fig.height = 3.6, fig.show = "hold", out.width = "49%", fig.alt = "Seven regulation playing surfaces drawn by sportyR: an NFL field, a college football field, NBA, WNBA and college basketball courts, an NHL rink and an MLB infield."}
surfaces <- c(
  nfl = "NFL field", cfb = "College football field", nba = "NBA court",
  wnba = "WNBA court", mbb = "College basketball court (men's and women's)",
  nhl = "NHL rink", mlb = "MLB field"
)
for (sport in names(surfaces)) {
  p <- sdv_surface(sport) + labs(title = surfaces[[sport]]) + titled
  # the baseball field paints the whole plot background grass green
  if (sport == "mlb") p <- p + theme(plot.title = element_text(colour = "white"))
  print(p)
}
```

`"wbb"` draws the same college court as `"mbb"`. sportyR's theme leaves no
margin around the surface, so each plot here adds one (`titled`, from the
setup chunk) to keep its title from being clipped.

## 2. The champion's field, with a logo at midfield

Given a `team`, the surface takes its colors (end zones, the paint and apron,
the center ice markings and boards); `center_logo = TRUE` adds its logo at
the center. The Super Bowl winner, read from nflverse's schedule:

```{r champion-field, fig.alt = "An NFL field with both end zones in the Seattle Seahawks' navy and the Seahawks logo at midfield, for the winner of the Super Bowl after the 2025 season."}
super_bowl <- nflreadr::load_schedules(2025) |>
  as.data.frame() |>
  filter(game_type == "SB")
champion <- with(super_bowl, if_else(home_score > away_score, home_team, away_team))
super_bowl[, c("away_team", "away_score", "home_team", "home_score")]

sdv_surface("nfl", champion, center_logo = TRUE) +
  labs(title = paste0("Super Bowl champions: ", champion), caption = "Data: nflreadr::load_schedules()") +
  titled
```

These are stylized surfaces from the team's two colors, not the stadium's real
paint; `color_updates =` (passed to sportyR) overrides any feature.

## 3. Every basketball court, in its league's top team

NBA, WNBA and college courts differ in their lines (the three-point arc, the
lane width), and sportyR draws each league's own. Each court here belongs to
its league's best team: the most regular-season wins for the pros, the best
adjusted efficiency margin in men's college basketball, and the best average
scoring margin in the women's game. The pros' regular seasons leave out the
All-Star and cup-final games, which ESPN also files as regular season.

```{r top-courts, fig.width = 6, fig.height = 3.6, fig.show = "hold", out.width = "49%", fig.alt = "Four basketball courts in the colors of each league's best 2025-26 or 2026 team, each with the team's logo at center court: the NBA, the WNBA, men's college basketball and women's college basketball."}
most_wins <- function(schedule, box) {
  standard <- schedule |>
    as.data.frame() |>
    filter(season_type == 2, type_abbreviation == "STD") |>
    pull(game_id)
  box |>
    as.data.frame() |>
    filter(game_id %in% standard) |>
    summarise(wins = sum(team_winner), .by = team_abbreviation) |>
    arrange(desc(wins), team_abbreviation) |>
    slice_head(n = 1)
}
nba_best <- most_wins(hoopR::load_nba_schedule(2026), hoopR::load_nba_team_box(2026))
wnba_best <- most_wins(wehoop::load_wnba_schedule(2026), wehoop::load_wnba_team_box(2026))

mbb_ref <- team_reference("mbb")
mbb_best <- hoopR::load_mbb_ratings(2026) |>
  as.data.frame() |>
  mutate(team = mbb_ref$team_abbr[match(team_id, mbb_ref$espn_team_id)]) |>
  filter(!is.na(team)) |>
  slice_max(adj_em, n = 1, with_ties = FALSE)

wbb_d1 <- wehoop::load_wbb_standings(2026) |>
  as.data.frame() |>
  distinct(team_id)
wbb_best <- wehoop::load_wbb_team_box(2026) |>
  as.data.frame() |>
  filter(season_type == 2, team_id %in% wbb_d1$team_id) |>
  summarise(margin = mean(team_score - opponent_team_score), .by = team_abbreviation) |>
  slice_max(margin, n = 1, with_ties = FALSE)

best <- c(
  nba = nba_best$team_abbreviation, wnba = wnba_best$team_abbreviation,
  mbb = mbb_best$team, wbb = wbb_best$team_abbreviation
)
best

for (sport in names(best)) {
  print(
    sdv_surface(sport, best[[sport]], center_logo = TRUE) +
      labs(title = paste(toupper(sport), best[[sport]])) +
      titled
  )
}
```

## 4. An NBA shot chart from the stats-API frame

hoopR's `load_nba_stats_shots()` reads the stats-API shot file the
SportsDataverse publishes on GitHub: `x_legacy` / `y_legacy` in tenths of a
foot with the hoop at the origin. sportyR's court has its origin at center
court, in feet. `sdv_court_coords()` converts one to the other and adds
`court_x` / `court_y`; every shot lands on the left half, so draw that half
with `display_range = "defense"`. This loader names a season by the year it
starts (`2025` is 2025-26), and a game id starting `002` is the regular
season. The season's top scorer from the floor:

```{r nba-shots, fig.alt = "Half-court shot chart of the 2025-26 NBA regular season's leading scorer from the field, every field goal attempt as a dot on his team's court: made shots filled in the team's primary color, misses in white, each with a dark outline."}
nba_shots <- hoopR::load_nba_stats_shots(2025) |>
  as.data.frame() |>
  filter(substr(game_id, 1, 3) == "002")
stopifnot(all(nba_shots$season == 2026)) # the file's own season column is the ending year

leader <- nba_shots |>
  summarise(
    points = sum(shot_value[shot_result == "Made"]),
    team = names(which.max(table(team_tricode))),
    .by = c(person_id, player_name)
  ) |>
  slice_max(points, n = 1, with_ties = FALSE)
leader

shots <- nba_shots |>
  filter(person_id == leader$person_id) |>
  sdv_court_coords()
head(shots[, c("x_legacy", "y_legacy", "court_x", "court_y")], 3)

team <- clean_team_abbrs(leader$team, sport = "nba")
sdv_surface("nba", team, display_range = "defense") +
  geom_point(aes(court_x, court_y, fill = shot_result), data = shots,
    shape = 21, colour = "grey15", stroke = 0.25, size = 1.6, alpha = 0.8
  ) +
  scale_fill_manual(
    values = c(Made = unname(sdv_team_colors("nba", team)), Missed = "white"),
    name = NULL
  ) +
  labs(
    title = paste0(leader$player_name, ", 2025-26 regular season"),
    subtitle = sprintf("%s field goal attempts", format(nrow(shots), big.mark = ",")),
    caption = "Data: hoopR::load_nba_stats_shots()"
  ) +
  titled +
  theme(legend.position = "bottom")
```

The paint is in the team's primary color too, so a made shot drawn in that
color would vanish there: an outline (`shape = 21`) keeps every dot visible.
The stats API's own tricodes (`OKC`, `GSW`, `SAS`) go through
`clean_team_abbrs()` to sdvplotR's keys. hoopR's `nba_shotchartdetail()` uses
the same frame in `LOC_X` / `LOC_Y`: pass those names to `sdv_court_coords()`.

## 5. The same for the WNBA

stats.wnba.com uses the same frame, so `sdv_court_coords()` works unchanged,
and `sdv_surface("wnba")` draws the WNBA's court, whose three-point arc is
closer to the basket than the NBA's. wehoop's loader takes the season's own
year, and a game id starting `1022` is the regular season:

```{r wnba-shots, fig.alt = "Half-court shot chart of the 2026 WNBA regular season's leading scorer from the field, every field goal attempt on her team's court: made shots filled in the team's primary color, misses in white, each with a dark outline."}
wnba_shots <- wehoop::load_wnba_stats_shots(2026) |>
  as.data.frame() |>
  filter(substr(game_id, 1, 4) == "1022")

wnba_leader <- wnba_shots |>
  summarise(
    points = sum(shot_value[shot_result == "Made"]),
    team = names(which.max(table(team_tricode))),
    .by = c(person_id, player_name)
  ) |>
  slice_max(points, n = 1, with_ties = FALSE)
wnba_leader

wnba_team <- clean_team_abbrs(wnba_leader$team, sport = "wnba")
wnba_player <- wnba_shots |>
  filter(person_id == wnba_leader$person_id) |>
  sdv_court_coords()

sdv_surface("wnba", wnba_team, display_range = "defense") +
  geom_point(aes(court_x, court_y, fill = shot_result), data = wnba_player,
    shape = 21, colour = "grey15", stroke = 0.25, size = 1.7, alpha = 0.8
  ) +
  scale_fill_manual(
    values = c(Made = unname(sdv_team_colors("wnba", wnba_team)), Missed = "white"),
    name = NULL
  ) +
  labs(
    title = paste0(wnba_leader$player_name, ", 2026 regular season"),
    subtitle = sprintf("%s field goal attempts", format(nrow(wnba_player), big.mark = ",")),
    caption = "Data: wehoop::load_wnba_stats_shots()"
  ) +
  titled +
  theme(legend.position = "bottom")
```

## 6. A vertical half court: rotate the surface and the data

sportyR's `rotation` turns the drawing; turn the points the same way. A
quarter turn counterclockwise (`rotation = 90`) maps each point `(x, y)` to
`(-y, x)` and puts the basket at the bottom, the way most shot charts are
drawn. `display_range` still names the half of the unrotated court. The NBA's
second-leading scorer from the floor, with a two-dimensional density of his
attempts over the dots (both in neutral colors, which read on the wood and on
the team-colored paint):

```{r vertical-court, fig.height = 6, fig.alt = "Vertical half-court shot chart, basket at the bottom, of the 2025-26 NBA regular season's second-leading scorer from the field: every attempt as a small white dot with black density contours over them, on a court in his team's colors."}
second <- nba_shots |>
  summarise(
    points = sum(shot_value[shot_result == "Made"]),
    team = names(which.max(table(team_tricode))),
    .by = c(person_id, player_name)
  ) |>
  arrange(desc(points), person_id) |>
  slice(2)
second_team <- clean_team_abbrs(second$team, sport = "nba")

turned <- nba_shots |>
  filter(person_id == second$person_id) |>
  sdv_court_coords() |>
  mutate(x = -court_y, y = court_x) # a quarter turn counterclockwise

sdv_surface("nba", second_team, display_range = "defense", rotation = 90) +
  geom_point(aes(x, y), data = turned, shape = 21, fill = "white", colour = "grey20",
    stroke = 0.2, alpha = 0.6, size = 1
  ) +
  geom_density_2d(aes(x, y), data = turned, colour = "black", linewidth = 0.4, bins = 8) +
  labs(
    title = paste0(second$player_name, ", 2025-26 regular season"),
    subtitle = "Every field goal attempt, with density contours",
    caption = "Data: hoopR::load_nba_stats_shots()"
  ) +
  titled
```

`rotation = 270` (or `-90`) puts the basket at the top, with points mapped to
`(y, -x)`.

## 7. A college shot chart from ESPN's locations

hoopR's and wehoop's ESPN shot files (`load_mbb_shots()`,
`load_wbb_shots()`) are in a different frame: `coordinate_x` / `coordinate_y`
are already feet from center court, with each team's shots at the basket it
attacks. Don't pass them to `sdv_court_coords()`. To put every shot at one
basket, turn the ones at the right-hand basket half a turn (flip both signs),
then draw the left half. A few rows hold placeholder values far off the court,
so keep the ones on it. The women's game's highest-volume shooter, 2025-26:

```{r college-shots, fig.alt = "Half-court shot chart, on a college court in her team's colors, of the women's college basketball player with the most field goal attempts in 2025-26: made shots filled in the team's primary color, misses in white, each with a dark outline."}
wbb_shots <- wehoop::load_wbb_shots(2026) |>
  as.data.frame() |>
  filter(
    !grepl("FreeThrow", type_text),
    abs(coordinate_x) <= 47, abs(coordinate_y) <= 25
  )

volume <- wbb_shots |>
  count(athlete_id_1, athlete_name_1, team_abbrev, name = "attempts") |>
  arrange(desc(attempts), athlete_id_1) |>
  slice_head(n = 1)
volume

college <- wbb_shots |>
  filter(athlete_id_1 == volume$athlete_id_1) |>
  mutate(
    flip = coordinate_x > 0,
    x = if_else(flip, -coordinate_x, coordinate_x),
    y = if_else(flip, -coordinate_y, coordinate_y),
    result = if_else(scoring_play, "Made", "Missed")
  )

sdv_surface("wbb", volume$team_abbrev, display_range = "defense") +
  geom_point(aes(x, y, fill = result), data = college,
    shape = 21, colour = "grey15", stroke = 0.25, size = 1.7, alpha = 0.8
  ) +
  scale_fill_manual(
    values = c(Made = unname(sdv_team_colors("wbb", volume$team_abbrev)), Missed = "white"),
    name = NULL
  ) +
  labs(
    title = paste0(volume$athlete_name_1, ", 2025-26"),
    subtitle = sprintf("%s field goal attempts, every game", format(nrow(college), big.mark = ",")),
    caption = "Data: wehoop::load_wbb_shots() (ESPN)"
  ) +
  titled +
  theme(legend.position = "bottom")
```

The college court and the NBA's share the 94-foot floor and the basket 5.25 ft
from the baseline, so these feet need no scaling.

## 8. A hockey goal map on a team-colored rink

fastRhockey's play-by-play `x` / `y` are feet on the NHL rink's frame, the
one sportyR draws, with the origin at center ice. Teams switch ends each
period, so turn every goal at the left-hand net half a turn (flip both signs)
to put them all at the right-hand one, then draw the attacking half with
`display_range = "offense"`. The season file is large (every event of every
game), so keep only the goals and let the rest go. The league's top goal
scorer, regular season, shootouts left out:

```{r goal-map, fig.alt = "Offensive half of an NHL rink in the colors of the 2025-26 regular season's top goal scorer's team, with each of his goals as a dot at the spot it was shot from."}
goals <- fastRhockey::load_nhl_pbp(2026) |>
  as.data.frame() |>
  filter(event_type == "GOAL", season_type == "R", period_type != "SHOOTOUT") |>
  select(game_id, event_team_abbr, event_player_1_id, event_player_1_name, x, y, empty_net)
invisible(gc())

scorer <- goals |>
  count(event_player_1_id, event_player_1_name, event_team_abbr, name = "goals") |>
  arrange(desc(goals), event_player_1_id) |>
  slice_head(n = 1)
scorer

his <- goals |>
  filter(event_player_1_id == scorer$event_player_1_id, !is.na(x)) |>
  mutate(
    x_net = if_else(x < 0, -x, x),
    y_net = if_else(x < 0, -y, y),
    net = if_else(empty_net, "Empty net", "Goalie in net")
  )
nhl_team <- clean_team_abbrs(scorer$event_team_abbr, sport = "nhl")

sdv_surface("nhl", nhl_team, display_range = "offense") +
  geom_point(aes(x_net, y_net, shape = net), data = his,
    size = 2.6, fill = sdv_team_colors("nhl", nhl_team, type = "secondary"), colour = "black", stroke = 0.4
  ) +
  scale_shape_manual(values = c(`Goalie in net` = 21, `Empty net` = 24), name = NULL) +
  labs(
    title = sprintf("%s's %d goals, 2025-26", scorer$event_player_1_name, scorer$goals),
    subtitle = "Regular season, from where each was shot",
    caption = "Data: fastRhockey::load_nhl_pbp()"
  ) +
  titled +
  theme(legend.position = "bottom")
```

`x_fixed` / `y_fixed` in the same file already put the home team's shots on
the right, but they mirror the away team's (only `x` changes sign); the half
turn here keeps each shot on the side it was taken from.

## 9. A baseball field with a logo in center field

Baseball fields have no team-colored lines, so `sdv_surface("mlb", team)`
draws the regulation field, and the origin is home plate, so a center logo
would sit on the plate (`center_logo` warns and skips it). The field also
paints the whole plot background green, so set titles in white. Draw the field,
then place the logo yourself, in center field about 330 ft out. The team with
MLB's best 2026 record:

```{r baseball-field, fig.alt = "A regulation MLB baseball field with the logo of the team with the best 2026 regular-season record in center field."}
mlb_ids <- baseballr::mlb_teams(season = 2026, sport_ids = 1) |>
  as.data.frame() |>
  select(team_records_team_id = team_id, team_abbreviation, team_full_name)
mlb_best <- baseballr::mlb_standings(season = 2026, league_id = "103,104") |>
  as.data.frame() |>
  inner_join(mlb_ids, by = "team_records_team_id") |>
  slice_max(team_records_wins, n = 1, with_ties = FALSE) |>
  transmute(
    team = clean_team_abbrs(team_abbreviation, sport = "mlb"),
    name = team_full_name, wins = team_records_wins, losses = team_records_losses
  )
mlb_best

sdv_surface("mlb") +
  geom_sdv_logos(aes(x = 0, y = 330, team = team), data = mlb_best, sport = "mlb", width = 0.16) +
  labs(
    title = sprintf("%s, %d-%d: MLB's best record in 2026", mlb_best$name, mlb_best$wins, mlb_best$losses),
    caption = "Data: baseballr::mlb_standings() (MLB Stats API)"
  ) +
  titled +
  theme(plot.title = element_text(colour = "white"), plot.caption = element_text(colour = "white"))
```

Any ggplot layer goes on a surface the same way, in the surface's units: here
feet from home plate, so a spray chart's hit locations drop straight in.

## 10. A football drive chart

A field is a canvas for drives: one arrow per possession, from where it
started to where it ended, the winner driving left to right and the loser
right to left. nflverse's play-by-play gives each drive's start and end as
`"SEA 25"` (a team's own 25) or `"NE 30"` (the opponent's 30); convert both to
yards from the offense's own goal line. A field's `x` runs from -50 (one goal
line) to 50 (the other) in yards. Super Bowl LX:

```{r drive-chart, fig.height = 5.5, fig.alt = "An NFL field with one arrow per drive of Super Bowl LX: the Seattle Seahawks' drives in the top half running left to right in navy and the New England Patriots' in the bottom half running right to left in red, each labelled with how it ended."}
pbp <- nflreadr::load_pbp(2025) |>
  as.data.frame() |>
  filter(game_id == super_bowl$game_id, !is.na(posteam), !is.na(fixed_drive))

own_yards <- function(spot, team) {
  side <- sub(" .*", "", spot)
  yards <- as.numeric(sub(".* ", "", spot))
  if_else(side == team, yards, 100 - yards) # "MID 50" is 50 either way
}

drives <- pbp |>
  summarise(
    start = first(drive_start_yard_line),
    end = first(drive_end_yard_line),
    result = first(fixed_drive_result),
    .by = c(fixed_drive, posteam)
  ) |>
  mutate(
    from = own_yards(start, posteam),
    to = if_else(result == "Touchdown", 100, own_yards(end, posteam)),
    winner = posteam == champion,
    x = if_else(winner, from - 50, 50 - from),
    xend = if_else(winner, to - 50, 50 - to)
  ) |>
  filter(!is.na(from), !is.na(to)) |>
  arrange(fixed_drive) |>
  mutate(lane = row_number(), .by = winner)
gap <- 25 / max(drives$lane) # spread each team's drives over its half of the field's width
drives <- mutate(drives, y = if_else(winner, gap * lane, -gap * lane))
count(drives, posteam, result)

# both teams' primaries are navy, so the loser takes its secondary
loser <- setdiff(drives$posteam, champion)
drive_colors <- c(sdv_team_colors("nfl", champion), sdv_team_colors("nfl", loser, type = "secondary"))
drive_colors

sdv_surface("nfl") +
  geom_segment(aes(x = x, xend = xend, y = y, yend = y, colour = posteam), data = drives,
    linewidth = 1.6, arrow = arrow(length = unit(0.12, "cm"), type = "closed")
  ) +
  geom_text(aes(x = xend, y = y, label = result, hjust = if_else(winner, -0.1, 1.1)),
    data = drives, size = 2.2, colour = "white"
  ) +
  scale_colour_manual(values = drive_colors, guide = "none") +
  labs(
    title = sprintf("Super Bowl LX: %s %d, %s %d",
      super_bowl$away_team, super_bowl$away_score, super_bowl$home_team, super_bowl$home_score
    ),
    subtitle = sprintf("Every drive, in order from the middle out; %s's run left to right in the top half", champion),
    caption = "Data: nflreadr::load_pbp()"
  ) +
  titled
```

`sdv_surface("nfl", team)` paints both end zones in one team's color, so a
two-team chart reads better on the plain field.
