---
title: "Interactive charts"
description: >
  Ten recipes for team logos on web charts: ggiraph tooltips, hover
  highlights and logo axes, plotly logos as layout images, a reactable table,
  self-contained HTML and static PNG exports for social posts.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

Ten recipes for team logos and headshots on charts you can hover, zoom and
share: [ggiraph](https://davidgohel.github.io/ggiraph/) makes a ggplot
interactive and keeps every sdvplotR layer, [plotly](https://plotly.com/r/)
draws logos as layout images, and [reactable](https://glin.github.io/reactable/)
makes a sortable table. The last two recipes save a chart as one HTML file and
as a PNG for a social post. The data is one season each from MLB (baseballr,
from the MLB Stats API), the NBA, the WNBA and the NHL (hoopR, wehoop and
fastRhockey), and the NFL and college football (nflreadr and cfbfastR), all
but MLB read from SportsDataverse release files.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
library(ggiraph)
library(plotly, warn.conflicts = FALSE)

season <- 2026 # MLB and WNBA: 2026; NBA and NHL: the 2025-26 season
```

## 1. ggiraph: logos with hover details

ggiraph's `girafe()` turns a ggplot into an interactive SVG, and every
sdvplotR layer comes along: logos are embedded in the SVG. Logos themselves
don't respond to the pointer, so put the tooltip on a nearly transparent
`geom_point_interactive()` at the same spot. MLB's 2026 season, runs scored
and allowed per game:

```{r mlb-data}
mlb_ids <- baseballr::mlb_teams(season = season, sport_ids = 1) |>
  as.data.frame() |>
  select(team_records_team_id = team_id, team_abbreviation, team_full_name)
mlb <- baseballr::mlb_standings(season = season, league_id = "103,104") |>
  as.data.frame() |>
  inner_join(mlb_ids, by = "team_records_team_id") |>
  transmute(
    team = clean_team_abbrs(team_abbreviation, sport = "mlb"),
    name = team_full_name,
    wins = team_records_wins,
    losses = team_records_losses,
    rs = team_records_runs_scored / team_records_games_played,
    ra = team_records_runs_allowed / team_records_games_played,
    run_diff = team_records_runs_scored - team_records_runs_allowed
  ) |>
  arrange(team)
nrow(mlb)
```

```{r ggiraph-logos}
p_mlb <- ggplot(mlb, aes(x = rs, y = ra)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = "dashed") +
  geom_sdv_logos(aes(team = team), sport = "mlb", width = 0.06) +
  geom_point_interactive(
    aes(tooltip = sprintf("%s<br>%d-%d<br>%.2f scored, %.2f allowed per game", name, wins, losses, rs, ra),
        data_id = team),
    size = 9, alpha = 0.01
  ) +
  scale_x_continuous(expand = expansion(mult = 0.08)) +
  scale_y_continuous(expand = expansion(mult = 0.08)) +
  labs(
    title = "MLB runs scored and allowed, 2026",
    subtitle = "Per game; teams below the dashed line outscored their opponents",
    x = "Runs scored per game", y = "Runs allowed per game",
    caption = "Data: baseballr::mlb_standings() (MLB Stats API)"
  ) +
  theme_minimal()

girafe(ggobj = p_mlb, width_svg = 7, height_svg = 5)
```

`alpha = 0.01`, not 0: a fully transparent point is dropped from the SVG and
catches nothing.

## 2. ggiraph: a tooltip with a picture

A tooltip is HTML, so it can carry an image. hoopR's player box score has
each player's ESPN headshot URL (`athlete_headshot_href`); the browser loads
it on hover, so the page stays light. The 2025-26 season's 25 top scorers
(50+ games), each point in his team's color:

```{r ggiraph-headshots}
# the regular season without the All-Star games and the NBA Cup final
nba_regular <- hoopR::load_nba_schedule(season) |>
  as.data.frame() |>
  filter(season_type == 2, type_abbreviation == "STD") |>
  pull(game_id)
players <- hoopR::load_nba_player_box(season) |>
  as.data.frame() |>
  filter(game_id %in% nba_regular, !did_not_play)

scorers <- players |>
  summarise(
    games = n(),
    ppg = mean(points),
    ts = sum(points) / (2 * (sum(field_goals_attempted) + 0.44 * sum(free_throws_attempted))),
    team = last(team_abbreviation, order_by = game_date),
    headshot = last(athlete_headshot_href, order_by = game_date),
    .by = c(athlete_id, athlete_display_name)
  ) |>
  filter(games >= 50) |>
  arrange(desc(ppg), athlete_id) |>
  slice_head(n = 25) |>
  mutate(tip = sprintf(
    "<img src='%s' height='80'><br><b>%s</b> (%s)<br>%.1f points, %.1f%% true shooting",
    headshot, athlete_display_name, team, ppg, 100 * ts
  ))

p_scorers <- ggplot(scorers, aes(x = ppg, y = ts)) +
  geom_point_interactive(aes(fill = team, tooltip = tip, data_id = athlete_id),
    shape = 21, size = 4, colour = "white", stroke = 0.6
  ) +
  scale_fill_sdv(sport = "nba") +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  labs(
    title = "The NBA's 25 top scorers, 2025-26",
    subtitle = "Hover a point for the player",
    x = "Points per game", y = "True shooting %",
    caption = "Data: hoopR::load_nba_player_box(); headshots: ESPN"
  ) +
  theme_minimal()

girafe(
  ggobj = p_scorers, width_svg = 7, height_svg = 4.5,
  options = list(opts_tooltip(css = "background:white;color:black;padding:6px;border-radius:6px;"))
)
```

A player traded mid-season shows his latest team here, the one in his last
game.

## 3. ggiraph: hover one line to pick it out

With fifteen teams, a line chart is a tangle. Give each line a `data_id` and
ggiraph highlights the one under the pointer; `opts_hover_inv()` fades the
rest. The WNBA's 2026 regular season as cumulative wins:

```{r ggiraph-lines}
# standard games only: no All-Star Game or Commissioner's Cup final
wnba_regular <- wehoop::load_wnba_schedule(season) |>
  as.data.frame() |>
  filter(season_type == 2, type_abbreviation == "STD") |>
  pull(game_id)
wnba_runs <- wehoop::load_wnba_team_box(season) |>
  as.data.frame() |>
  filter(game_id %in% wnba_regular) |>
  arrange(team_abbreviation, game_date, game_id) |>
  mutate(game_no = row_number(), wins = cumsum(team_winner), .by = team_abbreviation) |>
  rename(team = team_abbreviation)

p_wnba <- ggplot(wnba_runs, aes(x = game_no, y = wins, colour = team)) +
  geom_line_interactive(aes(data_id = team, tooltip = team_display_name), linewidth = 1) +
  scale_colour_sdv(sport = "wnba") +
  labs(
    title = "The WNBA's 2026 season, win by win",
    subtitle = "Hover a line to pick out a team",
    x = "Game", y = "Wins",
    caption = "Data: wehoop::load_wnba_team_box()"
  ) +
  theme_minimal()

girafe(
  ggobj = p_wnba, width_svg = 7, height_svg = 4.5,
  options = list(
    opts_hover(css = "stroke-width:3;"),
    opts_hover_inv(css = "opacity:0.15;")
  )
)
```

`opts_selection()` keeps a line highlighted on click, for comparing two teams.

## 4. ggiraph: interactive bars with logos on the axis

Axis logos from `scale_x_sdv()` and `theme_x_sdv()` work in ggiraph too, since
the SVG embeds the images. The bars carry the tooltip. NBA three-point volume,
2025-26:

```{r ggiraph-axis}
threes <- hoopR::load_nba_team_box(season) |>
  as.data.frame() |>
  filter(game_id %in% nba_regular) |>
  summarise(
    games = n(),
    tpa = mean(three_point_field_goals_attempted),
    tp_pct = sum(three_point_field_goals_made) / sum(three_point_field_goals_attempted),
    .by = team_abbreviation
  ) |>
  arrange(desc(tpa), team_abbreviation) |>
  mutate(team = factor(team_abbreviation, levels = team_abbreviation))

p_threes <- ggplot(threes, aes(x = team, y = tpa)) +
  geom_col_interactive(
    aes(fill = team_abbreviation, data_id = team_abbreviation,
        tooltip = sprintf("%s: %.1f attempts, %.1f%% made", team_abbreviation, tpa, 100 * tp_pct)),
    width = 0.75
  ) +
  scale_x_sdv(sport = "nba", size = 16) +
  scale_fill_sdv(sport = "nba") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "NBA three-point attempts per game, 2025-26",
    x = NULL, y = "3PA per game",
    caption = "Data: hoopR::load_nba_team_box()"
  ) +
  theme_minimal() +
  theme_x_sdv() +
  theme(panel.grid.major.x = element_blank())

girafe(ggobj = p_threes, width_svg = 8, height_svg = 4.5)
```

Every logo is embedded in the page at full size, so a chart with dozens of
them weighs a few hundred kilobytes; recipe 2's URL-in-tooltip approach is the
light one.

## 5. plotly: `ggplotly()` drops the logos, so add them back

`ggplotly()` converts most layers, but not sdvplotR's image geoms: it warns
and leaves them out. Draw the logos as plotly layout images instead, one per
team, positioned in data units. A small helper builds them from
`team_reference()`'s logo URLs. The NFL's 2025 offenses and defenses:

```{r plotly-helper}
logo_images <- function(teams, x, y, sport, sizex, sizey = sizex, xref = "x", yref = "y") {
  ref <- team_reference(sport)
  urls <- ref$logo_url[match(clean_team_abbrs(teams, sport = sport), ref$team_abbr)]
  lapply(seq_along(urls), function(i) {
    list(
      source = urls[i], xref = xref, yref = yref, x = x[i], y = y[i],
      sizex = sizex, sizey = sizey, xanchor = "center", yanchor = "middle", layer = "above"
    )
  })
}
```

```{r plotly-ggplotly, warning = FALSE, out.width = "100%"}
nfl_weeks <- nflreadr::load_team_stats(2025, summary_level = "week") |>
  as.data.frame() |>
  filter(season_type == "REG") |>
  mutate(plays = attempts + sacks_suffered + carries, epa = passing_epa + rushing_epa)
nfl <- inner_join(
  summarise(nfl_weeks, off_epa = sum(epa) / sum(plays), .by = team),
  nfl_weeks |>
    mutate(team = opponent_team) |>
    summarise(def_epa = sum(epa) / sum(plays), .by = team),
  by = "team"
) |>
  arrange(team)

p_nfl <- ggplot(nfl, aes(x = off_epa, y = def_epa)) +
  geom_point(aes(text = sprintf("%s<br>offense %+.3f<br>defense %+.3f", team, off_epa, def_epa)),
    alpha = 0
  ) +
  labs(
    title = "NFL offense vs defense, 2025",
    x = "Offense: EPA per play", y = "Defense: EPA per play allowed"
  ) +
  theme_minimal()

ggplotly(p_nfl, tooltip = "text") |>
  layout(
    images = logo_images(nfl$team, nfl$off_epa, nfl$def_epa, "nfl",
      sizex = 0.025, sizey = 0.025
    ),
    yaxis = list(autorange = "reversed") # good defense at the top
  )
```

`text` is an aesthetic only ggplotly reads, so ggplot2 warns that it is
unknown (the chunk hides that warning). `sizex` and `sizey` are in axis units,
and the image keeps its shape inside that box; here both axes run in EPA, so
one size fits. Reverse an axis in
`layout()` rather than with `scale_y_reverse()`, which ggplotly turns into
negative values.

## 6. plotly: logos as markers, built natively

Without ggplot2, `plot_ly()` draws a transparent marker trace that carries the
hover text, and the logos go on as layout images. Set the axis ranges yourself
and give the logos a margin, since plotly sizes its autorange from the markers
only. NHL goals for and against per game, 2025-26:

```{r plotly-native, out.width = "100%"}
nhl <- fastRhockey::load_nhl_team_box(season) |>
  as.data.frame() |>
  filter(substr(game_id, 5, 6) == "02") |>
  mutate(team = clean_team_abbrs(team_abbrev, sport = "nhl")) |>
  summarise(games = n(), gf = mean(goals), ga = mean(goals_against), .by = team) |>
  arrange(team)

pad <- 0.15
plot_ly(
  nhl, x = ~gf, y = ~ga, type = "scatter", mode = "markers",
  marker = list(size = 28, opacity = 0),
  text = ~sprintf("%s: %.2f for, %.2f against", team, gf, ga), hoverinfo = "text"
) |>
  layout(
    title = list(text = "NHL goals per game, 2025-26 regular season", x = 0.02),
    xaxis = list(title = "Goals for per game", range = range(nhl$gf) + c(-pad, pad)),
    yaxis = list(title = "Goals against per game", range = rev(range(nhl$ga) + c(-pad, pad))),
    images = logo_images(nhl$team, nhl$gf, nhl$ga, "nhl", sizex = 0.12, sizey = 0.12)
  )
```

A `range` given high-to-low reverses the axis, here so good defense is at the
top.

## 7. plotly: logos on a category axis

Put each logo under its bar by giving the image `yref = "paper"` (0 is the
bottom of the plot area) and `yanchor = "top"`, then hide the tick labels and
grow the bottom margin to make room. On a category axis, bars sit at
positions 0, 1, 2, and so on. MLB's 2026 run differential:

```{r plotly-axis, out.width = "100%"}
mlb_diff <- mlb |>
  arrange(desc(run_diff), team) |>
  mutate(fill = unname(sdv_team_colors("mlb", team)))

plot_ly(
  mlb_diff, x = ~factor(team, levels = team), y = ~run_diff, type = "bar",
  marker = list(color = ~fill),
  text = ~sprintf("%s: %+d runs", name, run_diff), hoverinfo = "text", textposition = "none"
) |>
  layout(
    title = list(text = "MLB run differential, 2026 regular season", x = 0.02),
    xaxis = list(title = "", showticklabels = FALSE),
    yaxis = list(title = "Run differential"),
    images = logo_images(mlb_diff$team, seq_len(nrow(mlb_diff)) - 1, rep(0, nrow(mlb_diff)), "mlb",
      sizex = 0.9, sizey = 0.08, yref = "paper"
    ) |> lapply(modifyList, list(yanchor = "top")),
    margin = list(b = 50, t = 50)
  )
```

`sizey` is now a share of the plot's height (paper units), and `sizex` a share
of one category's slot.

## 8. reactable: a sortable, searchable table

For a table rather than a chart, reactable sorts, searches and pages in the
browser, and sdvplotR supplies the cell renderers: `reactable_sdv_logos()` for
a logo column and `reactable_sdv_team_color_bar()` for a bar in each row's
team color. The SEC's 2025 football season from ESPN's schedule file:

```{r reactable}
sec <- team_reference("cfb") |>
  filter(conference == "SEC") |>
  select(team = team_abbr, school = team_location)

games <- cfbfastR::load_espn_cfb_schedules(2025) |> as.data.frame()
sec_table <- bind_rows(
  transmute(games, team = home_abbreviation, pf = home_score, pa = away_score),
  transmute(games, team = away_abbreviation, pf = away_score, pa = home_score)
) |>
  inner_join(sec, by = "team") |>
  summarise(
    W = sum(pf > pa), L = sum(pf < pa),
    PF = round(mean(pf), 1), PA = round(mean(pa), 1), Margin = round(mean(pf - pa), 1),
    .by = c(team, school)
  ) |>
  arrange(desc(Margin), team)

reactable::reactable(
  sec_table,
  searchable = TRUE, defaultPageSize = 8, highlight = TRUE, compact = TRUE,
  columns = list(
    team = reactable::colDef(name = "", cell = reactable_sdv_logos("cfb", height = 26), html = TRUE, width = 50),
    school = reactable::colDef(name = "School", minWidth = 120),
    Margin = reactable::colDef(
      style = reactable_sdv_team_color_bg(sec_table, "team", "cfb", alpha = 0.25)
    )
  )
)
```

The [reactable integration](reactable-integration.html) article covers the
other renderers: headshots, logo column headers and team-colored rows.

## 9. Share a chart as one HTML file

`htmlwidgets::saveWidget()` writes any of these widgets to a file.
`selfcontained = TRUE` inlines the JavaScript, so the file opens anywhere
with no folder beside it (it needs pandoc, which ships with RStudio).
ggiraph's logos travel inside the SVG; plotly's layout images and the reactable
logos stay links, so those need the network when the file is opened.

```{r save-html}
out <- tempfile("widgets-")
dir.create(out)
htmlwidgets::saveWidget(girafe(ggobj = p_mlb), file.path(out, "mlb.html"), selfcontained = TRUE)
htmlwidgets::saveWidget(girafe(ggobj = p_scorers), file.path(out, "scorers.html"), selfcontained = TRUE)
data.frame(
  file = c("mlb.html (30 logos inside)", "scorers.html (headshots by link)"),
  kb = round(file.size(file.path(out, c("mlb.html", "scorers.html"))) / 1024)
)
```

The gap between the two files is mostly the 30 embedded logos; the bulk of each is
ggiraph's own JavaScript and fonts, inlined. With `selfcontained = FALSE` the
libraries go in a folder beside the HTML instead, which a web server can share
across pages.

## 10. Export a static PNG for social

A ggiraph chart is a ggplot underneath, so `ggsave()` makes the static version
with no browser involved: 8 x 4.5 in at 150 dpi is a 1200 x 675 px card. The
interactive layers draw as their plain counterparts.

```{r static-png, fig.width = 8, fig.height = 4.5, dpi = 150, fig.alt = "Static version of the MLB runs scored and allowed chart for 2026: each team's logo placed by runs scored and allowed per game, with a dashed line where they are equal."}
png_file <- file.path(out, "mlb.png")
ggsave(png_file, p_mlb, width = 8, height = 4.5, dpi = 150, bg = "white")
magick::image_info(magick::image_read(png_file))[c("width", "height")]
p_mlb
```

plotly has no `ggsave()`; render its HTML in a headless browser with webshot2.
That needs Chrome or Chromium on the machine, and a browser started from a
scheduled build can stall, so this chunk is shown but not run when the site
builds:

```{r plotly-png, eval = FALSE}
p_plotly <- plot_ly(
  nhl, x = ~gf, y = ~ga, type = "scatter", mode = "markers", marker = list(opacity = 0)
) |>
  layout(
    title = "NHL goals per game, 2025-26",
    xaxis = list(range = range(nhl$gf) + c(-0.15, 0.15)),
    yaxis = list(range = rev(range(nhl$ga) + c(-0.15, 0.15))),
    images = logo_images(nhl$team, nhl$gf, nhl$ga, "nhl", sizex = 0.12, sizey = 0.12)
  ) |>
  config(staticPlot = TRUE)
htmlwidgets::saveWidget(p_plotly, file.path(out, "nhl.html"), selfcontained = TRUE)
webshot2::webshot(file.path(out, "nhl.html"), file.path(out, "nhl.png"),
  vwidth = 1200, vheight = 675, delay = 2
)
magick::image_info(magick::image_read(file.path(out, "nhl.png")))[c("width", "height")]
```

`delay` gives the logos time to load before the screenshot; raise it on a slow
connection.
