---
title: "Percentile Bars and Cut Lines"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

A table of numbers tells you the values. It does not tell you which ones stand
out, where the line falls between one group and the next, or how far a row has
moved. Three functions handle those jobs, and this article builds one table with
each:

1. **Percentile bars**: the 2025 NFL season's busiest quarterbacks, each one's
   EPA per dropback and completion percentage over expected drawn against every
   qualified passer with `gt_percentile_bar()`.
2. **Cut lines**: the NBA's 2025-26 Western Conference standings, with
   `gt_cutline()` marking where the playoff seeds end and where the play-in ends.
3. **Rank changes**: the final 2025 college football AP Top 25 with each team's
   move since the preseason poll, computed and colored by `gt_delta()`.

The quarterbacks come from nflverse's player stats via
[`nflreadr`](https://nflreadr.nflverse.com), the standings from ESPN via
[`hoopR`](https://hoopR.sportsdataverse.org) and the polls from
CollegeFootballData.com via [`cfbfastR`](https://cfbfastR.sportsdataverse.org)
(which needs a free API key). The three functions come from Andrew Weatherman's
[gtUtils](https://github.com/andreweatherman/gtUtils), and the first two
sections follow his delay-table example.

```{r setup}
library(sdvplotR)
library(gt)
library(gtExtras)
library(dplyr)
```

## 1. Percentile bars

We take every quarterback with at least 250 dropbacks (pass attempts plus sacks)
in the 2025 regular season and compute two rates: EPA per dropback and
completion percentage over expected (CPOE). Each gets a percentile among all the
qualified passers, so the fifteen busiest can be read against the whole group,
not just against each other.

```{r qb-data}
qbs <- nflreadr::load_player_stats(2025, summary_level = "reg") |>
  filter(position == "QB") |>
  mutate(dropbacks = attempts + sacks_suffered, epa_db = passing_epa / dropbacks) |>
  filter(dropbacks >= 250) |>
  mutate(
    epa_ptile = cume_dist(epa_db),
    cpoe_ptile = cume_dist(passing_cpoe)
  ) |>
  arrange(desc(dropbacks), player_id) |>
  slice_head(n = 15) |>
  select(player_id, player_display_name, recent_team, dropbacks, epa_db, epa_ptile, passing_cpoe, cpoe_ptile)
```

`cume_dist()` gives each value's percentile, on a 0-1 scale, which matters in a
moment. Sorting on `player_id` after `dropbacks` breaks any tie the same way on
every rebuild.

`gt_percentile_bar()` turns each percentile column into a filled track with a
numbered marker at the tip. `gt_sdv_headshots()` turns the GSIS ids into NFL.com
headshots and `gt_sdv_logos()` the team abbreviations into logos.

```{r qb-table}
qbs |>
  gt(id = "qb-bars") |>
  gt_theme_almanac() |>
  gt_sdv_headshots(player_id, sport = "nfl", height = 34) |>
  gt_sdv_logos(recent_team, sport = "nfl", height = 24) |>
  gt_percentile_bar(c(epa_ptile, cpoe_ptile), full_track = FALSE, width = 150) |>
  fmt_number(epa_db, decimals = 2, force_sign = TRUE) |>
  fmt_number(passing_cpoe, decimals = 1, force_sign = TRUE) |>
  tab_spanner(columns = c(epa_db, epa_ptile), "EPA per dropback") |>
  tab_spanner(columns = c(passing_cpoe, cpoe_ptile), "CPOE") |>
  cols_align(columns = -player_display_name, "center") |>
  gt_add_divider(dropbacks, color = "black", weight = px(1.5), include_labels = FALSE) |>
  cols_label(
    player_id = "", player_display_name = "Quarterback", recent_team = "",
    dropbacks = "Dropbacks", epa_db = "Value", epa_ptile = "Percentile",
    passing_cpoe = "Value", cpoe_ptile = "Percentile"
  ) |>
  tab_footnote(
    locations = cells_column_spanners(),
    footnote = "Percentile among every quarterback with at least 250 dropbacks in the 2025 regular season"
  ) |>
  tab_header(
    "The NFL's busiest passers, 2025",
    "The 15 quarterbacks with the most dropbacks, regular season"
  ) |>
  tab_source_note("Data: nflverse player stats via nflreadr")
```

The percentile columns are stored on a 0-1 scale, but the markers read as whole
numbers out of 100. That is `gt_percentile_bar()`'s `scale` argument, which defaults to
`"auto"`: a column whose values all fall in `[0, 1]` is treated as a proportion
and mapped onto the 0-100 axis, so a raw `0.71` prints as `71`. A column already
on a 0-100 scale is left alone, so the same code works either way. Set
`scale = "none"` to turn that off.

The other arguments cover the look:

- `domain`: the value range the bar spans, `c(0, 100)` by default.
- `palette`: colors mapped across the domain, used for both the fill and the
  marker. The default runs blue at the low end through grey to red at the high
  end.
- `full_track`: whether the unfilled track runs the full width or stops at the
  marker. We set it to `FALSE` so each bar ends at its own value.
- `width`, `marker_size`, `track_height`, `text_color`, `decimals`: the column
  width, the marker, the thickness of the track, the color of the number inside
  the marker, and its decimal places.
- `na_label`: what to show for a missing percentile. It defaults to an em dash
  and draws a broken track, so an empty row cannot be misread as a percentile of
  zero.

The whole bar is CSS, not an image, so it stays sharp at any export size.

## 2. Cut lines

The second table asks where the lines fall in a standings table. In the NBA,
seeds 1-6 in each conference go straight to the playoffs, seeds 7-10 play in
for the last two places, and the rest go to the lottery. ESPN's standings carry
each team's final seed, tiebreakers and the play-in applied, so the 7 and 8
seeds are the two play-in winners and the 8 seed can hold the better record.

```{r west-data}
nba_teams <- team_reference("nba")

west <- hoopR::espn_nba_standings(year = 2026) |>
  mutate(
    team_id = as.character(team_id),
    conference = nba_teams$conference[match(team_id, nba_teams$espn_team_id)],
    abbr = nba_teams$team_abbr[match(team_id, nba_teams$espn_team_id)]
  ) |>
  filter(conference == "Western") |>
  arrange(playoffseed) |>
  transmute(
    seed = playoffseed, team = abbr, wins, losses,
    pct = winpercent, diff = as.numeric(differential), last_ten = lasttengames
  )
```

`team_reference()` gives each ESPN team id its conference and abbreviation;
ESPN's standings leave the conference out.

```{r west-table}
west |>
  gt(id = "west") |>
  gt_theme_almanac(accent = "#1D428A") |>
  gt_sdv_logos(team, sport = "nba", height = 24) |>
  gt_cutline(after = 0, "Playoffs", color = "#1D428A") |>
  gt_cutline(after = 6, "Play-in", color = "#C8102E", gap = c(12, 0)) |>
  gt_cutline(after = 10, "Lottery", color = "#5A5A5A", gap = c(12, 0)) |>
  opt_row_striping(FALSE) |>
  fmt(pct, fns = function(x) sub("^0", "", sprintf("%.3f", x))) |>
  fmt_number(diff, decimals = 1, force_sign = TRUE) |>
  cols_label(
    seed = "Seed", team = "", wins = "W", losses = "L", pct = "Pct",
    diff = "Diff", last_ten = "Last 10"
  ) |>
  cols_align(columns = everything(), "center") |>
  tab_style(locations = cells_body(), style = cell_text(size = px(14))) |>
  tab_header(
    "Western Conference standings, 2025-26",
    "Regular-season records; final seeds, after the play-in. Diff is the average scoring margin."
  ) |>
  gt_538_caption(
    "Seeds 1-6 reach the playoffs; seeds 7-10 meet in the play-in for the last two places.",
    "Data: ESPN standings via hoopR"
  )
```

There are three cut lines, one per group. `gt_cutline(after = 6, ...)` draws the
rule between rows 6 and 7, where the playoff seeds give way to the play-in, and
`after = 10` does the same where the play-in ends. Each label sits on its line,
rendered in uppercase.

The first call, `gt_cutline(after = 0, ...)`, is the useful trick: `after = 0`
draws a line above the very first row, which labels the top section the same way
the others are labelled.

A few arguments shape the lines:

- `color`, `weight`, `style`: the color, thickness, and line style (`"dashed"`,
  `"solid"`, or `"dotted"`).
- `label`, `label_position`, `label_size`, `label_color`: the text on the line,
  whether it sits above or below, and how it looks.
- `gap`: extra space around the line, in pixels. A single number pads both sides;
  a length-2 vector `c(above, below)` sets them separately. Here `gap = c(12, 0)`
  opens room above each lower line without pushing the row below it down.

Note `cell_text(size = px(14))` on the body: the `gt_theme_*` functions set their
font sizes per cell, so `tab_options(table.font.size = ...)` will not move them.
Restyle the body cells directly instead.

Cut lines are positional, not data-driven. `gt::tab_row_group()` splits a table
on a value in the data and adds heading rows; a cut line leaves the table flowing
and marks a fixed row number, which is what a standings table wants.

## 3. Rank changes

The last table puts a final ranking next to where each team started. The
preseason AP poll is week 1 of the regular season in CollegeFootballData.com's
rankings, and the final poll is the only poll of the postseason.

```{r ap-data}
ap_poll <- function(...) {
  cfbfastR::cfbd_rankings(year = 2025, ...) |>
    filter(poll == "AP Top 25") |>
    select(school, conference, rank)
}

preseason <- ap_poll(week = 1, season_type = "regular")
final <- ap_poll(season_type = "postseason")

moves <- final |>
  left_join(select(preseason, school, preseason = rank), by = "school") |>
  arrange(rank) |>
  select(rank, school, conference, preseason)
```

`gt_delta(from, to)` adds a column holding `to - from`, signed and colored by
direction. For a ranking, a lower number is better, so passing the final rank as
`from` and the preseason rank as `to` gives the places a team climbed: positive
when it rose. A team that was unranked in the preseason has no starting rank, so
its change is left blank and its preseason cell reads `NR`.

```{r ap-table}
moves |>
  gt(id = "ap-moves") |>
  gt_theme_almanac(accent = "#9E1B32") |>
  gt_sdv_logos(school, sport = "cfb", height = 22, include_name = TRUE) |>
  gt_delta(rank, preseason, column_label = "Places climbed", decimals = 0, arrows = TRUE) |>
  sub_missing(preseason, missing_text = "NR") |>
  cols_label(rank = "Final", school = "Team", conference = "Conference", preseason = "Preseason") |>
  cols_align(columns = -c(school, conference), "center") |>
  tab_style(locations = cells_body(), style = cell_text(size = px(14))) |>
  tab_header(
    "The final 2025 AP Top 25, and where each team started",
    "Final poll after the national championship, against the preseason poll"
  ) |>
  tab_source_note("Data: CollegeFootballData.com via cfbfastR")
```

`arrows = TRUE` leads each value with a triangle in place of a sign, so a fall
reads as a down triangle in front of a positive number. With `percent = TRUE`
the change is shown as a percent of `from` instead, which suits values such as
points or yards better than ranks.
