---
title: "Creating Faceted Tables"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

Nobody likes a long table. Put 25 rows next to a handful of columns and you get
an image twice as tall as it is wide, mostly empty, with half the values
scrolled off the bottom of a feed. The fix is to fold that one tall table into
two shorter ones and set them side by side.

`gt_grid()` does that the direct way: build each block as a normal `gt` table,
put the tables in a list, and `gt_grid()` arranges the list under one shared
header and footer. Nothing gets reshaped, so every block is a real table you can
style with the rest of the package. This article builds a top 25 of men's college
basketball for 2025-26 that way, and uses two more functions along the route:
`gt_color_ranks()`, which colors a column by its values, and `gt_spotlight()`,
which lights a set of rows and dims the rest.

The games are [`hoopR`](https://hoopR.sportsdataverse.org)'s ESPN box scores.
`gt_grid()`, `gt_color_ranks()` and `gt_spotlight()` come from Andrew
Weatherman's [gtUtils](https://github.com/andreweatherman/gtUtils), and this
article follows his faceted-tables example.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr)
```

## 1. The data: a top 25 by efficiency margin

We rank every Division I team by efficiency margin, points scored minus points
allowed per 100 possessions, over its games against other Division I teams,
tournaments included. The margin is raw, not adjusted for opponents. Each team
also gets its national rank on offense and on defense.

```{r ratings}
d1 <- team_reference("mbb")

box <- hoopR::load_mbb_team_box(seasons = 2026) |>
  mutate(poss = field_goals_attempted - offensive_rebounds + total_turnovers + 0.475 * free_throws_attempted)

ratings <- box |>
  # Division I against Division I only
  filter(team_id %in% d1$espn_team_id, opponent_team_id %in% d1$espn_team_id) |>
  inner_join(
    select(box, game_id, opponent_team_id = team_id, opp_poss = poss),
    by = c("game_id", "opponent_team_id")
  ) |>
  summarise(
    off = 100 * sum(team_score) / sum(poss),
    def = 100 * sum(opponent_team_score) / sum(opp_poss),
    .by = team_id
  ) |>
  mutate(
    net = off - def,
    off_rank = min_rank(desc(off)),
    def_rank = min_rank(def),
    team = d1$team_location[match(team_id, d1$espn_team_id)],
    conf = d1$conference[match(team_id, d1$espn_team_id)]
  )

data <- ratings |>
  arrange(desc(net), team) |>
  mutate(rk = row_number()) |>
  filter(rk <= 25) |>
  select(rk, team, conf, off_rank, def_rank, net)
```

The team names are ESPN's school names (`"Michigan"`, `"Iowa State"`), and
`gt_sdv_logos()` resolves them as they are, so they go into the table untouched
and get their logos later. Sorting on `team` after `net` breaks any tie the same
way on every rebuild.

Before splitting anything, two values are computed once, off the full table.

```{r split}
rank_domain <- c(1, nrow(ratings))

sizes <- c(13, 12)
chunks <- split(data, rep(seq_along(sizes), sizes))
```

`rank_domain` is the full range of the national ranks, 1 to the number of
Division I teams, and it matters because of how coloring works. A color scale is
built from whatever data it sees. If each block builds its own scale, block one
maps its colors to the ranks it happens to hold and block two to its own, so the
same rank lands on a different shade depending on which block a team falls in.
Fixing the domain once and handing it to both blocks keeps one scale across the
whole table, and anchors it to the whole country rather than the top 25.

`sizes` sets the split: 13 rows in the first block and 12 in the second.
`rep(seq_along(sizes), sizes)` turns `c(13, 12)` into 13 `1`s followed by 12
`2`s, and `split()` cuts the data on that. For three blocks, use `c(9, 8, 8)`.

## 2. Coloring a column by rank

Inside each block we color the two national-rank columns with `gt_color_ranks()`.

```{r color-ranks, eval = FALSE}
gt_color_ranks(c(off_rank, def_rank), domain = rank_domain)
```

The arguments worth knowing:

- `palette`: either a vector of hex colors or a `paletteer` palette named as
  `"package::palette"`. The default is a 5-color green-to-red ramp, so a low
  value (a high rank) reads green.
- `domain`: the value range mapped onto the palette. Left `NULL`, it is taken
  from the column itself. We pass `rank_domain` so both blocks share a scale.
- `reverse`: flip the palette, for when high numbers should be green instead.
- `pal_type`: `"discrete"` or `"continuous"`, which only matters when you hand it
  a `paletteer` palette.
- `na_color`: fill for missing values, white by default.
- `autocolor_text`: sets the text color for contrast against the fill, so dark
  cells get light text automatically. On by default.

It colors values as they are and does not rank for you, so it works on any
numeric column. Reach for a `paletteer` palette when you want something other than
the default ramp:

```{r color-ranks-viridis, eval = FALSE}
gt_color_ranks(c(off_rank, def_rank),
  domain = rank_domain,
  palette = "viridis::mako", pal_type = "continuous", reverse = TRUE
)
```

## 3. One table, applied to each block

Every block is a `gt` table, so we write the recipe once and `lapply()` it over
the chunks. None of this is grid-specific; it is the same table you would build
for a single frame.

```{r blocks}
block <- function(x) {
  gt(x) |>
    gt_theme_swiss() |>
    cols_hide(conf) |>
    cols_align(columns = -team, align = "center") |>
    cols_align(columns = team, "left") |>
    cols_label(
      rk = "", team = "Team",
      off_rank = "Off.", def_rank = "Def.", net = "Margin"
    ) |>
    cols_width(team ~ px(140), c(off_rank, def_rank) ~ px(46), net ~ px(62)) |>
    tab_spanner(columns = c(off_rank, def_rank), label = "Nat'l rank") |>
    fmt_number(net, decimals = 1, force_sign = TRUE) |>
    gt_color_ranks(c(off_rank, def_rank), domain = rank_domain) |>
    gt_sdv_logos(team, sport = "mbb", height = 20, include_name = TRUE) |>
    gtExtras::gt_add_divider(team, color = "white", weight = px(10)) |>
    tab_options(
      data_row.padding = px(6),
      table_body.hlines.style = "solid",
      table_body.hlines.color = "black",
      table_body.hlines.width = px(1)
    )
}

tbls <- lapply(chunks, block)
```

`gt_color_ranks(..., domain = rank_domain)` is the shared scale from earlier.
`gt_sdv_logos(team, ..., include_name = TRUE)` puts each school's logo in front
of its name. `gt_add_divider()` (from `gtExtras`) adds a white gap after the team
column so the numbers do not crowd the logos. `cols_hide(conf)` keeps the
conference in the data but out of the table; the last section uses it.

## 4. Laying it out with gt_grid

Now the list goes to `gt_grid()`.

```{r grid}
gt_grid(tbls,
  ncol = 2, gap = 24,
  title = "The top 25 by efficiency margin",
  subtitle = "Men's college basketball, 2025-26, games between Division I teams",
  title_style = list(
    font = "Oswald", size = 34, transform = "uppercase",
    margin_bottom = -4
  ),
  subtitle_style = list(italic = TRUE, color = "#6A6A6A"),
  caption = "Margin is points scored minus points allowed per 100 possessions, not adjusted for opponents.",
  caption_style = list(align = "left"),
  source_note = "Data: ESPN box scores via hoopR",
  caption_rule = TRUE
)
```

`ncol` sets how many tables go across, and the number of rows follows from the
list length. `gap` is the space between them in pixels. There is also an `align`
argument (`"top"`, `"center"`, `"bottom"`) for when the blocks are different
heights, which is common with an uneven split like ours.

The header and footer are set on the grid, not on the individual tables, so one
title covers both blocks instead of a title stranded on each. `title`,
`subtitle`, `caption`, and `source_note` are the four text slots. Setting
`caption`, `source_note`, and `caption_rule = TRUE` together gives the
538-style split caption, and `caption` accepts `md()`, so links and bold text
work.

Each text slot has a matching `*_style` list, and the keys are the same across
all of them (see [Styling Headers, Legends and Captions](styling.html)): `font`,
`size`, `color`, `weight`, `italic`, `spacing`, `transform`, `align`,
`line_height`, and the `margin_*` and `padding_*` keys. Anything you leave out
keeps its default, so the `title_style` above changes only the font, size,
transform and margin. `margin_bottom = -4` pulls the subtitle up closer to the
title.

Two more arguments are worth knowing. `labels` puts a short caption above each
individual table (styled by `label_style`), which is handy for small multiples
like one table per conference or per season. `file` writes the grid straight to a
PNG, with `bg`, `whitespace` and `zoom` controlling that export.

## 5. Spotlighting rows across blocks

To make a point about one conference, we light its teams with `gt_spotlight()`
and dim the rest. The conference is whichever has the most teams in the top 25,
worked out from the data, with alphabetical order breaking a tie.

```{r top-conf}
top_conf <- data |>
  count(conf) |>
  arrange(desc(n), conf) |>
  slice(1)
top_conf
```

`gt_spotlight()` reads a filter expression against each block's own data, so it
drops right into the block function. The wrinkle is that the conference's teams
fall across both blocks, and a block with no match at all would be left fully
lit, so the highlight would look applied to one block and not the other. That is
what `if_none = "dim"` handles: a block that matches nobody dims completely.

```{r spotlight}
spot <- lapply(chunks, function(x) {
  block(x) |>
    gt_spotlight(
      rows = conf == top_conf$conf, if_none = "dim",
      accent_color = "darkred", dim_color = "lightgrey"
    )
})

gt_grid(spot,
  ncol = 2, gap = 24,
  title = paste(top_conf$conf, "has", top_conf$n, "of the top 25"),
  subtitle = "Men's college basketball, 2025-26, by efficiency margin",
  title_style = list(
    font = "Oswald", size = 34, transform = "uppercase",
    margin_bottom = -4
  ),
  subtitle_style = list(italic = TRUE, color = "#6A6A6A"),
  caption = "Margin is points scored minus points allowed per 100 possessions, not adjusted for opponents.",
  caption_style = list(align = "left"),
  source_note = "Data: ESPN box scores via hoopR",
  caption_rule = TRUE
)
```

This is why `conf` stayed hidden rather than dropped: `gt_spotlight()` filters on
it even though the column is never drawn. `rows` also accepts plain row numbers,
like `rows = 1:5`, when you want to highlight by position.

The rest of its arguments control the look:

- `accent_color`: draws a bar on the left edge of each lit row. Supplying a color
  is what turns the bar on. `accent_width` sets its thickness, and
  `accent_column` moves it to a different column.
- `dim_color`: the text color for the muted rows. Setting it to `NULL` emphasizes
  the chosen rows without dimming the others.
- `fill`, `text_color`, and `bold`: styling for the lit rows themselves.
- `columns`: narrows the spotlight to some columns, dimming the rest of the row
  too, so only those cells stay at full strength.

Outside of grids, `gt_spotlight()` is useful on a single table any time you want
one row or a small group highlighted.

## Wrapping up

The pattern is the same both times: build a normal table, `lapply()` it over your
chunks, and pass the list to `gt_grid()`. The only two faceting-specific habits
are fixing your color domain once so the blocks agree, and using
`if_none = "dim"` when a highlight lands in some blocks and not others. To keep a
long table as one table instead, `gt_snake()` folds it into blocks; the
[rolling-window article](window_wins.html) uses it.
