---
title: "College Football Weekly Leaderboard"
description: >
  The college football season to date, rebuilt every week: a top 25 by
  opponent-adjusted EPA per play, every FBS team by conference, and the top 25's
  offense against its defense.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 96
)
```

```{r season, include = FALSE}
library(sdvplotR)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(gt)

today <- Sys.Date()
# Week 0 is in late August, so before then the calendar points at last season
current <- as.integer(format(today, "%Y")) - (format(today, "%m") < "08")

# A season's play-by-play, or NULL when it is not published: load_cfb_pbp() errors
# for a season past cfbfastR's most recent one, and warns and returns no rows for
# one whose file is not up yet
play_by_play <- function(season) {
  pbp <- tryCatch(suppressWarnings(cfbfastR::load_cfb_pbp(season)), error = function(e) NULL)
  if (NROW(pbp) == 0) NULL else pbp
}

season <- current
pbp <- play_by_play(season)
if (is.null(pbp)) {
  season <- current - 1
  pbp <- play_by_play(season)
}
stopifnot(!is.null(pbp))

# A live API call, or NULL when the API does not answer: one bad week then shows
# a note instead of failing the whole site build
live <- function(expr) tryCatch(expr, error = function(e) NULL)

# CollegeFootballData.com's schedule (a live API: set CFBD_API_KEY, a free key)
schedule <- live(cfbfastR::load_cfb_schedules(season))
schedule_ok <- NROW(schedule) > 0
if (schedule_ok) {
  played <- schedule |>
    filter(completed) |>
    arrange(start_date, game_id)
  week <- max(played$week[played$season_type == "regular"])
  left <- filter(schedule, !completed)
} else {
  week <- max(pbp$week)
}
if (!schedule_ok) {
  status <- sprintf(
    "the %d season through week %d. CollegeFootballData.com's schedule was unavailable when this page was built, so records are missing this week; they return next week.",
    season, week
  )
  through <- paste("through week", week)
} else if (any(left$season_type == "regular")) {
  status <- sprintf("the %d season through week %d.", season, week)
  through <- paste("through week", week)
} else if (nrow(left) > 0) {
  status <- sprintf("the %d regular season is final; bowls and the playoff are under way.", season)
  through <- "regular season and finished bowls"
} else {
  status <- sprintf("the offseason, so this is the final %d season, bowls and playoff included.", season)
  through <- "final"
}
```

**Updated `r format(today)`:** `r status`

This page is rebuilt every week with the site. It rates every FBS team by
opponent-adjusted EPA per play, a simple cousin of SP+ built from the
play-by-play, for the latest season with data: the season to date during the
fall, the final season in the offseason. Data:
[cfbfastR](https://cfbfastR.sportsdataverse.org) play-by-play and
CollegeFootballData.com schedules.

## Picking the season

Week 0 is in late August, so before then the calendar points at last season.
`cfbfastR::load_cfb_pbp()` errors for a season past its most recent one and
warns, returning no rows, for a season whose file is not up yet; either way the
page steps back one season. The schedule says whether regular-season games are
left, which is what the status line reports. It is a live API call, so it goes
through `live()`: if CollegeFootballData.com does not answer, the page still
rates every team from the play-by-play and leaves the records blank for that
week instead of failing the site build.

```{r season-code, ref.label = "season", eval = FALSE}
```

The play-by-play carries each game's teams with their division and conference,
which picks out the FBS teams. Records come from the schedule's played games,
against FBS and FCS opponents alike.

```{r records}
sides <- bind_rows(
  distinct(pbp, game_id, week, school = home, conference = home_team_conference, division = home_team_division),
  distinct(pbp, game_id, week, school = away, conference = away_team_conference, division = away_team_division)
)
fbs <- sides |>
  filter(division %in% "fbs") |>
  arrange(week, game_id) |>
  group_by(school) |>
  summarise(conference = last(conference), .groups = "drop")

if (schedule_ok) {
  records <- bind_rows(
    transmute(played, school = home_team, win = home_points > away_points),
    transmute(played, school = away_team, win = away_points > home_points)
  ) |>
    group_by(school) |>
    summarise(record = paste(sum(win), sum(!win), sep = "-"), .groups = "drop")
  fbs <- left_join(fbs, records, by = "school")
} else {
  fbs$record <- NA_character_ # shown as a dash
}
nrow(fbs)
```

## 1. Top 25 by adjusted EPA per play

The ratings use FBS-against-FBS pass and run plays, minus garbage time (a lead of
more than 43 points in the first quarter, 37 in the second, 27 in the third or
21 in the fourth). One pass of opponent adjustment credits each offensive play
for the defense it faced (that defense's EPA allowed per play against the FBS
average), and each defensive play for the offense it faced. Net is adjusted
offense minus adjusted defense.

```{r ratings}
plays <- pbp |>
  filter(
    pass == 1 | rush == 1, !is.na(EPA),
    pos_team %in% fbs$school, def_pos_team %in% fbs$school
  ) |>
  mutate(margin = c(43, 37, 27, 21)[period]) |> # NA in overtime, which is never garbage time
  filter(!coalesce(abs(pos_score_diff_start) > margin, FALSE)) |>
  arrange(game_id, id_play) # a fixed order, so every mean sums the same way each week

avg <- mean(plays$EPA)
defense_allows <- plays |>
  group_by(def_pos_team) |>
  summarise(allowed = mean(EPA), .groups = "drop")
offense_gains <- plays |>
  group_by(pos_team) |>
  summarise(gained = mean(EPA), .groups = "drop")
adjusted <- plays |>
  left_join(defense_allows, by = "def_pos_team") |>
  left_join(offense_gains, by = "pos_team") |>
  mutate(adj_off = EPA - (allowed - avg), adj_def = EPA - (gained - avg))

ratings <- inner_join(
  adjusted |> group_by(school = pos_team) |> summarise(off = mean(adj_off), .groups = "drop"),
  adjusted |> group_by(school = def_pos_team) |> summarise(dfn = mean(adj_def), .groups = "drop"),
  by = "school"
) |>
  mutate(net = off - dfn) |>
  inner_join(fbs, by = "school") |>
  arrange(desc(net), school) |> # the school breaks a tie, so an unchanged week renders the same
  mutate(rank = row_number())

top25 <- ratings |>
  slice_head(n = 25) |>
  mutate(logo = school, conf_logo = conference) |>
  select(rank, logo, school, conf_logo, conference, record, off, dfn, net)
head(top25)
```

sdvplotR has a row for every FBS conference too, so `gt_sdv_logos()` draws the
conference marks from the play-by-play's conference names, next to the school
logos.

```{r top25-table}
top25 |>
  gt(id = "cfb-top25") |>
  tab_header(
    title = paste("College football top 25,", season),
    subtitle = paste("Opponent-adjusted EPA per play,", through)
  ) |>
  fmt_number(c(off, dfn, net), decimals = 3, force_sign = TRUE) |>
  sub_missing(record, missing_text = "—") |>
  data_color(net, palette = c("#f7f7f7", "#2e8b57"), domain = c(0, max(top25$net))) |>
  tab_spanner("Adjusted EPA per play", c(off, dfn, net)) |>
  cols_label(
    rank = "", logo = "", school = "Team", conf_logo = "", conference = "Conference",
    record = "Record", off = "Offense", dfn = "Defense", net = "Net"
  ) |>
  tab_source_note(paste(
    "Data: cfbfastR play-by-play and CollegeFootballData.com schedules.",
    "FBS vs FBS pass and run plays, garbage time removed; defense is EPA allowed",
    "(lower is better). Viz: sdvplotR"
  )) |>
  gt_sdv_logos(columns = logo, sport = "cfb", height = 26) |>
  gt_sdv_logos(columns = conf_logo, sport = "cfb", height = 20) |>
  gt_theme_ncaa()
```

## 2. Every FBS team, by conference

One row per conference, ordered by its average rating, each team's logo at its
net rating (nudged up and down in turn so neighbors overlap less). The bar in
the conference's color spans it from its lowest team to its highest, and the
tick marks its average. `scale_colour_sdv()` and `scale_y_sdv()` take the
conference names like any team key, so the colors and the conference marks on
the axis come from sdvplotR (`theme_y_sdv()` lets the axis draw images).

```{r by-conference, fig.height = 7.5, fig.alt = "Every FBS team's logo placed by its net opponent-adjusted EPA per play, one row per conference ordered by average rating, with each conference's logo on the vertical axis and a bar in its color spanning its teams."}
conf_order <- ratings |>
  group_by(conference) |>
  summarise(mean_net = mean(net), lo = min(net), hi = max(net), .groups = "drop") |>
  arrange(mean_net, conference) |>
  mutate(row = row_number(), conference = factor(conference, levels = conference))
teams <- ratings |>
  inner_join(select(conf_order, conference, row) |> mutate(conference = as.character(conference)), by = "conference") |>
  arrange(row, net, school) |>
  group_by(conference) |>
  mutate(y = row + if_else(row_number() %% 2 == 0, 0.17, -0.17)) |>
  ungroup()

ggplot() +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_segment(
    data = conf_order,
    aes(x = lo, xend = hi, y = conference, yend = conference, colour = conference),
    linewidth = 7, alpha = 0.35, lineend = "round"
  ) +
  geom_point(
    data = conf_order, aes(x = mean_net, y = conference, colour = conference),
    shape = "|", size = 9
  ) +
  # a discrete axis still takes numbers: row 3 is the third conference, so 3.17 sits just above it
  geom_sdv_logos(data = teams, aes(x = net, y = y, team = school), sport = "cfb", height = 0.04) +
  scale_colour_sdv(sport = "cfb", guide = "none") +
  scale_y_sdv(sport = "cfb", size = 30, expand = expansion(add = 0.6)) +
  scale_x_continuous(expand = expansion(mult = 0.04)) +
  labs(
    title = paste("FBS teams by conference,", season, through),
    subtitle = "Net opponent-adjusted EPA per play; the tick marks the conference average",
    x = "Net adjusted EPA per play", y = NULL,
    caption = "Data: cfbfastR | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  ) +
  theme_y_sdv()
```

## 3. Offense and defense of the top 25

The two halves of the rating for the top 25, one logo each. The defense axis is
reversed so the better defenses sit higher, and the dashed lines are the top
25's averages.

```{r off-def, fig.alt = "The top 25 college football teams' logos placed by adjusted offensive EPA per play (horizontal) and adjusted defensive EPA allowed per play (vertical, reversed so better defenses sit higher), with dashed lines at the group's averages."}
ggplot(top25, aes(x = off, y = dfn)) +
  geom_hline(yintercept = mean(top25$dfn), linetype = "dashed", colour = "grey55") +
  geom_vline(xintercept = mean(top25$off), linetype = "dashed", colour = "grey55") +
  geom_sdv_logos(aes(team = school), sport = "cfb", height = 0.075) +
  scale_x_continuous(expand = expansion(mult = 0.08)) +
  scale_y_reverse(expand = expansion(mult = 0.08)) +
  labs(
    title = paste("How the top 25 get there,", season),
    subtitle = paste("Opponent-adjusted EPA per play,", through),
    x = "Adjusted offense: EPA per play",
    y = "Adjusted defense: EPA allowed per play (reversed)",
    caption = "Data: cfbfastR | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

## Related

- [NFL weekly](leaderboard-nfl-weekly.html) runs the same EPA tables on nflverse
  data.
- [College Football](cfb-viz.html) walks through cfbfastR and sdvplotR more
  broadly.
- [Social graphics, automated](automation-social.html) posts tables like these on
  a schedule.
