---
title: "WNBA Leaderboard"
description: >
  The latest WNBA season, rebuilt every week: standings with net rating,
  offense against defense with logos, and the scoring leaders with headshots.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 96
)
```

```{r season, include = FALSE}
library(sdvplotR)
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(gt)

today <- Sys.Date()
day_label <- function(day, month = "%b") sub(" 0", " ", format(as.Date(day), paste(month, "%d"))) # "Oct 5"
# The season tips off in May, so until then the calendar points at last season
current <- as.integer(format(today, "%Y")) - (format(today, "%m") < "05")

# A season's team box scores, or NULL when it has no regular-season games yet:
# load_wnba_team_box() errors for a season past wehoop's most recent one, and a
# new season's file can hold only preseason games at first
team_games <- function(season) {
  box <- tryCatch(suppressWarnings(wehoop::load_wnba_team_box(seasons = season)), error = function(e) NULL)
  if (NROW(box) == 0 || !any(box$season_type == 2)) NULL else box
}

season <- current
box <- team_games(season)
if (is.null(box)) {
  season <- current - 1
  box <- team_games(season)
}
stopifnot(!is.null(box))

last_game <- max(box$game_date)
playoffs <- filter(box, season_type == 3)
if (season < current) {
  status <- sprintf(
    "the offseason, so this is the final %d regular season; the %d season has no games yet.",
    season, current
  )
  through <- "final regular season"
} else if (nrow(playoffs) == 0) {
  status <- sprintf("the %d season through %s.", season, day_label(last_game, "%B"))
  through <- paste("through", day_label(last_game))
} else if (as.numeric(today - max(playoffs$game_date)) <= 10) {
  status <- sprintf("the final %d regular season; the playoffs are under way.", season)
  through <- "final regular season"
} else {
  status <- sprintf("the offseason, so this is the final %d regular season.", season)
  through <- "final regular season"
}
```

**Updated `r format(today)`:** `r status`

This page is rebuilt every week with the site. It builds the standings with
offensive, defensive and net ratings, plots offense against defense and lists
the scoring leaders with their headshots, for the latest WNBA season with games:
the season to date from May to September, the final regular season once it is
over. Data: [wehoop](https://wehoop.sportsdataverse.org)'s ESPN box scores, read
from release files (no stats.wnba.com calls).

## Picking the season

The WNBA season tips off in May, so until then the calendar points at last
season. `wehoop::load_wnba_team_box()` errors for a season past wehoop's most
recent one, and a new season's file can hold only preseason games at first, so
the helper treats "no regular-season games" as "not published" and the page
steps back one season. The status line says what the numbers cover.

```{r season-code, ref.label = "season", eval = FALSE}
```

ESPN files the All-Star Game and the Commissioner's Cup final as regular-season
games, though neither counts in the standings; the schedule marks them in
`type_abbreviation` (`ALLSTAR`, `CC`), so keep the standard games (`STD`) only.
That also drops the All-Star teams, whose abbreviations are not franchises.

```{r ratings}
standard <- wehoop::load_wnba_schedule(seasons = season) |>
  filter(season_type == 2, type_abbreviation == "STD")
stopifnot(identical(class(box$game_id), class(standard$game_id)))

games <- box |>
  filter(season_type == 2, game_id %in% standard$game_id) |>
  mutate(poss = field_goals_attempted - offensive_rebounds + total_turnovers +
    0.44 * free_throws_attempted) |>
  group_by(game_id) |>
  mutate(poss = mean(poss, na.rm = TRUE)) |> # one side's box can miss a count; the other covers it
  ungroup()

teams <- team_reference("wnba") |>
  select(team = team_abbr, name = team_short_name, conference)
ratings <- games |>
  arrange(game_date, game_id) |>
  group_by(team = team_abbreviation) |>
  summarise(
    w = sum(team_winner), l = sum(!team_winner),
    l10 = sum(tail(team_winner, 10)),
    ortg = 100 * sum(team_score) / sum(poss),
    drtg = 100 * sum(opponent_team_score) / sum(poss),
    .groups = "drop"
  ) |>
  mutate(pct = w / (w + l), net = ortg - drtg) |>
  inner_join(teams, by = "team") |>
  arrange(desc(pct), desc(net), team) # the team breaks a tie, so an unchanged week renders the same
nrow(ratings)
```

## 1. Standings with net rating

One table for the league, ordered by winning percentage (the WNBA seeds its
playoffs league-wide), with the conference beside each team and its points
scored and allowed per 100 possessions. `gt_color_pills()` draws net rating on a
scale centred at zero.

```{r standings}
standings <- ratings |>
  mutate(
    seed = row_number(),
    record = paste(w, l, sep = "-"),
    last10 = paste(l10, pmin(10, w + l) - l10, sep = "-"),
    logo = team
  ) |>
  select(seed, logo, name, conference, record, pct, last10, ortg, drtg, net)

standings |>
  gt(id = "wnba-standings") |>
  tab_header(
    title = paste("WNBA standings and ratings,", season),
    subtitle = paste0(toupper(substr(through, 1, 1)), substring(through, 2))
  ) |>
  fmt_number(pct, decimals = 3) |>
  fmt_number(c(ortg, drtg), decimals = 1) |>
  tab_spanner("Per 100 possessions", c(ortg, drtg, net)) |>
  cols_label(
    seed = "", logo = "", name = "Team", conference = "Conf", record = "W-L",
    pct = "Pct", last10 = "Last 10", ortg = "Off", drtg = "Def", net = "Net"
  ) |>
  tab_source_note(paste(
    "Data: wehoop (ESPN box scores). Possessions estimated from the box score;",
    "ties ordered by net rating. Viz: sdvplotR"
  )) |>
  gt_color_pills(
    net,
    palette = c("#c84630", "#f7f7f7", "#2e8b57"),
    domain = c(-1, 1) * max(abs(standings$net)), digits = 1, pill_height = 22
  ) |>
  gt_sdv_logos(columns = logo, sport = "wnba", height = 26) |>
  gt_theme_pl()
```

## 2. Offense against defense

Points scored per 100 possessions against points allowed, one logo per team. The
y axis is reversed so the better defenses sit higher, and the faint diagonals
are lines of equal net rating, five points apart.

```{r off-def, fig.alt = "Every WNBA team's logo placed by offensive rating (horizontal) and defensive rating (vertical, reversed so better defenses sit higher), with dashed lines at the league averages and faint diagonals of equal net rating."}
ggplot(ratings, aes(x = ortg, y = drtg)) +
  geom_abline(slope = 1, intercept = seq(-30, 30, by = 5), colour = "grey88", linewidth = 0.3) +
  geom_hline(yintercept = mean(ratings$drtg), linetype = "dashed", colour = "grey55") +
  geom_vline(xintercept = mean(ratings$ortg), linetype = "dashed", colour = "grey55") +
  geom_sdv_logos(aes(team = team), sport = "wnba", width = 0.065) +
  # logos do not widen the limits, so leave room for the outermost ones
  scale_x_continuous(expand = expansion(mult = 0.08)) +
  scale_y_reverse(expand = expansion(mult = 0.12)) +
  labs(
    title = paste("WNBA offense and defense,", season),
    subtitle = paste("Points per 100 possessions,", through),
    x = "Offensive rating (points scored per 100 possessions)",
    y = "Defensive rating (points allowed, reversed)",
    caption = "Data: wehoop (ESPN box scores) | Viz: sdvplotR"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"), panel.grid.minor = element_blank())
```

## 3. Scoring leaders with headshots

Points per game from the player box scores (standard games only), for players
who appeared in at least half of the most games any team has played. ESPN
athlete ids are sdvplotR's default id for WNBA headshots, and the table is
dressed in the scoring leader's team colors by `gt_theme_sdv_team()`.

```{r leaders-data}
players <- wehoop::load_wnba_player_box(seasons = season) |>
  filter(season_type == 2, !did_not_play, !is.na(minutes), game_id %in% standard$game_id)
most_games <- max(count(distinct(players, team_abbreviation, game_id), team_abbreviation)$n)

leaders <- players |>
  arrange(game_date, game_id) |>
  group_by(athlete_id) |>
  summarise(
    name = last(athlete_display_name), team = last(team_abbreviation), gp = n(),
    ppg = mean(points), rpg = mean(rebounds), apg = mean(assists),
    fg_pct = sum(field_goals_made) / sum(field_goals_attempted),
    .groups = "drop"
  ) |>
  filter(gp >= most_games / 2) |>
  arrange(desc(ppg), athlete_id) |>
  slice_head(n = 10) |>
  mutate(rank = row_number(), headshot = athlete_id, logo = team)
head(leaders)
```

```{r leaders-table}
leaders |>
  select(rank, headshot, name, logo, gp, ppg, rpg, apg, fg_pct) |>
  gt(id = "wnba-leaders") |>
  tab_header(
    title = paste("WNBA scoring leaders,", season),
    subtitle = paste0(
      "Points per game, minimum ", ceiling(most_games / 2), " games, ", through
    )
  ) |>
  fmt_number(c(ppg, rpg, apg), decimals = 1) |>
  fmt_percent(fg_pct, decimals = 1) |>
  cols_label(
    rank = "", headshot = "", name = "Player", logo = "", gp = "GP",
    ppg = "PTS", rpg = "REB", apg = "AST", fg_pct = "FG%"
  ) |>
  tab_source_note("Data: wehoop (ESPN box scores) | Viz: sdvplotR") |>
  gt_sdv_headshots(columns = headshot, sport = "wnba", height = 40) |>
  gt_sdv_logos(columns = logo, sport = "wnba", height = 26) |>
  gt_theme_sdv_team(team = leaders$team[1], sport = "wnba")
```

## Related

- [WNBA](wnba-viz.html) walks through wehoop and sdvplotR more broadly.
- [NBA](leaderboard-nba.html) builds the same standings from hoopR.
- [Social graphics, automated](automation-social.html) posts leaderboards like
  these on a schedule.
