---
title: "NHL Shot Map"
description: >
  Map every unblocked shot of the Stanley Cup clincher on a rink in team
  colors, sized by expected goals, with the goals, a scoreboard and exports at
  1200 x 675 and 1600 x 900.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7.5, fig.height = 4, dpi = 96
)
```

**The brief:** the morning after the Stanley Cup clincher, post one image of the
game: every unblocked shot on a rink, each team in its colors, the goals called
out, and the score. It goes out at 1200 x 675 for X and Bluesky and 1600 x 900
for the blog recap. The play-by-play is the SportsDataverse NHL release through
fastRhockey, and `sdv_surface()` draws the rink.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2026 # the 2025-26 season: NHL seasons are named by the year they end
cup_final <- 2025030416 # Stanley Cup Final, Game 6
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

`load_nhl_pbp()` reads the whole season's play-by-play, one row per event, and
`load_nhl_schedule()` the games with their final scores. The game's own row says
who played where; the series rows confirm this game ended the Final.

```{r data}
pbp <- fastRhockey::load_nhl_pbp(season)
schedule <- fastRhockey::load_nhl_schedule(season)

game <- filter(schedule, game_id == cup_final)
final <- schedule |>
  filter(playoff_round == 4, game_id <= cup_final) |>
  mutate(winner = if_else(home_score > away_score, home_team_abbr, away_team_abbr))
count(final, winner)
stopifnot(max(count(final, winner)$n) == 4) # the winner's fourth win: the clincher

shots <- pbp |>
  filter(game_id == cup_final, event_type %in% c("SHOT", "MISSED_SHOT", "GOAL")) |>
  transmute(
    event_type, period, period_time,
    team = event_team_abbr, shooter = event_player_1_name,
    x, y, x_fixed, y_fixed, xg
  )
sprintf("%s %d at %s %d", game$away_team_abbr, game$away_score, game$home_team_abbr, game$home_score)
count(shots, team, event_type)
```

Blocked shots are left out: the feed records where a blocked shot was blocked,
not where it was taken.

## 2. The first draft

Scatter the coordinates as they come, one color per team.

```{r draft, fig.alt = "First draft: the game's unblocked shots as colored dots on a blank grid, with both teams' shots at both ends."}
ggplot(shots, aes(x, y, color = team)) +
  geom_point()
```

Both teams have shots at both ends: teams switch ends every period, and the
feed records where on the ice each shot happened. With no rink the dots float in
space, and the goals look like every other shot.

## 3. One end per team, on a rink

The release already carries `x_fixed` and `y_fixed`: every event rotated so
the home team shoots right and the visitors left for the whole game.
`sdv_surface("nhl")` draws a regulation rink in the same coordinates (feet,
center ice at 0, 0) and returns a ggplot, so the shots are just more layers.

```{r rink, fig.alt = "The shots on a regulation NHL rink, the visitors' shots all at the left end and the home team's at the right."}
sdv_surface("nhl") +
  geom_point(aes(x_fixed, y_fixed, color = team), data = shots)
```

## 4. Team colors, shot types and the goals

Each team's shots take its primary color through `scale_color_sdv()`. A shot on
goal is a filled dot, a miss an open ring, and each dot is sized by its expected
goals (`xg`, the release's chance that a shot like it scores), so the dangerous
chances stand out from the long-range ones. A goal is a large ringed dot with
the scorer's name and the time in a white box, so it reads over the shots and
the rink lines; the labels alternate above and below, so two goals close together
keep their labels apart. The drawing goes in a function so the final layout can
reuse it.

```{r shots, fig.alt = "The shot map in team colors: filled dots for shots on goal, open rings for misses, sized by expected goals, and the three goals as large ringed dots labelled with the scorer and time."}
goals <- shots |>
  filter(event_type == "GOAL") |>
  mutate(
    label = paste0(sub(".* ", "", shooter), " (P", period, " ", period_time, ")"),
    vjust = if_else(row_number() %% 2 == 1, -0.8, 1.8) # above, below, above: close goals stay apart
  )

draw_shots <- function(...) {
  sdv_surface("nhl", ...) +
    geom_point(
      aes(x_fixed, y_fixed, color = team, size = xg),
      data = filter(shots, event_type == "SHOT"), alpha = 0.8
    ) +
    geom_point(
      aes(x_fixed, y_fixed, color = team, size = xg),
      data = filter(shots, event_type == "MISSED_SHOT"), shape = 1, stroke = 0.9
    ) +
    geom_point(
      aes(x_fixed, y_fixed, fill = team),
      data = goals, shape = 21, size = 6, color = "white", stroke = 1.5
    ) +
    geom_label(
      aes(x_fixed, y_fixed, label = label, color = team, vjust = vjust),
      data = goals, size = 2.6, fontface = "bold",
      fill = "white", linewidth = 0, label.padding = unit(0.12, "lines")
    ) +
    scale_color_sdv(sport = "nhl", guide = "none") +
    scale_fill_sdv(sport = "nhl", guide = "none") +
    scale_size_area(max_size = 4.5, guide = "none")
}

draw_shots()
```

## 5. The frame: score, teams and story

A shot map needs a scoreboard. Each team gets a block on the side of the rink it
attacks: its logo, its score and its shots on goal (shots plus goals), placed
past the boards. `xlims` widens the rink's canvas to make room, and goes to
'sportyR' through `sdv_surface()`'s `...`. The title says what happened, the
subtitle how to read the rink, and the caption carries the source.

```{r story, fig.width = 8, fig.height = 4.5, fig.alt = "The finished shot map: the Stanley Cup clincher's unblocked shots on a rink in team colors, with each team's logo, score and shots on goal beside its end, a headline and a source line."}
club_names <- team_reference("nhl") |> select(team_abbr, team_name)
scoreboard <- tibble(
  team = c(game$away_team_abbr, game$home_team_abbr),
  x = c(-122, 122), # outside the boards, on the side each team attacks
  score = c(game$away_score, game$home_score)
) |>
  left_join(
    count(filter(shots, event_type %in% c("SHOT", "GOAL")), team, name = "sog"),
    by = "team"
  ) |>
  left_join(club_names, by = c("team" = "team_abbr"))
winner <- slice_max(scoreboard, score, n = 1)
loser <- slice_min(scoreboard, score, n = 1)

shot_map <- function() {
  draw_shots(xlims = c(-142, 142)) +
    geom_sdv_logos(aes(x = x, y = 14, team = team), data = scoreboard, sport = "nhl", height = 0.2) +
    geom_text(aes(x = x, y = -8, label = score), data = scoreboard, size = 11, fontface = "bold") +
    geom_text(
      aes(x = x, y = -21, label = paste(sog, "shots\non goal")),
      data = scoreboard, size = 2.7, color = "grey40", lineheight = 0.9
    ) +
    labs(
      title = sprintf(
        "The %s won the Stanley Cup, %d-%d in Game 6",
        winner$team_name, winner$score, loser$score
      ),
      subtitle = sprintf(
        "Every unblocked shot, %s. %s shoots left, %s right. Dots: shots on goal; rings: misses; size: expected goals.",
        game$venue, game$away_team_abbr, game$home_team_abbr
      ),
      caption = "Data: NHL via fastRhockey  |  Chart: sdvplotR"
    ) +
    theme(
      plot.title = element_text(face = "bold", size = 14, margin = margin(b = 3)),
      plot.subtitle = ggtext::element_textbox_simple(color = "grey40", size = 8.5, margin = margin(b = 4)),
      plot.caption = element_text(color = "grey45", size = 7),
      plot.margin = margin(10, 12, 6, 12),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

shot_map()
```

## 6. Export for X and the blog

1200 x 675 for X and Bluesky and 1600 x 900 for the blog are the same 16:9
shape, so one 8 x 4.5 in plot serves both; only the `dpi` changes (150 and 200).
Inches times dpi gives the exact pixels, and the logos, sized as a share of the
panel, scale with everything else.

```{r export, fig.alt = "The finished shot map exported at 1200 by 675 pixels for X and Bluesky."}
chart <- shot_map()
files <- c(
  file.path(out_dir, "nhl_shot_map_1200x675.png"),
  file.path(out_dir, "nhl_shot_map_1600x900.png")
)
ggsave(files[1], chart, width = 8, height = 4.5, dpi = 150)
ggsave(files[2], chart, width = 8, height = 4.5, dpi = 200)
magick::image_info(magick::image_read(files))[c("width", "height")]
magick::image_read(files[1])
```
