---
title: "Head-to-Head Card"
description: >
  Make a Super Bowl tale-of-the-tape card: two teams' seasons as mirrored
  league-rank bars in team colors with dark-background logos, exported at
  1200 x 675 and 1080 x 1080.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7.5, fig.height = 4.2, dpi = 96
)
```

**The brief:** Super Bowl week. The social team wants a "tale of the tape" card
for the two teams: their regular seasons side by side, in each team's colors,
with logos, at 1200 x 675 for X and 1080 x 1080 for Instagram. The numbers come
from nflverse play-by-play and schedules through nflreadr.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2025
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

Every stat is computed for all 32 teams, not just the two finalists, because a
comparison needs context: each one also gets a league rank (1 is best, whichever
direction "best" is for that stat). Points come from the schedule, efficiency
and turnovers from the play-by-play.

```{r data}
schedule <- nflreadr::load_schedules(season)
sb <- filter(schedule, game_type == "SB")
pair <- c(sb$away_team, sb$home_team) # left and right on the card

scoring <- schedule |>
  filter(game_type == "REG") |>
  nflreadr::clean_homeaway() |>
  group_by(team) |>
  summarise(
    w = sum(team_score > opponent_score),
    l = sum(team_score < opponent_score),
    ppg = mean(team_score),
    papg = mean(opponent_score)
  )

plays <- nflreadr::load_pbp(season) |>
  filter(season_type == "REG", play_type %in% c("pass", "run"), !is.na(epa))
offense <- plays |>
  group_by(team = posteam) |>
  summarise(
    off_epa = mean(epa),
    pass_epa = mean(epa[pass == 1]),
    rush_epa = mean(epa[rush == 1]),
    giveaways = sum(interception + fumble_lost)
  )
defense <- plays |>
  group_by(team = defteam) |>
  summarise(def_epa = mean(epa), takeaways = sum(interception + fumble_lost))

teams <- scoring |>
  inner_join(offense, by = "team") |>
  inner_join(defense, by = "team") |>
  mutate(to_margin = takeaways - giveaways)
stopifnot(nrow(teams) == 32)

stats <- tibble::tribble(
  ~stat,       ~label,                          ~fmt,     ~higher_is_better,
  "ppg",       "Points per game",               "%.1f",   TRUE,
  "papg",      "Points allowed per game",       "%.1f",   FALSE,
  "off_epa",   "Offense EPA per play",          "%+.3f",  TRUE,
  "def_epa",   "Defense EPA per play allowed",  "%+.3f",  FALSE,
  "pass_epa",  "EPA per dropback",              "%+.3f",  TRUE,
  "rush_epa",  "EPA per rush",                  "%+.3f",  TRUE,
  "to_margin", "Turnover margin",               "%+.0f",  TRUE
)

ranked <- teams |>
  select(team, all_of(stats$stat)) |>
  tidyr::pivot_longer(-team, names_to = "stat") |>
  inner_join(stats, by = "stat") |>
  group_by(stat) |>
  mutate(rank = if_else(higher_is_better, min_rank(desc(value)), min_rank(value))) |>
  ungroup()

filter(teams, team %in% pair) |>
  select(team, w, l, all_of(stats$stat))
```

## 2. The first draft

Both teams' numbers as grouped bars.

```{r draft, fig.alt = "First draft: grouped bars of the two Super Bowl teams' raw season numbers, where points per game dwarf every EPA value."}
two <- filter(ranked, team %in% pair)
ggplot(two, aes(value, label, fill = team)) +
  geom_col(position = "dodge")
```

Useless: points per game (about 25) dwarf EPA per play (about 0.1), "more" is
good for some rows and bad for others, and the default red and teal belong to
neither team.

## 3. One scale: league rank

Ranks put every stat on the same 1-32 scale with the same direction, so a longer
bar is always better. Mirroring the two teams around a center column of labels
(a "butterfly") makes each row a direct comparison; the actual value and the
rank sit at the end of each bar. The bars are drawn with `geom_rect()` on a
plain canvas: x runs from -1.35 to 1.35, the labels sit at 0, and row 1 is at
the top.

```{r butterfly, fig.alt = "A butterfly chart: the two teams' league ranks as mirrored bars around a center column of stat names, with each value and rank at the bar end, in ggplot2's default colors."}
ordinal <- function(n) {
  suffix <- ifelse(n %% 100 %in% 11:13, "th", c("th", "st", "nd", "rd", rep("th", 6))[n %% 10 + 1])
  paste0(n, suffix)
}

gap <- 0.42 # half the width of the label column
bars <- two |>
  mutate(
    side = if_else(team == pair[1], -1, 1),
    row = match(stat, stats$stat),
    start = side * gap,
    end = start + side * (33 - rank) / 32 * 0.55, # rank 1 is the longest bar
    value_label = if_else(
      side > 0,
      paste0(sprintf(fmt, value), "  (", ordinal(rank), ")"),
      paste0("(", ordinal(rank), ")  ", sprintf(fmt, value))
    )
  )

butterfly <- function(ink = "#1d1d1d", muted = "#6b6b6b") {
  ggplot(bars) +
    geom_rect(aes(xmin = pmin(start, end), xmax = pmax(start, end), ymin = row - 0.31, ymax = row + 0.31, fill = team)) +
    geom_text(aes(x = 0, y = row, label = label), data = distinct(bars, row, label), color = ink, size = 3.1) +
    geom_text(
      aes(x = end + side * 0.02, y = row, label = value_label, hjust = if_else(side > 0, 0, 1)),
      color = muted, size = 2.8
    ) +
    scale_y_reverse() +
    coord_cartesian(xlim = c(-1.35, 1.35)) +
    theme_void()
}

butterfly()
```

## 4. The teams' own colors

Swapping the defaults for team colors runs into a real-world snag: both teams'
primary color is the same navy, so the card would be one color on both sides.
Each team's secondary color tells them apart.

```{r colors, fig.alt = "The butterfly chart with each side in its team's secondary color, since both teams share the same navy primary."}
primary <- sdv_team_colors("nfl", pair)
secondary <- sdv_team_colors("nfl", pair, type = "secondary")
rbind(primary, secondary)
colors <- if (anyDuplicated(primary)) secondary else primary

butterfly() + scale_fill_manual(values = colors, guide = "none")
```

## 5. Make it a card

The card is dark, which most game-week graphics are, so the logos are the dark
variant: the mark drawn for a dark background, which keeps navy outlines from
disappearing. `geom_sdv_logos()` draws a team's primary mark; `team_reference()`
lists every team's `logo_dark_url`, and `geom_from_path()` (from ggpath,
re-exported by sdvplotR) draws any image URL. Each team gets its logo and record
above its side, the kicker names the game, and the footer the source. The
header is more rows on the same canvas, at negative y, so one function draws
both export sizes.

```{r card, fig.width = 8, fig.height = 4.5, fig.alt = "The finished tale-of-the-tape card on a dark background: the two Super Bowl teams' dark-variant logos and records above mirrored league-rank bars in their colors, with a kicker, a title and a source line."}
bg <- "#0f1923"
ink <- "#ffffff"
muted <- "#9fb0c3"

header <- team_reference("nfl") |>
  filter(team_abbr %in% pair) |>
  transmute(
    team = team_abbr, logo = logo_dark_url, name = team_mascot,
    x = if_else(team == pair[1], -0.85, 0.85)
  ) |>
  inner_join(select(teams, team, w, l), by = "team") |>
  mutate(record = paste0(name, "  ", w, "-", l))
stopifnot(!anyNA(header$logo))

card <- function(logo_width = 0.12) {
  butterfly(ink = ink, muted = muted) +
    scale_fill_manual(values = colors, guide = "none") +
    scale_color_manual(values = colors, guide = "none") +
    geom_from_path(aes(x = x, y = -1.75, path = logo), data = header, width = logo_width) +
    geom_text(aes(x = x, y = -0.45, label = record, color = team), data = header, size = 4, fontface = "bold") +
    annotate("text", x = 0, y = -2.45, label = "SUPER BOWL LX", color = muted, size = 3, fontface = "bold") +
    annotate("text", x = 0, y = -1.75, label = "Tale of the tape", color = ink, size = 5.6, fontface = "bold") +
    annotate(
      "text", x = 0, y = -1.1, label = paste(season, "regular season, rank among 32 teams"),
      color = muted, size = 2.9
    ) +
    labs(caption = "Data: nflverse via nflreadr  |  made with sdvplotR") +
    theme(
      plot.background = element_rect(fill = bg, color = NA),
      plot.caption = element_text(color = muted, size = 7.5, hjust = 0.5),
      plot.margin = margin(14, 14, 8, 14)
    )
}

card()
```

## 6. Export for X and Instagram

The same function draws both posts: 8 x 4.5 in and 7.2 x 7.2 in at 150 dpi are
exactly 1200 x 675 and 1080 x 1080. `ggsave()` needs the card's background
passed as `bg`, or the margins come out white. `width` is a share of the
panel's width, and the square's panel is narrower, so its logos get a slightly
larger share.

```{r export, fig.alt = "The finished tale-of-the-tape card exported at 1200 by 675 pixels for X."}
wide <- file.path(out_dir, "sb_tale_of_the_tape_1200x675.png")
square <- file.path(out_dir, "sb_tale_of_the_tape_1080x1080.png")
ggsave(wide, card(), width = 8, height = 4.5, dpi = 150, bg = bg)
ggsave(square, card(logo_width = 0.14), width = 7.2, height = 7.2, dpi = 150, bg = bg)
magick::image_info(magick::image_read(c(wide, square)))[c("width", "height")]
magick::image_read(wide)
```

The square cut:

```{r export-square, fig.alt = "The same card exported at 1080 by 1080 pixels for Instagram."}
magick::image_read(square)
```
