---
title: "CFB Conference Table"
description: >
  Build the final Big Ten standings table for a newsletter, with logos, records
  and game-by-game margin bars, exported 1600 px wide and as a square for
  social.
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

**The brief:** the season-review newsletter needs the final Big Ten standings as
an image: 1600 px wide for the email, plus a square cut for social. Indiana went
16-0 and won the national title, so the table should make that obvious. The
records are built from the CollegeFootballData.com schedule through cfbfastR,
and the table is gt with sdvplotR's logo, theme and export helpers.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr, warn.conflicts = FALSE)

season <- 2025
conference <- "Big Ten"
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

The schedule has one row per game. Stacking the home and away sides gives one
row per team per game, which makes every record a `group_by()`. The margins are
kept in date order as one comma-separated string per team, the form gt's
in-cell charts read, for a small chart later. `summarise()` sees its own earlier
results, so the margins and the last game's score are taken before `pf` and
`pa` become season averages.

```{r data}
schedule <- cfbfastR::load_cfb_schedules(season) |>
  filter(completed)

side <- function(me, opp) {
  transmute(
    schedule,
    start_date, season_type, conference_game, notes,
    team = .data[[paste0(me, "_team")]],
    conference = .data[[paste0(me, "_conference")]],
    opponent = .data[[paste0(opp, "_team")]],
    pf = .data[[paste0(me, "_points")]],
    pa = .data[[paste0(opp, "_points")]]
  )
}

games <- bind_rows(side("home", "away"), side("away", "home")) |>
  filter(conference == !!conference) |>
  arrange(start_date) |>
  mutate(won = pf > pa)

standings <- games |>
  group_by(team) |>
  summarise(
    margins = paste(pf - pa, collapse = ","),
    # the last game: where a bowl or the playoff shows up
    last_score = paste0(max(last(pf), last(pa)), "-", min(last(pf), last(pa))),
    last_type = last(season_type),
    last_won = last(won),
    last_opponent = last(opponent),
    last_event = last(notes),
    conf_w = sum(won & conference_game),
    conf_l = sum(!won & conference_game),
    w = sum(won),
    l = sum(!won),
    pf = mean(pf),
    pa = mean(pa)
  ) |>
  arrange(desc(conf_w), desc(w), team) # team breaks ties, so the order is the same every run
nrow(standings)
standings |>
  select(team, conf_w, conf_l, w, l, last_score, last_opponent, last_event) |>
  head(5)
```

## 2. The first draft

Hand the frame to gt as it is.

```{r draft}
gt(standings)
```

Every number is there, and none of it is readable: margins printed as text, a
dozen decimals, and column names only the analyst knows.

## 3. Shape it for a reader

Records read as "9-0", not two columns. Each team's last game becomes one short
line ("W 27-21 vs Miami, CFP National Championship"), which is where the
national title shows up. Columns get real labels, conference and overall
records sit under spanners, and the averages get one decimal.

```{r shape}
event <- function(x) {
  x |>
    sub(" Presented by.*", "", x = _) |>
    sub(" at the .*", "", x = _) |>
    sub("College Football Playoff", "CFP", x = _)
}

table <- standings |>
  mutate(
    conf = paste0(conf_w, "-", conf_l),
    overall = paste0(w, "-", l),
    postseason = if_else(
      last_type == "postseason",
      paste0(if_else(last_won, "W ", "L "), last_score, " vs ", last_opponent, ", ", event(last_event)),
      ""
    )
  ) |>
  select(team, conf, overall, pf, pa, margins, postseason)

draft <- gt(table) |>
  cols_hide(margins) |>
  cols_label(
    team = "Team", conf = "W-L", overall = "W-L",
    pf = "Pts/G", pa = "Opp/G", postseason = "Postseason"
  ) |>
  tab_spanner("Conference", conf) |>
  tab_spanner("Overall", c(overall, pf, pa)) |>
  fmt_number(c(pf, pa), decimals = 1) |>
  cols_align("center", c(conf, overall, pf, pa))
draft
```

## 4. Logos and the season at a glance

`gt_sdv_logos()` turns the team names into logos, and `include_name = TRUE`
keeps the name beside each one, so the column still reads as text. The margins
string becomes a nanoplot, gt's in-cell bar chart: one bar per game, green for a
win and red for a loss, so a perfect season is a solid green row.

```{r marks}
margin_bars <- nanoplot_options(
  data_bar_fill_color = "#2e8540",
  data_bar_negative_fill_color = "#c0392b",
  data_bar_stroke_color = "transparent",
  data_bar_negative_stroke_color = "transparent",
  show_data_points = FALSE,
  show_reference_line = FALSE,
  show_vertical_guides = FALSE,
  show_y_axis_guide = FALSE,
  interactive_data_values = TRUE # values on hover only, so the saved image stays clean
)

with_marks <- draft |>
  gt_sdv_logos(team, sport = "cfb", height = 24, include_name = TRUE) |>
  cols_nanoplot(
    columns = margins, plot_type = "bar", autoscale = TRUE,
    new_col_name = "bars", new_col_label = "Game by game",
    after = "pa", options = margin_bars
  ) |>
  tab_spanner("Margin", bars)
with_marks
```

## 5. Theme it and say what it means

A theme does the typography and rules in one call: `gt_theme_sdv()`, the
SportsDataverse house style. The title says the news, the subtitle how to read
the table, and the source note credits the data. The champion's row gets a soft
fill in its own color from `sdv_team_colors()`, and a footnote owns up to the
ordering: teams tied on conference record are listed by overall record, which is
not the conference's tiebreaker.

```{r final}
champ <- standings[1, ]
stopifnot(champ$l == 0, champ$last_won, grepl("National Championship", champ$last_event))
fill <- paste0(sdv_team_colors("cfb", champ$team), "1F") # the primary color at 12% opacity

final <- with_marks |>
  tab_header(
    title = sprintf(
      "%s ran the table: %d-0 in the %s, %d-0 overall and national champion",
      champ$team, champ$conf_w, conference, champ$w
    ),
    subtitle = html(paste0(
      "Final ", season, " ", conference, " standings. Bars are each game's margin, in date order: ",
      "<span style='color:#2e8540'><b>wins</b></span> and ",
      "<span style='color:#c0392b'><b>losses</b></span>."
    ))
  ) |>
  tab_source_note("Data: CollegeFootballData.com via cfbfastR  |  Table: sdvplotR + gt") |>
  tab_footnote(
    "Teams tied on conference record are listed by overall record, then by name.",
    locations = cells_column_labels(columns = conf)
  ) |>
  tab_style(style = cell_fill(color = fill), locations = cells_body(rows = team == champ$team)) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = team, rows = team == champ$team)
  ) |>
  gt_theme_sdv()
final
```

## 6. Export for the newsletter and for social

`gt_save_crop()` renders the table in headless Chrome, trims it with an even
border and, with `width`, scales it to the email's 1600 px. `gt_social_crop()`
centers the same table on a square canvas for Instagram, never cropping it: it
pads whichever way the table needs, side padding for a tall table and room above
and below for a wide one like this.

```{r export, fig.alt = "The finished Big Ten standings table exported 1600 pixels wide for the newsletter."}
newsletter <- gt_save_crop(final, file.path(out_dir, "big_ten_1600.png"), width = 1600)
square <- gt_social_crop(
  final, file.path(out_dir, "big_ten_1080x1080.png"),
  aspect_ratio = "1:1", width = 1080
)
magick::image_info(magick::image_read(c(newsletter, square)))[c("width", "height")]
magick::image_read(newsletter)
```

The square cut keeps the whole table and pads above and below it:

```{r export-square, fig.alt = "The same table centered on a 1080 by 1080 pixel square for Instagram."}
magick::image_read(square)
```
