---
title: "NBA Net Rating Quadrant"
description: >
  Chart every NBA team's offensive and defensive rating by conference, with
  logos, average lines and net-rating guides, for a blog post and a square
  social post.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7.5, fig.height = 4.5, dpi = 96
)
```

**The brief:** a season-review blog post needs one chart of where all 30 teams
finished on offense and defense, split by conference, at 1600 px wide, plus a
square version for Instagram. In ggplot2 each fix is one more layer: logos with
`geom_sdv_logos()`, per-conference averages with `geom_mean_lines()`, and
facets for the conferences. The ratings come from hoopR's ESPN team box scores.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2026 # the 2025-26 season: NBA seasons are named by the year they end
label <- paste0(season - 1, "-", season %% 100)
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

Offensive and defensive rating are points scored and allowed per 100
possessions. Possessions are estimated from the box score (FGA - OREB + TOV +
0.44 FTA) and averaged with the opponent's, so both teams in a game share one
count.

```{r data}
box <- hoopR::load_nba_team_box(season) |>
  filter(season_type == 2) # regular season

games <- box |>
  mutate(
    poss = field_goals_attempted - offensive_rebounds + total_turnovers +
      0.44 * free_throws_attempted
  )
filter(games, is.na(poss)) |>
  select(game_id, game_date, team_abbreviation, opponent_team_abbreviation)
```

One box score is missing a stat, so that game has no possession count. An `NA`
inside `sum()` would turn a whole season into `NA`, so the game is dropped, for
both teams together: after the join both rows of that game have no shared count.

```{r ratings}
opponents <- select(games, game_id, opponent_team_id = team_id, opp_poss = poss)
stopifnot(identical(class(games$opponent_team_id), class(opponents$opponent_team_id)))

ratings <- games |>
  inner_join(opponents, by = c("game_id", "opponent_team_id")) |>
  mutate(game_poss = (poss + opp_poss) / 2) |>
  filter(!is.na(game_poss)) |>
  group_by(team = team_abbreviation) |>
  summarise(
    ortg = 100 * sum(team_score) / sum(game_poss),
    drtg = 100 * sum(opponent_team_score) / sum(game_poss)
  ) |>
  mutate(net = ortg - drtg)
stopifnot(!anyNA(ratings$net))
nrow(ratings)
```

`r nrow(ratings)` teams in a 30-team league. `clean_team_abbrs()` finds the
extras: with `keep_non_matches = FALSE`, anything that is not an NBA team comes
back `NA`. They are the All-Star Game's teams, which ESPN files as
regular-season games. The conference comes from `team_reference()`, keyed by the
same cleaned abbreviation.

```{r resolve}
ratings <- ratings |>
  mutate(team_abbr = clean_team_abbrs(team, sport = "nba", keep_non_matches = FALSE))
filter(ratings, is.na(team_abbr))$team

conferences <- team_reference("nba") |>
  transmute(team_abbr, conference = sub("ern$", "", conference))
ratings <- ratings |>
  filter(!is.na(team_abbr)) |>
  inner_join(conferences, by = "team_abbr") |>
  arrange(desc(net), team_abbr)
stopifnot(nrow(ratings) == 30)
head(ratings)
```

## 2. The first draft

Points and abbreviations, ggplot2's defaults.

```{r draft, fig.alt = "First draft: a plain scatter of NBA teams' offensive rating against defensive rating, labelled with overlapping abbreviations."}
ggplot(ratings, aes(ortg, drtg, label = team_abbr)) +
  geom_point() +
  geom_text(nudge_y = 0.4, size = 3)
```

Three problems: the labels overlap, a low defensive rating is good but sits at
the bottom, and the two conferences (which only meet a third of the time) are
mixed together.

## 3. Logos, one panel per conference, and average lines

`geom_sdv_logos()` replaces the points, `facet_wrap()` splits the conferences,
and `geom_mean_lines()` draws each panel's own average offense and defense, so
every team is read against its conference.

```{r facets, fig.alt = "NBA teams drawn as logos in two panels, East and West, with dashed lines at each conference's average offensive and defensive rating."}
ggplot(ratings, aes(ortg, drtg)) +
  geom_mean_lines(aes(x0 = ortg, y0 = drtg), color = "grey60", linewidth = 0.5) +
  geom_sdv_logos(aes(team = team_abbr), sport = "nba", width = 0.1) +
  facet_wrap(vars(conference))
```

Better, but the best defenses are at the bottom, logos at the edges are cut off,
and the axis titles are column names.

## 4. Point the axes the right way and add net-rating guides

`scale_y_reverse()` puts the best defenses on top, so up and to the right is
good on both axes, and `expansion()` on both scales leaves room for the edge
logos. Net rating is the gap between the two numbers, so teams with the same net
rating sit on a diagonal: faint lines at -10, -5, 0, +5 and +10 show it without a
third axis. Each line is cut to the data's range and drawn as data
(`geom_segment()`), so the reversed scale moves it with the logos; a data frame
with no `conference` column repeats in both panels.

```{r guides, fig.alt = "The two-panel logo chart with the defensive axis reversed and faint diagonal lines marking net ratings of -10, -5, 0, +5 and +10."}
x_lim <- range(ratings$ortg) + c(-1, 1)
y_lim <- range(ratings$drtg) + c(-1, 1)
# the line drtg = ortg - net, cut to the box x_lim by y_lim
guides <- data.frame(net = c(-10, -5, 0, 5, 10)) |>
  mutate(
    x0 = pmax(x_lim[1], y_lim[1] + net),
    x1 = pmin(x_lim[2], y_lim[2] + net)
  ) |>
  filter(x0 < x1) |>
  mutate(
    y0 = x0 - net, y1 = x1 - net,
    lx = x0 + 0.88 * (x1 - x0), ly = lx - net, # a label near the lower-right end
    label = if_else(net > 0, paste0("+", net), as.character(net))
  )

net_guides <- list(
  geom_segment(
    aes(x = x0, y = y0, xend = x1, yend = y1),
    data = guides, inherit.aes = FALSE, color = "grey88", linewidth = 0.4
  ),
  geom_text(
    aes(x = lx, y = ly, label = label),
    data = guides, inherit.aes = FALSE, color = "grey60", size = 2.6
  ),
  geom_mean_lines(aes(x0 = ortg, y0 = drtg), color = "grey60", linewidth = 0.5)
)

p <- ggplot(ratings, aes(ortg, drtg)) +
  net_guides +
  geom_sdv_logos(aes(team = team_abbr), sport = "nba", width = 0.1) +
  facet_wrap(vars(conference)) +
  scale_x_continuous(expand = expansion(mult = 0.06)) +
  scale_y_reverse(expand = expansion(mult = 0.08)) +
  labs(
    x = "Offensive rating (points per 100 possessions)",
    y = "Defensive rating (allowed per 100)"
  )
p
```

## 5. Polish and tell the story

A theme strips the chart junk (minor grid, panel background) and sets the type;
the strip titles become plain bold labels on the left. The title states the
finding, the subtitle explains the reading, and the caption carries the source.
The two conference leaders get their net rating printed beside the logo.

```{r story, fig.alt = "Final two-panel chart of NBA offensive and defensive ratings by conference, with net-rating diagonals, conference average lines, and each conference leader's net rating labelled."}
leaders <- ratings |>
  group_by(conference) |>
  slice_max(net, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(note = sprintf("%s %+.1f", team_abbr, net))
top <- slice_max(leaders, net, n = 1)
east <- filter(leaders, conference == "East")

title <- if (top$conference == "West") {
  sprintf("%s stood alone at %+.1f; %s led the East", top$team_abbr, top$net, east$team_abbr)
} else {
  sprintf("%s led the league at %+.1f", top$team_abbr, top$net)
}

blog_theme <- theme_minimal(base_size = 10) +
  theme(
    plot.title.position = "plot",
    plot.title = ggtext::element_textbox_simple(face = "bold", size = 14, margin = margin(b = 4)),
    plot.subtitle = ggtext::element_textbox_simple(color = "grey40", size = 9, margin = margin(b = 8)),
    plot.caption = element_text(color = "grey45", size = 7.5),
    strip.text = element_text(face = "bold", size = 11, hjust = 0),
    axis.title = element_text(color = "grey40", size = 9),
    axis.text = element_text(color = "grey55"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey94"),
    plot.margin = margin(12, 14, 8, 12),
    plot.background = element_rect(fill = "white", color = NA)
  )

net_chart <- function(logo_width = 0.1, ncol = 2) {
  ggplot(ratings, aes(ortg, drtg)) +
    net_guides +
    geom_sdv_logos(aes(team = team_abbr), sport = "nba", width = logo_width) +
    geom_text(
      aes(label = note), data = leaders,
      hjust = 0, nudge_x = 1.1, size = 3, fontface = "bold", color = "grey20"
    ) +
    facet_wrap(vars(conference), ncol = ncol) +
    scale_x_continuous(expand = expansion(mult = 0.06)) +
    scale_y_reverse(expand = expansion(mult = 0.08)) +
    labs(
      title = title,
      subtitle = paste0(
        "Points scored and allowed per 100 possessions, ", label,
        " regular season; up and right is better. Dashed lines: each conference's ",
        "average. Diagonals: net rating, in steps of 5."
      ),
      x = "Offensive rating (points per 100 possessions)",
      y = "Defensive rating (allowed per 100)",
      caption = "Data: ESPN via hoopR  |  Chart: sdvplotR"
    ) +
    blog_theme
}

net_chart()
```

## 6. Export for the blog and for Instagram

`ggsave()` takes the size in inches and a `dpi`: 8 x 4.5 in at 200 dpi is the
blog's 1600 x 900. The square post stacks the conferences (`ncol = 1`). Each
panel is then about twice as wide as before, and `width` is a share of the
panel's width, so the square gets about half the share to keep the logos the
same size.

```{r export, fig.alt = "The finished NBA net rating chart exported at 1600 by 900 pixels for the blog."}
blog <- file.path(out_dir, "nba_net_rating_1600x900.png")
ggsave(blog, net_chart(), width = 8, height = 4.5, dpi = 200)

square <- file.path(out_dir, "nba_net_rating_1080x1080.png")
ggsave(square, net_chart(logo_width = 0.055, ncol = 1), width = 7.2, height = 7.2, dpi = 150)

magick::image_info(magick::image_read(c(blog, square)))[c("width", "height")]
magick::image_read(blog)
```

The square cut, stacked:

```{r export-square, fig.alt = "The same chart with the conferences stacked, exported at 1080 by 1080 pixels for Instagram."}
magick::image_read(square)
```
