---
title: "College Hoops Tiers"
description: >
  Turn a power rating into a Big Ten men's basketball tier list with
  sdv_team_tiers(), the tiers cut at the rating gaps, exported for social at
  1200 x 675 and 1080 x 1080.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7.5, fig.height = 4.6, dpi = 96
)
```

**The brief:** a college basketball account wants an end-of-season Big Ten
tier list, the kind fans argue about, but backed by a power rating instead of
vibes, as a 1200 x 675 post plus a 1080 x 1080 cut. The rating is the
SportsDataverse adjusted efficiency margin for men's college basketball through
hoopR, and `sdv_team_tiers()`, sdvplotR's Tiermaker, draws the list.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2026 # the 2025-26 season, named by the year it ends
conference <- "Big Ten"
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

One row per team: adjusted offense, defense and their margin (points per 100
possessions better than an average Division I team, adjusted for opponents).
The ratings carry ESPN team ids as text; `team_reference()` has the same ids as
integers, plus each team's conference and abbreviation. The id is cast once, on
the reference side, and the two types are checked before the join.

```{r data}
ratings <- hoopR::load_mbb_ratings(season)
teams <- team_reference("mbb") |>
  transmute(team_id = as.character(espn_team_id), team_abbr, team_short_name, conference)
stopifnot(identical(class(ratings$team_id), class(teams$team_id)))

league <- ratings |>
  inner_join(teams, by = "team_id") |>
  filter(conference == !!conference) |>
  arrange(desc(adj_em), team_abbr) |>
  select(team_abbr, team_short_name, rank, adj_em)
league
```

## 2. The first draft

Five tiers of equal size, straight into `sdv_team_tiers()`.

```{r draft, fig.alt = "First draft: the 18 Big Ten teams split into five equal tiers on the Tiermaker's dark background, with the default joke tier labels."}
draft <- league |>
  mutate(team = team_abbr, tier_no = ((row_number() - 1) * 5) %/% n() + 1) # five tiers of (nearly) equal size
sdv_team_tiers(draft, sport = "mbb")
```

It looks like a tier list, but the tiers are wrong. Equal-sized groups put the
best team in the conference in the same tier as a team more than ten points worse
per 100 possessions, and split neighbors less than two points apart. The default
labels ("Elite" ... "What are they doing?") are jokes, not information.

## 3. Let the ratings draw the lines

Tiers should break where the ratings do. Sorting by rating and measuring the gap
to the team above makes the widest gaps easy to find; cutting at the four widest
gives five tiers whose members are close to each other.

```{r gaps}
league <- league |>
  mutate(gap = lag(adj_em) - adj_em)
cut <- sort(league$gap, decreasing = TRUE)[4] # the fourth-widest gap
league <- league |>
  mutate(tier_no = cumsum(coalesce(gap, 0) >= cut) + 1)
league |>
  filter(gap >= cut) |>
  select(team_short_name, adj_em, gap, tier_no)
```

Each row above starts a new tier, and the size of its gap is the size of the
step down from the tier before.

## 4. Labels that say something

Each tier's label becomes the rating range of its members, so the list carries
its own evidence, and the title states the finding. `tier_rank` keeps teams in
rating order inside a tier. The subtitle explains the rating and the rule for
the cuts; the caption names the data.

```{r labels, fig.alt = "The Big Ten tier list with tiers cut at the four widest rating gaps, each tier labelled with its rating range, still on the dark background."}
ranges <- league |>
  group_by(tier_no) |>
  summarise(hi = max(adj_em), lo = min(adj_em)) |>
  mutate(desc = if_else(hi == lo, sprintf("%+.1f", hi), sprintf("%+.1f to %+.1f", hi, lo)))
tier_desc <- setNames(ranges$desc, ranges$tier_no)

top_tier <- filter(league, tier_no == 1)
title <- if (nrow(top_tier) == 1) {
  paste(top_tier$team_short_name, "stood alone in the", conference)
} else {
  paste(nrow(top_tier), "teams made up the", conference, "top tier")
}

tiers <- league |>
  group_by(tier_no) |>
  mutate(tier_rank = row_number()) |>
  ungroup() |>
  select(team = team_abbr, tier_no, tier_rank)

tier_list <- function(width = 0.075, alpha = 0.8, theme = "dark") {
  sdv_team_tiers(
    tiers,
    sport = "mbb",
    title = title,
    subtitle = paste0(
      "Tiers by adjusted efficiency margin, ", season - 1, "-", season %% 100,
      ". A new tier starts at each of the four widest gaps."
    ),
    caption = paste(
      "Margin: points per 100 possessions better than an average D-I team.",
      "Data: SportsDataverse ratings via hoopR  |  sdvplotR Tiermaker",
      sep = "\n"
    ),
    tier_desc = tier_desc,
    width = width,
    alpha = alpha,
    theme = theme
  )
}

tier_list()
```

## 5. Light or dark

On the Tiermaker's dark background, `sdv_team_tiers()` draws each school's
dark-background logo (its default `variant = "auto"`), so Iowa's hawk is gold,
Michigan State's Spartan white and Penn State's lion outlined instead of fading
into the background. A light background suits a page or feed that is white
anyway: `theme = "light"` draws the list on white, with the text and tier lines
turned dark, and each school's usual logo. The logos go to full opacity too
(`alpha = 1`); the default 0.8 softens them against the dark background but
washes them out on white. The exports below use the light one.

```{r light, fig.alt = "The finished tier list on a white background, logos at full strength, tiers labelled with rating ranges."}
tier_list(alpha = 1, theme = "light")
```

## 6. Export at social sizes

`ggsave()` re-lays the same plot at each export size. `width` is a share of the
panel's width, and the square's panel is narrower than the wide cut's, so the
square gets a slightly larger share to keep the logos the same size. The widest
tier holds the most logos side by side, which is why the wide 1200 x 675 cut is
the better post here.

```{r export, fig.alt = "The finished Big Ten tier list exported at 1200 by 675 pixels."}
wide <- file.path(out_dir, "big_ten_tiers_1200x675.png")
square <- file.path(out_dir, "big_ten_tiers_1080x1080.png")
ggsave(wide, tier_list(alpha = 1, theme = "light"), width = 8, height = 4.5, dpi = 150)
ggsave(square, tier_list(width = 0.085, alpha = 1, theme = "light"), width = 7.2, height = 7.2, dpi = 150)
magick::image_info(magick::image_read(c(wide, square)))[c("width", "height")]
magick::image_read(wide)
```

The 1080 x 1080 cut:

```{r export-square, fig.alt = "The same tier list exported at 1080 by 1080 pixels."}
magick::image_read(square)
```
