---
title: "Rank Bump Chart"
description: >
  Draw the college football season's AP Top 25 as a bump chart, with logos at
  the line ends and the biggest climbers called out, exported for a blog and
  for X.
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE,
  fig.width = 7.5, fig.height = 4.6, dpi = 96
)
```

**The brief:** a season-review piece wants the whole college football season in
one picture: how the final top 10 of the AP poll got there, poll by poll, with
the teams that climbed most called out and logos at the end of each line. It
runs 1600 x 900 on the blog and 1200 x 675 on X. The polls come from
CollegeFootballData.com through cfbfastR (its `cfbd_*()` functions need a free
API key in `CFBD_API_KEY`); ggplot2 draws the "bump chart" and sdvplotR the
logos and colors.

```{r setup}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)

season <- 2025
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

One row per team per poll. The AP poll comes out every week of the regular
season (the first one is the preseason poll) and once more after the bowls;
numbering the polls in order gives the x axis, with the final poll last.

```{r data}
ap <- cfbfastR::cfbd_rankings(year = season) |>
  filter(poll == "AP Top 25") |>
  mutate(poll_no = dense_rank(if_else(season_type == "postseason", 100L, week))) |>
  select(poll_no, season_type, week, rank, school, points)
n_polls <- max(ap$poll_no)
count(ap, poll_no, season_type, week) |> head(3)
n_polls

final <- filter(ap, poll_no == n_polls) |> arrange(rank)
preseason <- filter(ap, poll_no == 1) |> select(school, pre = rank)
final |>
  slice_head(n = 10) |>
  left_join(preseason, by = "school") |>
  select(rank, school, pre)
```

## 2. The first draft

One line per team that was ever ranked, ggplot2's defaults.

```{r draft, fig.alt = "First draft: a tangle of lines, one per team ever ranked in the AP poll, each in a different color, with a legend too long to read and No. 1 at the bottom."}
ggplot(ap, aes(poll_no, rank, color = school)) +
  geom_line()
```

Spaghetti: `r n_distinct(ap$school)` lines in near-identical hues, a legend
nobody can match to them, No. 1 at the bottom, and a line drawn straight across
the weeks a team dropped out of the poll, as if it had been ranked all along.

## 3. Flip it and pick the story

A poll reads top down, so the y axis is reversed, with a tick at every fifth
spot. The chart keeps only the final top 10, which also removes the gaps: all
ten were ranked in every poll (the check below confirms it), so every line is
whole. The champion and the two biggest climbers from their preseason spots are
drawn in their colors; the other seven are light grey for context.

Two teams in similar reds would read as one line, so each focus team's color is
checked against the ones already used and swapped for its secondary color when
the two are too close.

```{r focus, fig.alt = "The final AP top 10's paths through the season with No. 1 at the top, the champion and the two biggest climbers in team colors and the rest in grey."}
top10 <- slice_head(final, n = 10)
paths <- ap |>
  filter(school %in% top10$school) |>
  select(school, poll_no, rank)
stopifnot(all(count(paths, school)$n == n_polls)) # ranked in every poll: no gaps to draw across

climbers <- top10 |>
  slice(-1) |>
  inner_join(preseason, by = "school") |>
  mutate(climb = pre - rank) |>
  arrange(desc(climb), rank) |> # the final rank breaks a tie in the climb
  slice_head(n = 2)
focus <- c(top10$school[1], climbers$school)

distance <- function(a, b) sum(abs(grDevices::col2rgb(a) - grDevices::col2rgb(b))) / 255
colors <- c()
for (school in focus) {
  pick <- sdv_team_colors("cfb", school)
  if (any(vapply(colors, distance, numeric(1), b = pick) < 0.6)) {
    pick <- sdv_team_colors("cfb", school, type = "secondary")
  }
  colors[school] <- unname(pick)
}
rbind(primary = sdv_team_colors("cfb", focus), used = colors)

poll_labels <- c("Pre", 4, 8, 12, 16, "Final")
lines <- list(
  geom_line(
    aes(group = school),
    data = filter(paths, !school %in% focus), color = "grey85", linewidth = 0.8
  ),
  geom_line(
    aes(color = school),
    data = filter(paths, school %in% focus), linewidth = 1.6, lineend = "round"
  ),
  scale_color_manual(values = colors, guide = "none"),
  scale_y_reverse(breaks = c(1, 5, 10, 15, 20, 25)),
  scale_x_continuous(breaks = c(1, 4, 8, 12, 16, n_polls), labels = poll_labels)
)

ggplot(paths, aes(poll_no, rank)) +
  lines +
  labs(x = "AP poll", y = NULL) +
  theme_minimal()
```

The last focus team's red was too close to the champion's crimson, so it is
drawn in its secondary color: the `used` row above.

## 4. Logos at the line ends

Every line ends with its team's logo, in final-poll order, so the right edge
reads as the final top 10. The problem is room: ten teams finish on ten
neighboring rows, and a logo big enough to recognize is taller than a row. So
the logos alternate between two columns, odd ranks nearer the line ends and even
ranks one step further right, which gives each logo two rows of height. `width`
is a share of the panel's width. The three focus teams keep full-strength
logos; the rest are faded so the eye goes to the story.

```{r logos, fig.alt = "The bump chart with each of the final top 10's logos at the end of its line in two staggered columns, the three focus teams at full strength and the others faded."}
ends <- top10 |>
  mutate(
    x = n_polls + if_else(rank %% 2 == 1, 0.9, 1.9), # odd and even ranks in two columns
    alpha = if_else(school %in% focus, 1, 0.55)
  )

logo_ends <- function(width = 0.04) {
  list(
    geom_sdv_logos(aes(x = x, y = rank, team = school, alpha = alpha), data = ends, sport = "cfb", width = width),
    expand_limits(x = n_polls + 2.4)
  )
}

ggplot(paths, aes(poll_no, rank)) +
  lines +
  logo_ends() +
  labs(x = "AP poll", y = NULL) +
  theme_minimal()
```

## 5. Tell the story

Each focus team gets a label where its climb began, in its color, to the left of
the preseason poll so it never sits on a line; the x axis gets room for it. The
title states the finding and the subtitle what the lines are.

```{r story, fig.alt = "The finished AP poll bump chart: the final top 10's paths, the champion and two biggest climbers in team colors with labels at their preseason ranks, logos at the line ends, a headline and a source line."}
champion <- top10[1, ]
champion_pre <- preseason$pre[preseason$school == champion$school]
starts <- paths |>
  filter(school %in% focus, poll_no == 1) |>
  mutate(label = paste0(school, " (", rank, ")"))

bump <- function(logo_width = 0.04) {
  ggplot(paths, aes(poll_no, rank)) +
    lines +
    geom_text(
      aes(label = label, color = school),
      data = starts, hjust = 1, nudge_x = -0.35, size = 2.9, fontface = "bold"
    ) +
    expand_limits(x = -2.2) +
    logo_ends(logo_width) +
    labs(
      title = sprintf(
        "%s went from No. %d in the preseason to No. 1 in the final AP poll",
        champion$school, champion_pre
      ),
      subtitle = paste0(
        "AP Top 25 rank in every poll of the ", season,
        " season, preseason to final, for the teams that finished in the top 10."
      ),
      x = "AP poll", y = NULL,
      caption = "Data: CollegeFootballData.com via cfbfastR  |  Chart: sdvplotR"
    ) +
    theme_minimal(base_size = 10) +
    theme(
      plot.title.position = "plot",
      plot.title = ggtext::element_textbox_simple(face = "bold", size = 14, margin = margin(b = 4)),
      plot.subtitle = ggtext::element_textbox_simple(color = "grey40", size = 9, margin = margin(b = 8)),
      plot.caption = element_text(color = "grey45", size = 7.5),
      axis.title.x = element_text(color = "grey40", size = 9),
      axis.text = element_text(color = "grey50"),
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      plot.margin = margin(12, 14, 8, 12),
      plot.background = element_rect(fill = "white", color = NA)
    )
}

bump()
```

## 6. Export for the blog and X

The blog's 1600 x 900 and X's 1200 x 675 are both 16:9, so one 8 x 4.5 in plot
serves both, saved at 200 and 150 dpi. Inches times dpi gives the exact pixels;
the logos scale with their panel.

```{r export, fig.alt = "The finished AP poll bump chart exported at 1600 by 900 pixels for the blog."}
chart <- bump()
files <- c(
  file.path(out_dir, "ap_poll_bump_1600x900.png"),
  file.path(out_dir, "ap_poll_bump_1200x675.png")
)
ggsave(files[1], chart, width = 8, height = 4.5, dpi = 200)
ggsave(files[2], chart, width = 8, height = 4.5, dpi = 150)
magick::image_info(magick::image_read(files))[c("width", "height")]
magick::image_read(files[1])
```
