---
title: "WBB Conference Standings"
description: >
  Build the final SEC women's basketball standings table, with logos, every
  conference result as a win-loss strip and how each season ended, dressed in
  the champion's colors and exported for a newsletter and a 4:5 Instagram post.
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

**The brief:** the conference-recap newsletter wants the final SEC women's
basketball standings as an image, 1200 px wide, plus a 4:5 portrait cut
(1080 x 1350) for Instagram, where a tall table fills the screen better than a
square. Every team's conference season should be visible at a glance, along with
how its season ended in March. The games are ESPN's through wehoop, and the
table is gt with sdvplotR's logos, team colors, theme and export helpers.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr, warn.conflicts = FALSE)

season <- 2026 # the 2025-26 season, named by the year it ends
conference <- "SEC"
out_dir <- tempfile("sdvplotR-recipe-") # where the exports go; use your own folder
dir.create(out_dir)
```

## 1. Get the data

The schedule has one row per game; stacking the home and away sides gives one
row per team per game. Conference membership comes from `team_reference()`,
which lists each team's ESPN id and conference. The ids are integers on both
sides, checked before they are matched.

```{r data}
schedule <- wehoop::load_wbb_schedule(season) |>
  filter(status_type_completed)
members <- team_reference("wbb") |>
  filter(conference == !!conference) |>
  select(team_id = espn_team_id, team_abbr)
stopifnot(is.integer(schedule$home_id), is.integer(members$team_id))

side <- function(me, opp) {
  transmute(
    schedule,
    game_date,
    conf_flag = conference_competition,
    notes = coalesce(notes_headline, ""),
    team_id = .data[[paste0(me, "_id")]],
    team = .data[[paste0(me, "_location")]],
    opponent = .data[[paste0(opp, "_location")]],
    pf = .data[[paste0(me, "_score")]],
    pa = .data[[paste0(opp, "_score")]]
  )
}
games <- bind_rows(side("home", "away"), side("away", "home")) |>
  inner_join(members, by = "team_id") |>
  arrange(game_date) |>
  mutate(won = pf > pa)

games |>
  filter(conf_flag) |>
  count(notes)
```

ESPN's `conference_competition` flag is `TRUE` for any game between two
conference members, and the notes show what that sweeps in: the conference
tournament, and an early-season neutral-site final in which two SEC teams met.
The themed games (Play4Kay, We Back Pat) are real conference games. Dropping every game whose note names a tournament or a
championship leaves each team's conference schedule, which the check confirms
is the same length for everyone.

```{r standings}
games <- games |>
  mutate(is_conf = conf_flag & !grepl("Tournament|Championship", notes))
stopifnot(n_distinct(count(filter(games, is_conf), team)$n) == 1)

round_name <- function(notes) {
  round <- sub(".* - ", "", notes)
  if_else(round == "National Championship", "final", round)
}

standings <- games |>
  group_by(team_abbr, team) |>
  summarise(
    conf_results = list(as.integer(won[is_conf])), # in date order, for the strip
    conf_w = sum(won & is_conf),
    conf_l = sum(!won & is_conf),
    w = sum(won),
    l = sum(!won),
    last_notes = last(notes),
    last_result = paste0(
      if_else(last(won), "W ", "L "), max(last(pf), last(pa)), "-", min(last(pf), last(pa)), " vs ", last(opponent)
    ),
    .groups = "drop"
  ) |>
  mutate(
    march = case_when(
      startsWith(last_notes, "NCAA") ~ paste0("NCAA ", round_name(last_notes), ": ", last_result),
      startsWith(last_notes, "WBIT") ~ paste0("WBIT ", round_name(last_notes), ": ", last_result),
      grepl("Tournament", last_notes) ~ paste0(conference, " Tournament: ", last_result),
      .default = ""
    )
  ) |>
  arrange(desc(conf_w), conf_l, desc(w), team) # team breaks a full tie the same way every run
standings |>
  select(team, conf_w, conf_l, w, l, march)
```

## 2. The first draft

Hand the frame to gt as it is.

```{r draft}
gt(standings)
```

Every number is there, but the reader gets ids, a list printed as text, two
columns for every record and a raw ESPN note.

## 3. Shape it for a reader

Records read as "13-3", not two columns, and sit under spanners. The raw note
and the list go (the list comes back as a chart in the next step), and the March
column gets a real label.

```{r shape}
table <- standings |>
  mutate(
    conf = paste0(conf_w, "-", conf_l),
    overall = paste0(w, "-", l)
  ) |>
  select(team_abbr, team, conf, conf_results, overall, march)

draft <- gt(table) |>
  cols_hide(c(team_abbr, conf_results)) |>
  cols_label(team = "Team", conf = "W-L", overall = "W-L", march = "How the season ended") |>
  tab_spanner("Conference", conf) |>
  tab_spanner("Overall", overall) |>
  cols_align("center", c(conf, overall))
draft
```

## 4. Logos, team colors and the conference season

Three helpers carry the team layer. `gt_sdv_logos()` turns the abbreviation
column into logos. `gt_merge_stack_team_color()` stacks each team's name over its
overall record, the record in the team's own color, so the logo, name and
record read as one block. And gtExtras' `gt_plt_winloss()` draws the list of
conference results as a strip of pills, one per game in date order: a team's
conference season, wins up and losses down, in one glance.

```{r team-layer}
with_teams <- draft |>
  cols_unhide(c(team_abbr, conf_results)) |>
  gt_sdv_logos(team_abbr, sport = "wbb", height = 28) |>
  gt_merge_stack_team_color(team, overall, team_abbr, sport = "wbb") |>
  gtExtras::gt_plt_winloss(
    conf_results,
    max_wins = max(lengths(table$conf_results)),
    palette = c("#2e8540", "#c0392b", "grey70"),
    width = 45 # mm: wide enough for 16 readable pills
  ) |>
  cols_label(team_abbr = "", team = "Team", conf_results = "Game by game") |>
  rm_spanners("Conference") |> # widen the spanner over the new column
  tab_spanner("Conference", c(conf, conf_results))
with_teams
```

## 5. Dress it in the champion's colors

`gt_theme_sdv_team()` is the SportsDataverse table theme in one team's colors:
the title block takes the team's primary color and the rule under the column
labels its secondary. Here it is the regular-season champion's. The title states
the result, the subtitle how to read the strip, and the source note credits the
data; a footnote owns up to the ordering, which is by conference record and not
the conference's tiebreaker.

```{r final}
leaders <- filter(standings, conf_w == max(conf_w), conf_l == min(conf_l))
champ <- leaders[1, ]
title <- if (nrow(leaders) == 1) {
  sprintf("%s won the %s at %d-%d", champ$team, conference, champ$conf_w, champ$conf_l)
} else {
  sprintf("%s shared the %s title at %d-%d", paste(leaders$team, collapse = " and "), conference, champ$conf_w, champ$conf_l)
}

final <- with_teams |>
  tab_header(
    title = title,
    subtitle = paste0(
      "Final ", season - 1, "-", season %% 100, " ", conference,
      " standings. Each pill is a conference game, in date order: up for a win, down for a loss."
    )
  ) |>
  tab_source_note("Data: ESPN via wehoop  |  Table: sdvplotR + gt + gtExtras") |>
  tab_footnote(
    "Ordered by conference record, then overall record; not the conference's tiebreaker.",
    locations = cells_column_labels(columns = conf)
  ) |>
  gt_theme_sdv_team(team = champ$team_abbr, sport = "wbb")
final
```

## 6. Export for the newsletter and Instagram

`gt_save_crop()` renders the table in headless Chrome, trims it with an even
border and scales it to 1200 px. `gt_social_crop()` centers the same table on a
canvas of any shape without cropping it; `aspect_ratio = "4:5"` with
`width = 1080` gives Instagram's 1080 x 1350 portrait post.

```{r export, fig.alt = "The finished SEC women's basketball standings table exported 1200 pixels wide for the newsletter."}
newsletter <- gt_save_crop(final, file.path(out_dir, "sec_wbb_1200.png"), width = 1200)
portrait <- gt_social_crop(
  final, file.path(out_dir, "sec_wbb_1080x1350.png"),
  aspect_ratio = "4:5", width = 1080
)
magick::image_info(magick::image_read(c(newsletter, portrait)))[c("width", "height")]
magick::image_read(newsletter)
```

The 4:5 portrait cut for Instagram:

```{r export-portrait, fig.alt = "The same table centered on a 1080 by 1350 pixel portrait canvas for Instagram."}
magick::image_read(portrait)
```
