---
title: "Building a Schedule Matrix"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

A schedule matrix puts one team per row and one week per column, and draws each
opponent's logo in the cell where they met. This article builds two:

1. **A conference's season**: every SEC football team's 2025 schedule, each cell
   shaded by whether the game was at home, away or on a neutral field, with the
   byes marked.
2. **One team's season**: the 2025-26 NBA team with the best record, laid out as
   a calendar, each game shaded by whether it was won or lost.

The catch in both is the coloring. A cell's fill is a decision made cell by cell,
not column by column. `gt_highlight_cells()` fills individual cells by a
condition and `gt_legend_discrete()` draws the matching key; `gt_sdv_logos()`
turns the opponents into logos. The SEC schedule comes from
[`cfbfastR`](https://cfbfastR.sportsdataverse.org) (CollegeFootballData.com,
which needs a free API key) and the NBA games from
[`hoopR`](https://hoopR.sportsdataverse.org)'s ESPN team box scores.

`gt_highlight_cells()` and `gt_legend_discrete()` come from Andrew Weatherman's
[gtUtils](https://github.com/andreweatherman/gtUtils), and the SEC matrix follows
his schedule-matrix example.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr)
library(tidyr)
```

## 1. A conference's season, one row per team

`load_cfb_schedules()` holds one row per game. `nflreadr::clean_homeaway()`
doubles it into one row per team per game, renaming the `home_*` / `away_*` pairs
to `team_*` / `opponent_*` and adding a `location` of `"home"` or `"away"`. We
fold the schedule's `neutral_site` flag into that column as a third value, then
keep the SEC. Week 15 holds only the conference title games, so the matrix stops
at week 14.

```{r sec-data}
sec <- cfbfastR::load_cfb_schedules(2025) |>
  filter(season_type == "regular", week <= 14) |>
  select(week, home_team, away_team, home_conference, away_conference, neutral_site) |>
  nflreadr::clean_homeaway() |>
  filter(team_conference == "SEC") |>
  mutate(location = if_else(neutral_site %in% TRUE, "neutral", location)) |>
  select(team, week, location, opponent) |>
  arrange(team, week)
```

The cells will hold the opponents' school names, not image URLs.
`gt_sdv_logos()` resolves a name like `"Texas A&M"` or `"Central Arkansas"` to
the school's logo when the table renders, FBS and FCS alike.

```{r sec-pivot}
opponents <- sec |>
  pivot_wider(id_cols = team, names_from = week, values_from = opponent, names_sort = TRUE)

wk <- setdiff(names(opponents), "team")
```

`names_sort = TRUE` sorts the week columns. Because `week` is numeric they come
out `1` through `14` in order rather than `1, 10, 11, ...`.

## 2. Carrying the location as a parallel mask

`pivot_wider()` on the opponent loses the location. Pivot it into its own frame of
the same shape and turn each state into a logical mask.

```{r sec-masks}
locations <- sec |>
  pivot_wider(id_cols = team, names_from = week, values_from = location, names_sort = TRUE)
locations <- locations[match(opponents$team, locations$team), wk]

home_mask <- locations == "home"
away_mask <- locations == "away"
neutral_mask <- locations == "neutral"
```

The `match()` line holds `locations` to the row order of `opponents`, so the two
frames line up cell for cell. Each mask is a `TRUE`/`FALSE` grid the same shape as
the week columns, which is what `gt_highlight_cells()` takes. A bye is a missing
opponent, so it needs no mask: its cells are `NA` in every mask, which
`gt_highlight_cells()` reads as no match.

## 3. Filling cells by condition

`gt_highlight_cells()` fills the cells in a block of columns that meet a
condition: either a formula tested against each column, like `~ is.na(.x)`, or a
logical matrix the same shape as the columns, which is what the masks are. It is
the per-cell counterpart to `gtExtras::gt_highlight_cols()` and
`gt_highlight_rows()`, which fill whole columns or rows. Anything in `...` goes to
`gt::cell_text()` for the filled cells, which is how the byes get small grey type.

```{r sec-fills, eval = FALSE}
# ... |>
gt_highlight_cells(all_of(wk), condition = home_mask, fill = "#FBE7C6") |>
  gt_highlight_cells(all_of(wk), condition = away_mask, fill = "#FFFFFF") |>
  gt_highlight_cells(all_of(wk), condition = neutral_mask, fill = "#CFE0C3") |>
  gt_highlight_cells(
    all_of(wk),
    condition = ~ is.na(.x), fill = "#E4E4E4",
    text_color = "#565C64", size = px(9), weight = 600
  )
```

## 4. Logos for the opponents and the team

`gt_sdv_logos()` takes a whole block of columns, so one call draws every opponent;
each logo's `alt` text is the school's full name. On the team column,
`include_name = TRUE` keeps the name beside its logo. The byes are filled in with
`sub_missing()`; a cell that names no team passes through `gt_sdv_logos()` as
text.

```{r sec-logos, eval = FALSE}
# ... |>
sub_missing(columns = all_of(wk), missing_text = "BYE") |>
  gt_sdv_logos(columns = all_of(wk), sport = "cfb", height = 24) |>
  gt_sdv_logos(columns = team, sport = "cfb", height = 22, include_name = TRUE)
```

## 5. The legend

`gt_legend_discrete()` draws the key. Pass it a named vector of `label = color`
and it renders the swatches, using the same fills as the cells so the key matches
the table. `label_placement = "inside"` prints each label on its swatch, with the
text color chosen per swatch for contrast. The `heading_style` and `label_style`
lists take the same keys as every other style list in the package (see
[Styling Headers, Legends and Captions](styling.html)).

## The SEC matrix

```{r sec-table}
location_key <- c(
  "Home game" = "#FBE7C6", "Away game" = "#FFFFFF",
  "Neutral site" = "#CFE0C3", "Bye week" = "#E4E4E4"
)

opponents |>
  gt(id = "sec-schedule") |>
  gt_theme_scoreboard() |>
  gt_highlight_cells(all_of(wk), condition = home_mask, fill = "#FBE7C6") |>
  gt_highlight_cells(all_of(wk), condition = away_mask, fill = "#FFFFFF") |>
  gt_highlight_cells(all_of(wk), condition = neutral_mask, fill = "#CFE0C3") |>
  gt_highlight_cells(
    all_of(wk),
    condition = ~ is.na(.x), fill = "#E4E4E4",
    text_color = "#565C64", size = px(9), weight = 600
  ) |>
  sub_missing(columns = all_of(wk), missing_text = "BYE") |>
  gt_sdv_logos(columns = all_of(wk), sport = "cfb", height = 24) |>
  gt_sdv_logos(columns = team, sport = "cfb", height = 22, include_name = TRUE) |>
  cols_align(columns = all_of(wk), "center") |>
  cols_width(team ~ px(136), all_of(wk) ~ px(36)) |>
  cols_label(team = "") |>
  gt_border_grid(weight = 2) |>
  tab_options(table_body.hlines.width = px(2), data_row.padding = px(3)) |>
  tab_style(locations = cells_column_labels(), cell_text(size = px(13))) |>
  tab_style(locations = cells_body(columns = team), cell_text(size = px(13), weight = 600)) |>
  gt_legend_discrete(
    location_key,
    location = "top", label_placement = "inside", shape = "square",
    gap = 8, swatch_size = 20, border_width = 1.5, border_color = "black",
    heading = "2025 SEC football schedules",
    heading_style = list(
      font = "Oswald", size = 32, weight = 600,
      transform = "uppercase", margin_bottom = -2
    ),
    label_style = list(font = "Oswald", size = 13, margin_bottom = 6)
  ) |>
  gt_538_caption(
    "Regular season, weeks 1-14. The SEC title game (week 15) is left out.",
    "Data: CollegeFootballData.com via cfbfastR"
  )
```

`gt_border_grid()` draws the black rules between every cell, which is what makes
the fills read as a grid of games rather than a striped list.

## 6. One team's season as a calendar

The second matrix turns a single season into a calendar: one row per week, one
column per weekday, the opponent's logo on each game day. We find the team with
the most regular-season wins (alphabetical order breaks a tie, so a rebuild picks
the same team), then keep its games.

```{r nba-data}
# the NBA Cup final is filed as a regular-season game but counts in no record
cup_final <- hoopR::load_nba_schedule(2026) |>
  filter(type_abbreviation == "CC") |>
  pull(id)

nba <- hoopR::load_nba_team_box(2026) |>
  # regular season; the box scores also hold the All-Star games
  filter(season_type == 2, team_abbreviation %in% valid_team_names("nba"), !game_id %in% cup_final)

best <- nba |>
  summarise(wins = sum(team_winner), losses = sum(!team_winner), .by = team_abbreviation) |>
  arrange(desc(wins), team_abbreviation) |>
  slice(1)

days <- c("Mon", "Tue", "Wed", "Thu", "Fri", "Sat", "Sun")

games <- nba |>
  filter(team_abbreviation == best$team_abbreviation) |>
  mutate(
    weekday = as.integer(format(game_date, "%u")),
    day = days[weekday],
    week_of = game_date - (weekday - 1)
  ) |>
  select(week_of, day, opponent = opponent_team_abbreviation, won = team_winner)
```

`format(game_date, "%u")` gives the weekday as a number, Monday `1` to Sunday `7`,
whatever language the computer runs in, and subtracting it walks each game back
to the Monday of its week. Two pivots from the same frame give the logos and the
results, and `complete()` adds any week without a game (an All-Star break can
leave one), so the weeks run down the calendar without a gap.

```{r nba-calendar}
all_weeks <- seq(min(games$week_of), max(games$week_of), by = 7)

calendar <- games |>
  pivot_wider(id_cols = week_of, names_from = day, values_from = opponent) |>
  complete(week_of = all_weeks)
results <- games |>
  pivot_wider(id_cols = week_of, names_from = day, values_from = won) |>
  complete(week_of = all_weeks)

# every weekday as a column, in calendar order
calendar <- calendar[c("week_of", days)]
results <- results[days]

calendar <- calendar |>
  mutate(week = paste(month.abb[as.integer(format(week_of, "%m"))], as.integer(format(week_of, "%d")))) |>
  select(week, all_of(days))

win_mask <- results == TRUE
loss_mask <- results == FALSE
```

`month.abb` names the week in English whatever the system locale is, so the
labels match on every machine that builds this page.

`gt_theme_sdv_team()` dresses the calendar in the team's own colors: the title
block in its primary color and the line under the labels in its secondary. The
fills sit under the logos, so a glance down the grid shows the streaks.

```{r team-calendar}
team_name <- team_reference("nba")$team_name[team_reference("nba")$team_abbr == best$team_abbreviation]

calendar |>
  gt(id = "team-calendar") |>
  gt_highlight_cells(all_of(days), condition = win_mask, fill = "#CDEBD3") |>
  gt_highlight_cells(all_of(days), condition = loss_mask, fill = "#F6CFCB") |>
  sub_missing(columns = all_of(days), missing_text = "") |>
  gt_sdv_logos(columns = all_of(days), sport = "nba", height = 26) |>
  cols_label(week = "Week of") |>
  cols_align(columns = all_of(days), "center") |>
  cols_width(week ~ px(78), all_of(days) ~ px(52)) |>
  tab_header(
    title = paste0(team_name, ", ", best$wins, "-", best$losses),
    subtitle = "The 2025-26 regular season, game by game"
  ) |>
  gt_theme_sdv_team(team = best$team_abbreviation, sport = "nba", density = "compact") |>
  gt_border_grid(color = "#FFFFFF", weight = 2) |>
  gt_legend_discrete(
    c("Win" = "#CDEBD3", "Loss" = "#F6CFCB"),
    location = "bottom", label_placement = "inside", shape = "rounded",
    swatch_size = 18, gap = 6
  ) |>
  tab_source_note("Data: ESPN box scores via hoopR")
```

`gt_legend_discrete(location = "bottom")` sets the key under the table, where it
does not compete with the team-colored title block. A white `gt_border_grid()`
separates the fills without drawing a hard rule through the logos.
