---
title: "Shot Grid"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 6.5, dpi = 96
)
```

A shot chart with one dot per attempt shows where a player shoots, but not
whether he makes them more often than anyone else from there. This recipe cuts
the half court into 3 ft squares and colors each square by the player's field
goal percentage minus the league's from the same square. Busier squares are
drawn more solid, and **squares where the player took fewer than 5 attempts are
left out**, so no tile rests on one or two shots.

The court is drawn with `ggplot2::annotate()`, so the recipe needs no court
package. The colors are the team's own, from `sdv_team_colors()`, and the logo
in the title comes from `ggtitle_image()`.

```{r setup, message = FALSE}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```

## The data

hoopR's `load_nba_shots()` reads the SportsDataverse release of ESPN shot
locations, one row per shot, free throws included. Season `2026` is 2025-26.

```{r data}
shots <- as.data.frame(hoopR::load_nba_shots(2026))
dim(shots)
names(shots)
```

The file also holds the play-in and the playoffs. `load_nba_schedule()` marks
each game's `season_type` (`2` is the regular season), so keep those games,
drop free throws, and drop heaves from past half court.

```{r fga}
schedule <- as.data.frame(hoopR::load_nba_schedule(2026))
stopifnot(identical(class(shots$game_id), class(schedule$game_id)))
regular <- schedule$game_id[schedule$season_type == 2]

fga <- shots |>
  filter(
    game_id %in% regular,
    !grepl("Free Throw", type_text),
    coordinate_y_raw <= 41
  ) |>
  mutate(x = coordinate_x_raw - 25, y = coordinate_y_raw, made = scoring_play)
nrow(fga)
```

`coordinate_x_raw` runs 0 to 50 ft across the court, so the rim is at 25 and
`x` is feet left or right of it. `coordinate_y_raw` is feet from the rim toward
half court, with the baseline about 5 ft behind it. Two checks of that frame:
dunks should land about a foot from the rim, and no made three should be
closer than the 22 ft corner three (`score_value` is the points a shot scored,
so a made three has `score_value == 3`).

```{r frame}
dunks <- fga |> filter(grepl("Dunk", type_text))
threes <- fga |> filter(made, score_value == 3)
c(
  dunk_median_ft = median(sqrt(dunks$x^2 + dunks$y^2)),
  three_min_ft = min(sqrt(threes$x^2 + threes$y^2))
)
stopifnot(
  median(sqrt(dunks$x^2 + dunks$y^2)) < 2,
  min(sqrt(threes$x^2 + threes$y^2)) >= 22
)
```

## The grid

Round each shot to the centre of a 3 ft square, with one square centred on the
rim. The league's field goal percentage per square uses every regular-season
attempt, the player's included.

```{r grid}
bin <- 3
fga <- fga |> mutate(bx = round(x / bin) * bin, by = round(y / bin) * bin)

league <- fga |>
  summarise(league_fg = mean(made), .by = c(bx, by))

shooter <- "Shai Gilgeous-Alexander"
grid <- fga |>
  filter(athlete_name_1 == shooter) |>
  summarise(attempts = n(), fg = mean(made), .by = c(bx, by)) |>
  left_join(league, by = c("bx", "by")) |>
  mutate(diff = fg - league_fg)

shown <- grid |> filter(attempts >= 5)
stopifnot(all(shown$attempts >= 5))
c(
  attempts = sum(grid$attempts),
  squares = nrow(grid),
  squares_shown = nrow(shown),
  attempts_shown = sum(shown$attempts)
)
```

## The court

NBA dimensions in feet, with the rim at `(0, 0)`: the three-point arc is
23.75 ft from the rim and 22 ft in the corners, the lane is 16 ft wide, the
free-throw line is 15 ft from the backboard, and the restricted area is a 4 ft
arc. The function returns a list of layers, so it adds to a plot like any
other layer.

```{r court}
court <- function(colour = "grey40", linewidth = 0.4) {
  arc <- function(r, from, to, y0 = 0) {
    a <- seq(from, to, length.out = 120)
    data.frame(x = r * cos(a), y = y0 + r * sin(a))
  }
  corner <- sqrt(23.75^2 - 22^2)
  three <- arc(23.75, atan2(corner, 22), pi - atan2(corner, 22))
  free_throw <- arc(6, 0, pi, y0 = 13.75)
  restricted <- arc(4, 0, pi)
  rim <- arc(0.75, 0, 2 * pi)
  line <- function(geom, ...) {
    annotate(geom, ..., colour = colour, linewidth = linewidth)
  }
  list(
    line("rect", xmin = -25, xmax = 25, ymin = -5.25, ymax = 41.75, fill = NA),
    line("rect", xmin = -8, xmax = 8, ymin = -5.25, ymax = 13.75, fill = NA),
    line("path", x = free_throw$x, y = free_throw$y),
    line("path", x = restricted$x, y = restricted$y),
    line("path", x = three$x, y = three$y),
    line("segment", x = c(-22, 22), xend = c(-22, 22), y = -5.25, yend = corner),
    line("segment", x = -3, xend = 3, y = -1.25, yend = -1.25),
    line("path", x = rim$x, y = rim$y)
  )
}
```

## The plot

Tiles go under the court lines. The view stops 37 ft from the baseline
(31.75 ft past the rim), and the first line of the chunk checks that every
shown square fits inside it. The fill runs from the team's secondary color
(below the league) through grey (level with it) to its primary color (above
it), capped at 20 points either way so one hot corner does not wash out the
rest. Alpha follows attempts. If a team's secondary color is close to grey or
white, the "below" end would blur into the midpoint: pass a darker color as
`below` for that team.

```{r plot, fig.alt = "Half-court grid of Shai Gilgeous-Alexander's 2025-26 regular-season field goal attempts. Each 3 ft square is colored by his field goal percentage minus the league's from the same square, and drawn more solid where he took more shots. Squares with fewer than 5 attempts are left out."}
stopifnot(max(shown$by) + bin / 2 <= 31.75)
shooter_team <- unique(fga$team_abbrev[fga$athlete_name_1 %in% shooter])
shooter_team
# a player traded mid-season has two teams: pick one before taking its colors
stopifnot(length(shooter_team) == 1)
above <- sdv_team_colors("nba", shooter_team, type = "primary")
below <- sdv_team_colors("nba", shooter_team, type = "secondary")

ggplot(shown) +
  geom_tile(
    aes(bx, by, fill = diff, alpha = attempts),
    width = bin, height = bin
  ) +
  court() +
  scale_fill_gradient2(
    low = below, mid = "grey85", high = above, midpoint = 0,
    limits = c(-0.2, 0.2), oob = scales::squish,
    labels = function(v) sprintf("%+.0f", 100 * v),
    name = "FG% vs league\n(points)"
  ) +
  scale_alpha_continuous(range = c(0.35, 1), name = "Attempts") +
  coord_fixed(xlim = c(-25, 25), ylim = c(-5.25, 31.75), expand = FALSE) +
  ggtitle_image(
    title_image = shooter_team,
    title = shooter,
    subtitle = "2025-26 regular season, 3 ft squares with 5+ attempts",
    image_height = 22,
    sport = "nba"
  ) +
  labs(caption = "Data: hoopR::load_nba_shots()") +
  theme_void() +
  theme_title_image(size = 16) +
  theme(legend.position = "bottom", plot.margin = margin(8, 8, 8, 8))
```
