---
title: "Signature Ribbon"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", fig.width = 8, fig.height = 4.5, dpi = 96
)
```

Every shooter has a shape: how often he shoots from each distance, and how
well. This recipe draws that shape as a ribbon. The x axis is shot distance in
1 ft buckets, the ribbon's thickness is the share of his attempts from that
distance, and its fill is his field goal percentage minus the league's from the
same distance. **Buckets with fewer than 10 attempts are faded**, so a 3-for-4
from 31 ft does not read like a real strength.

```{r setup, message = FALSE}
library(sdvplotR)
library(ggplot2)
library(dplyr, warn.conflicts = FALSE)
```

## The data

The `nba_stats_metric_curves` release has one file per season and no R
loader, so read it by URL with `nflreadr::load_from_url()`. Its assets are
named by the season's ending year, so `2026` is 2025-26.

```{r data}
url <- paste0(
  "https://github.com/sportsdataverse/sportsdataverse-data/releases/download/",
  "nba_stats_metric_curves/metric_curves_2026.rds"
)
curves <- as.data.frame(nflreadr::load_from_url(url))
stopifnot(nrow(curves) > 0)
dim(curves)
count(curves, entity_type, metric)
```

Each row is one bucket of one curve: `attempts`, `successes` and `rate` between
`x_lo` and `x_hi` feet, for the league, a team or a player. The league curve
is the reference:

```{r league}
league <- curves |>
  filter(entity_type == "league", metric == "fg_pct_by_shot_distance") |>
  select(x_lo, x_hi, league_attempts = attempts, league_rate = rate)
league
```

The buckets are 1 ft wide out to 35 ft, then 35 to 50 and 50 to 95. Those last
two are mostly end-of-quarter heaves, so the ribbon stops at 35 ft.

## One shooter

Player rows carry a last name only (`"Brown"` names several players), so pick
the shooter by `entity_id`, the stats.nba.com person id. Buckets he never shot
from have no row; they join in from the league curve as zero attempts.

```{r shooter}
shooter_id <- "1629029"
shooter <- "Luka Dončić"

mine <- curves |>
  filter(
    entity_type == "player", entity_id == shooter_id,
    metric == "fg_pct_by_shot_distance"
  )
unique(mine$entity_name)
stopifnot(nrow(mine) > 0, !anyDuplicated(mine$x_lo))

ribbon <- league |>
  filter(x_hi <= 35) |>
  left_join(select(mine, x_lo, attempts, rate), by = "x_lo") |>
  mutate(
    attempts = coalesce(attempts, 0L),
    share = attempts / sum(attempts),
    league_share = league_attempts / sum(league_attempts),
    diff = rate - league_rate,
    sample = if_else(attempts < 10, "Under 10 attempts", "10+ attempts")
  )
sum(ribbon$attempts)
count(ribbon, sample)
```

The team colors come from the team rows, whose `entity_name` is the team's
abbreviation. The file gives each player one `team_id`, so a player traded
mid-season still maps to one team, but check that the lookup finds exactly
one.

```{r team}
shooter_team <- curves |>
  filter(entity_type == "team", entity_id %in% unique(mine$team_id)) |>
  pull(entity_name) |>
  unique()
stopifnot(length(shooter_team) == 1)
shooter_team
above <- sdv_team_colors("nba", shooter_team, type = "primary")
below <- sdv_team_colors("nba", shooter_team, type = "secondary")
```

## The ribbon

Each bucket is a rectangle centred on zero, `share` thick. The dashed outline is
the league's shape, for comparison. The fill runs from the team's secondary
color (below the league) through grey to its primary color (above the
league), capped at 20 points. If a team's secondary color is close to grey or
white, the "below" end blurs into the midpoint: pass a darker color as `below`
for that team.

```{r plot, fig.alt = "Ribbon of Luka Dončić's 2025-26 field goal attempts by shot distance, 0 to 35 feet. Thickness is his share of attempts from each 1 ft bucket, fill is his field goal percentage minus the league's from that distance, and buckets with fewer than 10 attempts are faded. A dashed outline shows the league's share."}
ggplot(ribbon) +
  geom_rect(
    aes(
      xmin = x_lo, xmax = x_hi, ymin = -share / 2, ymax = share / 2,
      fill = diff, alpha = sample
    )
  ) +
  geom_step(
    aes(x_lo, league_share / 2),
    linetype = "dashed", colour = "grey30"
  ) +
  geom_step(
    aes(x_lo, -league_share / 2),
    linetype = "dashed", colour = "grey30"
  ) +
  annotate(
    "text", x = 23.75, y = -max(ribbon$share) / 2, label = "3-pt arc",
    hjust = 1.1, vjust = 0, size = 3, colour = "grey30"
  ) +
  geom_vline(xintercept = 23.75, colour = "grey60", linewidth = 0.3) +
  scale_fill_gradient2(
    low = below, mid = "grey85", high = above, midpoint = 0,
    limits = c(-0.2, 0.2), oob = scales::squish, na.value = "grey95",
    labels = function(v) sprintf("%+.0f", 100 * v),
    name = "FG% vs league\n(points)"
  ) +
  scale_alpha_manual(
    values = c("10+ attempts" = 1, "Under 10 attempts" = 0.3), name = NULL
  ) +
  scale_x_continuous(breaks = seq(0, 35, 5), name = "Shot distance (ft)") +
  scale_y_continuous(
    labels = function(v) scales::percent(2 * abs(v), accuracy = 1),
    name = "Share of attempts\n(ribbon thickness)"
  ) +
  ggtitle_image(
    title_image = shooter_team,
    title = shooter,
    subtitle = "2025-26, 1 ft buckets to 35 ft, buckets under 10 attempts faded",
    image_height = 22,
    sport = "nba"
  ) +
  labs(caption = "Data: sportsdataverse-data release nba_stats_metric_curves") +
  theme_minimal() +
  theme_title_image(size = 16) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```
