---
title: "Creating Tier Lists"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

A tier list ranks teams into colored bands, best band on top, each band a row of
logos. `gt_tiers()` draws one from a `gt` table, but it needs the data in a
particular shape: one row per tier, a `tier` column naming it, and one column per
slot holding an image URL. This article builds two:

1. **The 2025 NFL season** in four tiers by EPA per play margin, from
   [`nflfastR`](https://www.nflfastr.com)'s play-by-play, on a light ground.
2. **The ACC's 2025-26 men's basketball season** in three tiers by efficiency
   margin, from [`hoopR`](https://hoopR.sportsdataverse.org)'s ESPN box scores, on
   the default dark ground.

`gt_tiers()` and `gt_theme_tier()` come from Andrew Weatherman's
[gtUtils](https://github.com/andreweatherman/gtUtils), and this article follows
his tier-list example.

```{r setup}
library(sdvplotR)
library(gt)
library(dplyr)
library(tidyr)
```

## 1. NFL teams by EPA per play margin

A team's EPA per play margin is the expected points it added per offensive snap
minus the expected points it allowed per defensive snap, over pass and run plays.
One season of play-by-play holds both ends.

```{r nfl-data}
pbp <- nflreadr::load_pbp(2025) |>
  filter(season_type == "REG", !is.na(epa), pass == 1 | rush == 1)

nfl <- inner_join(
  summarise(pbp, off_epa = mean(epa), .by = posteam),
  summarise(pbp, def_epa = mean(epa), .by = defteam),
  by = c(posteam = "defteam")
) |>
  rename(team = posteam) |>
  mutate(margin = off_epa - def_epa)
```

### Binning into tiers

Four tiers: the top eight, the next eight, and so on. `row_number()` ranks the
teams with the abbreviation as a tiebreaker, so a rebuild bins them the same way,
and integer division by eight turns the rank into a tier. `tier_order` is each
team's place within its tier, which becomes its column.

```{r nfl-tiers}
nfl_tiers <- c(Contenders = "#1B7837", Solid = "#7FBF7B", Shaky = "#F1B6A8", Rebuilding = "#B2182B")

nfl <- nfl |>
  arrange(desc(margin), team) |>
  mutate(
    rank = row_number(),
    tier = factor(names(nfl_tiers)[(rank - 1) %/% 8 + 1], levels = names(nfl_tiers)),
    tier_order = (rank - 1) %% 8 + 1
  )
```

`nfl_tiers` is a named vector of `tier = color`. `gt_tiers()` takes that shape
directly, and so does `gt_legend_discrete()`, so one object can drive both.

### The image columns

The images come from sdvplotR's own team table: `sdv_logo_url()` resolves the
same nflverse abbreviations the play-by-play uses to each team's logo, the
SportsDataverse logo archive's copy of it.

```{r nfl-wide}
nfl_wide <- nfl |>
  mutate(logo = sdv_logo_url(team, sport = "nfl")) |>
  pivot_wider(id_cols = tier, names_from = tier_order, values_from = logo, names_sort = TRUE) |>
  arrange(tier)
```

`arrange(tier)` sorts by the factor levels, so the best tier is on top rather
than whichever comes first alphabetically. Each row now holds a tier name and
eight logo URLs:

```{r nfl-peek}
nfl_wide[, 1:3]
```

### Plotting

`gt_tiers()` applies `gt_theme_tier()` (dark unless `style = "light"`), renders
every column other than `tier` as an image, clears the column labels, and fills
each tier cell with its color, with black or white text by whichever contrasts
more.

```{r nfl-tier-table}
nfl_wide |>
  gt() |>
  gt_tiers(nfl_tiers, style = "light", img_height = "46px") |>
  tab_header(
    title = "The 2025 NFL season in four tiers",
    subtitle = "Ranked by EPA per play margin: offense minus defense, pass and run plays, regular season"
  ) |>
  tab_source_note("Data: nflverse play-by-play via nflfastR")
```

## 2. The ACC by efficiency margin

For the second list we rank the ACC's men's basketball teams by efficiency
margin: points scored minus points allowed per 100 possessions over the 2025-26
season, tournaments included. `team_reference()` lists the conference's members,
with their ESPN ids and logos. The box scores hold one row per team per game;
joining each row to its opponent's row gives the possessions on both sides.

```{r acc-data}
acc <- team_reference("mbb") |>
  filter(conference == "ACC")

# possessions estimated from the box score
box <- hoopR::load_mbb_team_box(seasons = 2026) |>
  mutate(poss = field_goals_attempted - offensive_rebounds + total_turnovers + 0.475 * free_throws_attempted)

acc_margin <- box |>
  filter(team_id %in% acc$espn_team_id) |>
  inner_join(
    select(box, game_id, opponent_team_id = team_id, opp_poss = poss),
    by = c("game_id", "opponent_team_id")
  ) |>
  summarise(
    eff_margin = 100 * sum(team_score - opponent_team_score) / sum((poss + opp_poss) / 2),
    .by = team_id
  ) |>
  mutate(team = acc$team_abbr[match(team_id, acc$espn_team_id)])
```

This time the tiers are uneven: the top quarter, the middle half and the bottom
quarter. `cut()` divides the ranks at their 25th and 75th percentiles.

```{r acc-tiers}
acc_tiers <- c(Elite = "#FF7F7F", Average = "#FFDF7F", Poor = "#BFFF7F")

acc_margin <- acc_margin |>
  arrange(desc(eff_margin), team) |>
  mutate(
    rank = row_number(),
    tier = cut(rank,
      breaks = quantile(rank, probs = c(0, 0.25, 0.75, 1)),
      labels = names(acc_tiers),
      include.lowest = TRUE
    )
  ) |>
  mutate(tier_order = row_number(), .by = tier)
```

A dark table wants dark-mode logos: `variant = "dark"`, which gives the primary
for a team with no dark mark (every team has one; most conferences don't).

```{r acc-wide}
acc_wide <- acc_margin |>
  mutate(logo = sdv_logo_url(team, sport = "mbb", variant = "dark")) |>
  pivot_wider(id_cols = tier, names_from = tier_order, values_from = logo, names_sort = TRUE) |>
  arrange(tier)
```

```{r acc-tier-table}
acc_wide |>
  gt() |>
  gt_tiers(acc_tiers, img_height = "48px") |>
  tab_header(
    title = "The ACC in three tiers",
    subtitle = "Efficiency margin per 100 possessions, 2025-26 men's basketball"
  ) |>
  tab_source_note("Data: ESPN box scores via hoopR")
```

The middle tier holds half the conference, so it runs the widest. `gt_tiers()`
does not wrap a long row; shrink `img_height` or split the tier when a row gets
too wide for where the table is going.
